### 2024-07-18 ms.Rnw
set up

\documentclass[10pt]{article}
\usepackage{color}
\usepackage{graphicx} % Required for inserting 
%%\usepackage{algpseudocode} % Include the package for algorithmic environment
\usepackage{algorithm}
\usepackage{booktabs} % For better-quality horizontal lines
\usepackage{algpseudocode}
\usepackage{multirow}
\usepackage{threeparttable} % For table notes
\usepackage{array} % For table formatting

\usepackage{setspace}
\usepackage{float} % add this in your preamble
\usepackage{amsmath}
\usepackage{tabularx}

\usepackage[round]{natbib}

\usepackage{multirow}
\usepackage{fourier} 
\usepackage{booktabs}
\usepackage{array}
\usepackage{makecell}
\usepackage[font={small,it}]{caption}
\usepackage[utf8]{inputenc}
\usepackage{mathtools}
\usepackage{subcaption}
\usepackage{tabstackengine}
\usepackage[toc,page]{appendix}
\usepackage{graphicx, fullpage, verbatim, amsmath}
\usepackage{url, amsfonts, amssymb, amsthm,color, enumerate}
\usepackage{placeins, listings, textcomp, mathtools, multicol, tikz}
\usepackage{dsfont}
\usepackage{amsfonts}
\usepackage{bm}

\TABstackMath

\doublespacing
% Language setting
% Replace `english' with e.g. `spanish' to change the document language
\usepackage[english]{babel}

% Set page size and margins
% Replace `letterpaper' with `a4paper' for UK/EU standard size
\usepackage[a4paper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

% Useful packages
\usepackage{graphicx}
\bibliographystyle{apalike}
\usepackage[colorlinks=true,citecolor=blue]{hyperref}
\newcommand\choleraDeath{\delta_{C}}

% Using \doublespacing in the preamble 
% changes the text to double-line spacing
\doublespacing

\begin{document}

{\Large
\textbf\newline{Poisson Approximate Likelihood compared to the particle filter} 
}
\newline

\begin{abstract}
Recently, a novel approach (Poisson Approximate Likelihood, PAL) was introduced by \cite{wwr}, which employs a Poisson approximation to posterior densities, offering a fast and consistent approximation for the likelihood function for compatible POMP models. \cite{wwr} claimed that their method, along with its associated model, improved the maximum likelihood estimation compared to traditional sequential Monte Carlo (SMC) methods used in \cite{stocks} when applied to the German rotavirus, by approximately 3200 log-likelihood units (see Table 1. \cite{wwr}). However, our analysis of comparing two methods applied to the model for the same rotavirus dataset reveals that the improvement of 3200 log-likelihood units results from the use of two different datasets differed by a scaling factor. In simulations for a correctly specified model, a standard particle filter log-likelihood estimate, optimized using iteratered filtering, outperforms PAL. For the rotavirus data, once a poor specification of initial values is fixed and the same data are used for both methods, the particle filter also outperforms PAL. 
\end{abstract}

% R starts here

<<setup,echo=FALSE,message=FALSE,warning=FALSE>>=
rm(list = ls())     # clear objects

library("knitr")
opts_knit$set(concordance=TRUE)
opts_chunk$set(
    concordance = TRUE,
    tidy = TRUE,
    message = FALSE,
    warning = FALSE,
    tidy.opts = list(
        keep.blank.line = FALSE
    ),
    comment = "",
    echo = FALSE,
    results = FALSE,
    dev.args = list(
        bg = "transparent",
        pointsize = 9
    ),
    fig.path = "figure/"
)

myround <- function(x, digits = 1) {
  # taken from the broman package
  if (digits < 1)
    stop("This is intended for the case digits >= 1.")
  if (length(digits) > 1) {
    digits <- digits[1]
    warning("Using only digits[1]")
  }
  tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
  zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
  tmp[tmp == paste0("-", zero)] <- zero
  tmp
}

graphics.off()      # close graphics windows
library(pomp)
library(magrittr)
library(plyr)
library(reshape2)
library(ggplot2)
library(scales)
library(foreach)
library(doParallel)
registerDoParallel()
stopifnot(packageVersion("pomp") >= "1.7")

RUN_LEVEL = 1
NP_MIF       = switch(RUN_LEVEL, 4, 50000)
NMIF         = switch(RUN_LEVEL, 4,  100)
ncores       = switch(RUN_LEVEL, 4,  36)
@

% OvOv model is here

<<OvOv-model>>=
# measurement model 
dmeas <- Csnippet("
                  if (ISNA(cases1)) {
                  lik = (give_log) ? 0 : 1;
                  } else {
                        lik =  dbinom(cases1, H1, q1, 1) +
                        dbinom(cases2, H2, q2, 1) +
                        dbinom(cases3, H3, q3, 1);

                    lik = (give_log) ? lik : exp(lik);

                    }")
rmeas <-  Csnippet("
                    cases1 = rbinom(H1, q1);
                    cases2 = rbinom(H2, q2);
                    cases3 = rbinom(H3, q3);
                  ")

rproc <- Csnippet("
    int I = I1 + I2 + I3;
    int trans_S1[3], trans_S2[3], trans_S3[2], trans_I1[3], trans_I2[3], trans_I3[2], trans_R1[3], trans_R2[3], trans_R3[2];

    double prob_S1[3],prob_I1[3],prob_R1[3],prob_S2[3],prob_I2[3],prob_R2[3],prob_S3[2],prob_I3[2],prob_R3[2];

    double xi = rgamma(sigma_xi, 1/sigma_xi);

    double kappa = (1 + beta11*cos(2*3.141593*t/52 + phi)) * xi;

    // Define rate
    prob_S1[0] = 1-exp(-dt*beta1*kappa*I/N); // 0->1
    prob_S1[1] = 1-exp(-delta1*dt);
    prob_S1[2] = exp(-delta1*dt) + exp(-dt*beta1*kappa*I/N) - 1;

    prob_I1[0] = 1-exp(-gamma*dt);
    prob_I1[1] = 1-exp(-delta1*dt);
    prob_I1[2] = exp(-gamma*dt)+exp(-delta1*dt) - 1;

    prob_R1[0] = 1 - exp(-omega*dt);  // E_1,t this goes back to S_1,(t+1)
    prob_R1[1] = 1 - exp(-delta1*dt);
    prob_R1[2] = exp(-omega*dt) + exp(-delta1*dt) - 1;

    prob_S2[0] = 1-exp(-dt*beta2*kappa*I/N);
    prob_S2[1] = 1-exp(-delta2*dt);
    prob_S2[2] = exp(-delta2*dt) + exp(-dt*beta2*kappa*I/N) - 1;

    prob_I2[0] = 1-exp(-dt*gamma);
    prob_I2[1] = 1-exp(-dt*delta2);
    prob_I2[2] = exp(-dt*gamma)+exp(-dt*delta2) - 1;

    prob_R2[0] = 1 - exp(-dt*omega);  // E_1,t this goes back to S_1,(t+1)
    prob_R2[1] = 1 - exp(-dt*delta2);
    prob_R2[2] = exp(-dt*omega) + exp(-dt*delta2) - 1;

    // For Age Group (3): Die first before transition;

    int S3mD, I3mD, R3mD;

    S3mD = rbinom(S3, 1-dt*mu); // S3 minus Death: mu is the death rate, so it's 1-mu here
    I3mD = rbinom(I3, 1-dt*mu);
    R3mD = rbinom(R3, 1-dt*mu);

    prob_S3[0] = 1-exp(-dt*beta3*kappa*I/N);
    prob_S3[1] = exp(-dt*beta3*kappa*I/N);

    prob_I3[0] = 1 - exp(-dt*gamma);
    prob_I3[1] = exp(-dt*gamma);

    prob_R3[0] = 1 - exp(-dt*omega);
    prob_R3[1] = exp(-dt*omega);

    // Transition
    // B: S->I
    // C: I->R
    // F: Aging: (1)->(2)->(3)
    // E: R->S
    // D: Death
    //// Note: Here S_1, S_2... are all old value from (t-1)
    rmultinom(S1, &prob_S1, 3, &trans_S1); // B, F, S-B-F
    rmultinom(I1, &prob_I1, 3, &trans_I1); // C, F, I-C-F
    rmultinom(R1, &prob_R1, 3, &trans_R1); // E, F, R-E-F

    rmultinom(S2, &prob_S2, 3, &trans_S2); // B, F, S-B-F
    rmultinom(I2, &prob_I2, 3, &trans_I2); // C, F, I-C-F
    rmultinom(R2, &prob_R2, 3, &trans_R2); // E, F, R-E-F

    rmultinom(S3mD, &prob_S3, 2, &trans_S3); // B, (S-D)-B
    rmultinom(I3mD, &prob_I3, 2, &trans_I3); // C, (I-D)-C
    rmultinom(R3mD, &prob_R3, 2, &trans_R3); // E, (R-D)-E

    S1 = trans_S1[2] + trans_R1[0] + rpois(4*1025.7); // Include Birth
    I1 = trans_I1[2] + trans_S1[0];
    R1 = trans_R1[2] + trans_I1[0];

    S2 = trans_S2[2] + trans_R2[0] + trans_S1[1]; // Include Aging
    I2 = trans_I2[2] + trans_S2[0] + trans_I1[1];
    R2 = trans_R2[2] + trans_I2[0] + trans_R1[1];

    S3 = trans_S3[1] + trans_R3[0] + trans_S2[1]; // Include Aging
    I3 = trans_I3[1] + trans_S3[0] + trans_I2[1];
    R3 = trans_R3[1] + trans_I3[0] + trans_R2[1];

    //Accumvar
    H1 += trans_S1[0];
    H2 += trans_S2[0];
    H3 += trans_S3[0];

    q1 = -1; 
    while(q1 < 0 || q1 > 1){
      q1 = rnorm(0.07, sigma_q);
    }

    q2 = -1; 
    while(q2 < 0 || q2 > 1){
      q2 = rnorm(0.07, sigma_q);
    }

    q3 = -1; 
    while(q3 < 0 || q3 > 1){
      q3 = rnorm(0.07, sigma_q);
    }
")

# define parameters (without betas)
params_fixed <- c(gamma=1, delta1=1/(5*52),delta2=1/(55*52), alpha=1/(78.86912*52), 
                  mu=0, N=82372825, omega=1/(1*52))
# WWR's rinit
rinit <- Csnippet("
    double m = N/(S10+I10+R10+S20+I20+R20+S30+I30+R30);
    I1=nearbyint(m*I10);
    I2=nearbyint(m*I20);
    I3=nearbyint(m*I30);
    S1=nearbyint(m*S10);
    S2=nearbyint(m*S20);
    S3=nearbyint(m*S30);
    R1=nearbyint(m*R10);
    R2=nearbyint(m*R20);
    R3=nearbyint(m*R30);
    H1 = 0;
    H2 = 0;
    H3 = 0;
")

# Set to MLE
params_stocks_stst_mle <- params_fixed

params_stocks_stst_mle["beta1"] <- 11.48
params_stocks_stst_mle["beta2"] <- 0.25
params_stocks_stst_mle["beta3"] <- 0.35
params_stocks_stst_mle["phi"] <- 0.14
params_stocks_stst_mle["beta11"] <- 0.16
params_stocks_stst_mle["sigma_q"] <- 0.021
params_stocks_stst_mle["sigma_xi"] <- 66.89
params_stocks_stst_mle["S10"] <- 0.047061
params_stocks_stst_mle["I10"] <- 0.000368
params_stocks_stst_mle["R10"] <- 0.015967
params_stocks_stst_mle["S20"] <- 0.015967
params_stocks_stst_mle["I20"] <- 0.000011
params_stocks_stst_mle["R20"] <- 0.003677
params_stocks_stst_mle["S30"] <- 0.237624
params_stocks_stst_mle["I30"] <- 0.000031
params_stocks_stst_mle["R30"] <- 0.001591
@

% Maximization Round 1

<<maximization1>>=
pt <- pomp::parameter_trans(
  log = c("beta1","beta2","beta3","sigma_q","sigma_xi"),
  logit=c("beta11"),
  barycentric=c("S10","I10","R10",
                "S20","I20","R20",
                "S30","I30","R30"),
  toEst= pomp::Csnippet("T_phi = logit(phi/(M_2PI));"),
  fromEst= pomp::Csnippet("phi = M_2PI*expit(T_phi);")
)

read.table("real_rotavirus_metadata.txt") %>%
  rbind(data.frame(time=0,cases1=NA,cases2=NA,cases3=NA)) %>%
  arrange(time) -> dat

pomp(data = dat,
     times="time",
     t0=0,
     dmeasure = dmeas,
     rmeasure = rmeas,
     rprocess = discrete_time(step.fun = rproc, delta.t = 1/4),
     statenames = c("S1", "I1", "R1", "H1", 
                    "S2", "I2", "R2", "H2",
                    "S3", "I3", "R3", "H3", 
                    "q1", "q2", "q3"),
     paramnames = names(params_stocks_stst_mle),
     accumvars=c("H1", "H2", "H3"),
     rinit=rinit,
     partrans = pt,
     params = params_stocks_stst_mle
) -> sir

sir_panel <- panelPomp::panelPomp(list(unit1=sir),
                                  shared=NULL,
                                  specific=params_stocks_stst_mle |> 
                                    as.matrix() |>
                                    `colnames<-`("unit1")
)

require(doParallel)
cores <- ncores
registerDoParallel(cores)
mcopts <- list(set.seed=TRUE)

sir_box <- rbind(
  beta1=c(10,15),
  beta2=c(0.2,0.4),
  beta3=c(0.3,0.5),
  phi=c(0.01,0.3),
  beta11=c(0.1,0.2),
  sigma_q=c(0.001,0.1),
  sigma_xi=c(65,70),
  S10 = c(0.01,0.05),
  I10 = c(0.0001, 0.0005),
  R10 = c(0.01, 0.05),
  S20 = c(0.01,0.7),
  I20 = c(0.000001,0.00002),
  R20 = c(0.001,0.005),
  S30 = c(0.01,0.5),
  I30 = c(0.000001, 0.0001),
  R30 = c(0.0001,0.005)
)

## Make it panelPomp
c(apply(sir_box,1,function(x)runif(1,min=x[1],max=x[2])),
  params_fixed) |> 
  as.matrix() |>
  `colnames<-`("unit1") -> starting

bake(file = "output/round_01/ovov_mif_01.rds",{ 
  mifs_global <- foreach(i=1:cores,.packages='pomp', .options.multicore=mcopts) %dopar% {
    panelPomp::mif2(
      sir_panel,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01,beta2=0.01,beta3=0.01,
                    beta11=0.01,phi=0.01,sigma_q=0.01,sigma_xi=0.01,
                    S10=ivp(0.24),I10=ivp(0.24),R10=ivp(0.24),
                    S20=ivp(0.24),I20=ivp(0.24),R20=ivp(0.24),
                    S30=ivp(0.24),I30=ivp(0.24),R30=ivp(0.24)),
      cooling.type = "geometric",
      Nmif = NMIF,
      shared.start = numeric(0),
      specific.start = starting,
      block = F
    ) 
  }
  mifs_global
})

bake(file = "output/round_01/ovov_01_el.rds",{
  el <- measlespkg::eval_logLik(mifs_global, ncores=cores, np_pf = NP_MIF, nreps=cores)
  el
})
@

% Continue: Maximization Round 2

<<maximization2>>=
sd           = 2/3
top_n_fits   = 12
# measurement model 
dmeas <- Csnippet("
                  if (ISNA(cases1)) {
                  lik = (give_log) ? 0 : 1;
                  } else {
                        lik =  dbinom(cases1, H1, q1, 1) +
                        dbinom(cases2, H2, q2, 1) +
                        dbinom(cases3, H3, q3, 1);

                    lik = (give_log) ? lik : exp(lik);

                    }")
rmeas <-  Csnippet("
                    cases1 = rbinom(H1, q1);
                    cases2 = rbinom(H2, q2);
                    cases3 = rbinom(H3, q3);
                  ")

rproc <- Csnippet("
    int I = I1 + I2 + I3;
    int trans_S1[3], trans_S2[3], trans_S3[2], trans_I1[3], trans_I2[3], trans_I3[2], trans_R1[3], trans_R2[3], trans_R3[2];

    double prob_S1[3],prob_I1[3],prob_R1[3],prob_S2[3],prob_I2[3],prob_R2[3],prob_S3[2],prob_I3[2],prob_R3[2];

    double xi = rgamma(sigma_xi, 1/sigma_xi);

    double kappa = (1 + beta11*cos(2*3.141593*t/52 + phi)) * xi;

    // Define rate
    prob_S1[0] = 1-exp(-dt*beta1*kappa*I/N); // 0->1
    prob_S1[1] = 1-exp(-delta1*dt);
    prob_S1[2] = exp(-delta1*dt) + exp(-dt*beta1*kappa*I/N) - 1;

    prob_I1[0] = 1-exp(-gamma*dt);
    prob_I1[1] = 1-exp(-delta1*dt);
    prob_I1[2] = exp(-gamma*dt)+exp(-delta1*dt) - 1;

    prob_R1[0] = 1 - exp(-omega*dt);  // E_1,t this goes back to S_1,(t+1)
    prob_R1[1] = 1 - exp(-delta1*dt);
    prob_R1[2] = exp(-omega*dt) + exp(-delta1*dt) - 1;

    prob_S2[0] = 1-exp(-dt*beta2*kappa*I/N);
    prob_S2[1] = 1-exp(-delta2*dt);
    prob_S2[2] = exp(-delta2*dt) + exp(-dt*beta2*kappa*I/N) - 1;

    prob_I2[0] = 1-exp(-dt*gamma);
    prob_I2[1] = 1-exp(-dt*delta2);
    prob_I2[2] = exp(-dt*gamma)+exp(-dt*delta2) - 1;

    prob_R2[0] = 1 - exp(-dt*omega);  // E_1,t this goes back to S_1,(t+1)
    prob_R2[1] = 1 - exp(-dt*delta2);
    prob_R2[2] = exp(-dt*omega) + exp(-dt*delta2) - 1;

    // For Age Group (3): Die first before transition;

    int S3mD, I3mD, R3mD;

    S3mD = rbinom(S3, 1-dt*mu); // S3 minus Death: mu is the death rate, so it's 1-mu here
    I3mD = rbinom(I3, 1-dt*mu);
    R3mD = rbinom(R3, 1-dt*mu);

    prob_S3[0] = 1-exp(-dt*beta3*kappa*I/N);
    prob_S3[1] = exp(-dt*beta3*kappa*I/N);

    prob_I3[0] = 1 - exp(-dt*gamma);
    prob_I3[1] = exp(-dt*gamma);

    prob_R3[0] = 1 - exp(-dt*omega);
    prob_R3[1] = exp(-dt*omega);

    // Transition
    // B: S->I
    // C: I->R
    // F: Aging: (1)->(2)->(3)
    // E: R->S
    // D: Death
    //// Note: Here S_1, S_2... are all old value from (t-1)
    rmultinom(S1, &prob_S1, 3, &trans_S1); // B, F, S-B-F
    rmultinom(I1, &prob_I1, 3, &trans_I1); // C, F, I-C-F
    rmultinom(R1, &prob_R1, 3, &trans_R1); // E, F, R-E-F

    rmultinom(S2, &prob_S2, 3, &trans_S2); // B, F, S-B-F
    rmultinom(I2, &prob_I2, 3, &trans_I2); // C, F, I-C-F
    rmultinom(R2, &prob_R2, 3, &trans_R2); // E, F, R-E-F

    rmultinom(S3mD, &prob_S3, 2, &trans_S3); // B, (S-D)-B
    rmultinom(I3mD, &prob_I3, 2, &trans_I3); // C, (I-D)-C
    rmultinom(R3mD, &prob_R3, 2, &trans_R3); // E, (R-D)-E

    S1 = trans_S1[2] + trans_R1[0] + rpois(4*1025.7); // Include Birth
    I1 = trans_I1[2] + trans_S1[0];
    R1 = trans_R1[2] + trans_I1[0];

    S2 = trans_S2[2] + trans_R2[0] + trans_S1[1]; // Include Aging
    I2 = trans_I2[2] + trans_S2[0] + trans_I1[1];
    R2 = trans_R2[2] + trans_I2[0] + trans_R1[1];

    S3 = trans_S3[1] + trans_R3[0] + trans_S2[1]; // Include Aging
    I3 = trans_I3[1] + trans_S3[0] + trans_I2[1];
    R3 = trans_R3[1] + trans_I3[0] + trans_R2[1];

    //Accumvar
    H1 += trans_S1[0];
    H2 += trans_S2[0];
    H3 += trans_S3[0];

    q1 = -1; 
    while(q1 < 0 || q1 > 1){
      q1 = rnorm(0.07, sigma_q);
    }

    q2 = -1; 
    while(q2 < 0 || q2 > 1){
      q2 = rnorm(0.07, sigma_q);
    }

    q3 = -1; 
    while(q3 < 0 || q3 > 1){
      q3 = rnorm(0.07, sigma_q);
    }
")

# define parameters (without betas)
params_fixed <- c(gamma=1, delta1=1/(5*52),delta2=1/(55*52), alpha=1/(78.86912*52), 
                  mu=0, N=82372825, omega=1/(1*52))
# WWR's rinit
rinit <- Csnippet("
    double m = N/(S10+I10+R10+S20+I20+R20+S30+I30+R30);
    I1=nearbyint(m*I10);
    I2=nearbyint(m*I20);
    I3=nearbyint(m*I30);
    S1=nearbyint(m*S10);
    S2=nearbyint(m*S20);
    S3=nearbyint(m*S30);
    R1=nearbyint(m*R10);
    R2=nearbyint(m*R20);
    R3=nearbyint(m*R30);
    H1 = 0;
    H2 = 0;
    H3 = 0;
")

# Set to MLE
params_stocks_stst_mle <- params_fixed

params_stocks_stst_mle["beta1"] <- 11.48
params_stocks_stst_mle["beta2"] <- 0.25
params_stocks_stst_mle["beta3"] <- 0.35
params_stocks_stst_mle["phi"] <- 0.14
params_stocks_stst_mle["beta11"] <- 0.16
params_stocks_stst_mle["sigma_q"] <- 0.021
params_stocks_stst_mle["sigma_xi"] <- 66.89
params_stocks_stst_mle["S10"] <- 0.047061
params_stocks_stst_mle["I10"] <- 0.000368
params_stocks_stst_mle["R10"] <- 0.015967
params_stocks_stst_mle["S20"] <- 0.015967
params_stocks_stst_mle["I20"] <- 0.000011
params_stocks_stst_mle["R20"] <- 0.003677
params_stocks_stst_mle["S30"] <- 0.237624
params_stocks_stst_mle["I30"] <- 0.000031
params_stocks_stst_mle["R30"] <- 0.001591

pt <- pomp::parameter_trans(
  log = c("beta1","beta2","beta3","sigma_q","sigma_xi"),
  logit=c("beta11"),
  barycentric=c("S10","I10","R10",
                "S20","I20","R20",
                "S30","I30","R30"),
  toEst= pomp::Csnippet("T_phi = logit(phi/(M_2PI));"),
  fromEst= pomp::Csnippet("phi = M_2PI*expit(T_phi);")
)

read.table("real_rotavirus_metadata.txt") %>%
  rbind(data.frame(time=0,cases1=NA,cases2=NA,cases3=NA)) %>%
  arrange(time) -> dat

pomp(data = dat,
     times="time",
     t0=0,
     dmeasure = dmeas,
     rmeasure = rmeas,
     rprocess = discrete_time(step.fun = rproc, delta.t = 1/4),
     statenames = c("S1", "I1", "R1", "H1", 
                    "S2", "I2", "R2", "H2",
                    "S3", "I3", "R3", "H3", 
                    "q1", "q2", "q3"),
     paramnames = names(params_stocks_stst_mle),
     accumvars=c("H1", "H2", "H3"),
     rinit=rinit,
     partrans = pt,
     params = params_stocks_stst_mle
) -> sir

sir_panel <- panelPomp::panelPomp(list(unit1=sir),
                                  shared=NULL,
                                  specific=params_stocks_stst_mle |> 
                                    as.matrix() |>
                                    `colnames<-`("unit1")
)

require(doParallel)
cores <- ncores
registerDoParallel(cores)
mcopts <- list(set.seed=TRUE)

### Next-round code
el <- readRDS("output/round_01/ovov_01_el.rds")
x <- na.omit(el$fits)
score_total = x$logLik
ranking_total = order(score_total, decreasing = TRUE)[1:top_n_fits]

best_fits = dplyr::select(
  x[ranking_total,], -"logLik", -"se"
)

recycle_vec = sort(rep_len(1:top_n_fits, cores))
full_best_fit <- best_fits[recycle_vec, ]

coef_names <- colnames(full_best_fit)
colnames(full_best_fit) <- gsub(".{7}$","",coef_names)

starting_values <- vector(cores, mode="list")

for(i in 1:cores){
  t(full_best_fit[i, ])  |> 
    as.matrix() |>
    `colnames<-`("unit1") -> starting_values[[i]] 
}

mifs_global <- foreach(i=1:cores,.packages='pomp', .options.multicore=mcopts) %dopar% {
  panelPomp::mif2(
    sir_panel,
    Np = NP_MIF,
    cooling.fraction.50 = 0.5,
    rw.sd = rw_sd(beta1=0.01*sd,beta2=0.01*sd,beta3=0.01*sd,
                  beta11=0.01*sd,phi=0.01*sd,sigma_q=0.01*sd,sigma_xi=0.01*sd,
                  S10=ivp(0.24)*sd,I10=ivp(0.24)*sd,R10=ivp(0.24)*sd,
                  S20=ivp(0.24)*sd,I20=ivp(0.24)*sd,R20=ivp(0.24)*sd,
                  S30=ivp(0.24)*sd,I30=ivp(0.24)*sd,R30=ivp(0.24)*sd),
    cooling.type = "geometric",
    Nmif = NMIF,
    shared.start = numeric(0),
    specific.start = starting_values[[i]],
    block = F
  ) 
}

bake(file = "output/round_02/ovov_mif_02.rds",{ 
  mifs_global
})

bake(file = "output/round_02/ovov_02_el.rds",{
  el <- measlespkg::eval_logLik(mifs_global, ncores = cores, np_pf = NP_MIF, nreps=cores)
  el
})
@

% Continue: Maximization Round 3

<<maximization3>>=
sd           = (2/3)^2
top_n_fits   = 12
# measurement model 
dmeas <- Csnippet("
                  if (ISNA(cases1)) {
                  lik = (give_log) ? 0 : 1;
                  } else {
                        lik =  dbinom(cases1, H1, q1, 1) +
                        dbinom(cases2, H2, q2, 1) +
                        dbinom(cases3, H3, q3, 1);

                    lik = (give_log) ? lik : exp(lik);

                    }")
rmeas <-  Csnippet("
                    cases1 = rbinom(H1, q1);
                    cases2 = rbinom(H2, q2);
                    cases3 = rbinom(H3, q3);
                  ")

rproc <- Csnippet("
    int I = I1 + I2 + I3;
    int trans_S1[3], trans_S2[3], trans_S3[2], trans_I1[3], trans_I2[3], trans_I3[2], trans_R1[3], trans_R2[3], trans_R3[2];

    double prob_S1[3],prob_I1[3],prob_R1[3],prob_S2[3],prob_I2[3],prob_R2[3],prob_S3[2],prob_I3[2],prob_R3[2];

    double xi = rgamma(sigma_xi, 1/sigma_xi);

    double kappa = (1 + beta11*cos(2*3.141593*t/52 + phi)) * xi;

    // Define rate
    prob_S1[0] = 1-exp(-dt*beta1*kappa*I/N); // 0->1
    prob_S1[1] = 1-exp(-delta1*dt);
    prob_S1[2] = exp(-delta1*dt) + exp(-dt*beta1*kappa*I/N) - 1;

    prob_I1[0] = 1-exp(-gamma*dt);
    prob_I1[1] = 1-exp(-delta1*dt);
    prob_I1[2] = exp(-gamma*dt)+exp(-delta1*dt) - 1;

    prob_R1[0] = 1 - exp(-omega*dt);  // E_1,t this goes back to S_1,(t+1)
    prob_R1[1] = 1 - exp(-delta1*dt);
    prob_R1[2] = exp(-omega*dt) + exp(-delta1*dt) - 1;

    prob_S2[0] = 1-exp(-dt*beta2*kappa*I/N);
    prob_S2[1] = 1-exp(-delta2*dt);
    prob_S2[2] = exp(-delta2*dt) + exp(-dt*beta2*kappa*I/N) - 1;

    prob_I2[0] = 1-exp(-dt*gamma);
    prob_I2[1] = 1-exp(-dt*delta2);
    prob_I2[2] = exp(-dt*gamma)+exp(-dt*delta2) - 1;

    prob_R2[0] = 1 - exp(-dt*omega);  // E_1,t this goes back to S_1,(t+1)
    prob_R2[1] = 1 - exp(-dt*delta2);
    prob_R2[2] = exp(-dt*omega) + exp(-dt*delta2) - 1;

    // For Age Group (3): Die first before transition;

    int S3mD, I3mD, R3mD;

    S3mD = rbinom(S3, 1-dt*mu); // S3 minus Death: mu is the death rate, so it's 1-mu here
    I3mD = rbinom(I3, 1-dt*mu);
    R3mD = rbinom(R3, 1-dt*mu);

    prob_S3[0] = 1-exp(-dt*beta3*kappa*I/N);
    prob_S3[1] = exp(-dt*beta3*kappa*I/N);

    prob_I3[0] = 1 - exp(-dt*gamma);
    prob_I3[1] = exp(-dt*gamma);

    prob_R3[0] = 1 - exp(-dt*omega);
    prob_R3[1] = exp(-dt*omega);

    // Transition
    // B: S->I
    // C: I->R
    // F: Aging: (1)->(2)->(3)
    // E: R->S
    // D: Death
    //// Note: Here S_1, S_2... are all old value from (t-1)
    rmultinom(S1, &prob_S1, 3, &trans_S1); // B, F, S-B-F
    rmultinom(I1, &prob_I1, 3, &trans_I1); // C, F, I-C-F
    rmultinom(R1, &prob_R1, 3, &trans_R1); // E, F, R-E-F

    rmultinom(S2, &prob_S2, 3, &trans_S2); // B, F, S-B-F
    rmultinom(I2, &prob_I2, 3, &trans_I2); // C, F, I-C-F
    rmultinom(R2, &prob_R2, 3, &trans_R2); // E, F, R-E-F

    rmultinom(S3mD, &prob_S3, 2, &trans_S3); // B, (S-D)-B
    rmultinom(I3mD, &prob_I3, 2, &trans_I3); // C, (I-D)-C
    rmultinom(R3mD, &prob_R3, 2, &trans_R3); // E, (R-D)-E

    S1 = trans_S1[2] + trans_R1[0] + rpois(4*1025.7); // Include Birth
    I1 = trans_I1[2] + trans_S1[0];
    R1 = trans_R1[2] + trans_I1[0];

    S2 = trans_S2[2] + trans_R2[0] + trans_S1[1]; // Include Aging
    I2 = trans_I2[2] + trans_S2[0] + trans_I1[1];
    R2 = trans_R2[2] + trans_I2[0] + trans_R1[1];

    S3 = trans_S3[1] + trans_R3[0] + trans_S2[1]; // Include Aging
    I3 = trans_I3[1] + trans_S3[0] + trans_I2[1];
    R3 = trans_R3[1] + trans_I3[0] + trans_R2[1];

    //Accumvar
    H1 += trans_S1[0];
    H2 += trans_S2[0];
    H3 += trans_S3[0];

    q1 = -1; 
    while(q1 < 0 || q1 > 1){
      q1 = rnorm(0.07, sigma_q);
    }

    q2 = -1; 
    while(q2 < 0 || q2 > 1){
      q2 = rnorm(0.07, sigma_q);
    }

    q3 = -1; 
    while(q3 < 0 || q3 > 1){
      q3 = rnorm(0.07, sigma_q);
    }
")

# define parameters (without betas)
params_fixed <- c(gamma=1, delta1=1/(5*52),delta2=1/(55*52), alpha=1/(78.86912*52), 
                  mu=0, N=82372825, omega=1/(1*52))
# WWR's rinit
rinit <- Csnippet("
    double m = N/(S10+I10+R10+S20+I20+R20+S30+I30+R30);
    I1=nearbyint(m*I10);
    I2=nearbyint(m*I20);
    I3=nearbyint(m*I30);
    S1=nearbyint(m*S10);
    S2=nearbyint(m*S20);
    S3=nearbyint(m*S30);
    R1=nearbyint(m*R10);
    R2=nearbyint(m*R20);
    R3=nearbyint(m*R30);
    H1 = 0;
    H2 = 0;
    H3 = 0;
")

# Set to MLE
params_stocks_stst_mle <- params_fixed

params_stocks_stst_mle["beta1"] <- 11.48
params_stocks_stst_mle["beta2"] <- 0.25
params_stocks_stst_mle["beta3"] <- 0.35
params_stocks_stst_mle["phi"] <- 0.14
params_stocks_stst_mle["beta11"] <- 0.16
params_stocks_stst_mle["sigma_q"] <- 0.021
params_stocks_stst_mle["sigma_xi"] <- 66.89
params_stocks_stst_mle["S10"] <- 0.047061
params_stocks_stst_mle["I10"] <- 0.000368
params_stocks_stst_mle["R10"] <- 0.015967
params_stocks_stst_mle["S20"] <- 0.015967
params_stocks_stst_mle["I20"] <- 0.000011
params_stocks_stst_mle["R20"] <- 0.003677
params_stocks_stst_mle["S30"] <- 0.237624
params_stocks_stst_mle["I30"] <- 0.000031
params_stocks_stst_mle["R30"] <- 0.001591

pt <- pomp::parameter_trans(
  log = c("beta1","beta2","beta3","sigma_q","sigma_xi"),
  logit=c("beta11"),
  barycentric=c("S10","I10","R10",
                "S20","I20","R20",
                "S30","I30","R30"),
  toEst= pomp::Csnippet("T_phi = logit(phi/(M_2PI));"),
  fromEst= pomp::Csnippet("phi = M_2PI*expit(T_phi);")
)

read.table("real_rotavirus_metadata.txt") %>%
  rbind(data.frame(time=0,cases1=NA,cases2=NA,cases3=NA)) %>%
  arrange(time) -> dat

pomp(data = dat,
     times="time",
     t0=0,
     dmeasure = dmeas,
     rmeasure = rmeas,
     rprocess = discrete_time(step.fun = rproc, delta.t = 1/4),
     statenames = c("S1", "I1", "R1", "H1", 
                    "S2", "I2", "R2", "H2",
                    "S3", "I3", "R3", "H3", 
                    "q1", "q2", "q3"),
     paramnames = names(params_stocks_stst_mle),
     accumvars=c("H1", "H2", "H3"),
     rinit=rinit,
     partrans = pt,
     params = params_stocks_stst_mle
) -> sir

sir_panel <- panelPomp::panelPomp(list(unit1=sir),
                                  shared=NULL,
                                  specific=params_stocks_stst_mle |> 
                                    as.matrix() |>
                                    `colnames<-`("unit1")
)

require(doParallel)
cores <- ncores
registerDoParallel(cores)
mcopts <- list(set.seed=TRUE)

### Next-round code
el <- readRDS("output/round_02/ovov_02_el.rds")
x <- na.omit(el$fits)
score_total = x$logLik
ranking_total = order(score_total, decreasing = TRUE)[1:top_n_fits]

best_fits = dplyr::select(
  x[ranking_total,], -"logLik", -"se"
)

recycle_vec = sort(rep_len(1:top_n_fits, cores))
full_best_fit <- best_fits[recycle_vec, ]

coef_names <- colnames(full_best_fit)
colnames(full_best_fit) <- gsub(".{7}$","",coef_names)

starting_values <- vector(cores, mode="list")

for(i in 1:cores){
  t(full_best_fit[i, ])  |> 
    as.matrix() |>
    `colnames<-`("unit1") -> starting_values[[i]] 
}

mifs_global <- foreach(i=1:cores,.packages='pomp', .options.multicore=mcopts) %dopar% {
  panelPomp::mif2(
    sir_panel,
    Np = NP_MIF,
    cooling.fraction.50 = 0.5,
    rw.sd = rw_sd(beta1=0.01*sd,beta2=0.01*sd,beta3=0.01*sd,
                  beta11=0.01*sd,phi=0.01*sd,sigma_q=0.01*sd,sigma_xi=0.01*sd,
                  S10=ivp(0.24)*sd,I10=ivp(0.24)*sd,R10=ivp(0.24)*sd,
                  S20=ivp(0.24)*sd,I20=ivp(0.24)*sd,R20=ivp(0.24)*sd,
                  S30=ivp(0.24)*sd,I30=ivp(0.24)*sd,R30=ivp(0.24)*sd),
    cooling.type = "geometric",
    Nmif = NMIF,
    shared.start = numeric(0),
    specific.start = starting_values[[i]],
    block = F
  ) 
}

bake(file = "output/round_03/ovov_mif_03.rds",{ 
  mifs_global
})

bake(file = "output/round_03/ovov_03_el.rds",{
  el <- measlespkg::eval_logLik(mifs_global, ncores = cores, np_pf = NP_MIF, nreps=cores)
  el
})
@

\begin{table}[htbp]
  \centering
  \begin{threeparttable}
    \caption{Maximum likelihood estimation of parameters by PAL for three models \citep{wwr}, and by mif2 algorithm \citep{pomppackagepaper} for the OvOv model.}
    \label{tab:mlebywwr}
    \begin{tabular}{>{\raggedright\arraybackslash}p{2cm} *{3}{>{\centering\arraybackslash}p{1.5cm}} >{\centering\arraybackslash}p{2.5cm}}
      \toprule
      Parameter      & EqEq  & EqOv  & OvOv  & OvOv (\textbf{pomp})$^*$ \\
      \midrule
      \(\beta_1\)    & 12.15 & 12.74 & 11.48 & $\Sexpr{'test'}$\\
      \(\beta_2\)    & 0.22  & 0.21  & 0.25  & $\Sexpr{'test'}$\\
      \(\beta_3\)    & 0.34  & 0.31  & 0.35  & $\Sexpr{'test'}$\\
      \(\phi\)       & 0.017 & 0.14  & 0.14  & $\Sexpr{'test'}$\\
      \(\rho\)       & 0.022 & 0.19  & 0.16  & $\Sexpr{'test'}$\\
      \(\sigma^2_q\) & n/a   & 0.042 & 0.021 & $\Sexpr{'test'}$\\
      \(\sigma_\xi\) & n/a   & n/a   & 66.89 & $\Sexpr{'test'}$\\
      \addlinespace
      \multicolumn{1}{l}{\(S_{10}\)} & \multicolumn{3}{c}{$3876549^*$} & $5885201$\\
      \multicolumn{1}{l}{\(I_{10}\)} & \multicolumn{3}{c}{$30351$}     & $5041$\\
      \multicolumn{1}{l}{\(R_{10}\)} & \multicolumn{3}{c}{$1315221$}   & $721006$\\
      \multicolumn{1}{l}{\(S_{20}\)} & \multicolumn{3}{c}{$57139612$}  & $44546887$\\
      \multicolumn{1}{l}{\(I_{20}\)} & \multicolumn{3}{c}{$871$}       & $22$\\
      \multicolumn{1}{l}{\(R_{20}\)} & \multicolumn{3}{c}{$302852$}    & $26360811$\\
      \multicolumn{1}{l}{\(S_{30}\)} & \multicolumn{3}{c}{$19573727$}  & $4788420$\\
      \multicolumn{1}{l}{\(I_{30}\)} & \multicolumn{3}{c}{$2550$}      & $131$\\
      \multicolumn{1}{l}{\(R_{30}\)} & \multicolumn{3}{c}{$131092$}    & $65306$\\
      \midrule
      \addlinespace
      \textbf{AIC}  & $98866.65$ & $15154.75$ & $13778.08$ & $\Sexpr{'test'}$\\
      \addlinespace
      \bottomrule
    \end{tabular}
    \begin{tablenotes}
      \small
      \item[*] The initial distribution parameter \(\lambda_0 = (S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}, R_{30})\) are assumed to be \textbf{fixed} in \cite{wwr}. The results of the estimates of the OvOv (\textbf{pomp}) model are obtained by using the Iterated Filtering algorithm with 3 rounds and 100 iterations, 50,000 particles, and 36 replicates in each round with the top 12 best fits in terms of likelihood chosen to be the starting value for the next round. 
    \end{tablenotes}
  \end{threeparttable}
\end{table}

\bibliography{palcheck.bib}

\end{document}

### 2024-07-18 thesis.tex
set up

\documentclass[10pt]{article}
\usepackage{color}
\usepackage{graphicx} % Required for inserting 
%%\usepackage{algpseudocode} % Include the package for algorithmic environment
\usepackage{algorithm}
\usepackage{booktabs} % For better-quality horizontal lines
\usepackage{algpseudocode}
\usepackage{multirow}
\usepackage{threeparttable} % For table notes
\usepackage{array} % For table formatting

\usepackage{setspace}
\usepackage{float} % add this in your preamble
\usepackage{amsmath}
\usepackage{tabularx}

\usepackage[round]{natbib}

\usepackage{multirow}
\usepackage{fourier} 
\usepackage{booktabs}
\usepackage{array}
\usepackage{makecell}
\usepackage[font={small,it}]{caption}
\usepackage[utf8]{inputenc}
\usepackage{mathtools}
\usepackage{subcaption}
\usepackage{tabstackengine}
\usepackage[toc,page]{appendix}
\usepackage{graphicx, fullpage, verbatim, amsmath}
\usepackage{url, amsfonts, amssymb, amsthm,color, enumerate}
\usepackage{placeins, listings, textcomp, mathtools, multicol, tikz}
\usepackage{ dsfont }
\usepackage{bm}
\newcommand\fixed{^\dagger}

\TABstackMath

\doublespacing
% Language setting
% Replace `english' with e.g. `spanish' to change the document language
\usepackage[english]{babel}

% Set page size and margins
% Replace `letterpaper' with `a4paper' for UK/EU standard size
\usepackage[a4paper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

% Useful packages
\usepackage{graphicx}
\bibliographystyle{apalike}
\usepackage[colorlinks=true,citecolor=blue]{hyperref}
\newcommand\choleraDeath{\delta_{C}}

% Using \doublespacing in the preamble 
% changes the text to double-line spacing
\doublespacing

\title{%
  \textbf{\LARGE{PAL versus SMC: Two Approaches in Compartmental Modeling}} \\
  \large A case study with rotavirus in Germany during 2001-2008 \\}
\author{\large Yize Hao\\[0.5cm]{\median Supervisor: Prof. Edward Ionides \\ 
Graduate Student Mentor: Aaron Abkemeier}
    }
    \date{Submitted: April 23, 2024 \\ This version: May 31, 2024}

\begin{document}

\maketitle

\begin{abstract}
Partially Observed Markov Process (POMP) models have been extensively employed in epidemiological modeling over the past several decades to understand disease patterns and inform policy-making. Although the observation density is typically assumed to be known or at least can be evaluated point-wisely, the latent Markovian models lack closed-form expressions for initial density and transition kernels, making it challenging to compute the likelihood for POMP models with complex latent models. Recently, a novel approach (Poisson Approximate Likelihood, PAL) was introduced by \cite{wwr}, which employs a Poisson approximation to posterior densities, offering a fast and consistent approximation for the likelihood function. \cite{wwr} claimed that their method, along with its associated model, improved the maximum likelihood estimation compared to traditional sequential Monte Carlo (SMC) methods used in \cite{stocks} when applied to the German rotavirus, by approximately 3200 log-likelihood units. However, our analysis of comparing two methods applied to the model for the same rotavirus dataset reveals that the improvement of 3200 log-likelihood units results from the use of two different datasets differed by a scaling factor. Moreover, although PAL and SMC are two ways of approximating the likelihood, within the framework of models compatible with PAL, when computation time is preferred, PAL is recommended but it may suffer from a likelihood shortfall, which can't be overcome in general. When the model is misspecified for certain time points, SMC may fail, and PAL is possible to approximate the likelihood. When computational time is not a critical factor and the model is correctly specified, sequential Monte Carlo methods are recommended. 
\end{abstract}

\noindent\textbf{Keywords:} epidemiology; compartmental model; state-space model; sequential Monte Carlo; likelihood-based inference

\section{Introduction}

Understanding dynamic diseases is critical for public health and policy-making. Traditional time series models, such as the TSIR model \citep{TSIR1, TSIR_package}, are valuable for predicting and forecasting the transmission scale of specific infectious diseases, as well as for estimating critical parameters such as the basic reproduction number $R_0$ and the force of infection. However, these models still fail to provide a sufficiently clear and straightforward understanding of the disease's transmission dynamics. Therefore, modeling the epidemiology data differently is necessary. Here, we seek an alternative approach --- Partially Observed Markov Process (POMP) models with compartmental modeling in the latent process, which is one of the most widespread models for understanding the dynamics of infectious diseases and epidemiology data in the recent several decades \citep{mseh, HULIN2000197, daihai, plospomp, stocks, wastewater, daihai2,wwr}. However, when there is a desire for models to more closely resemble reality—namely, moving beyond linear Gaussian assumptions—traditional filtering methods such as the Kalman Filter \citep{kalman1960} and the Ensemble Kalman Filter \citep{ensembkal, ensemkalman2} become unsuitable. This complexity arises because many diseases are inherently more complicated; they are often modeled as nonlinear, non-Gaussian, high-dimensional, and highly stochastic POMP models with several compartments in the latent process.

POMP models are defined by their initial conditions, transition kernels, and observation densities. Though the observation density is always assumed to be mathematically evaluatable, the latent Markovian models have neither a closed-form initial density nor transition kernels. Therefore, the likelihood often is hard to retrieve when such a model is fairly complicated. However, as proposed by \cite{Kitagawa1987-fr, gordon1993, Kitagawa1996-wa}, Monte Carlo methods can be applied sequentially to approximate the posterior distribution, as well as the likelihood function. This is the so-called and widely-used sequential Monte Carlo method for filtering general models as long as one can simulate from the latent Markovian model, which is based on the idea of Sampling-Importance-Resampling \citep{Rubin1987}.

Owing to the unique characteristics of POMP models, which we will introduce subsequently, these models have gained significant importance and have been extensively applied in engineering and related disciplines, such as speech word identification \citep{rabiner1989}, target positioning and velocity estimation in radar systems \citep{gordon1993, avitzour1995}, volatility estimation in economic time series \citep{Pitt1999-volatility}, and structural type prediction in protein secondary structures \citep{Schmidler2000-protein}. In these models, the unobserved signal process $\{x_n: n \in \mathds{N}\}, x_n \in \mathcal{X}$, represents the true underlying signal, where $\mathcal{X}$ denotes the space where the state vectors live. A key example is the widely utilized SIR model for infectious disease modeling, which categorizes individuals into three compartments: S (susceptible), I (infectious), and R (recovered) \citep{daihai, seircov, SEIRmodel}. In this case, $\mathcal{X} = \mathds{N}^3$, where $\mathds{N}$ is the set of natural numbers. $x_n =  (S_n, I_n, R_n)$ where $S_n$, $I_n$, and $R_n$ indicate the number of individuals in each compartment at time $n$. However, what is observable is $\{y_n|n\in \mathds{N}\}, y_n \in \mathcal{Y}$, which depends only on $\{x_n: n\in\mathds{N}\}, x_n \in \mathcal{X}$ and always includes noise components. In the above SIR example, $\mathcal{Y} = \mathds{N}$ since $y_n$, our data, is the observed number of cases at time $n$. Therefore, accurately retrieving the values of $\{x_n: n\in\mathds{N}\}, x_n \in \mathcal{X}$ based on the observations $y_{1:N}$ is crucial, a challenge commonly referred to as a filtering problem. This entails filtering out the noise to recover the posterior distribution of $x_{1:n}$ given $y_{1:n}, n\in\mathds{N}$, a problem widely considered in Bayesian statistics. \citep{gordon1993, geweke1989, bayesiandoucet, pmcmc}

When inference and estimation of parameters are of interest, likelihood maximization methods are particularly challenging when the likelihood function is mathematically intractable, as detailed in \cite{Breto2019-sz, iteratedfiltering, daihai}. These methods often rely solely on the "plug-and-play" property, which depends only on simulations from the latent process model. Conversely, when the likelihood function is accessible, either through integration or approximation, methods such as the gradient ascent algorithm \citep{Lemarechal2012, Hadamard1908, Courant1943} and the Stochastic EM algorithm \citep{stochasticEM} can be effectively utilized.

Next, we introduce the POMP model in general and then review two different filtering methods that were proposed and compared in \cite{wwr}. Subsequently, we will discuss the model specifically used for the German rotavirus from 2001 to 2008 as a case study, which will enhance understanding of the two methods and, more importantly, provide guidance on whether and when one should use them.

\section{Methodology}

\begin{center}
\begin{tabular}{|c c|} 
 \hline
 Notation & Description  \\ [0.5ex] 
 \hline
 $\theta, \phi$ & Parameters, usually $\theta, \phi \in \mathds{R}^p, p\geq 1$\\ 
 \hline
 $f(\cdot; \theta),p(\cdot; \theta)$ & \makecell{Probability density functions parametrized by parameter $\theta$ of \\continuous and discrete random variables respectively}  \\
 \hline
 $X_{1:n}$ & State variable from time 1 to n.  \\
 \hline
 $Y_{1:n}$ & Observation variable from time 1 to n. \\
 \hline
 $x_{1:n}$ & The realization of the state variable from time 1 to n.  \\
 \hline
 $y_{1:n}$ & The realization of the observation variable from time 1 to n.  \\
 \hline
 $\mathcal{L}(\theta|\cdot)=f(\cdot|\theta)$ & The likelihood function at parameter $\theta$ given data.  \\
 \hline
 $\ell(\theta|\cdot)=logf(\cdot|\theta)$ & The log-likelihood function at parameter $\theta$ given data.  \\
 \hline
\end{tabular}
\end{center}

\subsection{The Partially Observed Markov Process (POMP) Model}
The Partially Observed Markov Process (POMP) model \citep{daihai}, also known as the state-space model (SSM), or the Hidden Markov Model (HMM), is a time series model consisting of processes \citep{Chopin, liujun, Doucet}: A latent process/ state model, written as $x_n \sim f_{X_n|X_{n-1}}(\cdot|x_{n-1},\theta)$ and an observation/ measurement model, written as $y_n \sim f_{Y_n|X_n}(\cdot|x_n,\theta)$. Here, we only consider the discrete-time representation. Where $\{x_n: n\in\mathds{N}\}, x_n \in \mathcal{X}$ denotes the hidden or latent states representing the unobserved signal. These states are treated as missing data within our model and are of interest, a point that will be elaborated upon subsequently. And $\{y_n|n\in \mathds{N}\}, y_n \in \mathcal{Y}$ are called observations and are our actual observed data. The latent process is modeled as a Markov process, i.e. $\{x_n: n\in\mathds{N}\}$ has the Markovian property: $f_{X_n|X_{1:n-1}}(x_n|x_{1:n-1}, \theta)=f_{X_n|X_{n-1}}(x_n|x_{n-1},\theta)$ with initial distribution $f_{X_0}(x_0;\theta)$ and the transition kernel $f_{X_n|X_{n-1}}(x_n|x_{n-1},\theta)$. $\{y_n|n\in \mathds{N}\}$ are conditionally independent given the state variables $\{x_n: n\in\mathds{N}\}$, and if $\{x_n: n\in\mathds{N}\}$ were given, $y_n$ only depends on $x_n$ in the way that $y_n \sim f_{Y_n|X_{n}}(y_n|x_n,\theta)$. And $f_{Y_n|X_{n}}(y_n|x_n,\theta)$ is called the observation/ measurement density. Such conditions are visualized by Figure \ref{fig1}.

Consequently, the likelihood function can be written as:
\vspace{-2mm}
\begin{align} 
\mathcal{L}(\theta|y_{1:N}) &= f_{Y_{1:N}}(y_{1:N}; \theta) \\
&= \int f_{Y_{1:N},X_{0:N}}(y_{1:N}, x_{0:N};\theta) dx_{0:N} \\
&= \int f_{X_0}(x_0;\theta) \prod_{n=1}^N
f_{Y_n|X_n}(y_n|x_n,\theta)
f_{X_n|X_{n-1}}(x_n|x_{n-1},\theta) dx_{0:N} 
\end{align}

\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth]{pic2.png}
\caption{\label{fig:hiddenmarkovmodel}Partially Observed Markov Process (POMP) Model \citep{gettingstartpomp}}
\label{fig1}
\end{figure}

However, $f_{X_0}(x_0;\theta)$, and $f_{X_n|X_{n-1}}(x_n|x_{n-1};\theta)$ doesn't always have a closed-form expression, especially for compartmental modeling for complex epidemiology data. Therefore, evaluating the likelihood by simply calculating the integral directly is nearly impossible. Here, for estimation and inference, we review two methods for calculating the likelihood.

\subsection{Sequential Monte Carlo/ Particle Filtering (SMC)}

The sequential Monte Carlo (SMC), also known as Particle Filtering, has been widely used with various variants developed over recent decades for posterior density approximation, including $f_{X_n|Y_{1:n-1}}(x_n|y_{1:n-1}, \theta)$ and $f_{X_n|Y_{1:n}}(x_n|y_{1:n},\theta)$, as well as for likelihood approximation. Theoretical foundations are discussed in \cite{Chopin, Doucet, liujun}, and R packages implementing SMC are detailed in \cite{pomppackagepaper}. Here, we introduce the basic concept of how SMC functions for likelihood approximation to serve our specific objectives.

Suppose we are interested in computing the likelihood for such POMP models. Recall that:
\begin{align*} 
\mathcal{L}(\theta|y_{1:N}) &= f_{Y_{1:N}}(y_{1:N}; \theta) \\
&=  \prod_{n=1}^N f_{Y_n|Y_{1:n-1}}(y_n|y_{1:n-1}, \theta) \\
&=  \prod_{n=1}^N \mathcal{L}_n(\theta) \\
\end{align*}
\vspace{-20mm}

where we let $f_{Y_n|Y_{1:n-1}}(y_n|y_{1:n-1}, \theta) = f_{Y_1}(y_1| \theta)$, when $n=1$, and $\mathcal{L}_n(\theta) := f_{Y_n|Y_{1:n-1}}(y_n|y_{1:n-1}, \theta)$.
Thus, we merely need to compute $\mathcal{L}_n(\theta)$. By the following formula:
\begin{align*}
    \mathcal{L}_n(\theta) &= f_{Y_n|Y_{1:n-1}}(y_n|y_{1:n-1}, \theta) \\
    &=  \int f_{Y_n,X_n|Y_{1:n-1}}(y_n, x_n|y_{1:n-1}, \theta)dx_n \\
    &=  \int f_{Y_n|X_{n},Y_{1:n-1}}(y_n| x_n,y_{1:n-1}, \theta)f_{X_n|Y_{1:n-1}}(x_n|y_{1:n-1}, \theta)dx_n \\
    & = \int f_{Y_n|X_{n}}(y_n|x_n, \theta)f_{X_n|Y_{1:n-1}}(x_n|y_{1:n-1}, \theta)dx_n \\
    &= \mathds{E}_{X_n|Y_{1:n-1}}[f_{Y_n|X_{n}}(y_n|X_n,\theta)|y_{1:n-1},\theta]
\end{align*}

by the conditional independency of $y_n$ given $x_n$. Then if we can sample $x^{(i)}_n \sim p(x_n|y_{1:n-1}, \theta)$, by Law of Large Numbers, we have 
\vspace{-1mm}
\begin{align*}
\frac{1}{J}\sum_{i=1}^{J}f_{Y_n|X_{n}}(y_n|x^{(i)}_n,\theta) \xrightarrow{\: J \to \infty \: } &\int f_{Y_n|X_{n}}(y_n|x_n, \theta)f_{X_n|Y_{1:n-1}}(x_n|y_{1:n-1}, \theta)dx_n \\
= &f_{Y_n|Y_{1:n-1}}(y_n|y_{1:n-1}, \theta) 
\end{align*}

Therefore, our objective shifts to sequentially sampling $x_n^{(i)}$ from $f_{X_n|Y_{1:n-1}}(x_n|y_{1:n-1}, \theta)$ for each $n \geq 1$. Sampling sequentially means that we don't want to sample a whole different set of $\{x_{n+1}^{(i)}\}_i$ whenever a new observation $y_{n+1}$ is available. This can be achieved by the sequential Monte Carlo technique as shown in Algorithm \ref{alg1}.

\vspace{3mm}

\begin{algorithm}
\caption{Sequential Monte Carlo (SMC, or particle filter)}
\label{alg1}
\begin{algorithmic}[1]
\Statex \textbf{Input: }Simulator for $f_{X_n|X_{n-1}}(x_n | x_{n-1}; \theta)$; evaluator for $f_{Y_n|X_n}(y_n | x_n; \theta)$; simulator for $f_{X_0}(x_0;\theta)$; parameter, $\theta$; data, $y_{1:N}$; number of particles, $J$.
\For{$n$ in $1:N$}
    \State Simulate for prediction: $X_{n,j}^P \sim f_{X_n|X_{n-1}}(\cdot | X_{n-1,j}^F;\theta)$ for $j$ in $1:J$.
    \State Evaluate weights: $w(n,j) = f_{Y_n|X_n}(y_n^*|X_{n,j}^P;\theta)$ for $j$ in $1:J$.
    \State Normalize weights: $\tilde{w}(n,j) = \frac{w(n,j)}{\sum_{m=1}^J w(n,m)}$.
    \State Apply select indices $k_{1:J}$ with $P[k_j = m] = \tilde{w}(n,m)$.
    \State Resample with replacement: set $X_{n,j}^F = X_{n,k_j}^P$ for $j$ in $1:J$.
    \State Compute conditional log likelihood: $\hat{l}_{n|1:n-1} = \log \left( J^{-1} \sum_{m=1}^J w(n,m) \right)$.
\EndFor
\end{algorithmic}
\textbf{output:} Log likelihood estimate, $\hat{l}(\theta) = \sum_{n=1}^N \hat{l}_{n|1:n-1}$; filter sample, $X_{n,1:J}^F$, for $n$ in $1:N$. \\
\textbf{complexity:} $O(J)$
\end{algorithm}

Moreover, one can show that $X_{n,j}^P, j=1,2...,J$ is an approximation of the prediction distribution $f_{X_{n}|Y_{1:n-1}}(x_{n}|y_{1:n-1})$. After reweighting, $X_{n,j}^F, j=1,2...,J$ is an approximation of the filtering distribution $f_{X_{n}|Y_{1:n}}(x_{n}|y_{1:n})$, see \cite{gordon1993, Smith1992-st}.

\vspace{3mm}
Another way of computing the likelihood is by noticing: 
\vspace{-3mm}
\begin{align}\label{eq4-6}
&\textit{Prediction Formula: }     &&f_{X_n|Y_{1:n-1}}(x_n|y_{1:n-1})=\int f_{X_n|X_{n-1}}(x_n|x_{n-1})f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y_{1:n-1})dx_{n-1}\\
&\textit{Filtering Formula: } &&f_{X_n|Y_{1:n}}(x_n|y_{1:n})=\frac{f_{Y_n|X_n}(y_n|x_n)f_{X_n|Y_{1:n-1}}(x_n|y_{1:n-1})}{f_{Y_n|Y_{1:n-1}}(y_n|y_{1:n-1})} \\
&\textit{Conditional Likelihood: } &&f_{Y_n|Y_{1:n-1}}(y_n|y_{1:n-1}) =\int f_{Y_n|X_n}(y_n|x_n)f_{X_n|Y_{1:n-1}}(x_n|y_{1:n-1})d{x_n}
\end{align}

When $f_{X_n|Y_{1:n}}(x_n \mid y_{1:n})$ and $f_{X_n|Y_{1:n-1}}(x_n \mid y_{1:n-1})$ are known functions of $n$, the likelihood can also be computed for all $n$, as these probabilities are inherently related to each other.

Since
\vspace{-3mm}
\begin{align*}
    f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y_{1:n-1}) \xrightarrow{\text{prediction}} f_{X_n|Y_{1:n-1}}(x_n|y_{1:n-1}) \xrightarrow{\text{update}} f_{X_n|Y_{1:n}}(x_{n}|y_{1:n}) 
\end{align*}

However, the evaluation of those integrals is almost impossible due to their complex and high-dimensional nature. Thus obtaining approximations for this distribution is more welcome \citep{Doucet}.

\subsection{Poisson Approximate Likelihood (PAL)}

Denote $p_{X_n|Y_{1:n-1}}(x_n|y_{1:n-1})$ as the prediction density; $p_{X_n|Y_{1:n}}(x_n|y_{1:n})$ as the filtering density. Suppose $\{X_n: n\in \mathds{N}\}$ and $\{Y_n: n\in \mathds{N}\}$ are discrete random variables, with $\mathcal{X}=\mathds{N}^m$,$\mathcal{Y}=\mathds{N}$, and $m \in \mathds{N}$ is the number of compartments, by very minimum algebra, we have:
\vspace{-3mm}
\begin{align*}
&\textit{Prediction Formula: }     &&p_{X_n|Y_{1:n-1}}(x_n|y_{1:n-1})=\sum_{x_{n-1}\in \mathds{N}_0^m}p_{X_n|X_{n-1}}(x_n|x_{n-1})p_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y_{1:n-1}) \\
&\textit{Filtering Formula: } &&p_{X_n|Y_{1:n}}(x_n|y_{1:n})=\frac{p_{Y_n|X_n}(y_n|x_n)p_{X_n|Y_{1:n-1}}(x_n|y_{1:n-1})}{p_{Y_n|Y_{1:n-1}}(y_n|y_{1:n-1})} \\
&\textit{Conditional Likelihood: } &&p_{Y_n|Y_{1:n-1}}(y_n|y_{1:n-1}) =\sum_{x_n\in \mathds{N}_0^m}f_{Y_n|X_n}(y_n|x_n)p_{X_n|Y_{1:n-1}}(x_n|y_{1:n-1})
\end{align*}
which is a discrete representation of Eq. (\ref{eq4-6})-(6).

\vspace{4mm}
Given $p_{X_0}(x_0), p_{X_n|X_{n-1}}(x_n|x_{n-1})$, and $p_{Y_n|X_{n}}(y_n|x_{n})$ for $n\geq1$, we can approximate $p_{X_n|Y_{1:n-1}}(x_n|y_{1:n-1})$, $p_{X_n|Y_{1:n}}(x_n|y_{1:n})$, and $p_{Y_n|Y_{1:n-1}}(y_n|y_{1:n-1})$ at any time $n\geq1$ and then use the above quantities to update $p_{X_n|X_{n-1}}(x_n|x_{n-1}), p_{X_n|Y_{1:n}}(x_n|y_{1:n})$, and $p_{X_n|Y_{1:n}}(x_n|y_{1:n})$ at the next given time $n+1$. The Poisson Approximate Likelihood (PAL) method assumes and approximates them by a vector Poisson approximation.

\newpage
\begin{enumerate}
    \item 
In the latent process model: 
\vspace{-2mm}
\begin{enumerate}
    \item $\bm{\delta_t}$ be the death rate vector;
    \item $\bm{\alpha_t}$ be the birth rate vector of m compartments;
    \item $\bm{K_{t, \eta}}$ be the row stochastic matrices;
\end{enumerate}
\vspace{-2mm}
Then the one-step transition of the latent process can be described as:

$\bm{x_t} = \bm{\tilde{x}_t} + \bm{\hat{x}_t}$, where $\bm{\tilde{x}}$ is obtained by $\bm{\bar{x}_{t-1}} \sim Binom(\bm{x}_{t-1}, \bm{1- \delta_t})$ and $\bm{\tilde{x}_t}$ is equal to the one-step transition of $\bm{\bar{x}_{t-1}}$ according to the transition kernel $\bm{K_{t-1, \eta}}$. And $\bm{\hat{x}_t} \sim Pois(\bm{\alpha}_t)$ is the importing birth, which is considered as stochastic noise. 
\vspace{-3mm}
\item In the observation model:
\vspace{-2mm}
    \begin{enumerate}
        \item $\bm{q_t}\in [0,1]^m$ be the observing probability, where $m$ is the number of compartments;
        \item $\bm{G_t}$ be the misreporting kernel, where non-zero off-diagonal entries indicate misreporting. 
        \item $\bm{\kappa_t}$ be the observation error rate.
    \end{enumerate}
\vspace{-2mm}
    Then $\bm{y_t} = \bm{\tilde{y}}_t + \bm{\hat{y}}_t$, where $\bm{\hat{y}}_t \sim Binom(\bm{x_t}, \bm{q_t})$. And $\bm{\hat{y}}_t \sim Pois(\bm{\kappa_t})$ is the observation error.

\end{enumerate}

A novel method suggested by \cite{wwr} is that the conditional likelihood of the above compartmental model can be approximated by Poisson densities by the following:

If we assume $p_{X_0}(\bm{x}_0) \approx \operatorname{Pois}(\bm{\lambda}_0)$, where $\bm{\lambda}_0$ is a parameter that characterizes the initial distribution of the numbers of individuals in each compartment.

Then for $t \geq 1$:

\begin{enumerate}
    \item $p(\bm{x}_t|\bm{y}_{1:t-1}) \approx \operatorname{Pois}(\bm{\lambda}_t)$, where $\bm{\lambda}_t$ :=$(\bm{\bar{\lambda}}_{t-1} \odot \bm{\delta_t})^\intercal \bm{K_{t,\eta(\bm{\bar{\lambda}}_{t-1} \odot \bm{\delta_t})}} + \bm{\alpha}_t$     
    \item

    $p\left(\mathbf{x}_t \mid \mathbf{y}_{1: t}\right) \approx \operatorname{Pois}\left(\bm{\bar{\lambda}}_t\right), \quad \bm{\bar{\lambda}}_t:=\left[\mathbf{1}_m-\mathbf{q}_t+\left(\left\{\mathbf{y}_t^{\top} \oslash\left[\left(\mathbf{q}_t \odot \bm{\lambda}_t\right)^{\top} \mathbf{G}_t+\boldsymbol{\kappa}_t^{\top}\right]\right\}\left[\left(\mathbf{1}_m \otimes \mathbf{q}_t\right) \odot \mathbf{G}_t^{\top}\right]\right)^{\top}\right] \odot \bm{\lambda}_t$

    \item $p(\mathbf{y}_t|\bm{y}_{1:t-1}) \approx \operatorname{Pois}\left(\bm{\mu}_t\right)$, where $\bm{\mu}_t = \left[(\bm{\lambda}_t \odot \mathbf{q}_t)^{\top} \mathbf{G}_t\right]^{\top}+\boldsymbol{\kappa}_t$.
\end{enumerate}

Therefore, the algorithm of the approximated likelihood is:
\vspace{2mm}
\begin{algorithm}
\caption{Poisson Approximate Likelihood (PAL)}
\begin{algorithmic}[1]
\Statex \textbf{Initialize: }$\bar{\lambda}_0 \leftarrow \lambda_0$.
\For{$t$ in $1:N$}
    \item \quad $\lambda_t \leftarrow\left[\left(\bar{\lambda}_{t-1} \odot \delta_t\right)^{\top} \mathbf{K}_{t, \boldsymbol{\eta}\left(\bar{\lambda}_{t-1} \odot \delta_t\right)}\right]^{\top}+\boldsymbol{\alpha}_t$
    \item \quad $\bar{\lambda}_t \leftarrow\left[\mathbf{1}_m-\mathbf{q}_t+\left(\left\{\mathbf{y}_t^{\top} \oslash\left[\left(\mathbf{q}_t \odot \lambda_t\right)^{\top} \mathbf{G}_t+\boldsymbol{\kappa}_t^{\top}\right]\right\}\left[\left(\mathbf{1}_m \otimes \mathbf{q}_t\right) \odot \mathbf{G}_t^{\top}\right]\right)^{\top}\right] \odot \lambda_t$
    \item \quad $\boldsymbol{\mu}_t \leftarrow\left[\left(\lambda_t \odot \mathrm{q}_t\right)^{\top} \mathbf{G}_t\right]^{\top}+\boldsymbol{\kappa}_t$
    \item \quad $\ell\left(\mathbf{y}_t \mid \mathbf{y}_{1: t-1}\right) \leftarrow-\boldsymbol{\mu}_t^{\top} \mathbf{1}_m+\mathbf{y}_t^{\top} \log \left(\boldsymbol{\mu}_t\right)-\mathbf{1}_m^{\top} \log \left(\mathbf{y}_{t} !\right)$
\EndFor
\end{algorithmic}
\end{algorithm}

\newpage

To account for overdispersion within the model, \cite{wwr}'s approach entails the introduction of a prior distribution over the parameters. If we separate parameters into two parts i.e. $\theta=[\vartheta, \bar{\theta}_{1:t}]$ where $\vartheta$ are either fixed or to be estimated and $\bar{\theta}_{1:t}$ are to be integrated out and $\bar{\theta}_{1:t} \sim f(\cdot|\phi)$, the prior distribution for $\bar{\theta}_{1:t}$, where $\phi$ are hyperparameters. These parameters $\bar{\theta}_{1:t}$ are then integrated out by importance sampling techniques with proposal distribution $\pi(\cdot|\cdot)$, yielding a marginal distribution of the data that reflects the overdispersion. The procedure is delineated in the following algorithm:

\vspace{3mm}

\begin{algorithm}[H]
\caption{PALSMC}
\begin{algorithmic}[1]
\Statex \textbf{Input: } Proposal distribution $\pi(\cdot|\cdot)$, number of particles J, parameter $[\vartheta, \phi]$
\Statex \textbf{Initialize: }$\bar{\lambda}^{(i)}_0 \leftarrow \lambda_0$, for $i =1,2,...,J$.
\For{$t$ in $1:N$}
    \For{$i= 1,2,...,J$}
    \State $\bar{\theta}^{(i)}_{1:t} \sim \pi(\cdot|\bar{\theta}^{(i)}_{1:t-1}, \bar{\lambda^{(i)}_{t-1}},y_{1:t})$.
    \State $\boldsymbol{\alpha}_t \leftarrow \boldsymbol{\alpha}_t(\vartheta,\bar{\theta}^{(i)}_{t})$, 
    $\mathbf{K}_{t, \boldsymbol{\eta}} \leftarrow \mathbf{K}_{t, \boldsymbol{\eta}}(\vartheta,\bar{\theta}^{(i)}_{t})$, 
    $\boldsymbol{\delta}_t \leftarrow \boldsymbol{\delta}_t(\vartheta,\bar{\theta}^{(i)}_{t})$, 
    $\boldsymbol{q}_t \leftarrow \boldsymbol{q}_t(\vartheta,\bar{\theta}^{(i)}_{t})$,
    \Statex
    \quad \quad \quad $\boldsymbol{\kappa}_t \leftarrow \boldsymbol{\kappa}_t(\vartheta,\bar{\theta}^{(i)}_{t})$,
    $\boldsymbol{G}_t \leftarrow \boldsymbol{G}_t(\vartheta,\bar{\theta}^{(i)}_{t})$
    \State $\lambda_t \leftarrow\left[\left(\bar{\lambda}_{t-1} \odot \delta_t\right)^{\top} \mathbf{K}_{t, \boldsymbol{\eta}\left(\bar{\lambda}_{t-1} \odot \delta_t\right)}\right]^{\top}+\boldsymbol{\alpha}_t$
    \State $\bar{\lambda}_t \leftarrow\left[\mathbf{1}_m-\mathbf{q}_t+\left(\left\{\mathbf{y}_t^{\top} \oslash\left[\left(\mathbf{q}_t \odot \lambda_t\right)^{\top} \mathbf{G}_t+\boldsymbol{\kappa}_t^{\top}\right]\right\}\left[\left(\mathbf{1}_m \otimes \mathbf{q}_t\right) \odot \mathbf{G}_t^{\top}\right]\right)^{\top}\right] \odot \lambda_t$
    \State $\boldsymbol{\mu}_t \leftarrow\left[\left(\lambda_t \odot \mathrm{q}_t\right)^{\top} \mathbf{G}_t\right]^{\top}+\boldsymbol{\kappa}_t$
    \State $\ell\left(\mathbf{y}_t \mid \mathbf{y}_{1: t-1}, \vartheta, \bar{\theta}^{(i)}_{1:t}\right) \leftarrow-\boldsymbol{\mu}_t^{\top} \mathbf{1}_m+\mathbf{y}_t^{\top} \log \left(\boldsymbol{\mu}_t\right)-\mathbf{1}_m^{\top} \log \left(\mathbf{y}_{t} !\right)$
    \State $\log{w^{(i)}_t} \leftarrow \ell\left(\mathbf{y}_t \mid \mathbf{y}_{1: t-1}, \vartheta, \bar{\theta}^{(i)}_{1:t}\right) + \log{f(\bar{\theta}^{(i)}_{t}|\bar{\theta}^{(i)}_{1:t-1}, \phi)}-\log{\pi(\bar{\theta}^{(i)}_{t}|\bar{\theta}^{(i)}_{1:t-1},\bar{\lambda}^{(i)}_t)}$
    \EndFor
    \State $\hat{\ell}\left(\mathbf{y}_t \mid \mathbf{y}_{1: t-1}, \vartheta, \phi\right) \leftarrow \log\left({\frac{1}{J}\sum_{i=1}^{J}w_{t}^{(i)}}\right)$
    \State $\bar{w}_{t}^{(i)}\leftarrow \frac{w_{t}^{(i)}}{\sum_{j=1}^{J}w_{t}^{(j)}}$
    \State Resample $\{\bar{\theta}^{(i)}_{1:t}, \bar{\lambda}_{t}^{(i)}\}$ according to weights $\{\bar{w}_{t}^{(i)}\}_{i=1}^{J}$
\EndFor
\end{algorithmic}
\end{algorithm}

\section{Model: A Case Study with Rotavirus in Germany: 2001-2008}

\begin{table}[ht]
\centering
\caption{Comparison of Model \& Method Performance \citep{wwr}}
\label{tab:model_performance}
% Increase row height
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{@{}lcc@{}}
\toprule
\textbf{Model} & \textbf{AIC} & \textbf{Ave. comp. time} \\
\midrule
EqEq & 98866.65 & 30 s \\
EqOv & 15154.75 & 2 hr \\
OvOv & 13778.08 & 3 hr \\
\cite{stocks} & 20134.38 & 11 hr \\
\bottomrule
\end{tabular}
\end{table}

Model selection is crucial in epidemiology modeling, as it determines the appropriateness of a model in capturing the complexities of disease transmission dynamics. The Akaike Information Criterion (AIC) represents a mathematical approach used in the selection of models and the evaluation of their structural parsimony. The formula for this criterion is given by.
$$
AIC = 2p - \ell(\hat{\theta})
$$
where $p$ is the number of parameters in the model and $\ell(\hat{\theta})$ is the maximized log-likelihood at the estimates.

When faced with multiple potential models for a dataset, the optimal choice is the model that yields the lowest AIC value. The AIC not only acknowledges the model's fit, evaluated by the likelihood function but also imposes a penalty that escalates with the number of parameters estimated within the model. This penalization is intended to deter overfitting, which is advantageous because augmenting the parameter count typically enhances the model's fit regardless of its true explanatory power.

Next, we introduce the POMP models used in the analysis, beginning with the latent process model, which follows a Markovian structure and includes nine compartments within $\mathcal{X}$.

\subsection{Latent Process Compartmental Model: The SIRSIRSIR model}

\begin{figure}[h]
\centering
\includegraphics[width=0.85\textwidth]{wwrlatentscheme.png}
\caption{\label{fig:SIRSIRSIR model}SIRSIRSIR model \citep{wwr}}
\end{figure}

Rotavirus disease primarily affects infants and young children, causing gastroenteritis in this age group, and is relatively rare in adults \citep{Lambert2009-it}. Nearly every child contracts rotavirus at least once by the age of five \citep{CDC}. Reinfection can occur since neither natural infection nor vaccination against rotavirus provides complete immunity against future infections \citep{stocks}. Consequently, it is practical to categorize the population into three age groups, leading to a model with nine compartments in total: $S_k$, susceptible; $I_k$, infected; $R_k$, recovered, where $k=1,2,3$ corresponds to the age classes 0-4, 5-59, and 60-99, respectively. The diagram of the latent model is shown in Figure \ref{fig:SIRSIRSIR model}, which is employed by both \cite{wwr} and \cite{stocks}. The model is explained in detail subsequently.

The evolution of the full age-stratified rotavirus latent process model at time $t$ is given by:
\begin{align*}
    S_{1,t+1} &= S_{1,t-1} + A_{1,t} + E_{1,t} - B_{1,t} - F^{(S)}_{1,t}, \\
    I_{1,t+1} &= I_{1,t} + B_{1,t} - C_{1,t} - F^{(I)}_{1,t}, \\
    R_{1,t+1} &= R_{1,t} + C_{1,t} - E_{1,t} - F^{(R)}_{1,t}, \\
    S_{2,t+1} &= S_{2,t} + F^{(S)}_{1,t} + E_{2,t} - B_{2,t} - F^{(S)}_{2,t}, \\
    I_{2,t+1} &= I_{2,t} + F^{(I)}_{1,t} + B_{2,t} - C_{2,t} - F^{(I)}_{2,t}, \\
    R_{2,t+1} &= R_{2,t} + F^{(R)}_{1,t} + C_{2,t} - E_{2,t} - F^{(R)}_{2,t}, \\
    S_{3,t+1} &= S_{3,t} + F^{(S)}_{2,t} + E_{3,t} - B_{3,t} - D^{(S)}_{t}, \\
    I_{3,t+1} &= I_{3,t} + F^{(I)}_{2,t} + B_{3,t} - C_{3,t} - D^{(I)}_{t}, \\
    R_{3,t+1} &= R_{3,t} + F^{(R)}_{2,t} + C_{3,t} - E_{3,t} - D^{(R)}_{t},
\end{align*}

where at time $t$: $A_{1,t} \sim \text{Pois}(\alpha_t)$, for some $\alpha_t \in \mathbb{R}$ represents new births, which is chosen according to historical birth record data; $B_{\cdot,t}$ represents new infectives; $C_{\cdot,t}$ represents recovering individuals; $D_{\cdot,t} \sim \text{Binom}(t-1, 1-\delta)$ represents emigrating (dying) individuals; $E_{\cdot,t}$ represents individuals experiencing waning immunity; and $F_{\cdot,t}$ represents aging individuals.

\[
\left[ \begin{array}{c}
    B_{1,t} \\
    F^{(S)}_{1,t} \\
    S_{1,t} - B_{1,t} - F^{(S)}_{1,t}
\end{array} \right]
\sim \text{Mult} 
\left( \begin{array}{c}
    S_{1,t} 
\end{array}, \left[ \begin{array}{c}
    p_{1,t} \\
    1 - e^{-hd1} \\
    e^{-hd1} - p_{1,t}
\end{array} \right] \right)
\]

\[
\left[ \begin{array}{c}
    C_{1,t} \\
    F^{(I)}_{1,t} \\
    I_{1,t} - C_{1,t} - F^{(I)}_{1,t}
\end{array} \right]
\sim \text{Mult} 
\left( \begin{array}{c}
    I_{1,t}
\end{array}, \left[ \begin{array}{c}
    1 - e^{-h\gamma} \\
    1 - e^{-hd1} \\
    e^{-h\gamma} + e^{-hd1} - 1
\end{array} \right] \right)
\]

\[
\left[ \begin{array}{c}
    E_{1,t} \\
    F^{(R)}_{1,t} \\
    R_{1,t} - E_{1,t} - F^{(R)}_{1,t}
\end{array} \right]
\sim \text{Mult} 
\left( \begin{array}{c}
    R_{1,t}
\end{array}, \left[ \begin{array}{c}
    1 - e^{-h\omega} \\
    1 - e^{-hd1} \\
    e^{-h\omega} + e^{-hd1} - 1
\end{array} \right] \right)
\]

\[
\left[ \begin{array}{c}
    B_{2,t} \\
    F^{(S)}_{2,t} \\
    S_{2,t} - B_{2,t} - F^{(S)}_{2,t}
\end{array} \right]
\sim \text{Mult} 
\left( \begin{array}{c}
    S_{2,t} 
\end{array}, \left[ \begin{array}{c}
    p_{2,t} \\
    1 - e^{-hd2} \\
    e^{-hd2} - p_{2,t}
\end{array} \right] \right)
\]

\[
\left[ \begin{array}{c}
    C_{2,t} \\
    F^{(I)}_{2,t} \\
    I_{2,t} - C_{2,t} - F^{(I)}_{2,t}
\end{array} \right]
\sim \text{Mult} 
\left( \begin{array}{c}
    I_{2,t}
\end{array}, \left[ \begin{array}{c}
    1 - e^{-h\gamma} \\
    1 - e^{-hd2} \\
    e^{-h\gamma} + e^{-hd2} - 1
\end{array} \right] \right)
\]

\[
\left[ \begin{array}{c}
    E_{2,t} \\
    F^{(R)}_{2,t} \\
    R_{2,t} - E_{2,t} - F^{(R)}_{2,t}
\end{array} \right]
\sim \text{Mult} 
\left( \begin{array}{c}
    R_{2,t}
\end{array}, \left[ \begin{array}{c}
    1 - e^{-h\omega} \\
    1 - e^{-hd2} \\
    e^{-h\omega} + e^{-hd2} - 1
\end{array} \right] \right)
\]

\[
\left[ \begin{array}{c}
    B_{3,t} \\
    S_{3,t} - D_{t}^{(S)} - B_{3,t}
\end{array} \right]
\sim \text{Mult} 
\left( \begin{array}{c}
    S_{3,t} - D_{t}^{(S)} 
\end{array}, \left[ \begin{array}{c}
    p_{3,t} \\
    1 - p_{3,t}
\end{array} \right] \right)
\]

\[
\left[ \begin{array}{c}
    C_{3,t} \\
    I_{2,t} - D_{t}^{(I)} - C_{2,t}
\end{array} \right]
\sim \text{Mult} 
\left( \begin{array}{c}
    I_{3,t} - D_{t}^{(I)} 
\end{array}, \left[ \begin{array}{c}
    1 - e^{-h\gamma} \\
    e^{-h\gamma}
\end{array} \right] \right)
\]

\[
\left[ \begin{array}{c}
    E_{3,t} \\
    R_{2,t} - D_{t}^{(R)}- E_{2,t}
\end{array} \right]
\sim \text{Mult} 
\left( \begin{array}{c}
    R_{3,t} - D_{t}^{(R)}
\end{array}, \left[ \begin{array}{c}
    1 - e^{-h\omega} \\
    e^{-h\omega}
\end{array} \right] \right)
\]

\[
\begin{aligned}
    \mathbf{K}_{t, \bm{\eta}} &= \scriptsize{
    \begin{bmatrix}
    -e^{-hd1} & p_{1,t} & 0 & 1 - e^{-hd1} & 0 & 0 & 0 & 0 & 0 \\
    0 & -e^{-h\gamma} + e^{-hd1} - 1 & 1 - e^{-h\gamma} & 1 - e^{-hd1} & 0 & 0 & 0 & 0 & 0 \\
    0 & 0 & 1 - e^{-h\omega} & 0 & e^{-h\omega} + e^{-hd1} - 1 & 1 - e^{-hd1} & 0 & 0 & 0 \\
    0 & 0 & 0 & -e^{-hd2} & p_{2,t} & 0 & 1 - e^{-hd2} & 0 & 0 \\
    0 & 0 & 0 & 0 & -e^{-h\gamma} + e^{-hd2} - 1 & 1 - e^{-h\gamma} & 1 - e^{-hd2} & 0 & 0 \\
    0 & 0 & 0 & 0 & 0 & 1 - e^{-h\omega} & 0 & e^{-h\omega} + e^{-hd2} - 1 & 1 - e^{-hd2} \\
    0 & 0 & 0 & 0 & 0 & 0 & -e^{-hd3} & p_{3,t} & 0 \\
    0 & 0 & 0 & 0 & 0 & 0 & 0 & -e^{-h\gamma} + e^{-hd3} - 1 & 1 - e^{-h\gamma} \\
    0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 - e^{-h\omega} \\
    \end{bmatrix}
    }
\end{aligned}
\]

where Mult denotes multinomial distribution. For models EqEq and EqOv we have \( p_{k,t} = 1 - \exp \left\{-\beta_k \frac{I_t}{n} \chi_t \right\} \) for \( k = 1, 2, 3 \), and for model OvOv we have \( p_{k,t} = 1 - \exp \left\{-\beta_k \frac{\tilde{I}_t}{n} \chi_t \xi_t \right\} \) for \( k = 1, 2, 3 \), \( \chi_t = 1+\rho \cos({2\pi t/p + \phi}) \), where $p=52$ is the fixed period corresponding to the annual outbreak of rotavirus.

\subsection{Observation Model}\label{observationmodel}

The observation model is defined as follows:
\vspace{-10mm}

\begin{align*}
    [Y_{kt} \mid X_{kt} = (S_{kt}, I_{kt}, R_{kt})] &= [Y_{kt} \mid I_{kt}], \quad k=1,2,3 \quad t=1,2,...,416 \\
    & \sim p(y_{kt} \mid x_{kt})
\end{align*}
\vspace{-10mm}

This model covers three age groups, \(k=1,2,3\), and specifies that the observable variable \(Y_{kt}\) is determined solely by \(I_{kt}\), which represents the number of infected individuals at time \(t\). The time variable \(t\) spans from week 0 to week 416, corresponding to an eight-year period from 2001 to 2008. This assumption is practical because most reported cases are confirmed either by laboratory tests or by clinical diagnosis in hospitals of infected individuals. The observed counts are modeled using either a binomial distribution, as assumed by \cite{wwr}, or a Poisson or Negative Binomial distribution, as assumed by \cite{stocks}. Both approaches use the conditional mean equal to the product of the number of infectious individuals and the reporting rate $q_t$. This model reflects the reality that researchers can only collect data on reported cases, which represent only a subset of actual infections due to underreporting.

For the observation models of \cite{stocks} and \cite{wwr} are defined as follows:

\begin{enumerate}
    \item \citep{stocks}:

    $$
    Y_{kt} \sim \text{Pois}(q_t \cdot H_{kt}) \;\;\;\; \text{or} \;\;\;\; Y_{kt} \sim \text{NegBinom}(q_t \cdot H_{kt}, \frac{1}{\theta})
    $$
    In the model proposed by \cite{stocks}, the number of observed cases \( Y_{kt} \) is modeled to follow either a Poisson distribution or a Negative Binomial distribution, to account for potential overdispersion in the observation model. The expected value of these distributions is the product of the reporting rate \( q_t \) and the number of infected individuals \( H_{kt} \).\( H_{kt} \) is modeled as an accumulated variable, in the sense that it's the total number of transition from $I_{k}$ to $R_{k}$ within a week, corresponding to the weekly data we use. The Negative Binomial distribution includes a dispersion parameter \( \theta \) to manage overdispersion relative to the Poisson model.

    However, \cite{stocks} assumed a non-underreporting observation model, implying 
    $q_t = 1$, which means the data they analyzed were effectively scaled up. This assumption led to an increase of 3200 log-units in the log-likelihood, as reported by \cite{wwr} and illustrated in Table \ref{tab:model_performance}.

    \item \citep{wwr}:

    $$
    Y_{kt} \sim \text{Binom}(H_{kt}, q_t)
    $$
    In \cite{wwr}'s model, \( Y_{kt} \) follows a Binomial distribution, where \( H_{kt} \) represents the trials and \( q_t \) the success probability in each trial, reflecting the proportion of infections that are actually detected and reported.
\end{enumerate}

Focusing on \cite{wwr}'s modeling, we consider three models:

\begin{enumerate}
    \item EqEq: a fully equi-dispersed model, in which \( q_{k,r} = 0.07 = \mu_q \in [0, 1] \), and \( \mu_q \) is assumed to be known;

    \item EqOv: an equi-dispersed latent compartmental model and an over-dispersed observation model, the same as EqEq except that \( q_{k,r} \overset{\text{iid}}{\sim} \mathcal{N}(\mu_{q}, \sigma^{2}_{q})_{\geq 0,\leq 1} \), where \( \sigma^{2}_{q} > 0 \) is to be estimated;

    \item OvOv: over-dispersion in both the latent and observation models, the same as EqOv except that we augment \( \chi_t \) to \( \chi_t \xi_{t,r} \), where for \( r \geq 1 \), \( \xi_{t,r} \overset{\text{iid}}{\sim} \text{Gamma}(\sigma^{2}_{\xi}, \sigma^{2}_{\xi}) \) are multiplicative disturbances with mean 1 and \( \sigma^{2}_{\xi} > 0 \) is to be estimated.

\end{enumerate}

The table of parameters can be found below. Note that some of them are assumed fixed while the rest are to be estimated.

\begin{table}[H]
\centering
\caption{Full-list of parameters}
\label{tab:para_table}
\begin{tabularx}{\textwidth}{@{}Xlll@{}} % Use X for the column that you want to wrap text
\toprule
Parameter & Explanation & Value & Unit \\
\midrule
$N$ & Population size & 82372825 & individuals \\
$\alpha$ & Individual birth rate & $1 /(52 \cdot 78.9)$ & week$^{-1}$ \\
$d_1$ & Aging from age group 1 (age 0-4) to 2 (age 5-59) & $1 /(52 \cdot 5)$ & week$^{-1}$ \\
$d_2$ & Aging from age group 2 (age 5-59) to 3 (age $60+$) & $1 /(52 \cdot 55)$ & week$^{-1}$ \\
$\delta$ & Death rate (from age group 3 (age $60+$)) & $1 /(52 \cdot 18.9)$ & week$^{-1}$ \\
$\beta_k$ & Susceptibility of age group $k \in \{1,2,3\}$ & To be estimated & individuals/week \\
$\omega$ & Immunity waning rate & $1 /(1 \cdot 52)$ & week$^{-1}$ \\
$\gamma$ & Recovery rate & 1 & week$^{-1}$ \\
$\phi$ & Phase shift of the seasonal forcing & To be estimated & 1 \\
$\rho$ & Amplitude of the seasonal forcing & To be estimated & 1 \\
$p_{k,t}$ & Force of infection of age group $k \in \{1,2,3\}$ & Defined above & week$^{-1}$ \\
$\chi(t)$ & Seasonal forcing function & Defined above & 1 \\
$\sigma_{q}^2$ & Overdispersion parameter of reporting rates & To be estimated & 1 \\
$\sigma_{\xi}$ & Scale parameter of the $\Gamma$-noise & To be estimated & 1 \\
\bottomrule
\end{tabularx}
\end{table}

\section{Results}

\cite{wwr} claimed that PAL is the first likelihood-based filtering method that can be applied to a broad range of POMP models with the finite-population assumption and consistent results, and requires much less computational time than SMC. To evaluate whether SMC performs as well as PAL in terms of both computational time and likelihood identifiability, we replicated the model using the R package \textbf{pomp} \citep{pomppackagepaper}. First, we verified that the two models are essentially identical. Here, we simulated data from the model 1000 times at the MLE shown in Table \ref{tab:mlebywwr} and compared the summary statistics of 1000 simulated time series from two models using the t-test. Summary statistics including the mean, median, and variance for all three age groups were tested. The results showed no significant differences between the two models from which the data were simulated, confirming that both the latent process model and the observation model are essentially consistent with each other, given that the observations were directly simulated. These results are presented in Appendix \ref{appendA}.

\begin{table}[htbp]
  \centering
  \begin{threeparttable}
    \caption{Maximum likelihood estimation of parameters by PAL for three models \citep{wwr}, and by mif2 algorithm \citep{pomppackagepaper} for the OvOv model.}
    \label{tab:mlebywwr}
    \begin{tabular}{>{\raggedright\arraybackslash}p{2cm} *{3}{>{\centering\arraybackslash}p{1.5cm}} >{\centering\arraybackslash}p{2.5cm}}
      \toprule
      Parameter & EqEq & EqOv & OvOv & OvOv (\textbf{pomp})$^*$ \\
      \midrule
      \(\beta_1\) & 12.15 & 12.74 & 11.48 & 10.30\\
      \(\beta_2\) & 0.22 & 0.21 & 0.25 & 0.21\\
      \(\beta_3\) & 0.34 & 0.31 & 0.35 & 0.86\\
      \(\phi\) & 0.017 & 0.14 & 0.14 & 0.03\\
      \(\rho\) & 0.022 & 0.19 & 0.16 & 0.14\\
      \(\sigma^2_q\) & n/a & 0.042 & 0.021 & 0.018\\
      \(\sigma_\xi\) & n/a & n/a & 66.89 & 14.85\\
      \addlinespace
      \multicolumn{1}{l}{\(S_{10}\)} & \multicolumn{3}{c}{$3876549^*$} & 5885201\\
      \multicolumn{1}{l}{\(I_{10}\)} & \multicolumn{3}{c}{$30351$} & 5041\\
      \multicolumn{1}{l}{\(R_{10}\)} & \multicolumn{3}{c}{$1315221$} & 721006\\
      \multicolumn{1}{l}{\(S_{20}\)} & \multicolumn{3}{c}{$57139612$} & 44546887\\
      \multicolumn{1}{l}{\(I_{20}\)} & \multicolumn{3}{c}{$871$} & 22\\
      \multicolumn{1}{l}{\(R_{20}\)} & \multicolumn{3}{c}{$302852$} & 26360811\\
      \multicolumn{1}{l}{\(S_{30}\)} & \multicolumn{3}{c}{$19573727$} & 4788420\\
      \multicolumn{1}{l}{\(I_{30}\)} & \multicolumn{3}{c}{$2550$} & 131\\
      \multicolumn{1}{l}{\(R_{30}\)} & \multicolumn{3}{c}{$131092$} & 65306\\
      \midrule
      \addlinespace
      \textbf{AIC}  & 98866.65 & 15154.75 &13778.08 & \textbf{13674.55}\\
      \addlinespace
      \bottomrule
    \end{tabular}
    \begin{tablenotes}
      \small
      \item[*] The initial distribution parameter \(\lambda_0 = (S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}, R_{30})\) are assumed to be \textbf{fixed} in \cite{wwr}. The results of the estimates of the OvOv (\textbf{pomp}) model are obtained by using the Iterated Filtering algorithm with 3 rounds and 100 iterations, 50,000 particles, and 36 replicates in each round with the top 12 best fits in terms of likelihood chosen to be the starting value for the next round. 
    \end{tablenotes}
  \end{threeparttable}
\end{table}

\subsection{Key Factors Contributing to the Change of AIC in Table \ref{tab:model_performance}}\label{section1}

From Table \ref{tab:model_performance}, it is evident that the OvOv model yields the lowest AIC, making it particularly noteworthy. Since one of \cite{stocks}'s main purposes was to do model selection, therefore it's exciting to see \cite{wwr} obtaining a much better model. The overdispersion model (OvOv) is often the best model in terms of likelihood, primarily because it allows for greater variation in the model. By incorporating more parameters, the OvOv model can capture subtleties and complexities in the dataset that simpler models may overlook, leading to a more accurate fit and higher likelihood values. \cite{wwr} claimed that it is the innovative observation model that contributed to an increase in log-likelihood by 3200 log-likelihood units. However, after comparing the results, we found that the increase in likelihood when implementing the OvOv model is attributable to the datasets used by \cite{wwr} and \cite{stocks}, where \cite{stocks}'s dataset has no underreporting. Therefore \cite{wwr}'s dataset can be viewed as a scaled-down version of \cite{stocks}'s dataset by the reporting rate, which is roughly 0.07. Within the framework of the same model, scaling the data downwards by a factor results in a corresponding rise in likelihood. This scaling effect accounts for the discrepancy observed in the increase of 3200 log-likelihood units.
\vspace{-2mm}
\subsection{PAL versus SMC: When the Model is Misspecified} \label{misspecified}

Our results demonstrate how SMC is sensitive to model misspecification, which consequently leads to a low likelihood estimate at the MLE, as shown in Table \ref{tab:ovovrealdata} by fitting the real German rotavirus dataset into the model.

\begin{table}[ht] % The placement specifier can be [h!tbp]
\centering % Centers the table
\caption{log-likelihood at the MLE of OvOv model for the real rotavirus data, computed using two filtering methods, employed 36 replicates and 50,000 particles} % This is effectively the table title
\label{tab:ovovrealdata} % For referencing this table elsewhere in your text
\begin{tabular}{|c c c|} 
 \hline
  & PAL & SMC  \\
  \hline
 OvOv & -6892.168 & -7200.866 \\ 
 \hline
\end{tabular}
\end{table}
\vspace{-3mm}

Table \ref{tab:ovovrealdata} indicates that the results from the two methods are quite similar but SMC obtained a bit lower log-likelihood estimate than PAL. One potential explanation is the model misspecification, leading to a sensitive detection of likelihood shortfalls for SMC. Since consistent datasets, when faced with model misspecification due to outliers in the data, challenge the SMC approach, leading to low weights being assigned to almost all particles (recall Algorithm \ref{alg1}), as the latent process model struggles to accommodate outliers and is misspecified for such data points, which results in low conditional likelihood at that specific data point. Moreover, for SMC, a situation where a few particles acquire a high weight while the majority are assigned negligible weights can lead to an imbalance in the resampling process. This scenario might result in these few heavily weighted particles being resampled repeatedly, thereby dominating the sample set. If, following this, the subsequent simulations are unable to track the model accurately, it is regarded as an SMC failure. However, such an occurrence was not observed in our analysis, indicating that the SMC method functioned without encountering this specific type of failure.

To examine the model misspecification at potential time points, as depicted in Figure \ref{fig:cond_SMC_PAL_realdat}, the majority of significant drops of the SMC occurred at time ($t=1, \,\,2, \,\,3, \,\,11, \,\,81, \,\,194, \,\,325-333$). The significant drops observed initially, specifically at times $t=1, 2, 3$, may suggest an improper initial distribution from which the first set of particles was simulated. At times $t=11, \,\,81, \,\,194$ and $t=325-333$, the shortfall approximated $50-100$ log-units. These failures can be attributed to outliers that the model cannot explain, as shown in Appendix \ref{appendB}. This suggests model misspecification, particularly in the latent process, which fails to generate suitable particles capable of producing a high likelihood of observing the true reported cases from the data. Subsequent issues observed at later times could indicate problems caused by fixed seasonality, which cannot be compensated for by adding more stochastic elements.

\vspace{-4.5mm}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{cond_SMC_PAL_realdat.png}
\caption{\label{fig:cond_SMC_PAL_realdat}Conditional log-likelihoods computed using two methods for the real rotavirus dataset, with the SMC method replicated 36 times due to its high variances. The blue line represents overlapping results from 36 SMC computations, while the red line is derived from PAL. The main sources of likelihood shortfall of SMC are at time points $t=1, 2, 3, 11, 81, 194$, and $325-333$.}
\end{figure}

\vspace{-8mm}
\subsection{PAL versus SMC: When the Model is Correctly Specified}\label{section3}

Here, we utilize 100 simulated datasets from the OvOv model with parameters set at the MLE, thereby eliminating model misspecification. We then examine the likelihood values computed by PAL and SMC.

\begin{table}[htbp] % The placement specifier can be [h!tbp]
\centering % Centers the table
\caption{Average log-likelihood at the MLE at the 100 simulated datasets from OvOv computed by two filtering methods, employed 36 replicates and 50,000 particles} % This is effectively the table title
\label{tab:ovovtrue} % For referencing this table elsewhere in your text
\begin{tabular}{|c c c|} 
 \hline
  & PAL & SMC  \\ 
 \hline
 OvOv & -6231.97 & -6224.245 \\ 
 \hline
\end{tabular}
\end{table}
\begin{figure}[H]
\centering
\vspace{-7mm}
\includegraphics[width=0.55\textwidth]{PAL_vs_SMC_ovov_100.png}
\caption{\label{fig:PAL_vs_SMC_ovov_100}Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model. The red line is the line $Lik(pfilter)=Lik(palfilter)$. We can see on the simulated data, SMC gives consistently higher likelihood estimates than PAL.}
\end{figure}

When the model is correctly specified, however, the log-likelihood values computed by PAL, as shown in Table \ref{tab:ovovtrue} and Figure \ref{fig:PAL_vs_SMC_ovov_100}, consistently exhibit a shortfall. This shortfall highlights the approximative nature of the PAL method compared to the more accurate SMC filter. It is posited that the model yielding the highest likelihood is, in theory, the true model, a concept we will explore in further detail in the discussion section \ref{dis}.

\subsection{Iterated Filtering Maximization of the Likelihood}

Besides proposing a novel filtering algorithm, by applying PAL to the model for rotavirus, \cite{wwr} claimed they have also identified a superior model, as a much lower AIC represented in Table \ref{tab:model_performance} indicates and the failure from getting a lower likelihood in Section \ref{misspecified}. However, as demonstrated in \ref{section3}, we recognize that SMC, functioning as an ideal filter, ought to yield a higher likelihood estimation when the model is correct. Notably, Section \ref{misspecified} reveals that when the model confronts actual rotavirus data, it encounters misspecification issues, particularly at times $t=1, \,\,2, \,\,3, \,\,11, \,\,81, \,\,194, \,\,325-333$, leading to shortfalls in likelihood. Appendix \ref{appendB} discusses this phenomenon, attributing it to certain data points that the model fails to adequately represent. This issue is especially pronounced at the initial time points. Particularly, in the initial time points $t=1,2,3$, the likelihood deviates from PAL estimates for about more than 100 log-likelihood units, suggesting an improper choice of initial distribution that might favor the PAL estimate. Consequently, we have opted to estimate the initial distribution \(\lambda_0 = (S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}, R_{30})\), where $S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}$, and $R_{30}$ represent the number of individuals in each compartment at time $t=0$. The results of this new estimation can be found in Table \ref{tab:mlebywwr}. Importantly, the maximization of the likelihood conducted via the Iterated Filtering (mif2) algorithm \citep{pomppackagepaper} procures a higher likelihood at the estimated MLE than that estimated by PAL, and a different yet comparable set of estimates is obtained. In terms of parameters pertinent to epidemiological interests, i.e. the first 7 rows in Table \ref{tab:mlebywwr}, we observe a lower force of infection parameter for the first age group, $\beta_1$, alongside a higher force of infection for the third age group, $\beta_3$. Additionally, we detect variations in the phase parameter $\phi$, which directly affects the seasonal time points, as well as in the stochastic parameter $\sigma_{\xi}$ and the initial compartmental distributions. The modifications to the initial distributions aim to rectify inadequate initialization, while the other adjustments address model misspecifications identified in Section \ref{misspecified}. Changes in the remaining parameter set are negligible. These findings indicate that when model misspecification is reduced to a minimal level, both PAL and SMC approaches tend to provide similar likelihood estimates, and the Iterated Filtering algorithm is capable of generating a superior maximum likelihood estimation than PAL equipped with the gradient ascent algorithm. Moreover, the results produced by SMC are more reliable due to the exact nature of the filter, despite PAL's computational efficiency.

\section{Discussion and Conclusion}\label{dis}

This paper examines two filtering approaches, with PAL specifically designed for certain types of compartmental POMP models. However, both methods have their limitations. Although this was not observed in our study, SMC is sensitive to model misspecification, which can lead to filtering failures. Conversely, PAL may experience a shortfall in likelihood estimation when the model is correctly specified.

PAL is an approximate filtering method used in scenarios where SMC is computationally prohibitive. It approximates the posterior distribution by a Poisson approximation. Due to its approximative nature, it does not perfectly capture the true posterior distribution. 
SMC, however, often referred to as the "perfect filter," uses a large set of particles and sophisticated resampling techniques to theoretically accurately approximate the posterior distribution. It is computationally more intensive than PAL but tends to provide a satisfying approximation, especially in complex models that are not compatible with PAL in general. When PAL and SMC are applied to a well-specified model meaning the model with the latent process model being correctly represented, SMC outperforms PAL because it better captures the nuances of the posterior distribution. This is why SMC might show a "likelihood shortfall" compared to PAL when the model is correctly specified. This is due to the approximation error of PAL can't be improved arbitrarily in general \citep{annurev:/content/journals/10.1146/annurev-control-042920-015119}, and likelihood being a proper scoring rule.

The concept of likelihood as a proper scoring rule \citep{gneiting1,gneiting2} in statistical models implies that using likelihood-based methods will naturally align with the true model of the data, assuming the model structure correctly represents the underlying process. When the likelihood function is used as a scoring rule, it is considered proper because maximizing the likelihood leads to parameter estimates that converge to their true values under large sample conditions, provided the model is specified correctly. A scoring rule is "strictly proper" if a forecaster maximizes the expected score by issuing a probabilistic forecast that exactly matches the true distribution of the outcomes. In other words, for a scoring rule to be strictly proper, the best strategy for the forecaster is to report the true probability distribution of the events they are forecasting. This promotes honest and accurate forecasting. This provides a theoretical explanation of why SMC outperforms PAL when the model is correct.

The "likelihood shortfall" in PAL when compared to SMC, particularly when the model is correct, highlights the trade-off between computational efficiency and accuracy in statistical approximations. While PAL may be faster and more computationally feasible in many scenarios, it sacrifices some accuracy, which becomes apparent when compared to the more exact SMC method, especially under the scrutiny of proper scoring rules as discussed in the context.

\section{Acknowledgement}

The code and results can be found on: \url{https://github.com/yhaoum/PAL-check.git}.

I would like to express my sincere gratitude to Prof. Ionides for selecting this interesting topic for me and providing consistent and immediate guidance throughout the entire process. I also want to extend my thanks to Aaron for dedicating his time to help review my code and results.

\newpage
\appendix
\section{Consistency of Models Built in \textbf{pomp} and \cite{wwr}}\label{appendA}

\begin{figure}[H]
\centering
\includegraphics[width=0.7\textwidth]{compare_density_plot_EqEq_old.png}
\caption{\label{fig:rotavirus} Density plot and t-test of mean, median, and variance of all three age-specific time series of 1000 Simulations from EqEq model built in WWR code and \textbf{pomp}}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.7\textwidth]{compare_density_plot_EqOv.png}
\caption{\label{fig:rotavirus} Density plot and t-test of mean, median, and variance of all three age-specific time series of 1000 Simulations from EqOv model built in WWR code and \textbf{pomp}}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=0.7\textwidth]{compare_density_plot_OvOv.png}
\caption{\label{fig:rotavirus} Density plot and t-test of mean, median, and variance of all three age-specific time series of 1000 Simulations from OvOv model built in WWR code and \textbf{pomp}}
\end{figure}

\section{Model Misspecification}\label{appendB}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{plot_zoom_png.png}
\caption{\label{fig:zoomedin_outlier_plot} Zoomed-in time-Cases plot at time $t=0-20, 75-85, 190-200$ and $320-340$. In the visualization, solid lines represent the true reported cases from the original rotavirus dataset, while the shaded strips with dashed boundaries-colored according to different groups—indicate enlarged "confidence intervals" for these cases. Specifically, the upper boundary of each strip corresponds to the upper 97.5\% percentile of 36 replicates of the filtered accumulated number of cases, $H_{kt}$, for $k=1,2,3$, and $t=1,2,...,416$, as introduced in Section \ref{observationmodel}. This boundary is further adjusted by the upper 97.5\% percentile of the reporting rate $q_t$, estimated by \cite{wwr} in Table \ref{tab:mlebywwr}. The reporting rate $q_t$ is assumed to follow a truncated normal distribution with a mean of 0.07 and a standard deviation of $\sigma_q = 0.021$, resulting in a lower 2.5\% percentile of 0.029 and an upper 97.5\% percentile of 0.111. The methodology for establishing the lower boundary of the strip mirrors that of the upper boundary. Although this approach is not entirely rigorous, each strip is intended to encompass our extended confidence interval for the true reported cases. Consequently, data points that fall outside these strips give us a intuition of where the model misspecifications may occur.}
\end{figure}

As illustrated in Figure \ref{fig:cond_SMC_PAL_realdat}, shortfalls are observed at times $t=1,\,\,2,\,\,3,\,\,11,\,\,81,\,\,194,$ and $325-333$. These discrepancies are primarily attributable to outliers that the model fails to accommodate. In the displayed results, it is immediately apparent that at the initial times $t=1, \,\,2, \,\,3$ all the true reported cases deviated from the expected range, falling outside the designated strip. Specifically, at time $t=11$, the model is misspecified for case 1, depicted by the solid red line. Similarly, at times $t=81,\,\, 82$, the model is misspecified for case 3, as indicated by the solid blue line that extended beyond the strip's boundaries. At $t=194$, the solid green line denotes case 2 as where the model is misspecified. During the period from $t=325-333$, all three observed cases are where the model misspecifications happen. These deviations provide a clear illustration of the likelihood shortfall of the sequential Monte Carlo (SMC) method when the model does not align with the data, suggesting a model misspecification.

\newpage
\bibliography{sample.bib}

\end{document}

### 2024-07-18 ms.Rnw
abstract rewritten

@@
 \usepackage{tabularx}
+
+\newcommand{\ed}[1]{\textcolor{red}{[EI:#1]}}
+\newcommand{\yize}[1]{\textcolor{green}{[YH:#1]}}
+\newcommand{\aaron}[1]{\textcolor{blue}{[AA:#1]}}
 
@@
 
-{\Large
-\textbf\newline{Poisson Approximate Likelihood compared to the particle filter} 
-}
-\newline
+\title{Poisson approximate likelihood compared to the particle filter}
+\author{Yize Hao, Aaron A. Abkemeier and Edward L. Ionides}
+\date{Draft compiled on \today}
+\maketitle
 
 \begin{abstract}
-Recently, a novel approach (Poisson Approximate Likelihood, PAL) was introduced by \cite{wwr}, which employs a Poisson approximation to posterior densities, offering a fast and consistent approximation for the likelihood function for compatible POMP models.
-\cite{wwr} claimed that their method, along with its associated model, improved the maximum likelihood estimation compared to traditional sequential Monte Carlo (SMC) methods used in \cite{stocks} when applied to the German rotavirus, by approximately 3200 log-likelihood units (see Table 1.
-\cite{wwr}).
-However, our analysis of comparing two methods applied to the model for the same rotavirus dataset reveals that the improvement of 3200 log-likelihood units results from the use of two different datasets differed by a scaling factor.
-In simulations for a correctly specified model, a standard particle filter log-likelihood estimate, optimized using iteratered filtering, outperforms PAL.
-For the rotavirus data, once a poor specification of initial values is fixed and the same data are used for both methods, the particle filter also outperforms PAL.
+Filtering algorithms are fundamental for inference on partially observed stochastic dynamic systems, since they provide access to the likelihood function and hence enable likelihood-based or Bayesian inference.
+Recently, a novel Poisson approximate likelihood (PAL) filter was introduced by \cite{wwr}.
+PAL employs a Poisson approximation to conditional densities, offering a fast and consistent approximation to the likelihood function for compatible POMP models.
+\cite{wwr} supported the importance of their contribution by theoretical analysis and numerical results.
+A central piece of evidence for PAL is the comparison in Table~1 of \cite{wwr}, which claims a large improvement for PAL over a standard particle filter algorithm.
+This evidence, based on a model and data from a previous scientific study, suggests at face value that researchers confronted with similar models would be irresponsible not to adopt PAL.
+We show that the evidence is flawed because the PAL calculations were carried out using a dataset differently scaled from the previous study.
+If PAL and the particle filter are used on this new scale, the superficial advantage of PAL largely disappears.
+On simulations where the model is correctly specified, the particle filter outperforms PAL.
+If a poorly fitting specification of initial value parameters is ammended, the particle filter also outperforms PAL on the actual data.
 \end{abstract}
@@
 @
+
+\section{Introduction}
+
+\ed{WE WILL NEED TO WORD THIS CAREFULLY.
+WE MUST BE FORCEFUL ABOUT THE VALUE OF CORRECTING THE SCIENTIFIC RECORD, WITHOUT SEEMING UNNECESSARILY CONFRONTATIONAL.
+ONE WAY TO DO THIS IS TO BE COMPLIMENTARY ABOUT PAL AS MUCH AS POSSIBLE.
+AFTER ALL, IF THE METHOD HAS SOME GOOD PROPERTIES THEN THIS ADDS VALUE TO UNDERSTANDING PROPERLY ITS LIMITATIONS.
+I'VE REWRITTEN THE ABSTRACT FROM THAT PERSPECTIVE.}
+
+\ed{AT THIS POINT, THE DRAFT IS QUITE PROVISIONAL, PUTTING THINGS IN PLACE SO IT CAN BE REFINED LATER.}
+
+\cite{wwr} claimed that PAL is the first likelihood-based filtering method that can be applied to a broad range of POMP models with the finite-population assumption and consistent results, and requires much less computational time than SMC.
+To evaluate whether SMC performs as well as PAL in terms of both computational time and likelihood identifiability, we replicated the model using the R package \textbf{pomp} \citep{pomppackagepaper}.
+First, we verified that the two models are essentially identical.
+Here, we simulated data from the model 1000 times at the MLE shown in Table \ref{tab:mlebywwr} and compared the summary statistics of 1000 simulated time series from two models using the t-test.
+Summary statistics including the mean, median, and variance for all three age groups were tested.
+The results showed no significant differences between the two models from which the data were simulated, confirming that both the latent process model and the observation model are essentially consistent with each other, given that the observations were directly simulated.
+These results are presented in Appendix \ref{appendA}.
 
@@
 
-\bibliography{palcheck.bib}
+\subsection{Key Factors Contributing to the Change of AIC in Table \ref{tab:model_performance}}\label{section1}
+
+From Table \ref{tab:model_performance}, it is evident that the OvOv model yields the lowest AIC, making it particularly noteworthy.
+Since one of \cite{stocks}'s main purposes was to do model selection, therefore it's exciting to see \cite{wwr} obtaining a much better model.
+The overdispersion model (OvOv) is often the best model in terms of likelihood, primarily because it allows for greater variation in the model.
+By incorporating more parameters, the OvOv model can capture subtleties and complexities in the dataset that simpler models may overlook, leading to a more accurate fit and higher likelihood values.
+\cite{wwr} claimed that it is the innovative observation model that contributed to an increase in log-likelihood by 3200 log-likelihood units.
+However, after comparing the results, we found that the increase in likelihood when implementing the OvOv model is attributable to the datasets used by \cite{wwr} and \cite{stocks}, where \cite{stocks}'s dataset has no underreporting.
+Therefore \cite{wwr}'s dataset can be viewed as a scaled-down version of \cite{stocks}'s dataset by the reporting rate, which is roughly 0.07.
+Within the framework of the same model, scaling the data downwards by a factor results in a corresponding rise in likelihood.
+This scaling effect accounts for the discrepancy observed in the increase of 3200 log-likelihood units.
+\vspace{-2mm}
+\subsection{PAL versus SMC: When the Model is Misspecified} \label{misspecified}
+
+Our results demonstrate how SMC is sensitive to model misspecification, which consequently leads to a low likelihood estimate at the MLE, as shown in Table \ref{tab:ovovrealdata} by fitting the real German rotavirus dataset into the model.
+
+\begin{table}[ht] % The placement specifier can be [h!tbp]
+\centering % Centers the table
+\caption{log-likelihood at the MLE of OvOv model for the real rotavirus data, computed using two filtering methods, employed 36 replicates and 50,000 particles} % This is effectively the table title
+\label{tab:ovovrealdata} % For referencing this table elsewhere in your text
+\begin{tabular}{|c c c|} 
+ \hline
+  & PAL & SMC  \\
+  \hline
+ OvOv & -6892.168 & -7200.866 \\ 
+ \hline
+\end{tabular}
+\end{table}
+\vspace{-3mm}
+
+Table \ref{tab:ovovrealdata} indicates that the results from the two methods are quite similar but SMC obtained a bit lower log-likelihood estimate than PAL.
+One potential explanation is the model misspecification, leading to a sensitive detection of likelihood shortfalls for SMC.
+Since consistent datasets, when faced with model misspecification due to outliers in the data, challenge the SMC approach, leading to low weights being assigned to almost all particles (recall Algorithm \ref{alg1}), as the latent process model struggles to accommodate outliers and is misspecified for such data points, which results in low conditional likelihood at that specific data point.
+Moreover, for SMC, a situation where a few particles acquire a high weight while the majority are assigned negligible weights can lead to an imbalance in the resampling process.
+This scenario might result in these few heavily weighted particles being resampled repeatedly, thereby dominating the sample set.
+If, following this, the subsequent simulations are unable to track the model accurately, it is regarded as an SMC failure.
+However, such an occurrence was not observed in our analysis, indicating that the SMC method functioned without encountering this specific type of failure.
+
+To examine the model misspecification at potential time points, as depicted in Figure \ref{fig:cond_SMC_PAL_realdat}, the majority of significant drops of the SMC occurred at time ($t=1, \,\,2, \,\,3, \,\,11, \,\,81, \,\,194, \,\,325-333$).
+The significant drops observed initially, specifically at times $t=1, 2, 3$, may suggest an improper initial distribution from which the first set of particles was simulated.
+At times $t=11, \,\,81, \,\,194$ and $t=325-333$, the shortfall approximated $50-100$ log-units.
+These failures can be attributed to outliers that the model cannot explain, as shown in Appendix \ref{appendB}.
+This suggests model misspecification, particularly in the latent process, which fails to generate suitable particles capable of producing a high likelihood of observing the true reported cases from the data.
+Subsequent issues observed at later times could indicate problems caused by fixed seasonality, which cannot be compensated for by adding more stochastic elements.
+
+\vspace{-4.5mm}
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{thesis-figs/cond_SMC_PAL_realdat.png}
+\caption{\label{fig:cond_SMC_PAL_realdat}Conditional log-likelihoods computed using two methods for the real rotavirus dataset, with the SMC method replicated 36 times due to its high variances.
+The blue line represents overlapping results from 36 SMC computations, while the red line is derived from PAL.
+The main sources of likelihood shortfall of SMC are at time points $t=1, 2, 3, 11, 81, 194$, and $325-333$.}
+\end{figure}
+
+\vspace{-8mm}
+\subsection{PAL versus SMC: When the Model is Correctly Specified}\label{section3}
+
+Here, we utilize 100 simulated datasets from the OvOv model with parameters set at the MLE, thereby eliminating model misspecification. We then examine the likelihood values computed by PAL and SMC.
+
+\begin{table}[htbp] % The placement specifier can be [h!tbp]
+\centering % Centers the table
+\caption{Average log-likelihood at the MLE at the 100 simulated datasets from OvOv computed by two filtering methods, employed 36 replicates and 50,000 particles} % This is effectively the table title
+\label{tab:ovovtrue} % For referencing this table elsewhere in your text
+\begin{tabular}{|c c c|} 
+ \hline
+  & PAL & SMC  \\ 
+ \hline
+ OvOv & -6231.97 & -6224.245 \\ 
+ \hline
+\end{tabular}
+\end{table}
+\begin{figure}[H]
+\centering
+\vspace{-7mm}
+\includegraphics[width=0.55\textwidth]{thesis-figs/PAL_vs_SMC_ovov_100.png}
+\caption{\label{fig:PAL_vs_SMC_ovov_100}Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model.
+The red line is the line $Lik(pfilter)=Lik(palfilter)$.
+We can see on the simulated data, SMC gives consistently higher likelihood estimates than PAL.}
+\end{figure}
+
+When the model is correctly specified, however, the log-likelihood values computed by PAL, as shown in Table \ref{tab:ovovtrue} and Figure \ref{fig:PAL_vs_SMC_ovov_100}, consistently exhibit a shortfall.
+This shortfall highlights the approximative nature of the PAL method compared to the more accurate SMC filter.
+It is posited that the model yielding the highest likelihood is, in theory, the true model, a concept we will explore in further detail in the discussion section \ref{dis}.
+
+\subsection{Iterated Filtering Maximization of the Likelihood}
+
+Besides proposing a novel filtering algorithm, by applying PAL to the model for rotavirus, \cite{wwr} claimed they have also identified a superior model, as a much lower AIC represented in Table \ref{tab:model_performance} indicates and the failure from getting a lower likelihood in Section \ref{misspecified}.
+However, as demonstrated in \ref{section3}, we recognize that SMC, functioning as an ideal filter, ought to yield a higher likelihood estimation when the model is correct.
+Notably, Section \ref{misspecified} reveals that when the model confronts actual rotavirus data, it encounters misspecification issues, particularly at times $t=1, \,\,2, \,\,3, \,\,11, \,\,81, \,\,194, \,\,325-333$, leading to shortfalls in likelihood.
+Appendix \ref{appendB} discusses this phenomenon, attributing it to certain data points that the model fails to adequately represent.
+This issue is especially pronounced at the initial time points.
+Particularly, in the initial time points $t=1,2,3$, the likelihood deviates from PAL estimates for about more than 100 log-likelihood units, suggesting an improper choice of initial distribution that might favor the PAL estimate.
+Consequently, we have opted to estimate the initial distribution \(\lambda_0 = (S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}, R_{30})\), where $S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}$, and $R_{30}$ represent the number of individuals in each compartment at time $t=0$.
+The results of this new estimation can be found in Table \ref{tab:mlebywwr}.
+Importantly, the maximization of the likelihood conducted via the Iterated Filtering (mif2) algorithm \citep{pomppackagepaper} procures a higher likelihood at the estimated MLE than that estimated by PAL, and a different yet comparable set of estimates is obtained.
+In terms of parameters pertinent to epidemiological interests, i.e.
+the first 7 rows in Table \ref{tab:mlebywwr}, we observe a lower force of infection parameter for the first age group, $\beta_1$, alongside a higher force of infection for the third age group, $\beta_3$.
+Additionally, we detect variations in the phase parameter $\phi$, which directly affects the seasonal time points, as well as in the stochastic parameter $\sigma_{\xi}$ and the initial compartmental distributions.
+The modifications to the initial distributions aim to rectify inadequate initialization, while the other adjustments address model misspecifications identified in Section \ref{misspecified}.
+Changes in the remaining parameter set are negligible.
+These findings indicate that when model misspecification is reduced to a minimal level, both PAL and SMC approaches tend to provide similar likelihood estimates, and the Iterated Filtering algorithm is capable of generating a superior maximum likelihood estimation than PAL equipped with the gradient ascent algorithm.
+Moreover, the results produced by SMC are more reliable due to the exact nature of the filter, despite PAL's computational efficiency.
+
+\section{Discussion and Conclusion}\label{dis}
+
+This paper examines two filtering approaches, with PAL specifically designed for certain types of compartmental POMP models.
+However, both methods have their limitations.
+Although this was not observed in our study, SMC is sensitive to model misspecification, which can lead to filtering failures.
+Conversely, PAL may experience a shortfall in likelihood estimation when the model is correctly specified.
+
+PAL is an approximate filtering method used in scenarios where SMC is computationally prohibitive.
+It approximates the posterior distribution by a Poisson approximation.
+Due to its approximative nature, it does not perfectly capture the true posterior distribution.
+SMC, however, often referred to as the "perfect filter," uses a large set of particles and sophisticated resampling techniques to theoretically accurately approximate the posterior distribution.
+It is computationally more intensive than PAL but tends to provide a satisfying approximation, especially in complex models that are not compatible with PAL in general.
+When PAL and SMC are applied to a well-specified model meaning the model with the latent process model being correctly represented, SMC outperforms PAL because it better captures the nuances of the posterior distribution.
+This is why SMC might show a "likelihood shortfall" compared to PAL when the model is correctly specified.
+This is due to the approximation error of PAL can't be improved arbitrarily in general \citep{annurev:/content/journals/10.1146/annurev-control-042920-015119}, and likelihood being a proper scoring rule.
+
+The concept of likelihood as a proper scoring rule \citep{gneiting1,gneiting2} in statistical models implies that using likelihood-based methods will naturally align with the true model of the data, assuming the model structure correctly represents the underlying process.
+When the likelihood function is used as a scoring rule, it is considered proper because maximizing the likelihood leads to parameter estimates that converge to their true values under large sample conditions, provided the model is specified correctly.
+A scoring rule is "strictly proper" if a forecaster maximizes the expected score by issuing a probabilistic forecast that exactly matches the true distribution of the outcomes.
+In other words, for a scoring rule to be strictly proper, the best strategy for the forecaster is to report the true probability distribution of the events they are forecasting.
+This promotes honest and accurate forecasting.
+This provides a theoretical explanation of why SMC outperforms PAL when the model is correct.
+
+The "likelihood shortfall" in PAL when compared to SMC, particularly when the model is correct, highlights the trade-off between computational efficiency and accuracy in statistical approximations.
+While PAL may be faster and more computationally feasible in many scenarios, it sacrifices some accuracy, which becomes apparent when compared to the more exact SMC method, especially under the scrutiny of proper scoring rules as discussed in the context.
+
+\section{Model Misspecification}\label{appendB}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=\textwidth]{thesis-figs/plot_zoom_png.png}
+\caption{\label{fig:zoomedin_outlier_plot} Zoomed-in time-Cases plot at time $t=0-20, 75-85, 190-200$ and $320-340$.
+In the visualization, solid lines represent the true reported cases from the original rotavirus dataset, while the shaded strips with dashed boundaries-colored according to different groups—indicate enlarged "confidence intervals" for these cases.
+Specifically, the upper boundary of each strip corresponds to the upper 97.5\% percentile of 36 replicates of the filtered accumulated number of cases, $H_{kt}$, for $k=1,2,3$, and $t=1,2,...,416$, as introduced in Section \ref{observationmodel}.
+This boundary is further adjusted by the upper 97.5\% percentile of the reporting rate $q_t$, estimated by \cite{wwr} in Table \ref{tab:mlebywwr}.
+The reporting rate $q_t$ is assumed to follow a truncated normal distribution with a mean of 0.07 and a standard deviation of $\sigma_q = 0.021$, resulting in a lower 2.5\% percentile of 0.029 and an upper 97.5\% percentile of 0.111.
+The methodology for establishing the lower boundary of the strip mirrors that of the upper boundary.
+Although this approach is not entirely rigorous, each strip is intended to encompass our extended confidence interval for the true reported cases.
+Consequently, data points that fall outside these strips give us a intuition of where the model misspecifications may occur.}
+\end{figure}
+
+As illustrated in Figure \ref{fig:cond_SMC_PAL_realdat}, shortfalls are observed at times $t=1,\,\,2,\,\,3,\,\,11,\,\,81,\,\,194,$ and $325-333$.
+These discrepancies are primarily attributable to outliers that the model fails to accommodate.
+In the displayed results, it is immediately apparent that at the initial times $t=1, \,\,2, \,\,3$ all the true reported cases deviated from the expected range, falling outside the designated strip.
+Specifically, at time $t=11$, the model is misspecified for case 1, depicted by the solid red line.
+Similarly, at times $t=81,\,\, 82$, the model is misspecified for case 3, as indicated by the solid blue line that extended beyond the strip's boundaries.
+At $t=194$, the solid green line denotes case 2 as where the model is misspecified.
+During the period from $t=325-333$, all three observed cases are where the model misspecifications happen.
+These deviations provide a clear illustration of the likelihood shortfall of the sequential Monte Carlo (SMC) method when the model does not align with the data, suggesting a model misspecification.
+
+\bibliography{bib-pal}
 

### 2024-07-19 ms.Rnw
some reorganization and rewriting

@@
 
-\title{Poisson approximate likelihood compared to the particle filter}
+\title{The Poisson approximate likelihood compared to the particle filter}
 \author{Yize Hao, Aaron A. Abkemeier and Edward L. Ionides}
@@
 Recently, a novel Poisson approximate likelihood (PAL) filter was introduced by \cite{wwr}.
-PAL employs a Poisson approximation to conditional densities, offering a fast and consistent approximation to the likelihood function for compatible POMP models.
+PAL employs a Poisson approximation to conditional densities, offering a fast and consistent approximation to the likelihood function for a certain subset of partially observed Markov process models.
 \cite{wwr} supported the importance of their contribution by theoretical analysis and numerical results.
@@
 
+%%%%%%%%%% 11111111 %%%%%%%%%%%%
+
 \section{Introduction}
 
-\ed{WE WILL NEED TO WORD THIS CAREFULLY.
+%% \ed{WE WILL NEED TO WORD THIS CAREFULLY.
 WE MUST BE FORCEFUL ABOUT THE VALUE OF CORRECTING THE SCIENTIFIC RECORD, WITHOUT SEEMING UNNECESSARILY CONFRONTATIONAL.
@@
 
-\ed{AT THIS POINT, THE DRAFT IS QUITE PROVISIONAL, PUTTING THINGS IN PLACE SO IT CAN BE REFINED LATER.}
-
-\cite{wwr} claimed that PAL is the first likelihood-based filtering method that can be applied to a broad range of POMP models with the finite-population assumption and consistent results, and requires much less computational time than SMC.
+%% \ed{AT THIS POINT, THE DRAFT IS QUITE PROVISIONAL, PUTTING THINGS IN PLACE SO IT CAN BE REFINED LATER.}
+
+This article results from an investigation of the results in Table~1 of \citet{wwr}.
+The authors of that paper were given the opportunity to submit a correction, after we shared the results of our investigation with them, but they declined.
+Our task here is to correct the error in Table~1 so that researchers considering whether to implement Poisson approximate likelihood (PAL) are appropriately informed about its benefits.
+Table~1 reanalyzes the model and data of \citet{stocks}, for which the likelihood was calculated using a particle filter.
+The theory developed by \cite{wwr} shows that PAL has some potentially useful scaling properties, but the numerical results in Table~1 appear to show much stronger performance than the particle filter on this example of moderate size.
+
+First, we show that most of this apparent improvement for PAL arises because \cite{wwr} used a different scaling of the data from \cite{stocks}.
+%% \section{Comparing PAL with the particle filter} 
+Two models for the same data can properly be compared by their likelihood, even if the models have entirely different structures.
+One can making allowance for the number of estimated parameters using a quantity such as Akaike's information criterion [REF].
+However, if data are rescaled, a correction is required to make likelihoods comparable.
+If one model describes a dataset in grams, and another describes it in kilograms, then the latter model will earn an increased log-likelihood of $\log(10^3)$ for each data point simply because of the change in scale.
+Presenting a direct comparison of a likelihood for the data in grams with a likelihood for the data in kilograms would evidently be inapprpriate.
+\cite{stocks} fitted their model to a dataset derived by dividing the original reported count data by an estimated reporting rate, to put their data on the scale of the actual number of cases in the population.
+\cite{wwr} fitted directly to the report data.
+The reporting rate used by \cite{stocks} varied over time, but was generally around $7\%$.
+On 1200 data points, this corresponds to a discrepancy of around $-1200\log(0.07)=3200$ log-likelihood units, largely explaining the difference reported in Table~1 and interpreted by \cite{wwr} as evidence supporting PAL.
+The comparison can be corrected either by applying the method of \cite{stocks} to the data of \cite{wwr} or vvice versa.
+Since the method of \cite{stocks} is applicable to a more general class of models, and supported by published software, it was more convenient to apply this method to the model and data of \cite{wwr}.
+The large discrepancy in log-likelihood disappears at this point (see our Table~\ref{tab:ovovrealdata} , column \ed{XXX}).
+This re-analysis does, however, show a small advantage for PAL.
+We continued our investigation to establish the cause of this.
+
+PAL provides an approximate filter, whereas the particle filter provides a Monte~Carlo estimate of an exact filter.
+Possible explanations for PAL obtaining a higher log-likelihood than a particle filter include:
+\begin{enumerate}
+\item The particle filter could have high Monte Carlo variance.
+This results in negative bias in its log-likilihood estimate due to Jensen's inequality, since the particle filter provides an unbiased likelihood estimate.
+\item If the model is misspecified, the approximation error in PAL could lead to a higher log-likelihood estimate than that of the actual misspecified model.
+\end{enumerate}
+Log-likelihood is a proper scoring rule for forecasts \cite{gneiting2}, and both the particle filter and PAL construct their log-likelihood estimates via a sequence of one-step forecasts.
+%% resulting in estimates of the conditional log-likelihood, $f_{Y
+Therefore, if the model is correctly specified, the approximation error in PAL can only decrease the expected log-likelihood.
+We tested this on simulated data, for which the model of \cite{wwr} is correctly specified.
+We found that, in this case, the particle filter out-performs PAL \ed{SHOW RESULTS}.
+This supports hypothesis 2, above.
+We looked for model misspecification by comparing the conditional log-likelihood estimates at each time point, as described in the Methods section.
+Figure~ \ref{fig:cond_SMC_PAL_realdat} indicates that the model provides a poor explanation of the data right at the start of the time series.
+This could be a problem with data collection, or some subtle issue with the science of the epidemiological system, but a simple possibility is that the initial values of the latent state process might be poorly specified.
+For this model, the initial values were fixed based on scientific reasoning, rather than being estimated.
+If we instead estimate the initial values, we obtain the results in \ed{XXXX}.
+
+\ed{WE CAN REVIEW EXACTLY WHICH NUMBERS TO SHOW. IDEALLY, AS LITTLE AS POSSIBLE IN ORDER TO MAKE A STRONG ARGUMENT.}
+
+We conclude that PAL is a potentially useful algorithm, with some favorable theoretical properties.
+However, the corrected evidence does not indicate an advantage for using PAL in situations where the particle filter is effective.
+
+\section{Methods}
+
+\ed{BRIEFLY DESCRIBE HERE ANY DETAILS NEEDED TO DESCRIBE THE RESULTS. COMMENT ON CODE AVAILABILITY.}
+
 To evaluate whether SMC performs as well as PAL in terms of both computational time and likelihood identifiability, we replicated the model using the R package \textbf{pomp} \citep{pomppackagepaper}.
@@
 The results showed no significant differences between the two models from which the data were simulated, confirming that both the latent process model and the observation model are essentially consistent with each other, given that the observations were directly simulated.
-These results are presented in Appendix \ref{appendA}.
+
+From Table \ref{tab:model_performance}, it is evident that the OvOv model, with overdispersion in both the dynamic system and the measurement model, yields the lowest AIC.
+We therefore focus on this version of the model.
+
+To examine the model misspecification at potential time points, as depicted in Figure \ref{fig:cond_SMC_PAL_realdat}, the majority of significant drops of the SMC occurred at time ($t=1, \,\,2, \,\,3, \,\,11, \,\,81, \,\,194, \,\,325-333$).
+The significant drops observed initially, specifically at times $t=1, 2, 3$, may suggest an improper initial distribution from which the first set of particles was simulated.
+At times $t=11, \,\,81, \,\,194$ and $t=325-333$, the shortfall approximated $50-100$ log-units.
+These failures can be attributed to outliers that the model cannot explain, as shown in \ed{XXXX}.
+This suggests model misspecification, particularly in the latent process, which fails to generate suitable particles capable of producing a high likelihood of observing the true reported cases from the data.
+Subsequent issues observed at later times could indicate problems caused by fixed seasonality, which cannot be compensated for by adding more stochastic elements.
+
+This article has focused on likelihood evaluation, but the inferences presented also require the likelihood estimate to maximized.
+\cite{stocks} used iterated filtering to maximize the particle filter likelihood estimate, whereas \cite{wwr} used direct numerical maximization.
+We follow the approaches adopted by these two papers.
+The optimization method is not directly pertinent to the comparison of the filtering methods, so we do not discuss optimization in further detail here.
 
@@
 
-\subsection{Key Factors Contributing to the Change of AIC in Table \ref{tab:model_performance}}\label{section1}
-
-From Table \ref{tab:model_performance}, it is evident that the OvOv model yields the lowest AIC, making it particularly noteworthy.
-Since one of \cite{stocks}'s main purposes was to do model selection, therefore it's exciting to see \cite{wwr} obtaining a much better model.
-The overdispersion model (OvOv) is often the best model in terms of likelihood, primarily because it allows for greater variation in the model.
-By incorporating more parameters, the OvOv model can capture subtleties and complexities in the dataset that simpler models may overlook, leading to a more accurate fit and higher likelihood values.
-\cite{wwr} claimed that it is the innovative observation model that contributed to an increase in log-likelihood by 3200 log-likelihood units.
-However, after comparing the results, we found that the increase in likelihood when implementing the OvOv model is attributable to the datasets used by \cite{wwr} and \cite{stocks}, where \cite{stocks}'s dataset has no underreporting.
-Therefore \cite{wwr}'s dataset can be viewed as a scaled-down version of \cite{stocks}'s dataset by the reporting rate, which is roughly 0.07.
-Within the framework of the same model, scaling the data downwards by a factor results in a corresponding rise in likelihood.
-This scaling effect accounts for the discrepancy observed in the increase of 3200 log-likelihood units.
-\vspace{-2mm}
-\subsection{PAL versus SMC: When the Model is Misspecified} \label{misspecified}
-
 Our results demonstrate how SMC is sensitive to model misspecification, which consequently leads to a low likelihood estimate at the MLE, as shown in Table \ref{tab:ovovrealdata} by fitting the real German rotavirus dataset into the model.
@@
 \end{table}
-\vspace{-3mm}
-
-Table \ref{tab:ovovrealdata} indicates that the results from the two methods are quite similar but SMC obtained a bit lower log-likelihood estimate than PAL.
-One potential explanation is the model misspecification, leading to a sensitive detection of likelihood shortfalls for SMC.
-Since consistent datasets, when faced with model misspecification due to outliers in the data, challenge the SMC approach, leading to low weights being assigned to almost all particles (recall Algorithm \ref{alg1}), as the latent process model struggles to accommodate outliers and is misspecified for such data points, which results in low conditional likelihood at that specific data point.
-Moreover, for SMC, a situation where a few particles acquire a high weight while the majority are assigned negligible weights can lead to an imbalance in the resampling process.
-This scenario might result in these few heavily weighted particles being resampled repeatedly, thereby dominating the sample set.
-If, following this, the subsequent simulations are unable to track the model accurately, it is regarded as an SMC failure.
-However, such an occurrence was not observed in our analysis, indicating that the SMC method functioned without encountering this specific type of failure.
-
-To examine the model misspecification at potential time points, as depicted in Figure \ref{fig:cond_SMC_PAL_realdat}, the majority of significant drops of the SMC occurred at time ($t=1, \,\,2, \,\,3, \,\,11, \,\,81, \,\,194, \,\,325-333$).
-The significant drops observed initially, specifically at times $t=1, 2, 3$, may suggest an improper initial distribution from which the first set of particles was simulated.
-At times $t=11, \,\,81, \,\,194$ and $t=325-333$, the shortfall approximated $50-100$ log-units.
-These failures can be attributed to outliers that the model cannot explain, as shown in Appendix \ref{appendB}.
-This suggests model misspecification, particularly in the latent process, which fails to generate suitable particles capable of producing a high likelihood of observing the true reported cases from the data.
-Subsequent issues observed at later times could indicate problems caused by fixed seasonality, which cannot be compensated for by adding more stochastic elements.
-
-\vspace{-4.5mm}
+
 \begin{figure}[H]
@@
 
-When the model is correctly specified, however, the log-likelihood values computed by PAL, as shown in Table \ref{tab:ovovtrue} and Figure \ref{fig:PAL_vs_SMC_ovov_100}, consistently exhibit a shortfall.
-This shortfall highlights the approximative nature of the PAL method compared to the more accurate SMC filter.
-It is posited that the model yielding the highest likelihood is, in theory, the true model, a concept we will explore in further detail in the discussion section \ref{dis}.
-
-\subsection{Iterated Filtering Maximization of the Likelihood}
-
-Besides proposing a novel filtering algorithm, by applying PAL to the model for rotavirus, \cite{wwr} claimed they have also identified a superior model, as a much lower AIC represented in Table \ref{tab:model_performance} indicates and the failure from getting a lower likelihood in Section \ref{misspecified}.
-However, as demonstrated in \ref{section3}, we recognize that SMC, functioning as an ideal filter, ought to yield a higher likelihood estimation when the model is correct.
-Notably, Section \ref{misspecified} reveals that when the model confronts actual rotavirus data, it encounters misspecification issues, particularly at times $t=1, \,\,2, \,\,3, \,\,11, \,\,81, \,\,194, \,\,325-333$, leading to shortfalls in likelihood.
+Notably, Section~\ref{misspecified} reveals that when the model confronts actual rotavirus data, it encounters misspecification issues, particularly at times $t=1, \,\,2, \,\,3, \,\,11, \,\,81, \,\,194, \,\,325-333$, leading to shortfalls in likelihood.
 Appendix \ref{appendB} discusses this phenomenon, attributing it to certain data points that the model fails to adequately represent.
@@
 Changes in the remaining parameter set are negligible.
-These findings indicate that when model misspecification is reduced to a minimal level, both PAL and SMC approaches tend to provide similar likelihood estimates, and the Iterated Filtering algorithm is capable of generating a superior maximum likelihood estimation than PAL equipped with the gradient ascent algorithm.
-Moreover, the results produced by SMC are more reliable due to the exact nature of the filter, despite PAL's computational efficiency.
-
-\section{Discussion and Conclusion}\label{dis}
-
-This paper examines two filtering approaches, with PAL specifically designed for certain types of compartmental POMP models.
-However, both methods have their limitations.
-Although this was not observed in our study, SMC is sensitive to model misspecification, which can lead to filtering failures.
-Conversely, PAL may experience a shortfall in likelihood estimation when the model is correctly specified.
 
@@
 This is due to the approximation error of PAL can't be improved arbitrarily in general \citep{annurev:/content/journals/10.1146/annurev-control-042920-015119}, and likelihood being a proper scoring rule.
-
-The concept of likelihood as a proper scoring rule \citep{gneiting1,gneiting2} in statistical models implies that using likelihood-based methods will naturally align with the true model of the data, assuming the model structure correctly represents the underlying process.
-When the likelihood function is used as a scoring rule, it is considered proper because maximizing the likelihood leads to parameter estimates that converge to their true values under large sample conditions, provided the model is specified correctly.
-A scoring rule is "strictly proper" if a forecaster maximizes the expected score by issuing a probabilistic forecast that exactly matches the true distribution of the outcomes.
-In other words, for a scoring rule to be strictly proper, the best strategy for the forecaster is to report the true probability distribution of the events they are forecasting.
-This promotes honest and accurate forecasting.
-This provides a theoretical explanation of why SMC outperforms PAL when the model is correct.
-
-The "likelihood shortfall" in PAL when compared to SMC, particularly when the model is correct, highlights the trade-off between computational efficiency and accuracy in statistical approximations.
-While PAL may be faster and more computationally feasible in many scenarios, it sacrifices some accuracy, which becomes apparent when compared to the more exact SMC method, especially under the scrutiny of proper scoring rules as discussed in the context.
-
-\section{Model Misspecification}\label{appendB}
 

### 2024-07-19 ms.Rnw
tidying

@@
 
-%% \ed{WE WILL NEED TO WORD THIS CAREFULLY.
-WE MUST BE FORCEFUL ABOUT THE VALUE OF CORRECTING THE SCIENTIFIC RECORD, WITHOUT SEEMING UNNECESSARILY CONFRONTATIONAL.
-ONE WAY TO DO THIS IS TO BE COMPLIMENTARY ABOUT PAL AS MUCH AS POSSIBLE.
-AFTER ALL, IF THE METHOD HAS SOME GOOD PROPERTIES THEN THIS ADDS VALUE TO UNDERSTANDING PROPERLY ITS LIMITATIONS.
-I'VE REWRITTEN THE ABSTRACT FROM THAT PERSPECTIVE.}
-
-%% \ed{AT THIS POINT, THE DRAFT IS QUITE PROVISIONAL, PUTTING THINGS IN PLACE SO IT CAN BE REFINED LATER.}
-
 This article results from an investigation of the results in Table~1 of \citet{wwr}.
@@
 
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.55\textwidth]{thesis-figs/PAL_vs_SMC_ovov_100.png}
+\caption{\label{fig:PAL_vs_SMC_ovov_100}Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model.
+The particle filter and PAL used 50,000 particles.
+The red line is the line $Lik(pfilter)=Lik(palfilter)$.
+On average, the particle filter likelihood is 7.7 log units higher than the PAL.
+%%  -6224.245 - (-6231.97) = 7.725
+}
+\end{figure}
+
 PAL provides an approximate filter, whereas the particle filter provides a Monte~Carlo estimate of an exact filter.
@@
 We tested this on simulated data, for which the model of \cite{wwr} is correctly specified.
-We found that, in this case, the particle filter out-performs PAL \ed{SHOW RESULTS}.
+We found that, in this case, the particle filter out-performs PAL (Figure~\ref{fig:PAL_vs_SMC_ovov_100}).
 This supports hypothesis 2, above.
-We looked for model misspecification by comparing the conditional log-likelihood estimates at each time point, as described in the Methods section.
-Figure~ \ref{fig:cond_SMC_PAL_realdat} indicates that the model provides a poor explanation of the data right at the start of the time series.
+Consequently, we looked for model misspecification by comparing the conditional log-likelihood estimates at each time point, as described in the Methods section.
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.9\textwidth]{thesis-figs/cond_SMC_PAL_realdat.png}
+\caption{\label{fig:cond_SMC_PAL_realdat}Conditional log-likelihoods computed using two methods for the real rotavirus dataset, with the SMC method replicated 36 times due to its high variances.
+The blue line represents overlapping results from 36 SMC computations, while the red line is derived from PAL.
+The main sources of likelihood shortfall of SMC are at time points $t=1, 2, 3, 11, 81, 194$, and $325-333$.}
+\end{figure}
+
+Figure~ \ref{fig:cond_SMC_PAL_realdat} indicates that the worst conditional log-likelihood values, estimated using the particle filter, arise at the start of the time series.
 This could be a problem with data collection, or some subtle issue with the science of the epidemiological system, but a simple possibility is that the initial values of the latent state process might be poorly specified.
 For this model, the initial values were fixed based on scientific reasoning, rather than being estimated.
-If we instead estimate the initial values, we obtain the results in \ed{XXXX}.
-
-\ed{WE CAN REVIEW EXACTLY WHICH NUMBERS TO SHOW. IDEALLY, AS LITTLE AS POSSIBLE IN ORDER TO MAKE A STRONG ARGUMENT.}
-
-We conclude that PAL is a potentially useful algorithm, with some favorable theoretical properties.
-However, the corrected evidence does not indicate an advantage for using PAL in situations where the particle filter is effective.
-
-\section{Methods}
-
-\ed{BRIEFLY DESCRIBE HERE ANY DETAILS NEEDED TO DESCRIBE THE RESULTS. COMMENT ON CODE AVAILABILITY.}
-
-To evaluate whether SMC performs as well as PAL in terms of both computational time and likelihood identifiability, we replicated the model using the R package \textbf{pomp} \citep{pomppackagepaper}.
-First, we verified that the two models are essentially identical.
-Here, we simulated data from the model 1000 times at the MLE shown in Table \ref{tab:mlebywwr} and compared the summary statistics of 1000 simulated time series from two models using the t-test.
-Summary statistics including the mean, median, and variance for all three age groups were tested.
-The results showed no significant differences between the two models from which the data were simulated, confirming that both the latent process model and the observation model are essentially consistent with each other, given that the observations were directly simulated.
-
-From Table \ref{tab:model_performance}, it is evident that the OvOv model, with overdispersion in both the dynamic system and the measurement model, yields the lowest AIC.
-We therefore focus on this version of the model.
-
-To examine the model misspecification at potential time points, as depicted in Figure \ref{fig:cond_SMC_PAL_realdat}, the majority of significant drops of the SMC occurred at time ($t=1, \,\,2, \,\,3, \,\,11, \,\,81, \,\,194, \,\,325-333$).
-The significant drops observed initially, specifically at times $t=1, 2, 3$, may suggest an improper initial distribution from which the first set of particles was simulated.
-At times $t=11, \,\,81, \,\,194$ and $t=325-333$, the shortfall approximated $50-100$ log-units.
-These failures can be attributed to outliers that the model cannot explain, as shown in \ed{XXXX}.
-This suggests model misspecification, particularly in the latent process, which fails to generate suitable particles capable of producing a high likelihood of observing the true reported cases from the data.
-Subsequent issues observed at later times could indicate problems caused by fixed seasonality, which cannot be compensated for by adding more stochastic elements.
-
-This article has focused on likelihood evaluation, but the inferences presented also require the likelihood estimate to maximized.
-\cite{stocks} used iterated filtering to maximize the particle filter likelihood estimate, whereas \cite{wwr} used direct numerical maximization.
-We follow the approaches adopted by these two papers.
-The optimization method is not directly pertinent to the comparison of the filtering methods, so we do not discuss optimization in further detail here.
+If we instead estimate the initial values, we obtain the results in Table~\ref{tab:mlebywwr}.
 
@@
   \begin{threeparttable}
-    \caption{Maximum likelihood estimation of parameters by PAL for three models \citep{wwr}, and by mif2 algorithm \citep{pomppackagepaper} for the OvOv model.}
+    \caption{Maximum likelihood estimation of parameters by PAL for three models \citep{wwr}, and by mif2 algorithm \citep{pomppackagepaper} for the OvOv model.
+    \ed{LET'S DISCUSS WHETHER WE WANT TO SHOW JUST OV-OV HERE. IT WOULD BE SIMPLER.}
+    }
     \label{tab:mlebywwr}
@@
 
-Our results demonstrate how SMC is sensitive to model misspecification, which consequently leads to a low likelihood estimate at the MLE, as shown in Table \ref{tab:ovovrealdata} by fitting the real German rotavirus dataset into the model.
+\ed{WE CAN REVIEW EXACTLY WHICH NUMBERS TO SHOW. IDEALLY, AS LITTLE AS POSSIBLE IN ORDER TO MAKE A STRONG ARGUMENT.}
+
+We conclude that PAL is a potentially useful algorithm, with some favorable theoretical properties.
+However, the corrected evidence does not indicate an advantage for using PAL in situations where the particle filter is effective.
+
+%%%%%%%%% 222222222 %%%%%%%%%%5
+
+\section*{Methods}
+
+\ed{BRIEFLY DESCRIBE HERE ANY DETAILS NEEDED TO DESCRIBE THE RESULTS. COMMENT ON CODE AVAILABILITY.}
+
+To evaluate whether SMC performs as well as PAL in terms of both computational time and likelihood identifiability, we replicated the model using the R package \textbf{pomp} \citep{pomppackagepaper}.
+First, we verified that the two models are essentially identical.
+Here, we simulated data from the model 1000 times at the MLE shown in Table \ref{tab:mlebywwr} and compared the summary statistics of 1000 simulated time series from two models using the t-test.
+Summary statistics including the mean, median, and variance for all three age groups were tested.
+The results showed no significant differences between the two models from which the data were simulated, confirming that both the latent process model and the observation model are essentially consistent with each other, given that the observations were directly simulated.
+
+From Table \ref{tab:model_performance}, it is evident that the OvOv model, with overdispersion in both the dynamic system and the measurement model, yields the lowest AIC.
+We therefore focus on this version of the model.
+
+This article has focused on likelihood evaluation, but the inferences presented also require the likelihood estimate to maximized.
+\cite{stocks} used iterated filtering to maximize the particle filter likelihood estimate, whereas \cite{wwr} used direct numerical maximization.
+We follow the approaches adopted by these two papers.
+The optimization method is not directly pertinent to the comparison of the filtering methods, so we do not discuss optimization in further detail here.
+
+An extended description of our reanalysis of \cite{wwr} is provided by \cite{hao24}.
+That thesis also contains some additional results reinforcing the conclusions presented in this article.
 
@@
 
-\begin{figure}[H]
-\centering
-\includegraphics[width=0.9\textwidth]{thesis-figs/cond_SMC_PAL_realdat.png}
-\caption{\label{fig:cond_SMC_PAL_realdat}Conditional log-likelihoods computed using two methods for the real rotavirus dataset, with the SMC method replicated 36 times due to its high variances.
-The blue line represents overlapping results from 36 SMC computations, while the red line is derived from PAL.
-The main sources of likelihood shortfall of SMC are at time points $t=1, 2, 3, 11, 81, 194$, and $325-333$.}
-\end{figure}
-
-\vspace{-8mm}
-\subsection{PAL versus SMC: When the Model is Correctly Specified}\label{section3}
-
-Here, we utilize 100 simulated datasets from the OvOv model with parameters set at the MLE, thereby eliminating model misspecification. We then examine the likelihood values computed by PAL and SMC.
-
-\begin{table}[htbp] % The placement specifier can be [h!tbp]
-\centering % Centers the table
-\caption{Average log-likelihood at the MLE at the 100 simulated datasets from OvOv computed by two filtering methods, employed 36 replicates and 50,000 particles} % This is effectively the table title
-\label{tab:ovovtrue} % For referencing this table elsewhere in your text
-\begin{tabular}{|c c c|} 
- \hline
-  & PAL & SMC  \\ 
- \hline
- OvOv & -6231.97 & -6224.245 \\ 
- \hline
-\end{tabular}
-\end{table}
-\begin{figure}[H]
-\centering
-\vspace{-7mm}
-\includegraphics[width=0.55\textwidth]{thesis-figs/PAL_vs_SMC_ovov_100.png}
-\caption{\label{fig:PAL_vs_SMC_ovov_100}Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model.
-The red line is the line $Lik(pfilter)=Lik(palfilter)$.
-We can see on the simulated data, SMC gives consistently higher likelihood estimates than PAL.}
-\end{figure}
-
-Notably, Section~\ref{misspecified} reveals that when the model confronts actual rotavirus data, it encounters misspecification issues, particularly at times $t=1, \,\,2, \,\,3, \,\,11, \,\,81, \,\,194, \,\,325-333$, leading to shortfalls in likelihood.
-Appendix \ref{appendB} discusses this phenomenon, attributing it to certain data points that the model fails to adequately represent.
-This issue is especially pronounced at the initial time points.
-Particularly, in the initial time points $t=1,2,3$, the likelihood deviates from PAL estimates for about more than 100 log-likelihood units, suggesting an improper choice of initial distribution that might favor the PAL estimate.
-Consequently, we have opted to estimate the initial distribution \(\lambda_0 = (S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}, R_{30})\), where $S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}$, and $R_{30}$ represent the number of individuals in each compartment at time $t=0$.
-The results of this new estimation can be found in Table \ref{tab:mlebywwr}.
-Importantly, the maximization of the likelihood conducted via the Iterated Filtering (mif2) algorithm \citep{pomppackagepaper} procures a higher likelihood at the estimated MLE than that estimated by PAL, and a different yet comparable set of estimates is obtained.
-In terms of parameters pertinent to epidemiological interests, i.e.
-the first 7 rows in Table \ref{tab:mlebywwr}, we observe a lower force of infection parameter for the first age group, $\beta_1$, alongside a higher force of infection for the third age group, $\beta_3$.
-Additionally, we detect variations in the phase parameter $\phi$, which directly affects the seasonal time points, as well as in the stochastic parameter $\sigma_{\xi}$ and the initial compartmental distributions.
-The modifications to the initial distributions aim to rectify inadequate initialization, while the other adjustments address model misspecifications identified in Section \ref{misspecified}.
-Changes in the remaining parameter set are negligible.
-
-PAL is an approximate filtering method used in scenarios where SMC is computationally prohibitive.
-It approximates the posterior distribution by a Poisson approximation.
-Due to its approximative nature, it does not perfectly capture the true posterior distribution.
-SMC, however, often referred to as the "perfect filter," uses a large set of particles and sophisticated resampling techniques to theoretically accurately approximate the posterior distribution.
-It is computationally more intensive than PAL but tends to provide a satisfying approximation, especially in complex models that are not compatible with PAL in general.
-When PAL and SMC are applied to a well-specified model meaning the model with the latent process model being correctly represented, SMC outperforms PAL because it better captures the nuances of the posterior distribution.
-This is why SMC might show a "likelihood shortfall" compared to PAL when the model is correctly specified.
-This is due to the approximation error of PAL can't be improved arbitrarily in general \citep{annurev:/content/journals/10.1146/annurev-control-042920-015119}, and likelihood being a proper scoring rule.
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=\textwidth]{thesis-figs/plot_zoom_png.png}
-\caption{\label{fig:zoomedin_outlier_plot} Zoomed-in time-Cases plot at time $t=0-20, 75-85, 190-200$ and $320-340$.
-In the visualization, solid lines represent the true reported cases from the original rotavirus dataset, while the shaded strips with dashed boundaries-colored according to different groups—indicate enlarged "confidence intervals" for these cases.
-Specifically, the upper boundary of each strip corresponds to the upper 97.5\% percentile of 36 replicates of the filtered accumulated number of cases, $H_{kt}$, for $k=1,2,3$, and $t=1,2,...,416$, as introduced in Section \ref{observationmodel}.
-This boundary is further adjusted by the upper 97.5\% percentile of the reporting rate $q_t$, estimated by \cite{wwr} in Table \ref{tab:mlebywwr}.
-The reporting rate $q_t$ is assumed to follow a truncated normal distribution with a mean of 0.07 and a standard deviation of $\sigma_q = 0.021$, resulting in a lower 2.5\% percentile of 0.029 and an upper 97.5\% percentile of 0.111.
-The methodology for establishing the lower boundary of the strip mirrors that of the upper boundary.
-Although this approach is not entirely rigorous, each strip is intended to encompass our extended confidence interval for the true reported cases.
-Consequently, data points that fall outside these strips give us a intuition of where the model misspecifications may occur.}
-\end{figure}
-
-As illustrated in Figure \ref{fig:cond_SMC_PAL_realdat}, shortfalls are observed at times $t=1,\,\,2,\,\,3,\,\,11,\,\,81,\,\,194,$ and $325-333$.
-These discrepancies are primarily attributable to outliers that the model fails to accommodate.
-In the displayed results, it is immediately apparent that at the initial times $t=1, \,\,2, \,\,3$ all the true reported cases deviated from the expected range, falling outside the designated strip.
-Specifically, at time $t=11$, the model is misspecified for case 1, depicted by the solid red line.
-Similarly, at times $t=81,\,\, 82$, the model is misspecified for case 3, as indicated by the solid blue line that extended beyond the strip's boundaries.
-At $t=194$, the solid green line denotes case 2 as where the model is misspecified.
-During the period from $t=325-333$, all three observed cases are where the model misspecifications happen.
-These deviations provide a clear illustration of the likelihood shortfall of the sequential Monte Carlo (SMC) method when the model does not align with the data, suggesting a model misspecification.
-
 \bibliography{bib-pal}

### 2024-07-22 ms.Rnw
compilation, variable naming, and table variable read-in

@@
 \begin{document}
+\SweaveOpts{concordance=TRUE}
 
@@
 library(magrittr)
-library(plyr)
-library(reshape2)
 library(ggplot2)
@@
 library(doParallel)
-registerDoParallel()
-stopifnot(packageVersion("pomp") >= "1.7")
+library(dplyr)
+source("choose_n_fits.R")
 
 RUN_LEVEL = 1
-NP_MIF       = switch(RUN_LEVEL, 4, 50000)
+NP_MIF       = switch(RUN_LEVEL, 5000, 50000)
 NMIF         = switch(RUN_LEVEL, 4,    100)
 ncores       = switch(RUN_LEVEL, 4,    36)
+top_n_fits   = switch(RUN_LEVEL, 1,    12)
 @
@@
 
-<<OvOv-model>>=
+<<OvOv-model, include=False, massage=False, warning=False>>=
 # measurement model 
@@
 
+# WWR's rinit
+rinit_wwr <- Csnippet("
+    S1=3876549;
+    S2=57139612;
+    S3=19573727;
+    I1=30351;
+    I2=871;
+    I3=2550;
+    R1=1315221;
+    R2=302852;
+    R3=131092;
+    H1 = 0;
+    H2 = 0;
+    H3 = 0;
+")
+
+# x <- c(3876549,57139612,19573727,30351,871,2550,1315221,302852,131092)
+# x/sum(x)
+
 # Set to MLE
-params_stocks_stst_mle <- params_fixed
+wwr_mle <- params_fixed
 
-params_stocks_stst_mle["beta1"] <- 11.48
-params_stocks_stst_mle["beta2"] <- 0.25
-params_stocks_stst_mle["beta3"] <- 0.35
-params_stocks_stst_mle["phi"] <- 0.14
-params_stocks_stst_mle["beta11"] <- 0.16
-params_stocks_stst_mle["sigma_q"] <- 0.021
-params_stocks_stst_mle["sigma_xi"] <- 66.89
-params_stocks_stst_mle["S10"] <- 0.047061
-params_stocks_stst_mle["I10"] <- 0.000368
-params_stocks_stst_mle["R10"] <- 0.015967
-params_stocks_stst_mle["S20"] <- 0.015967
-params_stocks_stst_mle["I20"] <- 0.000011
-params_stocks_stst_mle["R20"] <- 0.003677
-params_stocks_stst_mle["S30"] <- 0.237624
-params_stocks_stst_mle["I30"] <- 0.000031
-params_stocks_stst_mle["R30"] <- 0.001591
-@
+wwr_mle <- c(params_fixed, 
+             "beta1" = 11.48,
+             "beta2" = 0.25,
+             "beta3" = 0.35,
+             "phi" = 0.14,
+             "beta11" = 0.16,
+             "sigma_q" = 0.021,
+             "sigma_xi" = 66.89,
+             "S10" = 4.706102e-02,
+             "S20" = 6.936707e-01,
+             "S30" = 2.376236e-01,
+             "I10" = 3.684589e-04,
+             "I20" = 1.057388e-05,
+             "I30" = 3.095681e-05,
+             "R10" = 1.596669e-02,
+             "R20" = 3.676601e-03,
+             "R30" = 1.591447e-03)
 
-% Maximization Round 1
-
-<<maximization1>>=
 pt <- pomp::parameter_trans(
@@
                     "q1", "q2", "q3"),
-     paramnames = names(params_stocks_stst_mle),
+     paramnames = names(wwr_mle),
      accumvars=c("H1", "H2", "H3"),
@@
      partrans = pt,
-     params = params_stocks_stst_mle
+     params = wwr_mle
 ) -> sir
@@
                                   shared=NULL,
-                                  specific=params_stocks_stst_mle |> 
+                                  specific=wwr_mle |> 
                                     as.matrix() |>
@@
 )
+@
 
-require(doParallel)
-cores <- ncores
-registerDoParallel(cores)
-mcopts <- list(set.seed=TRUE)
+% Maximization Round 1
 
+<<maximization1, include=False, massage=False, warning=False>>=
+set.seed(123)
 sir_box <- rbind(
@@
 
+require(doParallel)
+registerDoParallel(ncores)
+mcopts <- list(set.seed=TRUE)
+
 bake(file = "output/round_01/ovov_mif_01.rds",{ 
-  mifs_global <- foreach(i=1:cores,.packages='pomp', .options.multicore=mcopts) %dopar% {
+  mifs_global_01 <- foreach(i=1:ncores,.packages='pomp', .options.multicore=mcopts) %dopar% {
+    set.seed(112*i)
     panelPomp::mif2(
@@
   }
-  mifs_global
+  mifs_global_01
 })
@@
 bake(file = "output/round_01/ovov_01_el.rds",{
-  el <- measlespkg::eval_logLik(mifs_global, ncores=cores, np_pf = NP_MIF, nreps=cores)
+  el <- measlespkg::eval_logLik(mifs_global_01, ncores=ncores, np_pf = NP_MIF, nreps=ncores)
   el
@@
 
-<<maximization2>>=
-sd           = 2/3
-top_n_fits   = 12
-# measurement model 
-dmeas <- Csnippet("
-                  if (ISNA(cases1)) {
-                  lik = (give_log) ? 0 : 1;
-                  } else {
-                        lik =  dbinom(cases1, H1, q1, 1) +
-                        dbinom(cases2, H2, q2, 1) +
-                        dbinom(cases3, H3, q3, 1);
+<<maximization2, include=False, massage=False, warning=False>>=
+set.seed(1234)
+sd_02           = 2/3
 
-                    lik = (give_log) ? lik : exp(lik);
-
-                    }")
-rmeas <-  Csnippet("
-                    cases1 = rbinom(H1, q1);
-                    cases2 = rbinom(H2, q2);
-                    cases3 = rbinom(H3, q3);
-                  ")
-
-rproc <- Csnippet("
-    int I = I1 + I2 + I3;
-    int trans_S1[3], trans_S2[3], trans_S3[2], trans_I1[3], trans_I2[3], trans_I3[2], trans_R1[3], trans_R2[3], trans_R3[2];
-
-    double prob_S1[3],prob_I1[3],prob_R1[3],prob_S2[3],prob_I2[3],prob_R2[3],prob_S3[2],prob_I3[2],prob_R3[2];
-
-    double xi = rgamma(sigma_xi, 1/sigma_xi);
-
-    double kappa = (1 + beta11*cos(2*3.141593*t/52 + phi)) * xi;
-
-    // Define rate
-    prob_S1[0] = 1-exp(-dt*beta1*kappa*I/N); // 0->1
-    prob_S1[1] = 1-exp(-delta1*dt);
-    prob_S1[2] = exp(-delta1*dt) + exp(-dt*beta1*kappa*I/N) - 1;
-
-    prob_I1[0] = 1-exp(-gamma*dt);
-    prob_I1[1] = 1-exp(-delta1*dt);
-    prob_I1[2] = exp(-gamma*dt)+exp(-delta1*dt) - 1;
-
-    prob_R1[0] = 1 - exp(-omega*dt);  // E_1,t this goes back to S_1,(t+1)
-    prob_R1[1] = 1 - exp(-delta1*dt);
-    prob_R1[2] = exp(-omega*dt) + exp(-delta1*dt) - 1;
-
-    prob_S2[0] = 1-exp(-dt*beta2*kappa*I/N);
-    prob_S2[1] = 1-exp(-delta2*dt);
-    prob_S2[2] = exp(-delta2*dt) + exp(-dt*beta2*kappa*I/N) - 1;
-
-    prob_I2[0] = 1-exp(-dt*gamma);
-    prob_I2[1] = 1-exp(-dt*delta2);
-    prob_I2[2] = exp(-dt*gamma)+exp(-dt*delta2) - 1;
-
-    prob_R2[0] = 1 - exp(-dt*omega);  // E_1,t this goes back to S_1,(t+1)
-    prob_R2[1] = 1 - exp(-dt*delta2);
-    prob_R2[2] = exp(-dt*omega) + exp(-dt*delta2) - 1;
-
-    // For Age Group (3): Die first before transition;
-
-    int S3mD, I3mD, R3mD;
-
-    S3mD = rbinom(S3, 1-dt*mu); // S3 minus Death: mu is the death rate, so it's 1-mu here
-    I3mD = rbinom(I3, 1-dt*mu);
-    R3mD = rbinom(R3, 1-dt*mu);
-
-    prob_S3[0] = 1-exp(-dt*beta3*kappa*I/N);
-    prob_S3[1] = exp(-dt*beta3*kappa*I/N);
-
-    prob_I3[0] = 1 - exp(-dt*gamma);
-    prob_I3[1] = exp(-dt*gamma);
-
-    prob_R3[0] = 1 - exp(-dt*omega);
-    prob_R3[1] = exp(-dt*omega);
-
-    // Transition
-    // B: S->I
-    // C: I->R
-    // F: Aging: (1)->(2)->(3)
-    // E: R->S
-    // D: Death
-    //// Note: Here S_1, S_2... are all old value from (t-1)
-    rmultinom(S1, &prob_S1, 3, &trans_S1); // B, F, S-B-F
-    rmultinom(I1, &prob_I1, 3, &trans_I1); // C, F, I-C-F
-    rmultinom(R1, &prob_R1, 3, &trans_R1); // E, F, R-E-F
-
-    rmultinom(S2, &prob_S2, 3, &trans_S2); // B, F, S-B-F
-    rmultinom(I2, &prob_I2, 3, &trans_I2); // C, F, I-C-F
-    rmultinom(R2, &prob_R2, 3, &trans_R2); // E, F, R-E-F
-
-    rmultinom(S3mD, &prob_S3, 2, &trans_S3); // B, (S-D)-B
-    rmultinom(I3mD, &prob_I3, 2, &trans_I3); // C, (I-D)-C
-    rmultinom(R3mD, &prob_R3, 2, &trans_R3); // E, (R-D)-E
-
-    S1 = trans_S1[2] + trans_R1[0] + rpois(4*1025.7); // Include Birth
-    I1 = trans_I1[2] + trans_S1[0];
-    R1 = trans_R1[2] + trans_I1[0];
-
-    S2 = trans_S2[2] + trans_R2[0] + trans_S1[1]; // Include Aging
-    I2 = trans_I2[2] + trans_S2[0] + trans_I1[1];
-    R2 = trans_R2[2] + trans_I2[0] + trans_R1[1];
-
-    S3 = trans_S3[1] + trans_R3[0] + trans_S2[1]; // Include Aging
-    I3 = trans_I3[1] + trans_S3[0] + trans_I2[1];
-    R3 = trans_R3[1] + trans_I3[0] + trans_R2[1];
-
-    //Accumvar
-    H1 += trans_S1[0];
-    H2 += trans_S2[0];
-    H3 += trans_S3[0];
-
-    q1 = -1; 
-    while(q1 < 0 || q1 > 1){
-      q1 = rnorm(0.07, sigma_q);
-    }
-
-    q2 = -1; 
-    while(q2 < 0 || q2 > 1){
-      q2 = rnorm(0.07, sigma_q);
-    }
-
-    q3 = -1; 
-    while(q3 < 0 || q3 > 1){
-      q3 = rnorm(0.07, sigma_q);
-    }
-")
-
-# define parameters (without betas)
-params_fixed <- c(gamma=1, delta1=1/(5*52),delta2=1/(55*52), alpha=1/(78.86912*52), 
-                  mu=0, N=82372825, omega=1/(1*52))
-# WWR's rinit
-rinit <- Csnippet("
-    double m = N/(S10+I10+R10+S20+I20+R20+S30+I30+R30);
-    I1=nearbyint(m*I10);
-    I2=nearbyint(m*I20);
-    I3=nearbyint(m*I30);
-    S1=nearbyint(m*S10);
-    S2=nearbyint(m*S20);
-    S3=nearbyint(m*S30);
-    R1=nearbyint(m*R10);
-    R2=nearbyint(m*R20);
-    R3=nearbyint(m*R30);
-    H1 = 0;
-    H2 = 0;
-    H3 = 0;
-")
-
-# Set to MLE
-params_stocks_stst_mle <- params_fixed
-
-params_stocks_stst_mle["beta1"] <- 11.48
-params_stocks_stst_mle["beta2"] <- 0.25
-params_stocks_stst_mle["beta3"] <- 0.35
-params_stocks_stst_mle["phi"] <- 0.14
-params_stocks_stst_mle["beta11"] <- 0.16
-params_stocks_stst_mle["sigma_q"] <- 0.021
-params_stocks_stst_mle["sigma_xi"] <- 66.89
-params_stocks_stst_mle["S10"] <- 0.047061
-params_stocks_stst_mle["I10"] <- 0.000368
-params_stocks_stst_mle["R10"] <- 0.015967
-params_stocks_stst_mle["S20"] <- 0.015967
-params_stocks_stst_mle["I20"] <- 0.000011
-params_stocks_stst_mle["R20"] <- 0.003677
-params_stocks_stst_mle["S30"] <- 0.237624
-params_stocks_stst_mle["I30"] <- 0.000031
-params_stocks_stst_mle["R30"] <- 0.001591
-
-pt <- pomp::parameter_trans(
-  log = c("beta1","beta2","beta3","sigma_q","sigma_xi"),
-  logit=c("beta11"),
-  barycentric=c("S10","I10","R10",
-                "S20","I20","R20",
-                "S30","I30","R30"),
-  toEst= pomp::Csnippet("T_phi = logit(phi/(M_2PI));"),
-  fromEst= pomp::Csnippet("phi = M_2PI*expit(T_phi);")
-)
-
-read.table("real_rotavirus_metadata.txt") %>%
-  rbind(data.frame(time=0,cases1=NA,cases2=NA,cases3=NA)) %>%
-  arrange(time) -> dat
-
-pomp(data = dat,
-     times="time",
-     t0=0,
-     dmeasure = dmeas,
-     rmeasure = rmeas,
-     rprocess = discrete_time(step.fun = rproc, delta.t = 1/4),
-     statenames = c("S1", "I1", "R1", "H1", 
-                    "S2", "I2", "R2", "H2",
-                    "S3", "I3", "R3", "H3", 
-                    "q1", "q2", "q3"),
-     paramnames = names(params_stocks_stst_mle),
-     accumvars=c("H1", "H2", "H3"),
-     rinit=rinit,
-     partrans = pt,
-     params = params_stocks_stst_mle
-) -> sir
-
-sir_panel <- panelPomp::panelPomp(list(unit1=sir),
-                                  shared=NULL,
-                                  specific=params_stocks_stst_mle |> 
-                                    as.matrix() |>
-                                    `colnames<-`("unit1")
-)
+el_01 <- readRDS("output/round_01/ovov_01_el.rds")
+starting_values_02 <- choose_n_fits(el_01, top_n_fits, ncores)
 
 require(doParallel)
-cores <- ncores
-registerDoParallel(cores)
+registerDoParallel(ncores)
 mcopts <- list(set.seed=TRUE)
 
-### Next-round code
-el <- readRDS("output/round_01/ovov_01_el.rds")
-x <- na.omit(el$fits)
-score_total = x$logLik
-ranking_total = order(score_total, decreasing = TRUE)[1:top_n_fits]
-
-best_fits = dplyr::select(
-  x[ranking_total,], -"logLik", -"se"
-)
-
-recycle_vec = sort(rep_len(1:top_n_fits, cores))
-full_best_fit <- best_fits[recycle_vec, ]
-
-coef_names <- colnames(full_best_fit)
-colnames(full_best_fit) <- gsub(".{7}$","",coef_names)
-
-starting_values <- vector(cores, mode="list")
-
-for(i in 1:cores){
-  t(full_best_fit[i, ])  |> 
-    as.matrix() |>
-    `colnames<-`("unit1") -> starting_values[[i]] 
-}
-
-mifs_global <- foreach(i=1:cores,.packages='pomp', .options.multicore=mcopts) %dopar% {
+bake(file = "output/round_02/ovov_mif_02.rds",{ 
+  mifs_global_02 <- foreach(i=1:ncores,.packages='pomp', .options.multicore=mcopts) %dopar% {
   panelPomp::mif2(
@@
     cooling.fraction.50 = 0.5,
-    rw.sd = rw_sd(beta1=0.01*sd,beta2=0.01*sd,beta3=0.01*sd,
-                  beta11=0.01*sd,phi=0.01*sd,sigma_q=0.01*sd,sigma_xi=0.01*sd,
-                  S10=ivp(0.24)*sd,I10=ivp(0.24)*sd,R10=ivp(0.24)*sd,
-                  S20=ivp(0.24)*sd,I20=ivp(0.24)*sd,R20=ivp(0.24)*sd,
-                  S30=ivp(0.24)*sd,I30=ivp(0.24)*sd,R30=ivp(0.24)*sd),
+    rw.sd = rw_sd(beta1=0.01*sd_02,beta2=0.01*sd_02,beta3=0.01*sd_02,
+                  beta11=0.01*sd_02,phi=0.01*sd_02,sigma_q=0.01*sd_02,sigma_xi=0.01*sd_02,
+                  S10=ivp(0.24)*sd_02,I10=ivp(0.24)*sd_02,R10=ivp(0.24)*sd_02,
+                  S20=ivp(0.24)*sd_02,I20=ivp(0.24)*sd_02,R20=ivp(0.24)*sd_02,
+                  S30=ivp(0.24)*sd_02,I30=ivp(0.24)*sd_02,R30=ivp(0.24)*sd_02),
     cooling.type = "geometric",
@@
     shared.start = numeric(0),
-    specific.start = starting_values[[i]],
+    specific.start = starting_values_02[[i]],
     block = F
@@
 }
-
-bake(file = "output/round_02/ovov_mif_02.rds",{ 
-  mifs_global
+  mifs_global_02
 })
@@
 bake(file = "output/round_02/ovov_02_el.rds",{
-  el <- measlespkg::eval_logLik(mifs_global, ncores = cores, np_pf = NP_MIF, nreps=cores)
+  el <- measlespkg::eval_logLik(mifs_global_02, ncores = ncores, np_pf = NP_MIF, nreps=ncores)
   el
@@
 
-<<maximization3>>=
-sd           = (2/3)^2
-top_n_fits   = 12
-# measurement model 
-dmeas <- Csnippet("
-                  if (ISNA(cases1)) {
-                  lik = (give_log) ? 0 : 1;
-                  } else {
-                        lik =  dbinom(cases1, H1, q1, 1) +
-                        dbinom(cases2, H2, q2, 1) +
-                        dbinom(cases3, H3, q3, 1);
+<<maximization3, include=False, massage=False, warning=False>>=
+set.seed(12345)
+sd_03           = (2/3)^2
 
-                    lik = (give_log) ? lik : exp(lik);
-
-                    }")
-rmeas <-  Csnippet("
-                    cases1 = rbinom(H1, q1);
-                    cases2 = rbinom(H2, q2);
-                    cases3 = rbinom(H3, q3);
-                  ")
-
-rproc <- Csnippet("
-    int I = I1 + I2 + I3;
-    int trans_S1[3], trans_S2[3], trans_S3[2], trans_I1[3], trans_I2[3], trans_I3[2], trans_R1[3], trans_R2[3], trans_R3[2];
-
-    double prob_S1[3],prob_I1[3],prob_R1[3],prob_S2[3],prob_I2[3],prob_R2[3],prob_S3[2],prob_I3[2],prob_R3[2];
-
-    double xi = rgamma(sigma_xi, 1/sigma_xi);
-
-    double kappa = (1 + beta11*cos(2*3.141593*t/52 + phi)) * xi;
-
-    // Define rate
-    prob_S1[0] = 1-exp(-dt*beta1*kappa*I/N); // 0->1
-    prob_S1[1] = 1-exp(-delta1*dt);
-    prob_S1[2] = exp(-delta1*dt) + exp(-dt*beta1*kappa*I/N) - 1;
-
-    prob_I1[0] = 1-exp(-gamma*dt);
-    prob_I1[1] = 1-exp(-delta1*dt);
-    prob_I1[2] = exp(-gamma*dt)+exp(-delta1*dt) - 1;
-
-    prob_R1[0] = 1 - exp(-omega*dt);  // E_1,t this goes back to S_1,(t+1)
-    prob_R1[1] = 1 - exp(-delta1*dt);
-    prob_R1[2] = exp(-omega*dt) + exp(-delta1*dt) - 1;
-
-    prob_S2[0] = 1-exp(-dt*beta2*kappa*I/N);
-    prob_S2[1] = 1-exp(-delta2*dt);
-    prob_S2[2] = exp(-delta2*dt) + exp(-dt*beta2*kappa*I/N) - 1;
-
-    prob_I2[0] = 1-exp(-dt*gamma);
-    prob_I2[1] = 1-exp(-dt*delta2);
-    prob_I2[2] = exp(-dt*gamma)+exp(-dt*delta2) - 1;
-
-    prob_R2[0] = 1 - exp(-dt*omega);  // E_1,t this goes back to S_1,(t+1)
-    prob_R2[1] = 1 - exp(-dt*delta2);
-    prob_R2[2] = exp(-dt*omega) + exp(-dt*delta2) - 1;
-
-    // For Age Group (3): Die first before transition;
-
-    int S3mD, I3mD, R3mD;
-
-    S3mD = rbinom(S3, 1-dt*mu); // S3 minus Death: mu is the death rate, so it's 1-mu here
-    I3mD = rbinom(I3, 1-dt*mu);
-    R3mD = rbinom(R3, 1-dt*mu);
-
-    prob_S3[0] = 1-exp(-dt*beta3*kappa*I/N);
-    prob_S3[1] = exp(-dt*beta3*kappa*I/N);
-
-    prob_I3[0] = 1 - exp(-dt*gamma);
-    prob_I3[1] = exp(-dt*gamma);
-
-    prob_R3[0] = 1 - exp(-dt*omega);
-    prob_R3[1] = exp(-dt*omega);
-
-    // Transition
-    // B: S->I
-    // C: I->R
-    // F: Aging: (1)->(2)->(3)
-    // E: R->S
-    // D: Death
-    //// Note: Here S_1, S_2... are all old value from (t-1)
-    rmultinom(S1, &prob_S1, 3, &trans_S1); // B, F, S-B-F
-    rmultinom(I1, &prob_I1, 3, &trans_I1); // C, F, I-C-F
-    rmultinom(R1, &prob_R1, 3, &trans_R1); // E, F, R-E-F
-
-    rmultinom(S2, &prob_S2, 3, &trans_S2); // B, F, S-B-F
-    rmultinom(I2, &prob_I2, 3, &trans_I2); // C, F, I-C-F
-    rmultinom(R2, &prob_R2, 3, &trans_R2); // E, F, R-E-F
-
-    rmultinom(S3mD, &prob_S3, 2, &trans_S3); // B, (S-D)-B
-    rmultinom(I3mD, &prob_I3, 2, &trans_I3); // C, (I-D)-C
-    rmultinom(R3mD, &prob_R3, 2, &trans_R3); // E, (R-D)-E
-
-    S1 = trans_S1[2] + trans_R1[0] + rpois(4*1025.7); // Include Birth
-    I1 = trans_I1[2] + trans_S1[0];
-    R1 = trans_R1[2] + trans_I1[0];
-
-    S2 = trans_S2[2] + trans_R2[0] + trans_S1[1]; // Include Aging
-    I2 = trans_I2[2] + trans_S2[0] + trans_I1[1];
-    R2 = trans_R2[2] + trans_I2[0] + trans_R1[1];
-
-    S3 = trans_S3[1] + trans_R3[0] + trans_S2[1]; // Include Aging
-    I3 = trans_I3[1] + trans_S3[0] + trans_I2[1];
-    R3 = trans_R3[1] + trans_I3[0] + trans_R2[1];
-
-    //Accumvar
-    H1 += trans_S1[0];
-    H2 += trans_S2[0];
-    H3 += trans_S3[0];
-
-    q1 = -1; 
-    while(q1 < 0 || q1 > 1){
-      q1 = rnorm(0.07, sigma_q);
-    }
-
-    q2 = -1; 
-    while(q2 < 0 || q2 > 1){
-      q2 = rnorm(0.07, sigma_q);
-    }
-
-    q3 = -1; 
-    while(q3 < 0 || q3 > 1){
-      q3 = rnorm(0.07, sigma_q);
-    }
-")
-
-# define parameters (without betas)
-params_fixed <- c(gamma=1, delta1=1/(5*52),delta2=1/(55*52), alpha=1/(78.86912*52), 
-                  mu=0, N=82372825, omega=1/(1*52))
-# WWR's rinit
-rinit <- Csnippet("
-    double m = N/(S10+I10+R10+S20+I20+R20+S30+I30+R30);
-    I1=nearbyint(m*I10);
-    I2=nearbyint(m*I20);
-    I3=nearbyint(m*I30);
-    S1=nearbyint(m*S10);
-    S2=nearbyint(m*S20);
-    S3=nearbyint(m*S30);
-    R1=nearbyint(m*R10);
-    R2=nearbyint(m*R20);
-    R3=nearbyint(m*R30);
-    H1 = 0;
-    H2 = 0;
-    H3 = 0;
-")
-
-# Set to MLE
-params_stocks_stst_mle <- params_fixed
-
-params_stocks_stst_mle["beta1"] <- 11.48
-params_stocks_stst_mle["beta2"] <- 0.25
-params_stocks_stst_mle["beta3"] <- 0.35
-params_stocks_stst_mle["phi"] <- 0.14
-params_stocks_stst_mle["beta11"] <- 0.16
-params_stocks_stst_mle["sigma_q"] <- 0.021
-params_stocks_stst_mle["sigma_xi"] <- 66.89
-params_stocks_stst_mle["S10"] <- 0.047061
-params_stocks_stst_mle["I10"] <- 0.000368
-params_stocks_stst_mle["R10"] <- 0.015967
-params_stocks_stst_mle["S20"] <- 0.015967
-params_stocks_stst_mle["I20"] <- 0.000011
-params_stocks_stst_mle["R20"] <- 0.003677
-params_stocks_stst_mle["S30"] <- 0.237624
-params_stocks_stst_mle["I30"] <- 0.000031
-params_stocks_stst_mle["R30"] <- 0.001591
-
-pt <- pomp::parameter_trans(
-  log = c("beta1","beta2","beta3","sigma_q","sigma_xi"),
-  logit=c("beta11"),
-  barycentric=c("S10","I10","R10",
-                "S20","I20","R20",
-                "S30","I30","R30"),
-  toEst= pomp::Csnippet("T_phi = logit(phi/(M_2PI));"),
-  fromEst= pomp::Csnippet("phi = M_2PI*expit(T_phi);")
-)
-
-read.table("real_rotavirus_metadata.txt") %>%
-  rbind(data.frame(time=0,cases1=NA,cases2=NA,cases3=NA)) %>%
-  arrange(time) -> dat
-
-pomp(data = dat,
-     times="time",
-     t0=0,
-     dmeasure = dmeas,
-     rmeasure = rmeas,
-     rprocess = discrete_time(step.fun = rproc, delta.t = 1/4),
-     statenames = c("S1", "I1", "R1", "H1", 
-                    "S2", "I2", "R2", "H2",
-                    "S3", "I3", "R3", "H3", 
-                    "q1", "q2", "q3"),
-     paramnames = names(params_stocks_stst_mle),
-     accumvars=c("H1", "H2", "H3"),
-     rinit=rinit,
-     partrans = pt,
-     params = params_stocks_stst_mle
-) -> sir
-
-sir_panel <- panelPomp::panelPomp(list(unit1=sir),
-                                  shared=NULL,
-                                  specific=params_stocks_stst_mle |> 
-                                    as.matrix() |>
-                                    `colnames<-`("unit1")
-)
+el_02 <- readRDS("output/round_02/ovov_02_el.rds")
+starting_values_03 <- choose_n_fits(el_02, top_n_fits, ncores)
 
 require(doParallel)
-cores <- ncores
-registerDoParallel(cores)
+registerDoParallel(ncores)
 mcopts <- list(set.seed=TRUE)
 
-### Next-round code
-el <- readRDS("output/round_02/ovov_02_el.rds")
-x <- na.omit(el$fits)
-score_total = x$logLik
-ranking_total = order(score_total, decreasing = TRUE)[1:top_n_fits]
-
-best_fits = dplyr::select(
-  x[ranking_total,], -"logLik", -"se"
-)
-
-recycle_vec = sort(rep_len(1:top_n_fits, cores))
-full_best_fit <- best_fits[recycle_vec, ]
-
-coef_names <- colnames(full_best_fit)
-colnames(full_best_fit) <- gsub(".{7}$","",coef_names)
-
-starting_values <- vector(cores, mode="list")
-
-for(i in 1:cores){
-  t(full_best_fit[i, ])  |> 
-    as.matrix() |>
-    `colnames<-`("unit1") -> starting_values[[i]] 
-}
-
-mifs_global <- foreach(i=1:cores,.packages='pomp', .options.multicore=mcopts) %dopar% {
+bake(file = "output/round_03/ovov_mif_03.rds",{ 
+  mifs_global_03 <- foreach(i=1:ncores,.packages='pomp', .options.multicore=mcopts) %dopar% {
   panelPomp::mif2(
@@
     cooling.fraction.50 = 0.5,
-    rw.sd = rw_sd(beta1=0.01*sd,beta2=0.01*sd,beta3=0.01*sd,
-                  beta11=0.01*sd,phi=0.01*sd,sigma_q=0.01*sd,sigma_xi=0.01*sd,
-                  S10=ivp(0.24)*sd,I10=ivp(0.24)*sd,R10=ivp(0.24)*sd,
-                  S20=ivp(0.24)*sd,I20=ivp(0.24)*sd,R20=ivp(0.24)*sd,
-                  S30=ivp(0.24)*sd,I30=ivp(0.24)*sd,R30=ivp(0.24)*sd),
+    rw.sd = rw_sd(beta1=0.01*sd_03,beta2=0.01*sd_03,beta3=0.01*sd_03,
+                  beta11=0.01*sd_03,phi=0.01*sd_03,sigma_q=0.01*sd_03,sigma_xi=0.01*sd_03,
+                  S10=ivp(0.24)*sd_03,I10=ivp(0.24)*sd_03,R10=ivp(0.24)*sd_03,
+                  S20=ivp(0.24)*sd_03,I20=ivp(0.24)*sd_03,R20=ivp(0.24)*sd_03,
+                  S30=ivp(0.24)*sd_03,I30=ivp(0.24)*sd_03,R30=ivp(0.24)*sd_03),
     cooling.type = "geometric",
@@
     shared.start = numeric(0),
-    specific.start = starting_values[[i]],
+    specific.start = starting_values_03[[i]],
     block = F
@@
 }
-
-bake(file = "output/round_03/ovov_mif_03.rds",{ 
-  mifs_global
+  mifs_global_03
 })
@@
 bake(file = "output/round_03/ovov_03_el.rds",{
-  el <- measlespkg::eval_logLik(mifs_global, ncores = cores, np_pf = NP_MIF, nreps=cores)
+  el <- measlespkg::eval_logLik(mifs_global_03, ncores = ncores, np_pf = NP_MIF, nreps=ncores)
   el
 })
+@
+
+<<estimated-mle, include=False, massage=False, warning=False>>=
+mif_el <- readRDS("output/round_03/ovov_03_el.rds") # Can be changed if more rounds needed
+est_mle_vec <- mif_el$fits[which.max(mif_el$fits$logLik),]
 @
@@
       \midrule
-      \(\beta_1\)    & 12.15 & 12.74 & 11.48 & $\Sexpr{'test'}$\\
-      \(\beta_2\)    & 0.22  & 0.21  & 0.25  & $\Sexpr{'test'}$\\
-      \(\beta_3\)    & 0.34  & 0.31  & 0.35  & $\Sexpr{'test'}$\\
-      \(\phi\)       & 0.017 & 0.14  & 0.14  & $\Sexpr{'test'}$\\
-      \(\rho\)       & 0.022 & 0.19  & 0.16  & $\Sexpr{'test'}$\\
-      \(\sigma^2_q\) & n/a   & 0.042 & 0.021 & $\Sexpr{'test'}$\\
-      \(\sigma_\xi\) & n/a   & n/a   & 66.89 & $\Sexpr{'test'}$\\
+      \(\beta_1\)    & 12.15 & 12.74 & 11.48 & $\Sexpr{est_mle_vec["beta1[unit1]"]}$\\
+      \(\beta_2\)    & 0.22  & 0.21  & 0.25  & $\Sexpr{est_mle_vec["beta2[unit1]"]}$\\
+      \(\beta_3\)    & 0.34  & 0.31  & 0.35  & $\Sexpr{est_mle_vec["beta3[unit1]"]}$\\
+      \(\phi\)       & 0.017 & 0.14  & 0.14  & $\Sexpr{est_mle_vec["phi[unit1]"]}$\\
+      \(\rho\)       & 0.022 & 0.19  & 0.16  & $\Sexpr{est_mle_vec["rho[unit1]"]}$\\
+      \(\sigma^2_q\) & n/a   & 0.042 & 0.021 & $\Sexpr{est_mle_vec["sigma_q[unit1]"]}$\\
+      \(\sigma_\xi\) & n/a   & n/a   & 66.89 & $\Sexpr{est_mle_vec["sigma_xi[unit1]"]}$\\
       \addlinespace
-      \multicolumn{1}{l}{\(S_{10}\)} & \multicolumn{3}{c}{$3876549^*$} & $5885201$\\
-      \multicolumn{1}{l}{\(I_{10}\)} & \multicolumn{3}{c}{$30351$}     & $5041$\\
-      \multicolumn{1}{l}{\(R_{10}\)} & \multicolumn{3}{c}{$1315221$}   & $721006$\\
-      \multicolumn{1}{l}{\(S_{20}\)} & \multicolumn{3}{c}{$57139612$}  & $44546887$\\
-      \multicolumn{1}{l}{\(I_{20}\)} & \multicolumn{3}{c}{$871$}       & $22$\\
-      \multicolumn{1}{l}{\(R_{20}\)} & \multicolumn{3}{c}{$302852$}    & $26360811$\\
-      \multicolumn{1}{l}{\(S_{30}\)} & \multicolumn{3}{c}{$19573727$}  & $4788420$\\
-      \multicolumn{1}{l}{\(I_{30}\)} & \multicolumn{3}{c}{$2550$}      & $131$\\
-      \multicolumn{1}{l}{\(R_{30}\)} & \multicolumn{3}{c}{$131092$}    & $65306$\\
+      \multicolumn{1}{l}{\(S_{10}\)} & \multicolumn{3}{c}{$3876549^*$} & $$\Sexpr{est_mle_vec["S10[unit1]"]}$$\\
+      \multicolumn{1}{l}{\(I_{10}\)} & \multicolumn{3}{c}{$30351$}     & $$\Sexpr{est_mle_vec["I10[unit1]"]}$$\\
+      \multicolumn{1}{l}{\(R_{10}\)} & \multicolumn{3}{c}{$1315221$}   & $$\Sexpr{est_mle_vec["R10[unit1]"]}$$\\
+      \multicolumn{1}{l}{\(S_{20}\)} & \multicolumn{3}{c}{$57139612$}  & $$\Sexpr{est_mle_vec["S20[unit1]"]}$$\\
+      \multicolumn{1}{l}{\(I_{20}\)} & \multicolumn{3}{c}{$871$}       & $$\Sexpr{est_mle_vec["sigma_xi[unit1]"]}$$\\
+      \multicolumn{1}{l}{\(R_{20}\)} & \multicolumn{3}{c}{$302852$}    & $$\Sexpr{est_mle_vec["sigma_xi[unit1]"]}$$\\
+      \multicolumn{1}{l}{\(S_{30}\)} & \multicolumn{3}{c}{$19573727$}  & $$\Sexpr{est_mle_vec["sigma_xi[unit1]"]}$$\\
+      \multicolumn{1}{l}{\(I_{30}\)} & \multicolumn{3}{c}{$2550$}      & $$\Sexpr{est_mle_vec["sigma_xi[unit1]"]}$$\\
+      \multicolumn{1}{l}{\(R_{30}\)} & \multicolumn{3}{c}{$131092$}    & $$\Sexpr{est_mle_vec["sigma_xi[unit1]"]}$$\\
       \midrule
       \addlinespace
-      \textbf{AIC}  & $98866.65$ & $15154.75$ & $13778.08$ & $\Sexpr{'test'}$\\
+      \textbf{AIC}  & $98866.65$ & $15154.75$ & $13778.08$ & $\Sexpr{(-2)*((length(wwr_mle) - length(params_fixed)) - est_mle_vec["logLik"])}$\\
       \addlinespace

### 2024-07-22 choose_n_fits.R

@@ -1,0 +1,28 @@
+# Choose top n fits
+
+choose_n_fits = function(el, top_n_fits, ncores){
+  ### Next-round code
+  fit_mat <- na.omit(el$fits)
+  score_total = fit_mat$logLik
+  ranking_total = order(score_total, decreasing = TRUE)[1:top_n_fits]
+  
+  best_fits = dplyr::select(
+    fit_mat[ranking_total,], -"logLik", -"se"
+  )
+  
+  recycle_vec = sort(rep_len(1:top_n_fits, ncores))
+  full_best_fit <- best_fits[recycle_vec, ] 
+  
+  coef_names <- colnames(full_best_fit)
+  colnames(full_best_fit) <- gsub(".{7}$","",coef_names)
+  
+  starting_values <- vector(ncores, mode="list")
+  
+  for(i in 1:ncores){
+    t(full_best_fit[i, ])  |> 
+      as.matrix() |>
+      `colnames<-`("unit1") -> starting_values[[i]] 
+  }
+  
+  return(starting_values)
+}

### 2024-07-22 ms.Rnw
additional naming

@@
 bake(file = "output/round_01/ovov_01_el.rds",{
-  el <- measlespkg::eval_logLik(mifs_global_01, ncores=ncores, np_pf = NP_MIF, nreps=ncores)
-  el
+  el_01 <- measlespkg::eval_logLik(mifs_global_01, ncores=ncores, np_pf = NP_MIF, nreps=ncores)
+  el_01
 })
@@
 bake(file = "output/round_02/ovov_02_el.rds",{
-  el <- measlespkg::eval_logLik(mifs_global_02, ncores = ncores, np_pf = NP_MIF, nreps=ncores)
-  el
+  el_02 <- measlespkg::eval_logLik(mifs_global_02, ncores = ncores, np_pf = NP_MIF, nreps=ncores)
+  el_02
 })
@@
 bake(file = "output/round_03/ovov_03_el.rds",{
-  el <- measlespkg::eval_logLik(mifs_global_03, ncores = ncores, np_pf = NP_MIF, nreps=ncores)
-  el
+  el_03 <- measlespkg::eval_logLik(mifs_global_03, ncores = ncores, np_pf = NP_MIF, nreps=ncores)
+  el_03
 })
@@
       \multicolumn{1}{l}{\(S_{20}\)} & \multicolumn{3}{c}{$57139612$}  & $$\Sexpr{est_mle_vec["S20[unit1]"]}$$\\
-      \multicolumn{1}{l}{\(I_{20}\)} & \multicolumn{3}{c}{$871$}       & $$\Sexpr{est_mle_vec["sigma_xi[unit1]"]}$$\\
-      \multicolumn{1}{l}{\(R_{20}\)} & \multicolumn{3}{c}{$302852$}    & $$\Sexpr{est_mle_vec["sigma_xi[unit1]"]}$$\\
-      \multicolumn{1}{l}{\(S_{30}\)} & \multicolumn{3}{c}{$19573727$}  & $$\Sexpr{est_mle_vec["sigma_xi[unit1]"]}$$\\
-      \multicolumn{1}{l}{\(I_{30}\)} & \multicolumn{3}{c}{$2550$}      & $$\Sexpr{est_mle_vec["sigma_xi[unit1]"]}$$\\
-      \multicolumn{1}{l}{\(R_{30}\)} & \multicolumn{3}{c}{$131092$}    & $$\Sexpr{est_mle_vec["sigma_xi[unit1]"]}$$\\
+      \multicolumn{1}{l}{\(I_{20}\)} & \multicolumn{3}{c}{$871$}       & $$\Sexpr{est_mle_vec["I20[unit1]"]}$$\\
+      \multicolumn{1}{l}{\(R_{20}\)} & \multicolumn{3}{c}{$302852$}    & $$\Sexpr{est_mle_vec["R20[unit1]"]}$$\\
+      \multicolumn{1}{l}{\(S_{30}\)} & \multicolumn{3}{c}{$19573727$}  & $$\Sexpr{est_mle_vec["S30[unit1]"]}$$\\
+      \multicolumn{1}{l}{\(I_{30}\)} & \multicolumn{3}{c}{$2550$}      & $$\Sexpr{est_mle_vec["I30[unit1]"]}$$\\
+      \multicolumn{1}{l}{\(R_{30}\)} & \multicolumn{3}{c}{$131092$}    & $$\Sexpr{est_mle_vec["R30[unit1]"]}$$\\
       \midrule

### 2024-07-22 ms.Rnw
set seeds

@@
 bake(file = "output/round_01/ovov_01_el.rds",{
+  set.seed(33124)
   el_01 <- measlespkg::eval_logLik(mifs_global_01, ncores=ncores, np_pf = NP_MIF, nreps=ncores)
@@
   mifs_global_02 <- foreach(i=1:ncores,.packages='pomp', .options.multicore=mcopts) %dopar% {
+    set.seed(202407222+1234*i)
     panelPomp::mif2(
@@
 bake(file = "output/round_02/ovov_02_el.rds",{
+  set.seed(4321)
   el_02 <- measlespkg::eval_logLik(mifs_global_02, ncores = ncores, np_pf = NP_MIF, nreps=ncores)
@@
   mifs_global_03 <- foreach(i=1:ncores,.packages='pomp', .options.multicore=mcopts) %dopar% {
+    set.seed(2207+4321*i)
     panelPomp::mif2(
@@
 bake(file = "output/round_03/ovov_03_el.rds",{
+  set.seed(1234)
   el_03 <- measlespkg::eval_logLik(mifs_global_03, ncores = ncores, np_pf = NP_MIF, nreps=ncores)

### 2024-07-24 ms.Rnw
compilation, dofuture

--- 2024-07-22 ms.Rnw
@@
 library(doParallel)
+library(doFuture)
 library(dplyr)
+library(Rcpp)
+library(RcppArmadillo)
+# install.packages('RcppDist')
+sourceCpp('rotavirus_normq.cpp')
 source("choose_n_fits.R")
+source("eval_logLik_pomp.R")
 
@@
 NMIF         = switch(RUN_LEVEL, 4,    100)
-ncores       = switch(RUN_LEVEL, 4,    36)
+ncores       = switch(RUN_LEVEL, 2,    36)
 top_n_fits   = switch(RUN_LEVEL, 1,    12)
+sim_rep      = switch(RUN_LEVEL, 1,    100) # Total number of simulations
+
+#population number
+pop <- 82372825
 @
@@
 
-<<OvOv-model, include=False, massage=False, warning=False>>=
+<<OvOv-model, echo=False, massage=False, warning=False, results=hide>>=
 # measurement model 
@@
                   mu=0, N=82372825, omega=1/(1*52))
-# WWR's rinit
+# full-estimate-rinit
 rinit <- Csnippet("
@@
 
-sir_panel <- panelPomp::panelPomp(list(unit1=sir),
-                                  shared=NULL,
-                                  specific=wwr_mle |> 
-                                    as.matrix() |>
-                                    `colnames<-`("unit1")
-)
+# 
+# sir_panel <- panelPomp::panelPomp(list(unit1=sir),
+#                                   shared=NULL,
+#                                   specific=wwr_mle |> 
+#                                     as.matrix() |>
+#                                     `colnames<-`("unit1")
+# )
 @
@@
 
-<<maximization1, include=False, massage=False, warning=False>>=
+<<maximization1, echo=False, massage=False, warning=False, results=hide>>=
 set.seed(123)
@@
 c(apply(sir_box,1,function(x)runif(1,min=x[1],max=x[2])),
-  params_fixed) |> 
-  as.matrix() |>
-  `colnames<-`("unit1") -> starting
+  params_fixed) -> starting
 
-require(doParallel)
-registerDoParallel(ncores)
-mcopts <- list(set.seed=TRUE)
-
-bake(file = "output/round_01/ovov_mif_01.rds",{ 
-  mifs_global_01 <- foreach(i=1:ncores,.packages='pomp', .options.multicore=mcopts) %dopar% {
-    set.seed(112*i)
-    panelPomp::mif2(
-      sir_panel,
+bake(file = paste0("output/round_01/RUN_LEVEL_",RUN_LEVEL,"/ovov_mif_01.rds"),{ 
+  registerDoFuture()
+  plan(multicore, workers=ncores)
+  mifs_global_01 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
+    pomp::mif2(
+      sir,
       Np = NP_MIF,
@@
       Nmif = NMIF,
-      shared.start = numeric(0),
-      specific.start = starting,
+      params = starting,
       block = F
@@
 
-bake(file = "output/round_01/ovov_01_el.rds",{
-  el_01 <- measlespkg::eval_logLik(mifs_global_01, ncores=ncores, np_pf = NP_MIF, nreps=ncores)
+bake(file = paste0("output/round_01/RUN_LEVEL_",RUN_LEVEL,"/ovov_el_01.rds"),{
+  el_01 <- eval_logLik_pomp(mifs_global_01, ncores, NP_MIF, 4248930)
   el_01
@@
 
-<<maximization2, include=False, massage=False, warning=False>>=
+<<maximization2, echo=False, massage=False, warning=False, results=hide>>=
 set.seed(1234)
@@
 
-el_01 <- readRDS("output/round_01/ovov_01_el.rds")
+el_01 <- readRDS(paste0("output/round_01/RUN_LEVEL_",RUN_LEVEL,"/ovov_el_01.rds"))
 starting_values_02 <- choose_n_fits(el_01, top_n_fits, ncores)
 
-require(doParallel)
-registerDoParallel(ncores)
-mcopts <- list(set.seed=TRUE)
-
-bake(file = "output/round_02/ovov_mif_02.rds",{ 
-  mifs_global_02 <- foreach(i=1:ncores,.packages='pomp', .options.multicore=mcopts) %dopar% {
-  panelPomp::mif2(
-    sir_panel,
+bake(file = paste0("output/round_02/RUN_LEVEL_",RUN_LEVEL,"/ovov_mif_02.rds"),{ 
+  registerDoFuture()
+  plan(multicore, workers=ncores)
+  mifs_global_02 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
+    pomp::mif2(
+      sir,
       Np = NP_MIF,
@@
       rw.sd = rw_sd(beta1=0.01*sd_02,beta2=0.01*sd_02,beta3=0.01*sd_02,
-                  beta11=0.01*sd_02,phi=0.01*sd_02,sigma_q=0.01*sd_02,sigma_xi=0.01*sd_02,
+                    beta11=0.01*sd_02,phi=0.01*sd_02,
+                    sigma_q=0.01*sd_02,sigma_xi=0.01*sd_02,
                     S10=ivp(0.24)*sd_02,I10=ivp(0.24)*sd_02,R10=ivp(0.24)*sd_02,
@@
       Nmif = NMIF,
-    shared.start = numeric(0),
-    specific.start = starting_values_02[[i]],
+      params = starting,
       block = F
@@
 
-bake(file = "output/round_02/ovov_02_el.rds",{
-  el_02 <- measlespkg::eval_logLik(mifs_global_02, ncores = ncores, np_pf = NP_MIF, nreps=ncores)
+bake(file = paste0("output/round_02/RUN_LEVEL_",RUN_LEVEL,"/ovov_el_02.rds"),{
+  el_02 <- eval_logLik_pomp(mifs_global_02, ncores, NP_MIF, seed=4248930)
   el_02
@@
 
-<<maximization3, include=False, massage=False, warning=False>>=
+<<maximization3, echo=False, massage=False, warning=False, results=hide>>=
 set.seed(12345)
@@
 
-el_02 <- readRDS("output/round_02/ovov_02_el.rds")
+el_02 <- readRDS(paste0("output/round_02/RUN_LEVEL_",RUN_LEVEL,"/ovov_el_02.rds"))
 starting_values_03 <- choose_n_fits(el_02, top_n_fits, ncores)
 
-require(doParallel)
-registerDoParallel(ncores)
-mcopts <- list(set.seed=TRUE)
-
-bake(file = "output/round_03/ovov_mif_03.rds",{ 
-  mifs_global_03 <- foreach(i=1:ncores,.packages='pomp', .options.multicore=mcopts) %dopar% {
-  panelPomp::mif2(
-    sir_panel,
+bake(file = paste0("output/round_03/RUN_LEVEL_",RUN_LEVEL,"/ovov_mif_03.rds"),{ 
+  registerDoFuture()
+  plan(multicore, workers=ncores)
+  mifs_global_03 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
+    pomp::mif2(
+      sir,
       Np = NP_MIF,
@@
       rw.sd = rw_sd(beta1=0.01*sd_03,beta2=0.01*sd_03,beta3=0.01*sd_03,
-                  beta11=0.01*sd_03,phi=0.01*sd_03,sigma_q=0.01*sd_03,sigma_xi=0.01*sd_03,
+                    beta11=0.01*sd_03,phi=0.01*sd_03,
+                    sigma_q=0.01*sd_03,sigma_xi=0.01*sd_03,
                     S10=ivp(0.24)*sd_03,I10=ivp(0.24)*sd_03,R10=ivp(0.24)*sd_03,
@@
       Nmif = NMIF,
-    shared.start = numeric(0),
-    specific.start = starting_values_03[[i]],
+      params = starting,
       block = F
@@
 
-bake(file = "output/round_03/ovov_03_el.rds",{
-  el_03 <- measlespkg::eval_logLik(mifs_global_03, ncores = ncores, np_pf = NP_MIF, nreps=ncores)
+bake(file = paste0("output/round_03/RUN_LEVEL_",RUN_LEVEL,"/ovov_el_03.rds"),{
+  el_03 <- eval_logLik_pomp(mifs_global_03, ncores, NP_MIF, seed=4248930)
   el_03
@@
 
-<<estimated-mle, include=False, massage=False, warning=False>>=
-mif_el <- readRDS("output/round_03/ovov_03_el.rds") # Can be changed if more rounds needed
-est_mle_vec <- mif_el$fits[which.max(mif_el$fits$logLik),]
+<<estimated-mle, echo=False, massage=False, warning=False>>=
+mif_el <- readRDS(paste0("output/round_03/RUN_LEVEL_",RUN_LEVEL,"/ovov_el_03.rds")) 
+# Can be changed if more rounds needed
+est_mle_vec <- mif_el[which.max(mif_el$logLik),]
 @
@@
 
+<<pf-sim-100, echo=False, results=hide, massage=False, warning=False>>=
+# Set WWR's MLE
+wwr_mle_raw <- c(params_fixed, 
+             "beta1" = 11.48,
+             "beta2" = 0.25,
+             "beta3" = 0.35,
+             "phi" = 0.14,
+             "beta11" = 0.16,
+             "sigma_q" = 0.021,
+             "sigma_xi" = 66.89)
+
+i <- 1
+
+bake(file=paste0("output/pfilter/ovov_sim_100_",RUN_LEVEL,".rds"),{
+  ovov_sim_100 <- list()
+  registerDoFuture()
+  plan(multicore, workers=ncores)
+
+  repeat {
+    # Simulation
+    sim1 <- simulate(t0 = 0,
+                     times = c(1:416),
+                     dmeasure = dmeas,
+                     rmeasure = rmeas,
+                     rprocess = discrete_time(rproc, delta.t = 1/4),
+                     statenames = c("S1", "I1", "R1", "H1", 
+                                    "S2", "I2", "R2", "H2",
+                                    "S3", "I3", "R3", "H3",
+                                    "q1", "q2", "q3"),
+                     obsnames = c("cases1", "cases2", "cases3"),
+                     paramnames = names(wwr_mle_raw),
+                     accumvars = c("H1", "H2", "H3"),
+                     rinit = rinit_wwr,
+                     params = wwr_mle_raw)
+
+    sim_data <- as.data.frame(sim1)
+    if (!any(sim_data == 0)) {
+      ovov_sim_100[[length(ovov_sim_100) + 1]] <- sim_data[, c("cases1", "cases2", "cases3")]
+      i <- i + 1  # Increment only if the condition is satisfied
+    } else {
+      print(paste("Condition not met for iteration", i, "- retrying"))
+    }
+
+    if (length(ovov_sim_100) >= sim_rep) break  # Exit the loop when the desired number of successful iterations is reached
+  }
+  ovov_sim_100
+})
+
+# load("ovov_sim_100.Rdata")
+@
+
+<<pfilter_sim_100, echo=False, massage=False, warning=False, results=hide>>=
+ovov_sim_100 <- readRDS(paste0("output/pfilter/ovov_sim_100_",RUN_LEVEL,".rds"))
+bake(file=paste0("output/pfilter/pfilter_ovov_sim_100_",RUN_LEVEL,".rds"),{
+  pfilter_ovov_sim_100 <- c()
+  registerDoFuture()
+  plan(multicore, workers=ncores)
+
+  for(i in 1:length(ovov_sim_100)){
+    dat <- data.frame(time=c(1:416), ovov_sim_100[[i]])
+    dat |>
+      pomp(t0=0,
+           time="time",
+           rprocess=discrete_time(rproc,delta.t=1/4), ## Obs time every week, but delta.t transition is actually everyday
+           rinit=rinit_wwr,
+           dmeasure=dmeas,
+           rmeasure=rmeas,
+           accumvars=c("H1","H2","H3"),
+           statenames=c("S1","S2","S3",
+                        "I1","I2","I3",
+                        "R1","R2","R3",
+                        "H1","H2","H3",
+                        "q1","q2","q3"),
+           paramnames = names(wwr_mle_raw)
+      ) -> sir_wwr
+
+    theta <- wwr_mle_raw
+
+    foreach(i=1:ncores, .combine=c,
+            .options.future=list(seed=998468235L)
+    ) %dofuture% {
+      pfilter(sir_wwr,Np=NP_MIF,save.states=T, params=theta)
+    } -> pfs
+
+    pfilter_ovov_sim_100[i] <-logmeanexp(logLik(pfs))  
+  }
+  pfilter_ovov_sim_100
+})
+@
+
+<<palfilter_sim_100, echo=False, massage=False, warning=False, results=hide>>=
+readRDS(paste0("output/pfilter/ovov_sim_100_",RUN_LEVEL,".rds"))
+prop <- 420
+
+init_dist <- c(3876549, 30351, 1315221, 57139612, 871, 302852, 19573727, 2550, 131092)
+# initial_guess <- c(runif(1,5,20),runif(1,0.05,0.5),runif(1,0.2,1),runif(1,0.05,0.4),runif(1,0.05,0.4),runif(1,0.05,0.2),runif(1,10,100))
+initial_guess <- c(11.48,0.25,0.35,0.14,0.16,0.021,66.89)
+
+### Just a filter
+Sys.time()
+
+bake(file=paste0("output/pfilter/palfilter_ovov_100_",RUN_LEVEL,".Rdata"),{
+  y <- c()
+  registerDoFuture()
+  plan(multicore, workers=ncores)
+  for(i in 1:length(ovov_sim_100)){
+    realdat <- ovov_sim_100[[i]] |> as.matrix()
+    lik_list <- rotavirus_SMC_qropxi(init_dist,realdat , 9, 
+                                     c(initial_guess[1],initial_guess[2],initial_guess[3],
+                                       initial_guess[4],initial_guess[5]), 
+                                     gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
+                                     norm_par = c(0.07,initial_guess[6]), 
+                                     t(prop), NP_MIF,1)
+    y[i] <- lik_list$log_lik
+  }
+  y
+})
+
+Sys.time()
+
+# lik_list$ll_storage |> sum()
+
+# x <- c()
+# for(i in 1:10){
+#   x[i] <- any(ovov_sim_100[[i]]==0)
+# }
+@
+
 \begin{figure}[h]
@@
 The red line is the line $Lik(pfilter)=Lik(palfilter)$.
-On average, the particle filter likelihood is 7.7 log units higher than the PAL.
+On average, the particle filter likelihood is $7.7$ log units higher than the PAL.
 %%  -6224.245 - (-6231.97) = 7.725
@@
       \midrule
-      \(\beta_1\)    & 12.15 & 12.74 & 11.48 & $\Sexpr{est_mle_vec["beta1[unit1]"]}$\\
-      \(\beta_2\)    & 0.22  & 0.21  & 0.25  & $\Sexpr{est_mle_vec["beta2[unit1]"]}$\\
-      \(\beta_3\)    & 0.34  & 0.31  & 0.35  & $\Sexpr{est_mle_vec["beta3[unit1]"]}$\\
-      \(\phi\)       & 0.017 & 0.14  & 0.14  & $\Sexpr{est_mle_vec["phi[unit1]"]}$\\
-      \(\rho\)       & 0.022 & 0.19  & 0.16  & $\Sexpr{est_mle_vec["rho[unit1]"]}$\\
-      \(\sigma^2_q\) & n/a   & 0.042 & 0.021 & $\Sexpr{est_mle_vec["sigma_q[unit1]"]}$\\
-      \(\sigma_\xi\) & n/a   & n/a   & 66.89 & $\Sexpr{est_mle_vec["sigma_xi[unit1]"]}$\\
+      \(\beta_1\)    & 12.15 & 12.74 & 11.48 & $\Sexpr{round(est_mle_vec["beta1"],2)}$\\
+      \(\beta_2\)    & 0.22  & 0.21  & 0.25  & $\Sexpr{round(est_mle_vec["beta2"],2)}$\\
+      \(\beta_3\)    & 0.34  & 0.31  & 0.35  & $\Sexpr{round(est_mle_vec["beta3"],2)}$\\
+      \(\phi\)       & 0.017 & 0.14  & 0.14  & $\Sexpr{round(est_mle_vec["phi"],2)}$\\
+      \(\rho\)       & 0.022 & 0.19  & 0.16  & $\Sexpr{round(est_mle_vec["beta11"],2)}$\\
+      \(\sigma^2_q\) & n/a   & 0.042 & 0.021 & $\Sexpr{round(est_mle_vec["sigma_q"],2)}$\\
+      \(\sigma_\xi\) & n/a   & n/a   & 66.89 & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$\\
       \addlinespace
-      \multicolumn{1}{l}{\(S_{10}\)} & \multicolumn{3}{c}{$3876549^*$} & $$\Sexpr{est_mle_vec["S10[unit1]"]}$$\\
-      \multicolumn{1}{l}{\(I_{10}\)} & \multicolumn{3}{c}{$30351$}     & $$\Sexpr{est_mle_vec["I10[unit1]"]}$$\\
-      \multicolumn{1}{l}{\(R_{10}\)} & \multicolumn{3}{c}{$1315221$}   & $$\Sexpr{est_mle_vec["R10[unit1]"]}$$\\
-      \multicolumn{1}{l}{\(S_{20}\)} & \multicolumn{3}{c}{$57139612$}  & $$\Sexpr{est_mle_vec["S20[unit1]"]}$$\\
-      \multicolumn{1}{l}{\(I_{20}\)} & \multicolumn{3}{c}{$871$}       & $$\Sexpr{est_mle_vec["I20[unit1]"]}$$\\
-      \multicolumn{1}{l}{\(R_{20}\)} & \multicolumn{3}{c}{$302852$}    & $$\Sexpr{est_mle_vec["R20[unit1]"]}$$\\
-      \multicolumn{1}{l}{\(S_{30}\)} & \multicolumn{3}{c}{$19573727$}  & $$\Sexpr{est_mle_vec["S30[unit1]"]}$$\\
-      \multicolumn{1}{l}{\(I_{30}\)} & \multicolumn{3}{c}{$2550$}      & $$\Sexpr{est_mle_vec["I30[unit1]"]}$$\\
-      \multicolumn{1}{l}{\(R_{30}\)} & \multicolumn{3}{c}{$131092$}    & $$\Sexpr{est_mle_vec["R30[unit1]"]}$$\\
+      \multicolumn{1}{l}{\(S_{10}\)} & \multicolumn{3}{c}{$3876549^*$} & $\Sexpr{round(pop*est_mle_vec["S10"],0)}$\\
+      \multicolumn{1}{l}{\(I_{10}\)} & \multicolumn{3}{c}{$30351$}     & $\Sexpr{round(pop*est_mle_vec["I10"],0)}$\\
+      \multicolumn{1}{l}{\(R_{10}\)} & \multicolumn{3}{c}{$1315221$}   & $\Sexpr{round(pop*est_mle_vec["R10"],0)}$\\
+      \multicolumn{1}{l}{\(S_{20}\)} & \multicolumn{3}{c}{$57139612$}  & $\Sexpr{round(pop*est_mle_vec["S20"],0)}$\\
+      \multicolumn{1}{l}{\(I_{20}\)} & \multicolumn{3}{c}{$871$}       & $\Sexpr{round(pop*est_mle_vec["I20"],0)}$\\
+      \multicolumn{1}{l}{\(R_{20}\)} & \multicolumn{3}{c}{$302852$}    & $\Sexpr{round(pop*est_mle_vec["R20"],0)}$\\
+      \multicolumn{1}{l}{\(S_{30}\)} & \multicolumn{3}{c}{$19573727$}  & $\Sexpr{round(pop*est_mle_vec["S30"],0)}$\\
+      \multicolumn{1}{l}{\(I_{30}\)} & \multicolumn{3}{c}{$2550$}      & $\Sexpr{round(pop*est_mle_vec["I30"],0)}$\\
+      \multicolumn{1}{l}{\(R_{30}\)} & \multicolumn{3}{c}{$131092$}    & $\Sexpr{round(pop*est_mle_vec["R30"],0)}$\\
       \midrule
       \addlinespace
-      \textbf{AIC}  & $98866.65$ & $15154.75$ & $13778.08$ & $\Sexpr{(-2)*((length(wwr_mle) - length(params_fixed)) - est_mle_vec["logLik"])}$\\
+      \textbf{AIC}  & $98866.65$ & $15154.75$ & $13778.08$ & $\Sexpr{round((-2)*((length(wwr_mle) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$\\
       \addlinespace

### 2024-07-24 choose_n_fits.R

@@ -2,7 +2,7 @@
 
 choose_n_fits = function(el, top_n_fits, ncores){
   ### Next-round code
-  fit_mat <- na.omit(el$fits)
+  fit_mat <- el
   score_total = fit_mat$logLik
   ranking_total = order(score_total, decreasing = TRUE)[1:top_n_fits]
   

### 2024-07-24 eval_logLik_pomp.R

@@ -1,0 +1,30 @@
+#pfilter for a list of mif2 results
+eval_logLik_pomp = function(model_obj_list, nreps, np_pf, seed){
+  pf_logLik_frame = data.frame(
+    logLik = rep(0, length(model_obj_list)),
+    se = rep(0, length(model_obj_list))
+  ) %>% cbind(
+    rbind(t(sapply(model_obj_list, pomp::coef)))
+  )
+  
+  for(i in seq_along(model_obj_list)){
+    registerDoFuture()
+    plan(multicore, workers = ncores)
+    
+    foreach::foreach(
+      j = 1:nreps,
+      .options.future=list(seed=seed),
+      .combine = rbind
+    ) %dofuture% {
+      pfilter_obj = pomp::pfilter(model_obj_list[[i]], Np = np_pf)
+      pomp::logLik(pfilter_obj)
+    } -> pf_logLik_matrix
+    
+    pf_logLik_frame[i, 1:2] = pomp::logmeanexp(
+      pf_logLik_matrix,
+      se = TRUE
+    )
+  }
+  
+  return(pf_logLik_frame)
+}

### 2024-07-24 rotavirus_normq.cpp

@@ -1,0 +1,326 @@
+// [[Rcpp::depends(RcppArmadillo)]]
+// [[Rcpp::depends(Rcpp)]]
+#include <RcppArmadillo.h>
+#include <RcppArmadilloExtensions/sample.h>
+#include <math.h>
+#define RCPPDIST_DONT_USE_ARMA
+#include <RcppDist.h>
+#include <vector>
+// [[Rcpp::depends(RcppDist)]]
+using namespace Rcpp;
+using namespace arma;
+
+
+void update_K(mat& K,
+              vec lambar,
+              vec betas,
+              double xi,
+              int N,
+              int t,
+              double phase,
+              double rho
+){
+  // seasonal component
+  
+  mat B(3,3, fill::zeros);
+  for(int i = 0; i<3;i++){
+    for(int j = 0; j<3;j++){
+      B(i,j) = xi*betas(i); 
+    }
+  }
+  //if(t==0){cout << B(0,0) << endl;}
+  
+  double kappa = (1 + rho*cos(2*3.141593*t/52 + phase));
+  
+  vec I;
+  I = {lambar(1),lambar(4),lambar(7)};
+  vec lambda = B*I*kappa/N;
+  K(0,1) = 1-exp(-0.25*lambda(0));
+  K(3,4) = 1-exp(-0.25*lambda(1));
+  K(6,7) = 1-exp(-0.25*lambda(2));
+  
+  
+  K(0,0)= 1 - K(0,1)-  K(0,2) - K(0,3) - K(0,4) - K(0,5) - K(0,6) - K(0,7) - K(0,8);
+  K(1,1)= 1 - K(1,0) - K(1,2) - K(1,3) - K(1,4) - K(1,5) - K(1,6) - K(1,7) - K(1,8);
+  K(2,2)= 1 - K(2,0) - K(2,1) - K(2,3) - K(2,4) - K(2,5) - K(2,6) - K(2,7) - K(2,8);
+  K(3,3)= 1 - K(3,0) - K(3,1) - K(3,2) - K(3,4) - K(3,5) - K(3,6) - K(3,7) - K(3,8);
+  K(4,4)= 1 - K(4,0) - K(4,1) - K(4,2) - K(4,3) - K(4,5) - K(4,6) - K(4,7) - K(4,8);
[…]

### 2024-07-24 eval_logLik_pomp.R
compilation, dofuture

@@ -9,7 +9,7 @@ eval_logLik_pomp = function(model_obj_list, nreps, np_pf, seed){
   
   for(i in seq_along(model_obj_list)){
     registerDoFuture()
-    plan(multicore, workers = ncores)
+    plan(multicore, workers = nreps)
     
     foreach::foreach(
       j = 1:nreps,

### 2024-07-24 ms.Rnw
correction on starting values

@@
       Nmif = NMIF,
-      params = starting,
+      params = starting_values_02[[i]],
       block = F
@@
       Nmif = NMIF,
-      params = starting,
+      params = starting_values_03[[i]],
       block = F

### 2024-07-24 choose_n_fits.R

@@ -13,15 +13,11 @@ choose_n_fits = function(el, top_n_fits, ncores){
   recycle_vec = sort(rep_len(1:top_n_fits, ncores))
   full_best_fit <- best_fits[recycle_vec, ] 
   
-  coef_names <- colnames(full_best_fit)
-  colnames(full_best_fit) <- gsub(".{7}$","",coef_names)
   
   starting_values <- vector(ncores, mode="list")
   
   for(i in 1:ncores){
-    t(full_best_fit[i, ])  |> 
-      as.matrix() |>
-      `colnames<-`("unit1") -> starting_values[[i]] 
+    full_best_fit[i, ] -> starting_values[[i]] 
   }
   
   return(starting_values)

### 2024-07-24 ms.Rnw
compile

--- 2024-07-24 ms.Rnw
@@
 \begin{document}
-\SweaveOpts{concordance=TRUE}
+%\SweaveOpts{concordance=TRUE}
 
@@
 RUN_LEVEL = 1
-NP_MIF       = switch(RUN_LEVEL, 5000, 50000)
+NP_MIF       = switch(RUN_LEVEL, 1000, 50000)
 NMIF         = switch(RUN_LEVEL, 4,    100)
@@
 
-<<OvOv-model, echo=False, massage=False, warning=False, results=hide>>=
+<<OvOv-model, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
 # measurement model 
@@
 
-<<maximization1, echo=False, massage=False, warning=False, results=hide>>=
+<<maximization1, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
 set.seed(123)
@@
       Nmif = NMIF,
-      params = starting,
-      block = F
+      params = starting
     ) 
@@
 
-<<maximization2, echo=False, massage=False, warning=False, results=hide>>=
+<<maximization2, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
 set.seed(1234)
@@
       Nmif = NMIF,
-      params = starting,
-      block = F
+      params = starting_values_02[[i]]
     ) 
@@
 
-<<maximization3, echo=False, massage=False, warning=False, results=hide>>=
+<<maximization3, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
 set.seed(12345)
@@
       Nmif = NMIF,
-      params = starting,
-      block = F
+      params = starting_values_03[[i]]
     ) 
@@
 
-<<estimated-mle, echo=False, massage=False, warning=False>>=
+<<estimated-mle, echo=FALSE, message=FALSE, warning=FALSE>>=
 mif_el <- readRDS(paste0("output/round_03/RUN_LEVEL_",RUN_LEVEL,"/ovov_el_03.rds")) 
@@
 
-<<pf-sim-100, echo=False, results=hide, massage=False, warning=False>>=
+<<pf-sim-100, echo=FALSE, results='hide', message=FALSE, warning=FALSE>>=
 # Set WWR's MLE
@@
 
-<<pfilter_sim_100, echo=False, massage=False, warning=False, results=hide>>=
+<<pfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
 ovov_sim_100 <- readRDS(paste0("output/pfilter/ovov_sim_100_",RUN_LEVEL,".rds"))
@@
 
-<<palfilter_sim_100, echo=False, massage=False, warning=False, results=hide>>=
+<<palfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
 readRDS(paste0("output/pfilter/ovov_sim_100_",RUN_LEVEL,".rds"))
@@
       \(\rho\)       & 0.022 & 0.19  & 0.16  & $\Sexpr{round(est_mle_vec["beta11"],2)}$\\
-      \(\sigma^2_q\) & n/a   & 0.042 & 0.021 & $\Sexpr{round(est_mle_vec["sigma_q"],2)}$\\
+      \(\sigma^2_q\) & n/a   & 0.042 & 0.021 & $\Sexpr{round(est_mle_vec["sigma_q"],3)}$\\
       \(\sigma_\xi\) & n/a   & n/a   & 66.89 & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$\\

### 2024-07-25 ms.Rnw
use runif_design, kintable

@@
 RUN_LEVEL = 1
-NP_MIF       = switch(RUN_LEVEL, 1000, 50000)
+NP_MIF       = switch(RUN_LEVEL, 10, 50000)
 NMIF         = switch(RUN_LEVEL, 4,    100)
@@
 pop <- 82372825
+
+out_dir <- paste0("output_0",RUN_LEVEL, "/")
 @
@@
 set.seed(123)
-sir_box <- rbind(
+sir_box <- list(
   beta1=c(10,15),
@@
   I30 = c(0.000001, 0.0001),
-  R30 = c(0.0001,0.005)
+  R30 = c(0.0001,0.005),
+  #fixed params
+  gamma = c(1,1),
+  delta1=c(1/(5*52),1/(5*52)),
+  delta2=c(1/(55*52),1/(55*52)),
+  alpha=c(1/(78.86912*52),1/(78.86912*52)),
+  mu=c(0,0),
+  N=c(82372825,82372825),
+  omega=c(1/(1*52),1/(1*52))
 )
 
-## Make it panelPomp
-c(apply(sir_box,1,function(x)runif(1,min=x[1],max=x[2])),
-  params_fixed) -> starting
+design_matrix <- runif_design(lower = sapply(sir_box, `[`, 1),
+                              upper = sapply(sir_box, `[`, 2),
+                              n = ncores)
 
-bake(file = paste0("output/round_01/RUN_LEVEL_",RUN_LEVEL,"/ovov_mif_01.rds"),{ 
+# c(apply(sir_box,1,function(x)runif(1,min=x[1],max=x[2])),
+#   params_fixed) -> starting
+
+bake(file = paste0(out_dir, "round_01/ovov_mif_01.rds"),{ 
   registerDoFuture()
@@
       Nmif = NMIF,
-      params = starting
+      params = design_matrix[i, ]
     ) 
@@
 
-bake(file = paste0("output/round_01/RUN_LEVEL_",RUN_LEVEL,"/ovov_el_01.rds"),{
+bake(file = paste0(out_dir, "round_01/ovov_el_01.rds"),{
   el_01 <- eval_logLik_pomp(mifs_global_01, ncores, NP_MIF, 4248930)
@@
 
-el_01 <- readRDS(paste0("output/round_01/RUN_LEVEL_",RUN_LEVEL,"/ovov_el_01.rds"))
+el_01 <- readRDS(paste0(out_dir,"round_01/ovov_el_01.rds"))
 starting_values_02 <- choose_n_fits(el_01, top_n_fits, ncores)
 
-bake(file = paste0("output/round_02/RUN_LEVEL_",RUN_LEVEL,"/ovov_mif_02.rds"),{ 
+bake(file = paste0(out_dir,"round_02/ovov_mif_02.rds"),{ 
   registerDoFuture()
@@
 
-bake(file = paste0("output/round_02/RUN_LEVEL_",RUN_LEVEL,"/ovov_el_02.rds"),{
+bake(file = paste0(out_dir, "round_02/ovov_el_02.rds"),{
   el_02 <- eval_logLik_pomp(mifs_global_02, ncores, NP_MIF, seed=4248930)
@@
 
-el_02 <- readRDS(paste0("output/round_02/RUN_LEVEL_",RUN_LEVEL,"/ovov_el_02.rds"))
+el_02 <- readRDS(paste0(out_dir, "round_02/ovov_el_02.rds"))
 starting_values_03 <- choose_n_fits(el_02, top_n_fits, ncores)
 
-bake(file = paste0("output/round_03/RUN_LEVEL_",RUN_LEVEL,"/ovov_mif_03.rds"),{ 
+bake(file = paste0(out_dir, "round_03/ovov_mif_03.rds"),{ 
   registerDoFuture()
@@
 
-bake(file = paste0("output/round_03/RUN_LEVEL_",RUN_LEVEL,"/ovov_el_03.rds"),{
+bake(file = paste0(out_dir,"round_03/ovov_el_03.rds"),{
   el_03 <- eval_logLik_pomp(mifs_global_03, ncores, NP_MIF, seed=4248930)
@@
 <<estimated-mle, echo=FALSE, message=FALSE, warning=FALSE>>=
-mif_el <- readRDS(paste0("output/round_03/RUN_LEVEL_",RUN_LEVEL,"/ovov_el_03.rds")) 
+mif_el <- readRDS(paste0(out_dir,"round_03/ovov_el_03.rds")) 
 # Can be changed if more rounds needed
@@
 
-bake(file=paste0("output/pfilter/ovov_sim_100_",RUN_LEVEL,".rds"),{
+bake(file=paste0(out_dir, "pfilter/ovov_sim_100.rds"),{
   ovov_sim_100 <- list()
@@
 <<pfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
-ovov_sim_100 <- readRDS(paste0("output/pfilter/ovov_sim_100_",RUN_LEVEL,".rds"))
-bake(file=paste0("output/pfilter/pfilter_ovov_sim_100_",RUN_LEVEL,".rds"),{
+ovov_sim_100 <- readRDS(paste0(out_dir, "pfilter/ovov_sim_100.rds"))
+bake(file=paste0(out_dir, "pfilter/pfilter_ovov_sim_100.rds"),{
   pfilter_ovov_sim_100 <- c()
@@
 <<palfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
-readRDS(paste0("output/pfilter/ovov_sim_100_",RUN_LEVEL,".rds"))
+ovov_sim_100 <- readRDS(paste0(out_dir, "pfilter/ovov_sim_100.rds"))
 prop <- 420
@@
 
-bake(file=paste0("output/pfilter/palfilter_ovov_100_",RUN_LEVEL,".Rdata"),{
+bake(file=paste0(out_dir, "pfilter/palfilter_ovov_100.rds"),{
   y <- c()

### 2024-07-25 ms.Rnw
some tidying

@@
 \usepackage{graphicx} % Required for inserting
-%%\usepackage{algpseudocode} % Include the package for algorithmic environment
-\usepackage{algorithm}
 \usepackage{booktabs} % For better-quality horizontal lines
-\usepackage{algpseudocode}
 \usepackage{multirow}
@@
 \usepackage{array} % For table formatting
-
 \usepackage{setspace}
@@
 \usepackage[round]{natbib}
-
 \usepackage{multirow}
-\usepackage{fourier} 
-\usepackage{booktabs}
 \usepackage{array}
@@
 \usepackage{subcaption}
-\usepackage{tabstackengine}
-\usepackage[toc,page]{appendix}
 \usepackage{graphicx, fullpage, verbatim, amsmath}
 \usepackage{url, amsfonts, amssymb, amsthm,color, enumerate}
-\usepackage{placeins, listings, textcomp, mathtools, multicol, tikz}
-\usepackage{dsfont}
 \usepackage{amsfonts}
@@
 
-\TABstackMath
-
 \doublespacing
-% Language setting
-% Replace `english' with e.g. `spanish' to change the document language
-\usepackage[english]{babel}
+% \usepackage[english]{babel}
 
@@
 \begin{document}
-%\SweaveOpts{concordance=TRUE}
 
@@
 opts_chunk$set(
-    concordance = TRUE,
+#    concordance = TRUE,
     tidy = TRUE,
-    message = FALSE,
-    warning = FALSE,
+#    message = FALSE,
+#    warning = FALSE,
     tidy.opts = list(
@@
 
-<<OvOv-model, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
+<<OvOv-model>>=
 # measurement model 
@@
 
-<<maximization1, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
+<<maximization1>>=
 set.seed(123)
@@
 
-<<maximization2, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
+<<maximization2>>=
 set.seed(1234)
@@
 
-<<maximization3, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
+<<maximization3>>=
 set.seed(12345)
@@
 
-<<estimated-mle, echo=FALSE, message=FALSE, warning=FALSE>>=
+<<estimated-mle>>=
 mif_el <- readRDS(paste0(out_dir,"round_03/ovov_el_03.rds")) 
@@
 
-<<pf-sim-100, echo=FALSE, results='hide', message=FALSE, warning=FALSE>>=
+<<pf-sim-100>>=
 # Set WWR's MLE
@@
 
-<<pfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
+<<pfilter_sim_100>>=
 ovov_sim_100 <- readRDS(paste0(out_dir, "pfilter/ovov_sim_100.rds"))
@@
 
-<<palfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
+<<palfilter_sim_100>>=
 ovov_sim_100 <- readRDS(paste0(out_dir, "pfilter/ovov_sim_100.rds"))

### 2024-07-25 rotavirus_normq.cpp

@@ -10,6 +10,9 @@
 using namespace Rcpp;
 using namespace arma;
 
+// This file is adapted from
+// https://github.com/LorenzoRimella/PAL/RealDataExperiments/rotavirus
+// Copied on November, 2023
 
 void update_K(mat& K,
               vec lambar,

### 2024-07-25 ms.Rnw
add plot

--- 2024-07-25 ms.Rnw
@@
 \usepackage{algorithm}
-\usepackage{booktabs} % For better-quality horizontal lines
 \usepackage{algpseudocode}
-\usepackage{multirow}
 \usepackage{threeparttable} % For table notes
@@
 \usepackage{placeins, listings, textcomp, mathtools, multicol, tikz}
-\usepackage{dsfont}
-\usepackage{amsfonts}
 \usepackage{bm}
@@
 \usepackage[colorlinks=true,citecolor=blue]{hyperref}
-\newcommand\choleraDeath{\delta_{C}}
 
@@
 
+wwr_mle_raw <- c(params_fixed, 
+             "beta1" = 11.48,
+             "beta2" = 0.25,
+             "beta3" = 0.35,
+             "phi" = 0.14,
+             "beta11" = 0.16,
+             "sigma_q" = 0.021,
+             "sigma_xi" = 66.89)
+
 pt <- pomp::parameter_trans(
@@
 
-bake(file = paste0(out_dir, "round_01/ovov_mif_01.rds"),{ 
+bake(file = paste0(out_dir, "mif2_maximization/round_01/ovov_mif_01.rds"),{ 
   registerDoFuture()
@@
 
-bake(file = paste0(out_dir, "round_01/ovov_el_01.rds"),{
+bake(file = paste0(out_dir, "mif2_maximization/round_01/ovov_el_01.rds"),{
   el_01 <- eval_logLik_pomp(mifs_global_01, ncores, NP_MIF, 4248930)
@@
 
-el_01 <- readRDS(paste0(out_dir,"round_01/ovov_el_01.rds"))
+el_01 <- readRDS(paste0(out_dir,"mif2_maximization/round_01/ovov_el_01.rds"))
 starting_values_02 <- choose_n_fits(el_01, top_n_fits, ncores)
 
-bake(file = paste0(out_dir,"round_02/ovov_mif_02.rds"),{ 
+bake(file = paste0(out_dir,"mif2_maximization/round_02/ovov_mif_02.rds"),{ 
   registerDoFuture()
@@
 
-bake(file = paste0(out_dir, "round_02/ovov_el_02.rds"),{
+bake(file = paste0(out_dir, "mif2_maximization/round_02/ovov_el_02.rds"),{
   el_02 <- eval_logLik_pomp(mifs_global_02, ncores, NP_MIF, seed=4248930)
@@
 
-el_02 <- readRDS(paste0(out_dir, "round_02/ovov_el_02.rds"))
+el_02 <- readRDS(paste0(out_dir, "mif2_maximization/round_02/ovov_el_02.rds"))
 starting_values_03 <- choose_n_fits(el_02, top_n_fits, ncores)
 
-bake(file = paste0(out_dir, "round_03/ovov_mif_03.rds"),{ 
+bake(file = paste0(out_dir, "mif2_maximization/round_03/ovov_mif_03.rds"),{ 
   registerDoFuture()
@@
 
-bake(file = paste0(out_dir,"round_03/ovov_el_03.rds"),{
+bake(file = paste0(out_dir,"mif2_maximization/round_03/ovov_el_03.rds"),{
   el_03 <- eval_logLik_pomp(mifs_global_03, ncores, NP_MIF, seed=4248930)
@@
 <<estimated-mle, echo=FALSE, message=FALSE, warning=FALSE>>=
-mif_el <- readRDS(paste0(out_dir,"round_03/ovov_el_03.rds")) 
+mif_el <- readRDS(paste0(out_dir,"mif2_maximization/round_03/ovov_el_03.rds")) 
 # Can be changed if more rounds needed
@@
 
-bake(file=paste0(out_dir, "pfilter/ovov_sim_100.rds"),{
+bake(file=paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"),{
   ovov_sim_100 <- list()
@@
 <<pfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
-ovov_sim_100 <- readRDS(paste0(out_dir, "pfilter/ovov_sim_100.rds"))
-bake(file=paste0(out_dir, "pfilter/pfilter_ovov_sim_100.rds"),{
+ovov_sim_100 <- readRDS(paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"))
+bake(file=paste0(out_dir, "pfilter/simulated_dataset/pfilter_ovov_sim_100.rds"),{
   pfilter_ovov_sim_100 <- c()
@@
 <<palfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
-ovov_sim_100 <- readRDS(paste0(out_dir, "pfilter/ovov_sim_100.rds"))
+ovov_sim_100 <- readRDS(paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"))
 prop <- 420
@@
 
-bake(file=paste0(out_dir, "pfilter/palfilter_ovov_100.rds"),{
-  y <- c()
+bake(file=paste0(out_dir, "pfilter/simulated_dataset/palfilter_ovov_100.rds"),{
+  pal_log_lik <- c()
   registerDoFuture()
@@
                                      t(prop), NP_MIF,1)
-    y[i] <- lik_list$log_lik
+    pal_log_lik[i] <- lik_list$log_lik
   }
-  y
+  pal_log_lik
 })
@@
 @
+
+<<pfilter_org, echo=FALSE, message=FALSE, warning=FALSE>>=
+dat[-1, ] |>
+  pomp(t0=0,
+       time="time",
+       rprocess=discrete_time(rproc,delta.t=1/4), 
+       rinit=rinit_wwr,
+       dmeasure=dmeas,
+       rmeasure=rmeas,
+       accumvars=c("H1","H2","H3"),
+       statenames=c("S1","S2","S3",
+                    "I1","I2","I3",
+                    "R1","R2","R3",
+                    "H1","H2","H3",
+                    "q1","q2","q3"),
+       paramnames = names(wwr_mle_raw)
+  ) -> m1
+
+bake(file=paste0(out_dir,"pfilter/original_dataset/pfilter_org.rds"),{
+  registerDoFuture()
+  plan(multicore, workers = ncores)
+  foreach(i=1:ncores, .combine=c,
+          .options.future=list(seed=998468235L)
+  ) %dofuture% {
+    pfilter(m1,Np=NP_MIF,params=wwr_mle_raw)
+  } -> pfs
+  pfs
+})
+
+bake(file=paste0(out_dir,"pfilter/original_dataset/pfilter_list_cond_lik.rds"),{
+  list_cond_lik <- list()
+  registerDoFuture()
+  plan(multicore, workers = ncores)
+  for(k in 1:ncores){
+    list_cond_lik[[k]] <- pfs[[k]]@cond.logLik
+  }
+  list_cond_lik
+})
+@
+
+<<palfilter_org, echo=FALSE>>=
+load("real_rotavirus_metadata.Rdata")
+prop <- 420
+
+init_dist <- c(3876549, 30351, 1315221, 57139612, 871, 302852, 19573727, 2550, 131092)
+# initial_guess <- c(runif(1,5,20),runif(1,0.05,0.5),runif(1,0.2,1),runif(1,0.05,0.4),runif(1,0.05,0.4),runif(1,0.05,0.2),runif(1,10,100))
+initial_guess <- c(11.48,0.25,0.35,0.14,0.16,0.021,66.89)
+
+### palfiter
+registerDoFuture()
+plan(multicore, workers=ncores)
+
+Sys.time()
+bake(file=paste0(out_dir,"pfilter/original_dataset/palfilter_org.rds"),{
+  pal_lik_list_org <- rotavirus_SMC_qropxi(init_dist,realdat , 9, 
+                                   c(initial_guess[1],initial_guess[2],
+                                     initial_guess[3],initial_guess[4],initial_guess[5]), 
+                                   gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
+                                   norm_par = c(0.07,initial_guess[6]), 
+                                   t(prop), NP_MIF,1)
+  pal_lik_list_org
+})
+Sys.time()
+@
+
+<<plot1, echo=FALSE, fig.height=4, fig.width=5.5>>=
+# Create your data
+time <- 1:416
+ncores <- length(list_cond_lik)
+
+# Combine the data into a single dataframe for SMC
+data_list <- lapply(1:ncores, function(i) {
+  data.frame(time = time, cond_loglik = list_cond_lik[[i]], method = "SMC")
+})
+
+# Combine all the SMC data frames into one
+df_smc <- bind_rows(data_list)
+
+# Create the PAL data frame
+df_pal <- data.frame(time = time, cond_loglik = pal_lik_list_org$ll_storage, method = "PAL")
+
+# Combine both data frames
+df <- bind_rows(df_smc, df_pal)
+
+# Plot with ggplot
+ggplot(df, aes(x = time, y = cond_loglik, group = interaction(method, factor(1:ncores)), color = method)) +
+  geom_line(data = filter(df, method == "SMC"), color = "blue") +
+  geom_line(data = filter(df, method == "PAL"), color = "red") +
+  theme_minimal() +
+  labs(title = "Conditional Log Likelihood Over Time",
+       x = "Time",
+       y = "Conditional Log Likelihood",
+       color = "Method") +
+  theme(legend.position = "bottomright")
+
+@
+
+\includegraphics[width=4.5in]{plot1}
 
@@
 Consequently, we looked for model misspecification by comparing the conditional log-likelihood estimates at each time point, as described in the Methods section.
+
+\begin{table}[ht] % The placement specifier can be [h!tbp]
+\centering % Centers the table
+\caption{log-likelihood at the MLE of OvOv model for the real rotavirus data, computed using two filtering methods, employed 36 replicates and 50,000 particles} % This is effectively the table title
+\label{tab:ovovrealdata} % For referencing this table elsewhere in your text
+\begin{tabular}{|c c c|} 
+ \hline
+  & PAL & SMC  \\
+  \hline
+ OvOv & \Sexpr{pal_lik_list_org$log_lik} & $\Sexpr{logmeanexp(sapply(pfs, logLik))}$ \\ 
+ \hline
+\end{tabular}
+\end{table}
 
@@
 
-\begin{table}[ht] % The placement specifier can be [h!tbp]
-\centering % Centers the table
-\caption{log-likelihood at the MLE of OvOv model for the real rotavirus data, computed using two filtering methods, employed 36 replicates and 50,000 particles} % This is effectively the table title
-\label{tab:ovovrealdata} % For referencing this table elsewhere in your text
-\begin{tabular}{|c c c|} 
- \hline
-  & PAL & SMC  \\
-  \hline
- OvOv & -6892.168 & -7200.866 \\ 
- \hline
-\end{tabular}
-\end{table}
-
 \bibliography{bib-pal}

### 2024-07-25 ms.Rnw

@@
 \usepackage{graphicx} % Required for inserting
-%%\usepackage{algpseudocode} % Include the package for algorithmic environment
-\usepackage{algorithm}
-\usepackage{algpseudocode}
+\usepackage{booktabs} % For better-quality horizontal lines
+\usepackage{multirow}
 \usepackage{threeparttable} % For table notes
 \usepackage{array} % For table formatting
-
 \usepackage{setspace}
@@
 \usepackage{subcaption}
-\usepackage{tabstackengine}
-\usepackage[toc,page]{appendix}
 \usepackage{graphicx, fullpage, verbatim, amsmath}
 \usepackage{url, amsfonts, amssymb, amsthm,color, enumerate}
-\usepackage{placeins, listings, textcomp, mathtools, multicol, tikz}
+\usepackage{amsfonts}
 \usepackage{bm}
 
-\TABstackMath
-
 \doublespacing
-% Language setting
-% Replace `english' with e.g. `spanish' to change the document language
-\usepackage[english]{babel}
+% \usepackage[english]{babel}
 

### 2024-07-26 ms.Rnw
enable cond-lik plot

@@
 
-<<OvOv-model>>=
+<<OvOv-model, echo=F, warning=F, message=F>>=
 # measurement model 
@@
 
-<<maximization1>>=
+<<maximization1, echo=F, warning=F, message=F>>=
 set.seed(123)
@@
 
-<<maximization2>>=
+<<maximization2, echo=F, warning=F, message=F>>=
 set.seed(1234)
@@
 
-<<maximization3>>=
+<<maximization3, echo=F, warning=F, message=F>>=
 set.seed(12345)
@@
 
-<<pf-sim-100>>=
+<<pf-sim-100, echo=F, warning=F, message=F>>=
 # Set WWR's MLE
@@
 bake(file=paste0(out_dir, "pfilter/simulated_dataset/pfilter_ovov_sim_100.rds"),{
+  sir_wwr <- list()
   pfilter_ovov_sim_100 <- c()
@@
   for(i in 1:length(ovov_sim_100)){
-    dat <- data.frame(time=c(1:416), ovov_sim_100[[i]])
-    dat |>
-      pomp(t0=0,
+  data.frame(time=c(1:416), ovov_sim_100[[i]]) |>  pomp(t0=0,
            time="time",
-           rprocess=discrete_time(rproc,delta.t=1/4), ## Obs time every week, but delta.t transition is actually everyday
+           rprocess=discrete_time(rproc,delta.t=1/4),
            rinit=rinit_wwr,
@@
            paramnames = names(wwr_mle_raw)
-      ) -> sir_wwr
+      ) -> sir_wwr[[i]]
 
-    theta <- wwr_mle_raw
-
-    foreach(i=1:ncores, .combine=c,
+    foreach(j=1:ncores, .combine=c,
             .options.future=list(seed=998468235L)
     ) %dofuture% {
-      pfilter(sir_wwr,Np=NP_MIF,save.states=T, params=theta)
+      pfilter(sir_wwr[[i]],Np=NP_MIF,save.states=T, params=wwr_mle_raw)
     } -> pfs
@@
   for(i in 1:length(ovov_sim_100)){
-    realdat <- ovov_sim_100[[i]] |> as.matrix()
-    lik_list <- rotavirus_SMC_qropxi(init_dist,realdat , 9, 
+    pal_sim_dat <- ovov_sim_100[[i]] |> as.matrix()
+    lik_list <- rotavirus_SMC_qropxi(init_dist,pal_sim_dat , 9, 
                                      c(initial_guess[1],initial_guess[2],initial_guess[3],
@@
 
-<<plot1, echo=FALSE, fig.height=4, fig.width=5.5>>=
+<<plot1, echo=FALSE, fig.height=4, fig.width=5.5, fig.cap='Conditional log-likelihoods computed using two methods for the real rotavirus dataset, with the SMC method replicated 36 times due to its high variances.
+The blue line represents overlapping results from 36 SMC computations, while the red line is derived from PAL.
+The main sources of likelihood shortfall of SMC are at time points $t=1, 2, 3, 11, 81, 194$, and $325-333$.'>>=
 # Create your data
@@
 ggplot(df, aes(x = time, y = cond_loglik, group = interaction(method, factor(1:ncores)), color = method)) +
-  geom_line(data = filter(df, method == "SMC"), color = "blue") +
-  geom_line(data = filter(df, method == "PAL"), color = "red") +
+  geom_line() +
   theme_minimal() +
@@
        color = "Method") +
-  theme(legend.position = "bottomright")
-
+  scale_color_manual(values = c("SMC" = "blue", "PAL" = "red")) +
+  theme(legend.position = "bottom")
 @

### 2024-07-26 ms.Rnw
add sim 100 comparison plot

@@
 
-bake(file=paste0(out_dir, "pfilter/simulated_dataset/palfilter_ovov_100.rds"),{
-  pal_log_lik <- c()
+bake(file=paste0(out_dir, "pfilter/simulated_dataset/palfilter_ovov_sim_100.rds"),{
+  palfilter_ovov_sim_100 <- c()
   registerDoFuture()
@@
                                      t(prop), NP_MIF,1)
-    pal_log_lik[i] <- lik_list$log_lik
+    palfilter_ovov_sim_100[i] <- lik_list$log_lik
   }
-  pal_log_lik
+  palfilter_ovov_sim_100
 })
@@
 # }
+@
+
+<<plot1,echo=FALSE, fig.height=4, fig.width=5.5, fig.cap='Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model.The particle filter and PAL used 50,000 particles.The red line is the line $Lik(pfilter)=Lik(palfilter)$.On average, the particle filter likelihood is $7.7$ log units higher than the PAL.
+{This may not be wrong, since the likelihood evaluated by pfilter is not fully optimized by employing many particles.}'>>=
+data <- data.frame(pfilter = pfilter_ovov_sim_100, palfilter = palfilter_ovov_sim_100)
+
+ggplot(data, aes(x = pfilter, y = palfilter)) +
+  geom_point(shape = 20, size = 0.3) +  # pch=20, cex=0.3 equivalent
+  geom_abline(intercept = 0, slope = 1, color = "red") +
+  labs(title = "PAL vs SMC: OvOv with 100 simulations",
+       x = "pfilter",
+       y = "palfilter") +
+  theme_minimal()
 @
@@
 
-<<plot1, echo=FALSE, fig.height=4, fig.width=5.5, fig.cap='Conditional log-likelihoods computed using two methods for the real rotavirus dataset, with the SMC method replicated 36 times due to its high variances.
+<<plot2, echo=FALSE, fig.height=4, fig.width=5.5, fig.cap='Conditional log-likelihoods computed using two methods for the real rotavirus dataset, with the SMC method replicated 36 times due to its high variances.
 The blue line represents overlapping results from 36 SMC computations, while the red line is derived from PAL.
@@
 @
-
-\includegraphics[width=4.5in]{plot1}
 

### 2024-07-26 ms.Rnw
a few edits, but not addressing points in my email

@@
 Filtering algorithms are fundamental for inference on partially observed stochastic dynamic systems, since they provide access to the likelihood function and hence enable likelihood-based or Bayesian inference.
-Recently, a novel Poisson approximate likelihood (PAL) filter was introduced by \cite{wwr}.
+A novel Poisson approximate likelihood (PAL) filter was introduced by \cite{wwr}.
 PAL employs a Poisson approximation to conditional densities, offering a fast and consistent approximation to the likelihood function for a certain subset of partially observed Markov process models.
@@
 A central piece of evidence for PAL is the comparison in Table~1 of \cite{wwr}, which claims a large improvement for PAL over a standard particle filter algorithm.
-This evidence, based on a model and data from a previous scientific study, suggests at face value that researchers confronted with similar models would be irresponsible not to adopt PAL.
-We show that the evidence is flawed because the PAL calculations were carried out using a dataset differently scaled from the previous study.
+This evidence, based on a model and data from a previous scientific study, suggests at face value that researchers confronted with similar models should use PAL rather than previous particle filter methods.
+We show that the evidence is flawed because the PAL calculations were carried out using a dataset scaled differently from the previous study.
 If PAL and the particle filter are used on this new scale, the superficial advantage of PAL largely disappears.
@@
 opts_chunk$set(
-#    concordance = TRUE,
     tidy = TRUE,
@@
 ) -> sir
-
-# 
-# sir_panel <- panelPomp::panelPomp(list(unit1=sir),
-#                                   shared=NULL,
-#                                   specific=wwr_mle |> 
-#                                     as.matrix() |>
-#                                     `colnames<-`("unit1")
-# )
 @
@@
 bake(file=paste0(out_dir,"pfilter/original_dataset/palfilter_org.rds"),{
+  realdat <- ovov_sim_100[[i]] |> as.matrix()
   pal_lik_list_org <- rotavirus_SMC_qropxi(init_dist,realdat , 9, 

### 2024-07-27 ms.Rnw
addressing the email issues, correction on AIC

@@
 NP_MIF       = switch(RUN_LEVEL, 10, 50000)
-NMIF         = switch(RUN_LEVEL, 4,    100)
-ncores       = switch(RUN_LEVEL, 2,    36)
+NMIF         = switch(RUN_LEVEL, 5,    100)
+ncores       = switch(RUN_LEVEL, 4,    36)
 top_n_fits   = switch(RUN_LEVEL, 1,    12)
-sim_rep      = switch(RUN_LEVEL, 1,    100) # Total number of simulations
+sim_rep      = switch(RUN_LEVEL, 5,    100) # Total number of simulations
 
@@
 dmeas <- Csnippet("
-                  if (ISNA(cases1)) {
-                  lik = (give_log) ? 0 : 1;
-                  } else {
   lik =  dbinom(cases1, H1, q1, 1) +
   dbinom(cases2, H2, q2, 1) +
-                        dbinom(cases3, H3, q3, 1);
-
-                    lik = (give_log) ? lik : exp(lik);
-
-                    }")
+  dbinom(cases3, H3, q3, 1);"
+)
 rmeas <-  Csnippet("
@@
 
-# define parameters (without betas)
-params_fixed <- c(gamma=1, delta1=1/(5*52),delta2=1/(55*52), alpha=1/(78.86912*52), 
-                  mu=0, N=82372825, omega=1/(1*52))
 # full-estimate-rinit
-rinit <- Csnippet("
+rinit_full <- Csnippet("
   double m = N/(S10+I10+R10+S20+I20+R20+S30+I30+R30);
@@
 
+# define parameters (without betas)
+params_fixed <- c(gamma=1, delta1=1/(5*52),delta2=1/(55*52), alpha=1/(78.86912*52), 
+                  mu=0, N=82372825, omega=1/(1*52))
 # Set to MLE
-wwr_mle <- params_fixed
-
-wwr_mle <- c(params_fixed, 
+params_wwr <- c(params_fixed, 
   "beta1" = 11.48,
@@
   "sigma_q" = 0.021,
-             "sigma_xi" = 66.89,
+  "sigma_xi" = 66.89)
+
+params_full <- c(params_wwr,
   "S10" = 4.706102e-02,
@@
 
-wwr_mle_raw <- c(params_fixed, 
-             "beta1" = 11.48,
-             "beta2" = 0.25,
-             "beta3" = 0.35,
-             "phi" = 0.14,
-             "beta11" = 0.16,
-             "sigma_q" = 0.021,
-             "sigma_xi" = 66.89)
-
-pt <- pomp::parameter_trans(
+pt_full <- pomp::parameter_trans(
   log = c("beta1","beta2","beta3","sigma_q","sigma_xi"),
@@
 
+pt_wwr <- pomp::parameter_trans(
+  log = c("beta1","beta2","beta3","sigma_q","sigma_xi"),
+  logit=c("beta11"),
+  toEst= pomp::Csnippet("T_phi = logit(phi/(M_2PI));"),
+  fromEst= pomp::Csnippet("phi = M_2PI*expit(T_phi);")
+)
+
 read.table("real_rotavirus_metadata.txt") %>%
-  rbind(data.frame(time=0,cases1=NA,cases2=NA,cases3=NA)) %>%
   arrange(time) -> dat
 
-pomp(data = dat,
+statenames <- c(
+  "S1", "I1", "R1", "H1",
+  "S2", "I2", "R2", "H2",
+  "S3", "I3", "R3", "H3",
+  "q1", "q2", "q3"
+)
+
+sir_wwr <- pomp(data = dat,
   times="time",
@@
   rprocess = discrete_time(step.fun = rproc, delta.t = 1/4),
-     statenames = c("S1", "I1", "R1", "H1", 
-                    "S2", "I2", "R2", "H2",
-                    "S3", "I3", "R3", "H3", 
-                    "q1", "q2", "q3"),
-     paramnames = names(wwr_mle),
-     accumvars=c("H1", "H2", "H3"),
-     rinit=rinit,
-     partrans = pt,
-     params = wwr_mle
-) -> sir
+  statenames = statenames,
+  paramnames = names(params_wwr),
+  accumvars= c("H1", "H2", "H3"),
+  rinit= rinit_wwr,
+  partrans = pt_wwr,
+  params = params_wwr
+)
+
+sir_full <- pomp(sir_wwr,
+  rinit = rinit_full,
+  params = params_full,
+  partrans = pt_full,
+  statenames = statenames,
+  paramnames = names(params_full)
+)
 @
@@
   R30 = c(0.0001,0.005),
-  #fixed params
   gamma = c(1,1),
@@
 
-# c(apply(sir_box,1,function(x)runif(1,min=x[1],max=x[2])),
-#   params_fixed) -> starting
-
-bake(file = paste0(out_dir, "mif2_maximization/round_01/ovov_mif_01.rds"),{ 
+bake(file = paste0(out_dir, "mif2_maximization/round_01/mifs_global_01.rds"),{ 
   registerDoFuture()
@@
     pomp::mif2(
-      sir,
+      sir_full,
       Np = NP_MIF,
@@
 
-bake(file = paste0(out_dir, "mif2_maximization/round_01/ovov_el_01.rds"),{
+bake(file = paste0(out_dir, "mif2_maximization/round_01/el_01.rds"),{
   el_01 <- eval_logLik_pomp(mifs_global_01, ncores, NP_MIF, 4248930)
@@
 
-el_01 <- readRDS(paste0(out_dir,"mif2_maximization/round_01/ovov_el_01.rds"))
+el_01 <- readRDS(paste0(out_dir,"mif2_maximization/round_01/el_01.rds"))
 starting_values_02 <- choose_n_fits(el_01, top_n_fits, ncores)
 
-bake(file = paste0(out_dir,"mif2_maximization/round_02/ovov_mif_02.rds"),{ 
+bake(file = paste0(out_dir,"mif2_maximization/round_02/mifs_global_02.rds"),{ 
   registerDoFuture()
@@
     pomp::mif2(
-      sir,
-      Np = NP_MIF,
-      cooling.fraction.50 = 0.5,
+      mifs_global_01[[1]],
       rw.sd = rw_sd(beta1=0.01*sd_02,beta2=0.01*sd_02,beta3=0.01*sd_02,
@@
                 S30=ivp(0.24)*sd_02,I30=ivp(0.24)*sd_02,R30=ivp(0.24)*sd_02),
-      cooling.type = "geometric",
-      Nmif = NMIF,
       params = starting_values_02[[i]]
@@
 
-bake(file = paste0(out_dir, "mif2_maximization/round_02/ovov_el_02.rds"),{
+bake(file = paste0(out_dir, "mif2_maximization/round_02/el_02.rds"),{
   el_02 <- eval_logLik_pomp(mifs_global_02, ncores, NP_MIF, seed=4248930)
@@
 
-el_02 <- readRDS(paste0(out_dir, "mif2_maximization/round_02/ovov_el_02.rds"))
+el_02 <- readRDS(paste0(out_dir, "mif2_maximization/round_02/el_02.rds"))
 starting_values_03 <- choose_n_fits(el_02, top_n_fits, ncores)
 
-bake(file = paste0(out_dir, "mif2_maximization/round_03/ovov_mif_03.rds"),{ 
+bake(file = paste0(out_dir, "mif2_maximization/round_03/mifs_global_03.rds"),{ 
   registerDoFuture()
@@
     pomp::mif2(
-      sir,
-      Np = NP_MIF,
-      cooling.fraction.50 = 0.5,
+      mifs_global_02[[1]],
       rw.sd = rw_sd(beta1=0.01*sd_03,beta2=0.01*sd_03,beta3=0.01*sd_03,
@@
                 S30=ivp(0.24)*sd_03,I30=ivp(0.24)*sd_03,R30=ivp(0.24)*sd_03),
-      cooling.type = "geometric",
-      Nmif = NMIF,
       params = starting_values_03[[i]]
@@
 
-bake(file = paste0(out_dir,"mif2_maximization/round_03/ovov_el_03.rds"),{
+bake(file = paste0(out_dir,"mif2_maximization/round_03/el_03.rds"),{
   el_03 <- eval_logLik_pomp(mifs_global_03, ncores, NP_MIF, seed=4248930)
@@
 <<estimated-mle, echo=FALSE, message=FALSE, warning=FALSE>>=
-mif_el <- readRDS(paste0(out_dir,"mif2_maximization/round_03/ovov_el_03.rds")) 
+el_03 <- readRDS(paste0(out_dir,"mif2_maximization/round_03/el_03.rds")) 
 # Can be changed if more rounds needed
-est_mle_vec <- mif_el[which.max(mif_el$logLik),]
+est_mle_vec <- el_03[which.max(el_03$logLik),]
 @
@@
 
-<<pf-sim-100, echo=F, warning=F, message=F>>=
-# Set WWR's MLE
-wwr_mle_raw <- c(params_fixed, 
-             "beta1" = 11.48,
-             "beta2" = 0.25,
-             "beta3" = 0.35,
-             "phi" = 0.14,
-             "beta11" = 0.16,
-             "sigma_q" = 0.021,
-             "sigma_xi" = 66.89)
-
-i <- 1
-
+<<sim-100, echo=F, warning=F, message=F>>=
 bake(file=paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"),{
@@
 
+  i <- 1
   repeat {
     # Simulation
-    sim1 <- simulate(t0 = 0,
-                     times = c(1:416),
-                     dmeasure = dmeas,
-                     rmeasure = rmeas,
-                     rprocess = discrete_time(rproc, delta.t = 1/4),
-                     statenames = c("S1", "I1", "R1", "H1", 
-                                    "S2", "I2", "R2", "H2",
-                                    "S3", "I3", "R3", "H3",
-                                    "q1", "q2", "q3"),
-                     obsnames = c("cases1", "cases2", "cases3"),
-                     paramnames = names(wwr_mle_raw),
-                     accumvars = c("H1", "H2", "H3"),
-                     rinit = rinit_wwr,
-                     params = wwr_mle_raw)
+    sim_data <- simulate(sir_wwr, format="data.frame") |>
+      subset(select=c("time","cases1","cases2","cases3"))
 
-    sim_data <- as.data.frame(sim1)
     if (!any(sim_data == 0)) {
-      ovov_sim_100[[length(ovov_sim_100) + 1]] <- sim_data[, c("cases1", "cases2", "cases3")]
-      i <- i + 1  # Increment only if the condition is satisfied
+      ovov_sim_100[[length(ovov_sim_100) + 1]] <- sim_data
+      i <- i + 1  
     } else {
@@
 
-    if (length(ovov_sim_100) >= sim_rep) break  # Exit the loop when the desired number of successful iterations is reached
+    if (length(ovov_sim_100) >= sim_rep) break  
   }
@@
 })
-
-# load("ovov_sim_100.Rdata")
 @
@@
 bake(file=paste0(out_dir, "pfilter/simulated_dataset/pfilter_ovov_sim_100.rds"),{
-  sir_wwr <- list()
+  pfilter_obj <- list()
   pfilter_ovov_sim_100 <- c()
@@
   for(i in 1:length(ovov_sim_100)){
-  data.frame(time=c(1:416), ovov_sim_100[[i]]) |>  pomp(t0=0,
-           time="time",
-           rprocess=discrete_time(rproc,delta.t=1/4),
-           rinit=rinit_wwr,
-           dmeasure=dmeas,
-           rmeasure=rmeas,
-           accumvars=c("H1","H2","H3"),
-           statenames=c("S1","S2","S3",
-                        "I1","I2","I3",
-                        "R1","R2","R3",
-                        "H1","H2","H3",
-                        "q1","q2","q3"),
-           paramnames = names(wwr_mle_raw)
-      ) -> sir_wwr[[i]]
+    pomp(data = ovov_sim_100[[i]],
+    times="time",
+    t0=0,
+    dmeasure = dmeas,
+    rmeasure = rmeas,
+    rprocess = discrete_time(step.fun = rproc, delta.t = 1/4),
+    statenames = statenames,
+    paramnames = names(params_wwr),
+    accumvars= c("H1", "H2", "H3"),
+    rinit= rinit_wwr,
+    partrans = pt_wwr,
+    params = params_wwr
+    )  -> pfilter_obj[[i]]
 
@@
     ) %dofuture% {
-      pfilter(sir_wwr[[i]],Np=NP_MIF,save.states=T, params=wwr_mle_raw)
+      pfilter(pfilter_obj[[i]],Np=NP_MIF,save.states=T, params=params_wwr)
     } -> pfs
@@
 init_dist <- c(3876549, 30351, 1315221, 57139612, 871, 302852, 19573727, 2550, 131092)
-# initial_guess <- c(runif(1,5,20),runif(1,0.05,0.5),runif(1,0.2,1),runif(1,0.05,0.4),runif(1,0.05,0.4),runif(1,0.05,0.2),runif(1,10,100))
 initial_guess <- c(11.48,0.25,0.35,0.14,0.16,0.021,66.89)
-
-### Just a filter
-Sys.time()
 
@@
   for(i in 1:length(ovov_sim_100)){
-    pal_sim_dat <- ovov_sim_100[[i]] |> as.matrix()
+    pal_sim_dat <- ovov_sim_100[[i]][,-1] |> as.matrix()
     lik_list <- rotavirus_SMC_qropxi(init_dist,pal_sim_dat , 9, 
-                                     c(initial_guess[1],initial_guess[2],initial_guess[3],
-                                       initial_guess[4],initial_guess[5]), 
+      c(initial_guess[1],initial_guess[2],initial_guess[3],initial_guess[4],initial_guess[5]), 
       gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
@@
 
-Sys.time()
-
 # lik_list$ll_storage |> sum()
-
-# x <- c()
-# for(i in 1:10){
-#   x[i] <- any(ovov_sim_100[[i]]==0)
-# }
 @
 
-<<plot1,echo=FALSE, fig.height=4, fig.width=5.5, fig.cap='Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model.The particle filter and PAL used 50,000 particles.The red line is the line $Lik(pfilter)=Lik(palfilter)$.On average, the particle filter likelihood is $7.7$ log units higher than the PAL.
+<<plot1-100-sim-compare,echo=FALSE, fig.height=4, fig.width=5.5, fig.cap='Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model.The particle filter and PAL used 50,000 particles.The red line is the line $Lik(pfilter)=Lik(palfilter)$.On average, the particle filter likelihood is $7.7$ log units higher than the PAL.
 {This may not be wrong, since the likelihood evaluated by pfilter is not fully optimized by employing many particles.}'>>=
@@
 
-<<pfilter_org, echo=FALSE, message=FALSE, warning=FALSE>>=
-dat[-1, ] |>
-  pomp(t0=0,
-       time="time",
-       rprocess=discrete_time(rproc,delta.t=1/4), 
-       rinit=rinit_wwr,
-       dmeasure=dmeas,
-       rmeasure=rmeas,
-       accumvars=c("H1","H2","H3"),
-       statenames=c("S1","S2","S3",
-                    "I1","I2","I3",
-                    "R1","R2","R3",
-                    "H1","H2","H3",
-                    "q1","q2","q3"),
-       paramnames = names(wwr_mle_raw)
-  ) -> m1
-
-bake(file=paste0(out_dir,"pfilter/original_dataset/pfilter_org.rds"),{
+<<pfilter_wwr, echo=FALSE, message=FALSE, warning=FALSE>>=
+bake(file=paste0(out_dir,"pfilter/original_dataset/pfilter_wwr.rds"),{
   registerDoFuture()
@@
   ) %dofuture% {
-    pfilter(m1,Np=NP_MIF,params=wwr_mle_raw)
-  } -> pfs
-  pfs
+    pfilter(sir_wwr,Np=NP_MIF,params=params_wwr)
+  } -> pfilter_wwr
+  pfilter_wwr
 })
 
-bake(file=paste0(out_dir,"pfilter/original_dataset/pfilter_list_cond_lik.rds"),{
-  list_cond_lik <- list()
+bake(file=paste0(out_dir,"pfilter/original_dataset/pfilter_wwr_cond_lik.rds"),{
+  pfilter_wwr_cond_lik <- list()
   registerDoFuture()
@@
   for(k in 1:ncores){
-    list_cond_lik[[k]] <- pfs[[k]]@cond.logLik
+    pfilter_wwr_cond_lik[[k]] <- pfilter_wwr[[k]]@cond.logLik
   }
-  list_cond_lik
+  pfilter_wwr_cond_lik
 })
@@
 
-<<palfilter_org, echo=FALSE>>=
-load("real_rotavirus_metadata.Rdata")
+<<palfilter_wwr, echo=FALSE>>=
+realdat <- as.matrix(dat[,-1])
 prop <- 420
@@
 init_dist <- c(3876549, 30351, 1315221, 57139612, 871, 302852, 19573727, 2550, 131092)
-# initial_guess <- c(runif(1,5,20),runif(1,0.05,0.5),runif(1,0.2,1),runif(1,0.05,0.4),runif(1,0.05,0.4),runif(1,0.05,0.2),runif(1,10,100))
 initial_guess <- c(11.48,0.25,0.35,0.14,0.16,0.021,66.89)
@@
 
-Sys.time()
-bake(file=paste0(out_dir,"pfilter/original_dataset/palfilter_org.rds"),{
-  realdat <- ovov_sim_100[[i]] |> as.matrix()
-  pal_lik_list_org <- rotavirus_SMC_qropxi(init_dist,realdat , 9, 
-                                   c(initial_guess[1],initial_guess[2],
-                                     initial_guess[3],initial_guess[4],initial_guess[5]), 
-                                   gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
+bake(file=paste0(out_dir,"pfilter/original_dataset/palfilter_lik_list_wwr.rds"),{
+  palfilter_lik_list_wwr <- rotavirus_SMC_qropxi(init_dist,realdat , 9, 
+    c(initial_guess[1],initial_guess[2],initial_guess[3],initial_guess[4],initial_guess[5]),     gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
     norm_par = c(0.07,initial_guess[6]), 
     t(prop), NP_MIF,1)
-  pal_lik_list_org
+  palfilter_lik_list_wwr
 })
-Sys.time()
 @
 
-<<plot2, echo=FALSE, fig.height=4, fig.width=5.5, fig.cap='Conditional log-likelihoods computed using two methods for the real rotavirus dataset, with the SMC method replicated 36 times due to its high variances.
+<<plot2-cond-loglik, echo=FALSE, fig.height=4, fig.width=5.5, fig.cap='Conditional log-likelihoods computed using two methods for the real rotavirus dataset, with the SMC method replicated 36 times due to its high variances.
 The blue line represents overlapping results from 36 SMC computations, while the red line is derived from PAL.
@@
 time <- 1:416
-ncores <- length(list_cond_lik)
 
@@
 data_list <- lapply(1:ncores, function(i) {
-  data.frame(time = time, cond_loglik = list_cond_lik[[i]], method = "SMC")
+  data.frame(time = time, cond_loglik = pfilter_wwr_cond_lik[[i]], method = "SMC")
 })
@@
 # Create the PAL data frame
-df_pal <- data.frame(time = time, cond_loglik = pal_lik_list_org$ll_storage, method = "PAL")
+df_pal <- data.frame(time = time, cond_loglik = palfilter_lik_list_wwr$ll_storage, method = "PAL")
 
@@
   \hline
- OvOv & \Sexpr{pal_lik_list_org$log_lik} & $\Sexpr{logmeanexp(sapply(pfs, logLik))}$ \\ 
+ OvOv & \Sexpr{palfilter_lik_list_wwr$log_lik} & $\Sexpr{logmeanexp(logLik(pfilter_wwr))}$ \\ 
  \hline
@@
       \addlinespace
-      \textbf{AIC}  & $98866.65$ & $15154.75$ & $13778.08$ & $\Sexpr{round((-2)*((length(wwr_mle) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$\\
+      \textbf{AIC}  & $98866.65$ & $15154.75$ & $13778.08$ & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$\\
       \addlinespace

### 2024-07-27 ms.Rnw
correcting bake syntax and various other edits

@@
 \usepackage{multirow}
+\usepackage{makecell}
 \usepackage{threeparttable} % For table notes
 \usepackage{array} % For table formatting
-\usepackage{setspace}
-\usepackage{float} % add this in your preamble
-\usepackage{amsmath}
+\usepackage{subcaption}
+\usepackage{setspace} % For double-line spacing
+\usepackage{float} 
 \usepackage{tabularx}
+\usepackage[round]{natbib}
+\usepackage[font={small,it}]{caption}
+\usepackage[utf8]{inputenc}
+\usepackage{mathtools}
+\usepackage{fullpage}
+\usepackage{verbatim}
+\usepackage{amsfonts,amssymb,amsthm}
+\usepackage{url}
+\usepackage[colorlinks=true,citecolor=blue]{hyperref}
+% Using \doublespacing in the preamble 
+% changes the text to double-line spacing
+\doublespacing
+
+% Set page size and margins
+% Replace `letterpaper' with `a4paper' for UK/EU standard size
+\usepackage[a4paper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}
 
@@
 
-\usepackage[round]{natbib}
-\usepackage{multirow}
-\usepackage{array}
-\usepackage{makecell}
-\usepackage[font={small,it}]{caption}
-\usepackage[utf8]{inputenc}
-\usepackage{mathtools}
-\usepackage{subcaption}
-\usepackage{graphicx, fullpage, verbatim, amsmath}
-\usepackage{url, amsfonts, amssymb, amsthm,color, enumerate}
-\usepackage{amsfonts}
-\usepackage{bm}
-
-\doublespacing
-% \usepackage[english]{babel}
-
-% Set page size and margins
-% Replace `letterpaper' with `a4paper' for UK/EU standard size
-\usepackage[a4paper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}
-
-% Useful packages
-\usepackage{graphicx}
 \bibliographystyle{apalike}
-\usepackage[colorlinks=true,citecolor=blue]{hyperref}
-
-% Using \doublespacing in the preamble 
-% changes the text to double-line spacing
-\doublespacing
 
@@
 A central piece of evidence for PAL is the comparison in Table~1 of \cite{wwr}, which claims a large improvement for PAL over a standard particle filter algorithm.
-This evidence, based on a model and data from a previous scientific study, suggests at face value that researchers confronted with similar models should use PAL rather than previous particle filter methods.
-We show that the evidence is flawed because the PAL calculations were carried out using a dataset scaled differently from the previous study.
+This evidence, based on a model and data from a previous scientific study, might suggests that researchers confronted with similar models should use PAL rather than previous particle filter methods.
+Alternatively, the improvement in likelihood demonstrated over the results presented by \cite{stocks} might indicate an error in that previous work.
+We show that the comparison of log-likelihood values made by \cite{wwr} is flawed because the PAL calculations were carried out using a dataset scaled differently from the previous study.
 If PAL and the particle filter are used on this new scale, the superficial advantage of PAL largely disappears.
@@
 library(pomp)
-library(magrittr)
 library(ggplot2)
@@
 library(RcppArmadillo)
-# install.packages('RcppDist')
 sourceCpp('rotavirus_normq.cpp')
@@
   plan(multicore, workers=ncores)
-  mifs_global_01 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
+  foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
     pomp::mif2(
@@
   }
-  mifs_global_01
-})
+}) -> mifs_global_01
 
 bake(file = paste0(out_dir, "mif2_maximization/round_01/el_01.rds"),{
-  el_01 <- eval_logLik_pomp(mifs_global_01, ncores, NP_MIF, 4248930)
-  el_01
-})
+  eval_logLik_pomp(mifs_global_01, ncores, NP_MIF, 4248930)
+}) ->  el_01
+
 @
@@
 sd_02 = 2/3
-
-el_01 <- readRDS(paste0(out_dir,"mif2_maximization/round_01/el_01.rds"))
 starting_values_02 <- choose_n_fits(el_01, top_n_fits, ncores)
@@
   plan(multicore, workers=ncores)
-  mifs_global_02 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
+  foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
     pomp::mif2(
@@
   }
-  mifs_global_02
-})
+}) ->  mifs_global_02
 
 bake(file = paste0(out_dir, "mif2_maximization/round_02/el_02.rds"),{
-  el_02 <- eval_logLik_pomp(mifs_global_02, ncores, NP_MIF, seed=4248930)
-  el_02
-})
+  eval_logLik_pomp(mifs_global_02, ncores, NP_MIF, seed=4248930)
+}) ->  el_02
 @
@@
 sd_03 = (2/3)^2
-
-el_02 <- readRDS(paste0(out_dir, "mif2_maximization/round_02/el_02.rds"))
 starting_values_03 <- choose_n_fits(el_02, top_n_fits, ncores)
@@
   plan(multicore, workers=ncores)
-  mifs_global_03 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
+  foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
     pomp::mif2(
       mifs_global_02[[1]],
-      rw.sd = rw_sd(beta1=0.01*sd_03,beta2=0.01*sd_03,beta3=0.01*sd_03,
+      rw.sd = rw_sd(
+        beta1=0.01*sd_03,beta2=0.01*sd_03,beta3=0.01*sd_03,
         beta11=0.01*sd_03,phi=0.01*sd_03,
@@
         S20=ivp(0.24)*sd_03,I20=ivp(0.24)*sd_03,R20=ivp(0.24)*sd_03,
-                S30=ivp(0.24)*sd_03,I30=ivp(0.24)*sd_03,R30=ivp(0.24)*sd_03),
+        S30=ivp(0.24)*sd_03,I30=ivp(0.24)*sd_03,R30=ivp(0.24)*sd_03
+      ),
       params = starting_values_03[[i]]
@@
   }
-  mifs_global_03
-})
+}) -> mifs_global_03
 
 bake(file = paste0(out_dir,"mif2_maximization/round_03/el_03.rds"),{
-  el_03 <- eval_logLik_pomp(mifs_global_03, ncores, NP_MIF, seed=4248930)
-  el_03
-})
+  eval_logLik_pomp(mifs_global_03, ncores, NP_MIF, seed=4248930)
+}) ->  el_03
 @
@@
 <<estimated-mle, echo=FALSE, message=FALSE, warning=FALSE>>=
-el_03 <- readRDS(paste0(out_dir,"mif2_maximization/round_03/el_03.rds")) 
 # Can be changed if more rounds needed
@@
 bake(file=paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"),{
-  ovov_sim_100 <- list()
+  ovov_sim_computation <- list()
   registerDoFuture()
@@
     if (!any(sim_data == 0)) {
-      ovov_sim_100[[length(ovov_sim_100) + 1]] <- sim_data
+      ovov_sim_computation[[length(ovov_sim_computation) + 1]] <- sim_data
       i <- i + 1  
@@
 
-    if (length(ovov_sim_100) >= sim_rep) break  
+    if (length(ovov_sim_computation) >= sim_rep) break  
   }
-  ovov_sim_100
-})
+  ovov_sim_computation
+}) -> ovov_sim_100
 @
@@
 <<pfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
-ovov_sim_100 <- readRDS(paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"))
 bake(file=paste0(out_dir, "pfilter/simulated_dataset/pfilter_ovov_sim_100.rds"),{
   pfilter_obj <- list()
-  pfilter_ovov_sim_100 <- c()
+  pfilter_ovov_sim_computation <- c()
   registerDoFuture()
@@
 
-    pfilter_ovov_sim_100[i] <-logmeanexp(logLik(pfs))  
+    pfilter_ovov_sim_computation[i] <- logmeanexp(logLik(pfs))  
   }
-  pfilter_ovov_sim_100
-})
+  pfilter_ovov_sim_computation
+}) ->  pfilter_ovov_sim_100
 @
@@
 <<palfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
-ovov_sim_100 <- readRDS(paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"))
 prop <- 420
-
 init_dist <- c(3876549, 30351, 1315221, 57139612, 871, 302852, 19573727, 2550, 131092)
@@
 bake(file=paste0(out_dir, "pfilter/simulated_dataset/palfilter_ovov_sim_100.rds"),{
-  palfilter_ovov_sim_100 <- c()
+  palfilter_ovov_sim_computation <- c()
   registerDoFuture()
@@
     lik_list <- rotavirus_SMC_qropxi(init_dist,pal_sim_dat , 9, 
-      c(initial_guess[1],initial_guess[2],initial_guess[3],initial_guess[4],initial_guess[5]), 
-      gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
+      c(initial_guess[1],initial_guess[2],initial_guess[3],
+        initial_guess[4],initial_guess[5]), 
+        gamma_par = c(initial_guess[7],  1/initial_guess[7]
+      ), 
       norm_par = c(0.07,initial_guess[6]), 
       t(prop), NP_MIF,1)
-    palfilter_ovov_sim_100[i] <- lik_list$log_lik
+    palfilter_ovov_sim_computation[i] <- lik_list$log_lik
   }
-  palfilter_ovov_sim_100
-})
+  palfilter_ovov_sim_computation
+}) -> palfilter_ovov_sim_100
 
@@
     pfilter(sir_wwr,Np=NP_MIF,params=params_wwr)
-  } -> pfilter_wwr
-  pfilter_wwr
-})
+  } 
+}) ->  pfilter_wwr
 
 bake(file=paste0(out_dir,"pfilter/original_dataset/pfilter_wwr_cond_lik.rds"),{
-  pfilter_wwr_cond_lik <- list()
+  pfilter_wwr_cond_lik_computation <- list()
   registerDoFuture()
@@
   for(k in 1:ncores){
-    pfilter_wwr_cond_lik[[k]] <- pfilter_wwr[[k]]@cond.logLik
+    pfilter_wwr_cond_lik_computation[[k]] <- pfilter_wwr[[k]]@cond.logLik
   }
-  pfilter_wwr_cond_lik
-})
+  pfilter_wwr_cond_lik_computation
+}) -> pfilter_wwr_cond_lik
 @
@@
 bake(file=paste0(out_dir,"pfilter/original_dataset/palfilter_lik_list_wwr.rds"),{
-  palfilter_lik_list_wwr <- rotavirus_SMC_qropxi(init_dist,realdat , 9, 
-    c(initial_guess[1],initial_guess[2],initial_guess[3],initial_guess[4],initial_guess[5]),     gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
+  rotavirus_SMC_qropxi(
+    init_dist,realdat , 9, 
+    c(initial_guess[1],initial_guess[2],initial_guess[3],initial_guess[4],
+      initial_guess[5]),     gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
     norm_par = c(0.07,initial_guess[6]), 
-    t(prop), NP_MIF,1)
-  palfilter_lik_list_wwr
-})
+    t(prop), NP_MIF,1
+  )
+}) -> palfilter_lik_list_wwr
 @
@@
 Log-likelihood is a proper scoring rule for forecasts \cite{gneiting2}, and both the particle filter and PAL construct their log-likelihood estimates via a sequence of one-step forecasts.
-%% resulting in estimates of the conditional log-likelihood, $f_{Y
 Therefore, if the model is correctly specified, the approximation error in PAL can only decrease the expected log-likelihood.

### 2024-07-29 ms.Rnw
another RUN_LEVEL added

@@
 
-RUN_LEVEL = 1
-NP_MIF       = switch(RUN_LEVEL, 10, 50000)
-NMIF         = switch(RUN_LEVEL, 5,    100)
-ncores       = switch(RUN_LEVEL, 4,    36)
-top_n_fits   = switch(RUN_LEVEL, 1,    12)
-sim_rep      = switch(RUN_LEVEL, 5,    100) # Total number of simulations
+RUN_LEVEL = 2
+NP_MIF       = switch(RUN_LEVEL, 10,   5000,  50000)
+NMIF         = switch(RUN_LEVEL, 5,    20,    100)
+ncores       = switch(RUN_LEVEL, 4,    10,    36)
+top_n_fits   = switch(RUN_LEVEL, 1,    2,     12)
+sim_rep      = switch(RUN_LEVEL, 5,    20,    100) #sim_ovov
 

### 2024-07-29 ms.Rnw
Improve code formatting

@@
 On simulations where the model is correctly specified, the particle filter outperforms PAL.
-If a poorly fitting specification of initial value parameters is ammended, the particle filter also outperforms PAL on the actual data.
+If a poorly fitting specification of initial value parameters is amended, the particle filter also outperforms PAL on the actual data.
 \end{abstract}
@@
   int I = I1 + I2 + I3;
-  int trans_S1[3], trans_S2[3], trans_S3[2], trans_I1[3], trans_I2[3], trans_I3[2], trans_R1[3], trans_R2[3], trans_R3[2];
+  int trans_S1[3], trans_S2[3], trans_S3[2], trans_I1[3], trans_I2[3], 
+  trans_I3[2], trans_R1[3], trans_R2[3], trans_R3[2];
 
-  double prob_S1[3],prob_I1[3],prob_R1[3],prob_S2[3],prob_I2[3],prob_R2[3],prob_S3[2],prob_I3[2],prob_R3[2];
+  double prob_S1[3], prob_I1[3], prob_R1[3], prob_S2[3], prob_I2[3], prob_R2[3],
+  prob_S3[2], prob_I3[2], prob_R3[2];
 
@@
   // Define rate
-  prob_S1[0] = 1-exp(-dt*beta1*kappa*I/N); // 0->1
-  prob_S1[1] = 1-exp(-delta1*dt);
+  prob_S1[0] = 1 - exp(-dt*beta1*kappa*I/N); // 0->1
+  prob_S1[1] = 1 - exp(-delta1*dt);
   prob_S1[2] = exp(-delta1*dt) + exp(-dt*beta1*kappa*I/N) - 1;
 
-  prob_I1[0] = 1-exp(-gamma*dt);
-  prob_I1[1] = 1-exp(-delta1*dt);
+  prob_I1[0] = 1 - exp(-gamma*dt);
+  prob_I1[1] = 1 - exp(-delta1*dt);
   prob_I1[2] = exp(-gamma*dt)+exp(-delta1*dt) - 1;
@@
 
-  prob_S2[0] = 1-exp(-dt*beta2*kappa*I/N);
-  prob_S2[1] = 1-exp(-delta2*dt);
+  prob_S2[0] = 1 - exp(-dt*beta2*kappa*I/N);
+  prob_S2[1] = 1 - exp(-delta2*dt);
   prob_S2[2] = exp(-delta2*dt) + exp(-dt*beta2*kappa*I/N) - 1;
 
-  prob_I2[0] = 1-exp(-dt*gamma);
-  prob_I2[1] = 1-exp(-dt*delta2);
+  prob_I2[0] = 1 - exp(-dt*gamma);
+  prob_I2[1] = 1 - exp(-dt*delta2);
   prob_I2[2] = exp(-dt*gamma)+exp(-dt*delta2) - 1;
@@
   // D: Death
-  //// Note: Here S_1, S_2... are all old value from (t-1)
+  //// Note: Here S_1, S_2... are all old values from (t-1)
   rmultinom(S1, &prob_S1, 3, &trans_S1); // B, F, S-B-F
@@
 # define parameters (without betas)
-params_fixed <- c(gamma=1, delta1=1/(5*52),delta2=1/(55*52), alpha=1/(78.86912*52), 
-                  mu=0, N=82372825, omega=1/(1*52))
+params_fixed <- c(
+  gamma = 1, 
+  delta1 = 1/(5*52),
+  delta2 = 1/(55*52), 
+  alpha = 1/(78.86912*52), 
+  mu = 0, 
+  N = 82372825, 
+  omega = 1/(1*52)
+)
 # Set to MLE
-params_wwr <- c(params_fixed, 
+params_wwr <- c(
+  params_fixed, 
   "beta1" = 11.48,
@@
   "sigma_q" = 0.021,
-  "sigma_xi" = 66.89)
+  "sigma_xi" = 66.89
+)
 
-params_full <- c(params_wwr,
+params_full <- c(
+  params_wwr,
   "S10" = 4.706102e-02,
@@
   "R20" = 3.676601e-03,
-  "R30" = 1.591447e-03)
+  "R30" = 1.591447e-03
+)
 
 pt_full <- pomp::parameter_trans(
-  log = c("beta1","beta2","beta3","sigma_q","sigma_xi"),
-  logit=c("beta11"),
-  barycentric=c("S10","I10","R10",
+  log = c("beta1", "beta2", "beta3", "sigma_q", "sigma_xi"),
+  logit = c("beta11"),
+  barycentric = c("S10","I10","R10",
                   "S20","I20","R20",
                   "S30","I30","R30"),
-  toEst= pomp::Csnippet("T_phi = logit(phi/(M_2PI));"),
-  fromEst= pomp::Csnippet("phi = M_2PI*expit(T_phi);")
+  toEst = pomp::Csnippet("T_phi = logit(phi/(M_2PI));"),
+  fromEst = pomp::Csnippet("phi = M_2PI*expit(T_phi);")
 )
@@
 pt_wwr <- pomp::parameter_trans(
-  log = c("beta1","beta2","beta3","sigma_q","sigma_xi"),
-  logit=c("beta11"),
-  toEst= pomp::Csnippet("T_phi = logit(phi/(M_2PI));"),
-  fromEst= pomp::Csnippet("phi = M_2PI*expit(T_phi);")
+  log = c("beta1", "beta2", "beta3", "sigma_q", "sigma_xi"),
+  logit = c("beta11"),
+  toEst = pomp::Csnippet("T_phi = logit(phi/(M_2PI));"),
+  fromEst = pomp::Csnippet("phi = M_2PI*expit(T_phi);")
 )
 
-read.table("real_rotavirus_metadata.txt") %>%
-  arrange(time) -> dat
+dat <- read.table("real_rotavirus_metadata.txt") |>
+  arrange(time)
 
@@
 
-sir_wwr <- pomp(data = dat,
-  times="time",
-  t0=0,
+sir_wwr <- pomp(
+  data = dat,
+  times = "time",
+  t0 = 0,
   dmeasure = dmeas,
@@
   paramnames = names(params_wwr),
-  accumvars= c("H1", "H2", "H3"),
-  rinit= rinit_wwr,
+  accumvars = c("H1", "H2", "H3"),
+  rinit = rinit_wwr,
   partrans = pt_wwr,
@@
 sir_box <- list(
-  beta1=c(10,15),
-  beta2=c(0.2,0.4),
-  beta3=c(0.3,0.5),
-  phi=c(0.01,0.3),
-  beta11=c(0.1,0.2),
-  sigma_q=c(0.001,0.1),
-  sigma_xi=c(65,70),
-  S10 = c(0.01,0.05),
+  beta1 = c(10, 15),
+  beta2 = c(0.2, 0.4),
+  beta3 = c(0.3, 0.5),
+  phi = c(0.01, 0.3),
+  beta11 = c(0.1, 0.2),
+  sigma_q = c(0.001, 0.1),
+  sigma_xi = c(65, 70),
+  S10 = c(0.01, 0.05),
   I10 = c(0.0001, 0.0005),
   R10 = c(0.01, 0.05),
-  S20 = c(0.01,0.7),
-  I20 = c(0.000001,0.00002),
-  R20 = c(0.001,0.005),
-  S30 = c(0.01,0.5),
+  S20 = c(0.01, 0.7),
+  I20 = c(0.000001, 0.00002),
+  R20 = c(0.001, 0.005),
+  S30 = c(0.01, 0.5),
   I30 = c(0.000001, 0.0001),
-  R30 = c(0.0001,0.005),
-  gamma = c(1,1),
-  delta1=c(1/(5*52),1/(5*52)),
-  delta2=c(1/(55*52),1/(55*52)),
-  alpha=c(1/(78.86912*52),1/(78.86912*52)),
-  mu=c(0,0),
-  N=c(82372825,82372825),
-  omega=c(1/(1*52),1/(1*52))
+  R30 = c(0.0001, 0.005),
+  gamma = c(1, 1),
+  delta1 = c(1/(5*52), 1/(5*52)),
+  delta2 = c(1/(55*52), 1/(55*52)),
+  alpha = c(1/(78.86912*52), 1/(78.86912*52)),
+  mu = c(0, 0),
+  N = c(82372825, 82372825),
+  omega = c(1/(1*52), 1/(1*52))
 )
 
-design_matrix <- runif_design(lower = sapply(sir_box, `[`, 1),
+design_matrix <- runif_design(
+  lower = sapply(sir_box, `[`, 1),
   upper = sapply(sir_box, `[`, 2),
-                              n = ncores)
+  n = ncores
+)
 
@@
   plan(multicore, workers=ncores)
-  foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
+  foreach(i = 1:ncores, .options.future = list(seed=652643293)) %dofuture% {
     pomp::mif2(
@@
       cooling.fraction.50 = 0.5,
-      rw.sd = rw_sd(beta1=0.01,beta2=0.01,beta3=0.01,
-        beta11=0.01,phi=0.01,sigma_q=0.01,sigma_xi=0.01,
-        S10=ivp(0.24),I10=ivp(0.24),R10=ivp(0.24),
-        S20=ivp(0.24),I20=ivp(0.24),R20=ivp(0.24),
-        S30=ivp(0.24),I30=ivp(0.24),R30=ivp(0.24)),
+      rw.sd = rw_sd(
+        beta1 = 0.01, beta2=0.01, beta3=0.01, beta11 = 0.01, 
+        phi = 0.01, sigma_q = 0.01, sigma_xi = 0.01,
+        S10 = ivp(0.24), I10 = ivp(0.24), R10 = ivp(0.24),
+        S20 = ivp(0.24), I20 = ivp(0.24), R20 = ivp(0.24),
+        S30 = ivp(0.24), I30 = ivp(0.24), R30 = ivp(0.24)
+      ),
       cooling.type = "geometric",
@@
   registerDoFuture()
-  plan(multicore, workers=ncores)
-  foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
+  plan(multicore, workers = ncores)
+  foreach(i = 1:ncores, .options.future = list(seed = 652643293)) %dofuture% {
     pomp::mif2(
       mifs_global_01[[1]],
-      rw.sd = rw_sd(beta1=0.01*sd_02,beta2=0.01*sd_02,beta3=0.01*sd_02,
-                beta11=0.01*sd_02,phi=0.01*sd_02,
-                sigma_q=0.01*sd_02,sigma_xi=0.01*sd_02,
-                S10=ivp(0.24)*sd_02,I10=ivp(0.24)*sd_02,R10=ivp(0.24)*sd_02,
-                S20=ivp(0.24)*sd_02,I20=ivp(0.24)*sd_02,R20=ivp(0.24)*sd_02,
-                S30=ivp(0.24)*sd_02,I30=ivp(0.24)*sd_02,R30=ivp(0.24)*sd_02),
+      rw.sd = rw_sd(
+        beta1 = 0.01*sd_02, beta2 = 0.01*sd_02, beta3 = 0.01*sd_02,
+        beta11 = 0.01*sd_02, phi = 0.01*sd_02,
+        sigma_q = 0.01*sd_02, sigma_xi = 0.01*sd_02,
+        S10 = ivp(0.24)*sd_02, I10 = ivp(0.24)*sd_02, R10 = ivp(0.24)*sd_02,
+        S20 = ivp(0.24)*sd_02, I20 = ivp(0.24)*sd_02, R20 = ivp(0.24)*sd_02,
+        S30 = ivp(0.24)*sd_02, I30 = ivp(0.24)*sd_02, R30 = ivp(0.24)*sd_02
+      ),
       params = starting_values_02[[i]]
@@
   registerDoFuture()
-  plan(multicore, workers=ncores)
-  foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
+  plan(multicore, workers = ncores)
+  foreach(i = 1:ncores, .options.future = list(seed = 652643293)) %dofuture% {
     pomp::mif2(
@@
       rw.sd = rw_sd(
-        beta1=0.01*sd_03,beta2=0.01*sd_03,beta3=0.01*sd_03,
-        beta11=0.01*sd_03,phi=0.01*sd_03,
-        sigma_q=0.01*sd_03,sigma_xi=0.01*sd_03,
-        S10=ivp(0.24)*sd_03,I10=ivp(0.24)*sd_03,R10=ivp(0.24)*sd_03,
-        S20=ivp(0.24)*sd_03,I20=ivp(0.24)*sd_03,R20=ivp(0.24)*sd_03,
-        S30=ivp(0.24)*sd_03,I30=ivp(0.24)*sd_03,R30=ivp(0.24)*sd_03
+        beta1 = 0.01*sd_03, beta2 = 0.01*sd_03, beta3 = 0.01*sd_03,
+        beta11 = 0.01*sd_03, phi = 0.01*sd_03,
+        sigma_q = 0.01*sd_03, sigma_xi = 0.01*sd_03,
+        S10 = ivp(0.24)*sd_03, I10 = ivp(0.24)*sd_03, R10 = ivp(0.24)*sd_03,
+        S20 = ivp(0.24)*sd_03, I20 = ivp(0.24)*sd_03, R20 = ivp(0.24)*sd_03,
+        S30 = ivp(0.24)*sd_03, I30 = ivp(0.24)*sd_03, R30 = ivp(0.24)*sd_03
       ),
@@
 <<sim-100, echo=F, warning=F, message=F>>=
-bake(file=paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"),{
+bake(file = paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"),{
   ovov_sim_computation <- list()
   registerDoFuture()
-  plan(multicore, workers=ncores)
+  plan(multicore, workers = ncores)
 
@@
     # Simulation
-    sim_data <- simulate(sir_wwr, format="data.frame") |>
-      subset(select=c("time","cases1","cases2","cases3"))
+    sim_data <- simulate(sir_wwr, format = "data.frame") |>
+      subset(select=c("time", "cases1", "cases2", "cases3"))
 
@@
 <<pfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
-bake(file=paste0(out_dir, "pfilter/simulated_dataset/pfilter_ovov_sim_100.rds"),{
+bake(file = paste0(out_dir, "pfilter/simulated_dataset/pfilter_ovov_sim_100.rds"),{
   pfilter_obj <- list()
@@
     pomp(data = ovov_sim_100[[i]],
-      times="time",
-      t0=0,
+      times = "time",
+      t0 = 0,
       dmeasure = dmeas,
@@
       paramnames = names(params_wwr),
-      accumvars= c("H1", "H2", "H3"),
-      rinit= rinit_wwr,
+      accumvars = c("H1", "H2", "H3"),
+      rinit = rinit_wwr,
       partrans = pt_wwr,
@@
 
-    foreach(j=1:ncores, .combine=c,
-      .options.future=list(seed=998468235L)
+    foreach(j = 1:ncores, .combine = c,
+      .options.future = list(seed = 998468235L)
     ) %dofuture% {
-      pfilter(pfilter_obj[[i]],Np=NP_MIF,save.states=T, params=params_wwr)
+      pfilter(
+        pfilter_obj[[i]], Np = NP_MIF, save.states = T, params = params_wwr
+      )
     } -> pfs
-
     pfilter_ovov_sim_computation[i] <- logmeanexp(logLik(pfs))  
@@
 prop <- 420
-init_dist <- c(3876549, 30351, 1315221, 57139612, 871, 302852, 19573727, 2550, 131092)
-initial_guess <- c(11.48,0.25,0.35,0.14,0.16,0.021,66.89)
+init_dist <- c(
+  3876549, 30351, 1315221, 57139612, 871, 302852, 19573727, 2550, 131092
+)
+initial_guess <- c(11.48, 0.25, 0.35, 0.14, 0.16, 0.021, 66.89)
 
-bake(file=paste0(out_dir, "pfilter/simulated_dataset/palfilter_ovov_sim_100.rds"),{
+bake(file = paste0(out_dir, "pfilter/simulated_dataset/palfilter_ovov_sim_100.rds"),{
   palfilter_ovov_sim_computation <- c()
   registerDoFuture()
-  plan(multicore, workers=ncores)
+  plan(multicore, workers = ncores)
   for(i in 1:length(ovov_sim_100)){
     pal_sim_dat <- ovov_sim_100[[i]][,-1] |> as.matrix()
-    lik_list <- rotavirus_SMC_qropxi(init_dist,pal_sim_dat , 9, 
-      c(initial_guess[1],initial_guess[2],initial_guess[3],
-        initial_guess[4],initial_guess[5]), 
-        gamma_par = c(initial_guess[7],  1/initial_guess[7]
+    lik_list <- rotavirus_SMC_qropxi(
+      init_dist = init_dist, 
+      y_obs = pal_sim_dat, 
+      m = 9, 
+      regular_params = c(
+        initial_guess[1], initial_guess[2], initial_guess[3], initial_guess[4],
+        initial_guess[5]
       ), 
-      norm_par = c(0.07,initial_guess[6]), 
-      t(prop), NP_MIF,1)
+      gamma_par = c(initial_guess[7], 1/initial_guess[7]), 
+      norm_par = c(0.07, initial_guess[6]), 
+      prop = t(prop), 
+      n_particles = NP_MIF,
+      ncores = 1
+    )
     palfilter_ovov_sim_computation[i] <- lik_list$log_lik
@@
 {This may not be wrong, since the likelihood evaluated by pfilter is not fully optimized by employing many particles.}'>>=
-data <- data.frame(pfilter = pfilter_ovov_sim_100, palfilter = palfilter_ovov_sim_100)
+data <- data.frame(
+  pfilter = pfilter_ovov_sim_100, 
+  palfilter = palfilter_ovov_sim_100
+)
 
@@
   geom_abline(intercept = 0, slope = 1, color = "red") +
-  labs(title = "PAL vs SMC: OvOv with 100 simulations",
+  labs(
+    title = "PAL vs SMC: OvOv with 100 simulations",
     x = "pfilter",
-       y = "palfilter") +
+    y = "palfilter"
+  ) +
   theme_minimal()
@@
 <<pfilter_wwr, echo=FALSE, message=FALSE, warning=FALSE>>=
-bake(file=paste0(out_dir,"pfilter/original_dataset/pfilter_wwr.rds"),{
+bake(file = paste0(out_dir, "pfilter/original_dataset/pfilter_wwr.rds"),{
   registerDoFuture()
   plan(multicore, workers = ncores)
-  foreach(i=1:ncores, .combine=c,
-          .options.future=list(seed=998468235L)
+  foreach(
+    i = 1:ncores, .combine = c, .options.future = list(seed = 998468235L)
   ) %dofuture% {
-    pfilter(sir_wwr,Np=NP_MIF,params=params_wwr)
+    pfilter(sir_wwr, Np = NP_MIF, params = params_wwr)
   } 
@@
 
-bake(file=paste0(out_dir,"pfilter/original_dataset/pfilter_wwr_cond_lik.rds"),{
+bake(file = paste0(out_dir, "pfilter/original_dataset/pfilter_wwr_cond_lik.rds"),{
   pfilter_wwr_cond_lik_computation <- list()
@@
 
-init_dist <- c(3876549, 30351, 1315221, 57139612, 871, 302852, 19573727, 2550, 131092)
-initial_guess <- c(11.48,0.25,0.35,0.14,0.16,0.021,66.89)
+init_dist <- c(
+  3876549, 30351, 1315221, 57139612, 871, 302852, 19573727, 2550, 131092
+)
+initial_guess <- c(11.48, 0.25, 0.35, 0.14, 0.16, 0.021, 66.89)
 
@@
 registerDoFuture()
-plan(multicore, workers=ncores)
+plan(multicore, workers = ncores)
 
-bake(file=paste0(out_dir,"pfilter/original_dataset/palfilter_lik_list_wwr.rds"),{
+bake(file = paste0(out_dir,"pfilter/original_dataset/palfilter_lik_list_wwr.rds"),{
   rotavirus_SMC_qropxi(
-    init_dist,realdat , 9, 
-    c(initial_guess[1],initial_guess[2],initial_guess[3],initial_guess[4],
-      initial_guess[5]),     gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
-    norm_par = c(0.07,initial_guess[6]), 
-    t(prop), NP_MIF,1
+    init_dist = init_dist, 
+    y_obs = realdat, 
+    m = 9, 
+    regular_params = c(
+      initial_guess[1], initial_guess[2], initial_guess[3], initial_guess[4],
+      initial_guess[5]
+    ), 
+    gamma_par = c(initial_guess[7], 1/initial_guess[7]), 
+    norm_par = c(0.07, initial_guess[6]), 
+    prop = t(prop), 
+    n_particles = NP_MIF, 
+    ncores = 1
   ) 
@@
 data_list <- lapply(1:ncores, function(i) {
-  data.frame(time = time, cond_loglik = pfilter_wwr_cond_lik[[i]], method = "SMC")
+  data.frame(
+    time = time, 
+    cond_loglik = pfilter_wwr_cond_lik[[i]], 
+    method = "SMC"
+  )
 })
@@
 # Create the PAL data frame
-df_pal <- data.frame(time = time, cond_loglik = palfilter_lik_list_wwr$ll_storage, method = "PAL")
+df_pal <- data.frame(
+  time = time, 
+  cond_loglik = palfilter_lik_list_wwr$ll_storage, 
+  method = "PAL"
+)
 
@@
 # Plot with ggplot
-ggplot(df, aes(x = time, y = cond_loglik, group = interaction(method, factor(1:ncores)), color = method)) +
+ggplot(
+  df, 
+  aes(
+    x = time, y = cond_loglik, 
+    group = interaction(method, factor(1:ncores)), 
+    color = method
+  )
+) +
   geom_line() +
   theme_minimal() +
-  labs(title = "Conditional Log Likelihood Over Time",
+  labs(
+    title = "Conditional Log Likelihood Over Time",
     x = "Time",
     y = "Conditional Log Likelihood",
-       color = "Method") +
+    color = "Method"
+  ) +
   scale_color_manual(values = c("SMC" = "blue", "PAL" = "red")) +

### 2024-07-29 ms.Rnw
Improve code formatting a little more

@@
 
-sir_full <- pomp(sir_wwr,
+sir_full <- pomp(
+  sir_wwr,
   rinit = rinit_full,
@@
   registerDoFuture()
-  plan(multicore, workers=ncores)
-  foreach(i = 1:ncores, .options.future = list(seed=652643293)) %dofuture% {
+  plan(multicore, workers = ncores)
+  foreach(i = 1:ncores, .options.future = list(seed = 652643293)) %dofuture% {
     pomp::mif2(
@@
       rw.sd = rw_sd(
-        beta1 = 0.01, beta2=0.01, beta3=0.01, beta11 = 0.01, 
+        beta1 = 0.01, beta2 = 0.01, beta3 = 0.01, beta11 = 0.01, 
         phi = 0.01, sigma_q = 0.01, sigma_xi = 0.01,
@@
 bake(file = paste0(out_dir,"mif2_maximization/round_03/el_03.rds"),{
-  eval_logLik_pomp(mifs_global_03, ncores, NP_MIF, seed=4248930)
+  eval_logLik_pomp(mifs_global_03, ncores, NP_MIF, seed = 4248930)
 }) ->  el_03
@@
   registerDoFuture()
-  plan(multicore, workers=ncores)
+  plan(multicore, workers = ncores)
 
   for(i in 1:length(ovov_sim_100)){
-    pomp(data = ovov_sim_100[[i]],
+    pomp(
+      data = ovov_sim_100[[i]],
       times = "time",
@@
       pfilter(
-        pfilter_obj[[i]], Np = NP_MIF, save.states = T, params = params_wwr
+        pfilter_obj[[i]], Np = NP_MIF, save.states = TRUE, params = params_wwr
       )

### 2024-08-01 ms.Rnw
start editting$

@@
 Two models for the same data can properly be compared by their likelihood, even if the models have entirely different structures.
-One can making allowance for the number of estimated parameters using a quantity such as Akaike's information criterion [REF].
+One can make allowance for the number of estimated parameters using a quantity such as Akaike's information criterion [REF].
 However, if data are rescaled, a correction is required to make likelihoods comparable.
@@
 The reporting rate used by \cite{stocks} varied over time, but was generally around $7\%$.
-On 1200 data points, this corresponds to a discrepancy of around $-1200\log(0.07)=3200$ log-likelihood units, largely explaining the difference reported in Table~1 and interpreted by \cite{wwr} as evidence supporting PAL.
-The comparison can be corrected either by applying the method of \cite{stocks} to the data of \cite{wwr} or vvice versa.
+On $\approx 1200$ data points ($3*416$), this corresponds to a discrepancy of around $-1200\log(0.07)=3200$ log-likelihood units, largely explaining the difference reported in Table~1 and interpreted by \cite{wwr} as evidence supporting PAL.
+The comparison can be corrected either by applying the method of \cite{stocks} to the data of \cite{wwr} or vice versa.
 Since the method of \cite{stocks} is applicable to a more general class of models, and supported by published software, it was more convenient to apply this method to the model and data of \cite{wwr}.

### 2024-08-05 ms.Rnw
more dofuture, one bib, edit, gl

@@
 If a poorly fitting specification of initial value parameters is amended, the particle filter also outperforms PAL on the actual data.
+Nevertheless, PAL still yields reliable results and demonstrates superiority in computational time.
 \end{abstract}
@@
 
-RUN_LEVEL = 2
+RUN_LEVEL = 1
 NP_MIF       = switch(RUN_LEVEL, 10,   5000,  50000)
@@
 pop <- 82372825
+time <- c(1:416)
 
@@
 
-# define parameters (without betas)
+# define parameters (fixed)
 params_fixed <- c(
@@
 Two models for the same data can properly be compared by their likelihood, even if the models have entirely different structures.
-One can make allowance for the number of estimated parameters using a quantity such as Akaike's information criterion [REF].
+One can make allowance for the number of estimated parameters using a quantity such as Akaike's information criterion \citep{Akaike1974ANL}.
 However, if data are rescaled, a correction is required to make likelihoods comparable.
 If one model describes a dataset in grams, and another describes it in kilograms, then the latter model will earn an increased log-likelihood of $\log(10^3)$ for each data point simply because of the change in scale.
-Presenting a direct comparison of a likelihood for the data in grams with a likelihood for the data in kilograms would evidently be inapprpriate.
-\cite{stocks} fitted their model to a dataset derived by dividing the original reported count data by an estimated reporting rate, to put their data on the scale of the actual number of cases in the population.
-\cite{wwr} fitted directly to the report data.
+Presenting a direct comparison of a likelihood for the data in grams with a likelihood for the data in kilograms would evidently be inappropriate.
+\cite{stocks} fitted their model to a dataset derived by dividing the original reported count data by an estimated reporting rate, to put their data on the scale of the actual number of cases in the population, whereas \cite{wwr} fitted directly to the report data.
 The reporting rate used by \cite{stocks} varied over time, but was generally around $7\%$.
-On $\approx 1200$ data points ($3*416$), this corresponds to a discrepancy of around $-1200\log(0.07)=3200$ log-likelihood units, largely explaining the difference reported in Table~1 and interpreted by \cite{wwr} as evidence supporting PAL.
+On approximately 1200 data points, this corresponds to a discrepancy of around $-1200\log(0.07) \approx 3200$ log-likelihood units, largely explaining the difference reported in Table~1 and interpreted by \cite{wwr} as evidence supporting PAL.
 The comparison can be corrected either by applying the method of \cite{stocks} to the data of \cite{wwr} or vice versa.
 Since the method of \cite{stocks} is applicable to a more general class of models, and supported by published software, it was more convenient to apply this method to the model and data of \cite{wwr}.
-The large discrepancy in log-likelihood disappears at this point (see our Table~\ref{tab:ovovrealdata} , column \ed{XXX}).
-This re-analysis does, however, show a small advantage for PAL.
+The large discrepancy in log-likelihood disappears at this point by recomputing the likelihood of the model using PAL and particle filter separately (see Table~\ref{tab:ovovrealdata}).
+This re-analysis does, however, show a discrepancy between two methods.
 We continued our investigation to establish the cause of this.
@@
   plan(multicore, workers = ncores)
-  for(i in 1:length(ovov_sim_100)){
+  foreach(
+    i = 1:length(ovov_sim_100), .combine = c, .options.future = list(seed = 998468235L)
+  ) %dofuture% {
     pal_sim_dat <- ovov_sim_100[[i]][,-1] |> as.matrix()
@@
       n_particles = NP_MIF,
-      ncores = 1
+      ncores = ncores
     )
@@
 
-<<plot1-100-sim-compare,echo=FALSE, fig.height=4, fig.width=5.5, fig.cap='Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model.The particle filter and PAL used 50,000 particles.The red line is the line $Lik(pfilter)=Lik(palfilter)$.On average, the particle filter likelihood is $7.7$ log units higher than the PAL.
-{This may not be wrong, since the likelihood evaluated by pfilter is not fully optimized by employing many particles.}'>>=
+<<plot1-100-sim-compare,echo=FALSE, fig.height=4, fig.width=5.5, fig.cap='Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model.The particle filter and PAL used 50,000 particles.The red line is the line $Lik(pfilter)=Lik(palfilter)$.On average, the particle filter likelihood is $7.7$ log units higher than the PAL.'>>=
 data <- data.frame(
@@
 realdat <- as.matrix(dat[,-1])
-prop <- 420
-
-init_dist <- c(
-  3876549, 30351, 1315221, 57139612, 871, 302852, 19573727, 2550, 131092
-)
-initial_guess <- c(11.48, 0.25, 0.35, 0.14, 0.16, 0.021, 66.89)
-
-### palfiter
-registerDoFuture()
-plan(multicore, workers = ncores)
 
 bake(file = paste0(out_dir,"pfilter/original_dataset/palfilter_lik_list_wwr.rds"),{
+ set.seed(1223)
  rotavirus_SMC_qropxi(
@@
     n_particles = NP_MIF, 
-    ncores = 1
+    ncores = ncores
   ) 
@@
 The main sources of likelihood shortfall of SMC are at time points $t=1, 2, 3, 11, 81, 194$, and $325-333$.'>>=
-# Create your data
-time <- 1:416
 
-# Combine the data into a single dataframe for SMC
+# Combine the data into a single dataframe for SMC with grouping variable
 data_list <- lapply(1:ncores, function(i) {
@@
     cond_loglik = pfilter_wwr_cond_lik[[i]], 
-    method = "SMC"
+    method = "SMC",
+    group = i
   )
@@
   cond_loglik = palfilter_lik_list_wwr$ll_storage, 
-  method = "PAL"
+  method = "PAL",
+  group = 1  # Single group for PAL
 )
@@
 # Plot with ggplot
-ggplot(
-  df, 
-  aes(
-    x = time, y = cond_loglik, 
-    group = interaction(method, factor(1:ncores)), 
-    color = method
-  )
-) +
-  geom_line() +
+ggplot(df, aes(x = time, y = cond_loglik, color = method)) +
+  geom_line(data = df %>% filter(method == "SMC"), aes(group = group), color = "blue", alpha = 0.7) +
+  geom_line(data = df %>% filter(method == "PAL"), aes(group = group), color = "red") +
   theme_minimal() +
@@
   ) +
-  scale_color_manual(values = c("SMC" = "blue", "PAL" = "red")) +
   theme(legend.position = "bottom")
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.55\textwidth]{thesis-figs/PAL_vs_SMC_ovov_100.png}
-\caption{\label{fig:PAL_vs_SMC_ovov_100}Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model.
-The particle filter and PAL used 50,000 particles.
-The red line is the line $Lik(pfilter)=Lik(palfilter)$.
-On average, the particle filter likelihood is $7.7$ log units higher than the PAL.
-%%  -6224.245 - (-6231.97) = 7.725
-}
-\end{figure}
+% \begin{figure}[h]
+% \centering
+% \includegraphics[width=0.55\textwidth]{thesis-figs/PAL_vs_SMC_ovov_100.png}
+% \caption{\label{fig:PAL_vs_SMC_ovov_100}Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model.
+% The particle filter and PAL used 50,000 particles.
+% The red line is the line $Lik(pfilter)=Lik(palfilter)$.
+% On average, the particle filter likelihood is $7.7$ log units higher than the PAL.
+% %%  -6224.245 - (-6231.97) = 7.725
+% }
+% \end{figure}
 
@@
 We tested this on simulated data, for which the model of \cite{wwr} is correctly specified.
-We found that, in this case, the particle filter out-performs PAL (Figure~\ref{fig:PAL_vs_SMC_ovov_100}).
+We found that, in this case, the particle filter out-performs PAL (Figure~1).
 This supports hypothesis 2, above.
@@
   \hline
- OvOv & \Sexpr{palfilter_lik_list_wwr$log_lik} & $\Sexpr{logmeanexp(logLik(pfilter_wwr))}$ \\ 
+ OvOv & \Sexpr{round(palfilter_lik_list_wwr$log_lik,2)} & $\Sexpr{round(logmeanexp(logLik(pfilter_wwr),2))}$ \\ 
  \hline
@@
 
-\begin{figure}[H]
-\centering
-\includegraphics[width=0.9\textwidth]{thesis-figs/cond_SMC_PAL_realdat.png}
-\caption{\label{fig:cond_SMC_PAL_realdat}Conditional log-likelihoods computed using two methods for the real rotavirus dataset, with the SMC method replicated 36 times due to its high variances.
+% \begin{figure}[H]
+% \centering
+% \includegraphics[width=0.9\textwidth]{thesis-figs/cond_SMC_PAL_realdat.png}
+% \caption{\label{fig:cond_SMC_PAL_realdat}Conditional log-likelihoods computed using two methods for the real rotavirus dataset, with the SMC method replicated 36 times due to its high variances.
 The blue line represents overlapping results from 36 SMC computations, while the red line is derived from PAL.
 The main sources of likelihood shortfall of SMC are at time points $t=1, 2, 3, 11, 81, 194$, and $325-333$.}
-\end{figure}
+% \end{figure}
 
-Figure~ \ref{fig:cond_SMC_PAL_realdat} indicates that the worst conditional log-likelihood values, estimated using the particle filter, arise at the start of the time series.
+Figure~2 indicates that the worst conditional log-likelihood values, estimated using the particle filter, arise at the start of the time series.
 This could be a problem with data collection, or some subtle issue with the science of the epidemiological system, but a simple possibility is that the initial values of the latent state process might be poorly specified.
@@
 First, we verified that the two models are essentially identical.
-Here, we simulated data from the model 1000 times at the MLE shown in Table \ref{tab:mlebywwr} and compared the summary statistics of 1000 simulated time series from two models using the t-test.
+Here, we simulated data from the two models 1000 times at the MLE shown in Table \ref{tab:mlebywwr} and compared the summary statistics of 1000 simulated time series from two models using the t-test.
 Summary statistics including the mean, median, and variance for all three age groups were tested.
 The results showed no significant differences between the two models from which the data were simulated, confirming that both the latent process model and the observation model are essentially consistent with each other, given that the observations were directly simulated.
-
-From Table \ref{tab:model_performance}, it is evident that the OvOv model, with overdispersion in both the dynamic system and the measurement model, yields the lowest AIC.
+From Table \ref{tab:mlebywwr}, it is evident that the OvOv model, with overdispersion in both the dynamic system and the measurement model, yields the lowest AIC.
 We therefore focus on this version of the model.
 
-This article has focused on likelihood evaluation, but the inferences presented also require the likelihood estimate to maximized.
+This article has focused on likelihood evaluation, but the inferences presented also require the likelihood estimate to be maximized.
 \cite{stocks} used iterated filtering to maximize the particle filter likelihood estimate, whereas \cite{wwr} used direct numerical maximization.
@@
 
-An extended description of our reanalysis of \cite{wwr} is provided by \cite{hao24}.
+Our workflow is fully reproducible. An extended description of our reanalysis of \cite{wwr} is provided by \cite{hao24}.
 That thesis also contains some additional results reinforcing the conclusions presented in this article.

### 2024-08-09 ms.Rnw
edits to the intro

@@
 A central piece of evidence for PAL is the comparison in Table~1 of \cite{wwr}, which claims a large improvement for PAL over a standard particle filter algorithm.
-This evidence, based on a model and data from a previous scientific study, might suggests that researchers confronted with similar models should use PAL rather than previous particle filter methods.
-Alternatively, the improvement in likelihood demonstrated over the results presented by \cite{stocks} might indicate an error in that previous work.
+This evidence, based on a model and data from a previous scientific study by \citet{stocks}, might suggests that researchers confronted with similar models should use PAL rather than previous particle filter methods.
+Alternatively, the improvement in likelihood found by \cite{wwr} compared to \cite{stocks} might indicate a weakness in that previous work.
 We show that the comparison of log-likelihood values made by \cite{wwr} is flawed because the PAL calculations were carried out using a dataset scaled differently from the previous study.
@@
 If a poorly fitting specification of initial value parameters is amended, the particle filter also outperforms PAL on the actual data.
-Nevertheless, PAL still yields reliable results and demonstrates superiority in computational time.
 \end{abstract}
@@
 The authors of that paper were given the opportunity to submit a correction, after we shared the results of our investigation with them, but they declined.
+The theory developed by \cite{wwr} shows that PAL has some potentially useful scaling properties, but the numerical results in Table~1 appear to show much stronger performance than a standard particle filter on this example of moderate size.
 Our task here is to correct the error in Table~1 so that researchers considering whether to implement Poisson approximate likelihood (PAL) are appropriately informed about its benefits.
 Table~1 reanalyzes the model and data of \citet{stocks}, for which the likelihood was calculated using a particle filter.
-The theory developed by \cite{wwr} shows that PAL has some potentially useful scaling properties, but the numerical results in Table~1 appear to show much stronger performance than the particle filter on this example of moderate size.
+\citet{stocks} found strong evidence for the importance of overdispersion in a stochastic dynamic model for their epidemiological data.
+This is significant because earlier research on population dynamics largely avoided consideration of overdispersion, perhaps because of the lack of statistical methodology to fit such models. 
+The conclusions of \cite{stocks} hinge on a comparison of likelihoods, and so the results of \citet{wwr} discredit those conclusions by indicating that \citet{stocks} used an inaccurate filter to calculate their likelihoods.
+An important consequence of correcting Table~1 is that the results of \cite{stocks} stand undiminished.
 
-First, we show that most of this apparent improvement for PAL arises because \cite{wwr} used a different scaling of the data from \cite{stocks}.
-%% \section{Comparing PAL with the particle filter} 
+First, we show that most of the apparent advantage for PAL, compared to a particle filter, arises because \cite{wwr} used a different scaling of the data from \cite{stocks}.
 Two models for the same data can properly be compared by their likelihood, even if the models have entirely different structures.

### 2024-08-10 ms.Rnw
some more edits, but an error running at run level 1

@@
 However, if data are rescaled, a correction is required to make likelihoods comparable.
-If one model describes a dataset in grams, and another describes it in kilograms, then the latter model will earn an increased log-likelihood of $\log(10^3)$ for each data point simply because of the change in scale.
+For example, if one model describes a dataset in grams and another describes it in kilograms, then the latter model will earn an increased log-likelihood of $\log(10^3)$ for each data point simply because of the change in scale.
 Presenting a direct comparison of a likelihood for the data in grams with a likelihood for the data in kilograms would evidently be inappropriate.
@@
 
-# lik_list$ll_storage |> sum()
 @
 
-<<plot1-100-sim-compare,echo=FALSE, fig.height=4, fig.width=5.5, fig.cap='Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model.The particle filter and PAL used 50,000 particles.The red line is the line $Lik(pfilter)=Lik(palfilter)$.On average, the particle filter likelihood is $7.7$ log units higher than the PAL.'>>=
+<<plot1-100-sim-compare,echo=FALSE, fig.height=4, fig.width=5.5, fig.cap='Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model.
+The particle filter and PAL used 50,000 particles.
+The red line corresponds to equality of the two estimates.
+On average, the particle filter likelihood estimate is $7.7$ log units higher than the PAL.'>>=
 data <- data.frame(
@@
 We tested this on simulated data, for which the model of \cite{wwr} is correctly specified.
-We found that, in this case, the particle filter out-performs PAL (Figure~1).
-This supports hypothesis 2, above.
+We focus on the version of the model with overdispersion in both the population dynamics and the measurement process.
+This version of the model, called OvOv, was found by both \cite{wwr} and \cite{stocks} to be the best choice to fit the data.
+For this simulation study, the particle filter out-performs PAL (Figure~1) which supports hypothesis 2, above.
 Consequently, we looked for model misspecification by comparing the conditional log-likelihood estimates at each time point, as described in the Methods section.
@@
 \centering % Centers the table
-\caption{log-likelihood at the MLE of OvOv model for the real rotavirus data, computed using two filtering methods, employed 36 replicates and 50,000 particles} % This is effectively the table title
+\caption{log-likelihood at the MLE of OvOv model for the real rotavirus data, computed using two filtering methods, employed 36 replicates and 50,000 particles} 
 \label{tab:ovovrealdata} % For referencing this table elsewhere in your text
@@
   \begin{threeparttable}
-    \caption{Maximum likelihood estimation of parameters by PAL for three models \citep{wwr}, and by mif2 algorithm \citep{pomppackagepaper} for the OvOv model.
-    \ed{LET'S DISCUSS WHETHER WE WANT TO SHOW JUST OV-OV HERE. IT WOULD BE SIMPLER.}
+    \caption{Maximum likelihood estimation of parameters by PAL \citep{wwr}, and by mif2 algorithm \citep{pomppackagepaper} for the OvOv model.}
     }
     \label{tab:mlebywwr}
-    \begin{tabular}{>{\raggedright\arraybackslash}p{2cm} *{3}{>{\centering\arraybackslash}p{1.5cm}} >{\centering\arraybackslash}p{2.5cm}}
+    \begin{tabular}
       \toprule
-      Parameter      & EqEq  & EqOv  & OvOv  & OvOv (\textbf{pomp})$^*$ \\
+      Parameter      & OvOv  & OvOv (\textbf{pomp})$^*$ \\
       \midrule
-      \(\beta_1\)    & 12.15 & 12.74 & 11.48 & $\Sexpr{round(est_mle_vec["beta1"],2)}$\\
-      \(\beta_2\)    & 0.22  & 0.21  & 0.25  & $\Sexpr{round(est_mle_vec["beta2"],2)}$\\
-      \(\beta_3\)    & 0.34  & 0.31  & 0.35  & $\Sexpr{round(est_mle_vec["beta3"],2)}$\\
-      \(\phi\)       & 0.017 & 0.14  & 0.14  & $\Sexpr{round(est_mle_vec["phi"],2)}$\\
-      \(\rho\)       & 0.022 & 0.19  & 0.16  & $\Sexpr{round(est_mle_vec["beta11"],2)}$\\
-      \(\sigma^2_q\) & n/a   & 0.042 & 0.021 & $\Sexpr{round(est_mle_vec["sigma_q"],3)}$\\
-      \(\sigma_\xi\) & n/a   & n/a   & 66.89 & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$\\
-      \addlinespace
-      \multicolumn{1}{l}{\(S_{10}\)} & \multicolumn{3}{c}{$3876549^*$} & $\Sexpr{round(pop*est_mle_vec["S10"],0)}$\\
-      \multicolumn{1}{l}{\(I_{10}\)} & \multicolumn{3}{c}{$30351$}     & $\Sexpr{round(pop*est_mle_vec["I10"],0)}$\\
-      \multicolumn{1}{l}{\(R_{10}\)} & \multicolumn{3}{c}{$1315221$}   & $\Sexpr{round(pop*est_mle_vec["R10"],0)}$\\
-      \multicolumn{1}{l}{\(S_{20}\)} & \multicolumn{3}{c}{$57139612$}  & $\Sexpr{round(pop*est_mle_vec["S20"],0)}$\\
-      \multicolumn{1}{l}{\(I_{20}\)} & \multicolumn{3}{c}{$871$}       & $\Sexpr{round(pop*est_mle_vec["I20"],0)}$\\
-      \multicolumn{1}{l}{\(R_{20}\)} & \multicolumn{3}{c}{$302852$}    & $\Sexpr{round(pop*est_mle_vec["R20"],0)}$\\
-      \multicolumn{1}{l}{\(S_{30}\)} & \multicolumn{3}{c}{$19573727$}  & $\Sexpr{round(pop*est_mle_vec["S30"],0)}$\\
-      \multicolumn{1}{l}{\(I_{30}\)} & \multicolumn{3}{c}{$2550$}      & $\Sexpr{round(pop*est_mle_vec["I30"],0)}$\\
-      \multicolumn{1}{l}{\(R_{30}\)} & \multicolumn{3}{c}{$131092$}    & $\Sexpr{round(pop*est_mle_vec["R30"],0)}$\\
+      \(\beta_1\)    & 11.48 & $\Sexpr{round(est_mle_vec["beta1"],2)}$\\
+      \(\beta_2\)    & 0.25  & $\Sexpr{round(est_mle_vec["beta2"],2)}$\\
+      \(\beta_3\)    & 0.35  & $\Sexpr{round(est_mle_vec["beta3"],2)}$\\
+      \(\phi\)       & 0.14  & $\Sexpr{round(est_mle_vec["phi"],2)}$\\
+      \(\rho\)       & 0.16  & $\Sexpr{round(est_mle_vec["beta11"],2)}$\\
+      \(\sigma^2_q\) & 0.021 & $\Sexpr{round(est_mle_vec["sigma_q"],3)}$\\
+      \(\sigma_\xi\) & 66.89 & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$\\
+      \(S_{10}\) & 3876549 & $\Sexpr{round(pop*est_mle_vec["S10"],0)}$\\
+      \(I_{10}\) & 30351     & $\Sexpr{round(pop*est_mle_vec["I10"],0)}$\\
+      \(R_{10}\) & 1315221   & $\Sexpr{round(pop*est_mle_vec["R10"],0)}$\\
+      \(S_{20}\) & 57139612  & $\Sexpr{round(pop*est_mle_vec["S20"],0)}$\\
+      \(I_{20}\) & 871     & $\Sexpr{round(pop*est_mle_vec["I20"],0)}$\\
+      \(R_{20}\) & 302852    & $\Sexpr{round(pop*est_mle_vec["R20"],0)}$\\
+     \(S_{30}\) & 19573727  & $\Sexpr{round(pop*est_mle_vec["S30"],0)}$\\
+      \(I_{30}\) & 2550    & $\Sexpr{round(pop*est_mle_vec["I30"],0)}$\\
+      \(R_{30}\) & 131092   & $\Sexpr{round(pop*est_mle_vec["R30"],0)}$\\
       \midrule
       \addlinespace
-      \textbf{AIC}  & $98866.65$ & $15154.75$ & $13778.08$ & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$\\
+      \textbf{AIC}  & $13778.08$ & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$\\
       \addlinespace

### 2024-08-17 ms.Rnw
fixed table, unused some foreach's, doFurture modified

@@
 bake(file = paste0(out_dir, "mif2_maximization/round_01/mifs_global_01.rds"),{ 
-  registerDoFuture()
   plan(multicore, workers = ncores)
@@
 bake(file = paste0(out_dir,"mif2_maximization/round_02/mifs_global_02.rds"),{ 
-  registerDoFuture()
   plan(multicore, workers = ncores)
@@
 bake(file = paste0(out_dir, "mif2_maximization/round_03/mifs_global_03.rds"),{ 
-  registerDoFuture()
   plan(multicore, workers = ncores)
@@
   ovov_sim_computation <- list()
-  registerDoFuture()
-  plan(multicore, workers = ncores)
 
@@
   pfilter_ovov_sim_computation <- c()
-  registerDoFuture()
-  plan(multicore, workers = ncores)
 
@@
     )  -> pfilter_obj[[i]]
+
+    plan(multicore, workers = ncores)
 
@@
 bake(file = paste0(out_dir, "pfilter/simulated_dataset/palfilter_ovov_sim_100.rds"),{
-  palfilter_ovov_sim_computation <- c()
-  registerDoFuture()
   plan(multicore, workers = ncores)
@@
     )
-    palfilter_ovov_sim_computation[i] <- lik_list$log_lik
+    lik_list$log_lik
   }
-  palfilter_ovov_sim_computation
 }) -> palfilter_ovov_sim_100
 
+# lik_list$ll_storage |> sum()
 @
@@
 bake(file = paste0(out_dir, "pfilter/original_dataset/pfilter_wwr.rds"),{
-registerDoFuture()
   plan(multicore, workers = ncores)
@@
   pfilter_wwr_cond_lik_computation <- list()
-  registerDoFuture()
-  plan(multicore, workers = ncores)
   for(k in 1:ncores){
@@
     \caption{Maximum likelihood estimation of parameters by PAL \citep{wwr}, and by mif2 algorithm \citep{pomppackagepaper} for the OvOv model.}
-    }
     \label{tab:mlebywwr}
-    \begin{tabular}
+    \begin{tabular}{lcc}  % Left-aligned for the first column, centered for the other two columns
       \toprule
@@
       \midrule
-      \addlinespace
-      \textbf{AIC}  & $13778.08$ & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$\\
-      \addlinespace
+      \textbf{AIC}  & 13778.08  & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$\\
       \bottomrule

### 2024-08-19 ms.Rnw

@@
 \usepackage{makecell}
-\usepackage{threeparttable} % For table notes
+% \usepackage{threeparttable} % For table notes
 \usepackage{array} % For table formatting
@@
 PAL employs a Poisson approximation to conditional densities, offering a fast and consistent approximation to the likelihood function for a certain subset of partially observed Markov process models.
-\cite{wwr} supported the importance of their contribution by theoretical analysis and numerical results.
 A central piece of evidence for PAL is the comparison in Table~1 of \cite{wwr}, which claims a large improvement for PAL over a standard particle filter algorithm.
 This evidence, based on a model and data from a previous scientific study by \citet{stocks}, might suggests that researchers confronted with similar models should use PAL rather than previous particle filter methods.
-Alternatively, the improvement in likelihood found by \cite{wwr} compared to \cite{stocks} might indicate a weakness in that previous work.
-We show that the comparison of log-likelihood values made by \cite{wwr} is flawed because the PAL calculations were carried out using a dataset scaled differently from the previous study.
+Alternatively, the improvement in likelihood found by \cite{wwr} compared to \cite{stocks} might indicate a flaw in the model or numerical methods for that previous work.
+We show that neither of these conclusions is valid, and the comparison of log-likelihood values made by \cite{wwr} is flawed because the PAL calculations were carried out using a dataset scaled differently from the previous study.
 If PAL and the particle filter are used on this new scale, the superficial advantage of PAL largely disappears.
@@
 
-This article results from an investigation of the results in Table~1 of \citet{wwr}.
-The authors of that paper were given the opportunity to submit a correction, after we shared the results of our investigation with them, but they declined.
-The theory developed by \cite{wwr} shows that PAL has some potentially useful scaling properties, but the numerical results in Table~1 appear to show much stronger performance than a standard particle filter on this example of moderate size.
+This article results from an investigation of the results presented by \citet{wwr} (henceforth, WWR) in their Table~1.
+WWR were given the opportunity to submit a correction, after we shared the results of our investigation with them, but they declined.
+The theory developed by \cite{wwr} shows that PAL has some potentially useful scaling properties, but the numerical results in Table~1 appear to show much stronger performance than a standard particle filter on an example of scientific interest but moderate size.
 Our task here is to correct the error in Table~1 so that researchers considering whether to implement Poisson approximate likelihood (PAL) are appropriately informed about its benefits.
-Table~1 reanalyzes the model and data of \citet{stocks}, for which the likelihood was calculated using a particle filter.
-\citet{stocks} found strong evidence for the importance of overdispersion in a stochastic dynamic model for their epidemiological data.
+
+Table~1 of WWR reanalyzes the model and data of \citet{stocks} (henceforth, SBH) for which the likelihood was calculated using a particle filter.
+SBH found strong evidence for the importance of overdispersion in a stochastic dynamic model for their epidemiological data.
 This is significant because earlier research on population dynamics largely avoided consideration of overdispersion, perhaps because of the lack of statistical methodology to fit such models. 
-The conclusions of \cite{stocks} hinge on a comparison of likelihoods, and so the results of \citet{wwr} discredit those conclusions by indicating that \citet{stocks} used an inaccurate filter to calculate their likelihoods.
-An important consequence of correcting Table~1 is that the results of \cite{stocks} stand undiminished.
+The conclusions of SBH hinge on a comparison of likelihoods, and so the results of WWR discredit those conclusions by indicating that SBH used an inaccurate filter to calculate their likelihoods.
+An important consequence of correcting Table~1 is that the results of SBH stand undiminished.
 
-First, we show that most of the apparent advantage for PAL, compared to a particle filter, arises because \cite{wwr} used a different scaling of the data from \cite{stocks}.
+SBH and WWR each fitted three different rotavirus models.
+The first has equidispersion (i.e., no overdispersion) in the measurement model and the dynamic model, and is called
+EqEq.
+The second, EqOv, includes overdispersion in only the measurement model.
+The third, OvOv, includes overdispersion in both these model components.
+We focus on OvOv, which both SBH and WWR find to be the best fitting model. 
+However, WWR report in Table~1 that their EqOv model fits better than the SBH OvOv model, from which one might conclude that the use of PAL is more important than the decision of whether to include overdispersion in the process model.
+
+We show that most of the apparent advantage for PAL, compared to a particle filter, arises because \cite{wwr} used a different scaling of the data from \cite{stocks}.
 Two models for the same data can properly be compared by their likelihood, even if the models have entirely different structures.
@@
 Presenting a direct comparison of a likelihood for the data in grams with a likelihood for the data in kilograms would evidently be inappropriate.
+
 \cite{stocks} fitted their model to a dataset derived by dividing the original reported count data by an estimated reporting rate, to put their data on the scale of the actual number of cases in the population, whereas \cite{wwr} fitted directly to the report data.
@@
 On approximately 1200 data points, this corresponds to a discrepancy of around $-1200\log(0.07) \approx 3200$ log-likelihood units, largely explaining the difference reported in Table~1 and interpreted by \cite{wwr} as evidence supporting PAL.
-The comparison can be corrected either by applying the method of \cite{stocks} to the data of \cite{wwr} or vice versa.
-Since the method of \cite{stocks} is applicable to a more general class of models, and supported by published software, it was more convenient to apply this method to the model and data of \cite{wwr}.
+The comparison can be corrected either by applying the method of SBH to the data of WWR or vice versa.
+Since the method of SBH is applicable to a more general class of models, and supported by published software, it was convenient to apply the SBH method to the model and data of WWR.
 The large discrepancy in log-likelihood disappears at this point by recomputing the likelihood of the model using PAL and particle filter separately (see Table~\ref{tab:ovovrealdata}).
@@
 We continued our investigation to establish the cause of this.
+We discovered that WWR (unlike SBH) used a poorly fitting choice of initial conditions for the latent process in January 2001 (their time 0).
+
+\begin{table}[ht] % The placement specifier can be [h!tbp]
+\centering % Centers the table
+\caption{log-likelihood at the MLE of OvOv model for the real rotavirus data, computed using two filtering methods, employed 36 replicates and 50,000 particles} 
+\label{tab:ovovrealdata} % For referencing this table elsewhere in your text
+\begin{tabular}{|c c c|} 
+ \hline
+  & PAL & SMC  \\
+  \hline
+ OvOv & \Sexpr{round(palfilter_lik_list_wwr$log_lik,2)} & $\Sexpr{round(logmeanexp(logLik(pfilter_wwr),2))}$ \\ 
+ \hline
+\end{tabular}
+\end{table}
 
@@
 
-% \begin{figure}[h]
-% \centering
-% \includegraphics[width=0.55\textwidth]{thesis-figs/PAL_vs_SMC_ovov_100.png}
-% \caption{\label{fig:PAL_vs_SMC_ovov_100}Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model.
-% The particle filter and PAL used 50,000 particles.
-% The red line is the line $Lik(pfilter)=Lik(palfilter)$.
-% On average, the particle filter likelihood is $7.7$ log units higher than the PAL.
-% %%  -6224.245 - (-6231.97) = 7.725
-% }
-% \end{figure}
-
-PAL provides an approximate filter, whereas the particle filter provides a Monte~Carlo estimate of an exact filter.
-Possible explanations for PAL obtaining a higher log-likelihood than a particle filter include:
-\begin{enumerate}
-\item The particle filter could have high Monte Carlo variance.
-This results in negative bias in its log-likilihood estimate due to Jensen's inequality, since the particle filter provides an unbiased likelihood estimate.
-\item If the model is misspecified, the approximation error in PAL could lead to a higher log-likelihood estimate than that of the actual misspecified model.
-\end{enumerate}
-Log-likelihood is a proper scoring rule for forecasts \cite{gneiting2}, and both the particle filter and PAL construct their log-likelihood estimates via a sequence of one-step forecasts.
-Therefore, if the model is correctly specified, the approximation error in PAL can only decrease the expected log-likelihood.
-We tested this on simulated data, for which the model of \cite{wwr} is correctly specified.
-We focus on the version of the model with overdispersion in both the population dynamics and the measurement process.
-This version of the model, called OvOv, was found by both \cite{wwr} and \cite{stocks} to be the best choice to fit the data.
-For this simulation study, the particle filter out-performs PAL (Figure~1) which supports hypothesis 2, above.
-Consequently, we looked for model misspecification by comparing the conditional log-likelihood estimates at each time point, as described in the Methods section.
-
-\begin{table}[ht] % The placement specifier can be [h!tbp]
-\centering % Centers the table
-\caption{log-likelihood at the MLE of OvOv model for the real rotavirus data, computed using two filtering methods, employed 36 replicates and 50,000 particles} 
-\label{tab:ovovrealdata} % For referencing this table elsewhere in your text
-\begin{tabular}{|c c c|} 
- \hline
-  & PAL & SMC  \\
-  \hline
- OvOv & \Sexpr{round(palfilter_lik_list_wwr$log_lik,2)} & $\Sexpr{round(logmeanexp(logLik(pfilter_wwr),2))}$ \\ 
- \hline
-\end{tabular}
-\end{table}
-
-% \begin{figure}[H]
-% \centering
-% \includegraphics[width=0.9\textwidth]{thesis-figs/cond_SMC_PAL_realdat.png}
-% \caption{\label{fig:cond_SMC_PAL_realdat}Conditional log-likelihoods computed using two methods for the real rotavirus dataset, with the SMC method replicated 36 times due to its high variances.
-The blue line represents overlapping results from 36 SMC computations, while the red line is derived from PAL.
-The main sources of likelihood shortfall of SMC are at time points $t=1, 2, 3, 11, 81, 194$, and $325-333$.}
-% \end{figure}
-
 Figure~2 indicates that the worst conditional log-likelihood values, estimated using the particle filter, arise at the start of the time series.
@@
 \begin{table}[htbp]
-  \centering
-  \begin{threeparttable}
-    \caption{Maximum likelihood estimation of parameters by PAL \citep{wwr}, and by mif2 algorithm \citep{pomppackagepaper} for the OvOv model.}
+    \caption{Maximum likelihood estimation of parameters by PAL \citep{wwr}, and by mif2 algorithm \citep{pomppackagepaper} for the OvOv model.
+The initial distribution parameter \(\lambda_0 = (S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}, R_{30})\) are assumed to be \textbf{fixed} in \cite{wwr}.
+The iterated filtering estimates were obtained by using 3 rounds of refinement, with 100 filtering iterations, 50,000 particles, and 36 replicates in each round with the top 12 best fits in terms of likelihood chosen to be the starting value for the next round.}
     \label{tab:mlebywwr}
+\begin{center}    
     \begin{tabular}{lcc}  % Left-aligned for the first column, centered for the other two columns
       \toprule
-      Parameter      & OvOv      & OvOv (\textbf{pomp})$^*$ \\
+      Parameter      & PAL      & iterated filtering \\
       \midrule
@@
     \end{tabular}
-    \begin{tablenotes}
-      \small
-      \item[*] The initial distribution parameter \(\lambda_0 = (S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}, R_{30})\) are assumed to be \textbf{fixed} in \cite{wwr}.
-The results of the estimates of the OvOv (\textbf{pomp}) model are obtained by using the Iterated Filtering algorithm with 3 rounds and 100 iterations, 50,000 particles, and 36 replicates in each round with the top 12 best fits in terms of likelihood chosen to be the starting value for the next round.
-    \end{tablenotes}
-  \end{threeparttable}
+\end{center}
 \end{table}
 
-\ed{WE CAN REVIEW EXACTLY WHICH NUMBERS TO SHOW. IDEALLY, AS LITTLE AS POSSIBLE IN ORDER TO MAKE A STRONG ARGUMENT.}
+\section*{Discussion}
+
+PAL provides an approximate filter, whereas the particle filter provides a Monte~Carlo estimate of an exact filter.
+Possible explanations for PAL obtaining a higher log-likelihood than a particle filter include:
+\begin{enumerate}
+\item The particle filter could have high Monte Carlo variance.
+This results in negative bias in its log-likilihood estimate due to Jensen's inequality, since the particle filter provides an unbiased likelihood estimate.
+\item If the model is misspecified, the approximation error in PAL could lead to a higher log-likelihood estimate than that of the actual misspecified model.
+\end{enumerate}
+Log-likelihood is a proper scoring rule for forecasts \cite{gneiting2}, and both the particle filter and PAL construct their log-likelihood estimates via a sequence of one-step forecasts.
+Therefore, if the model is correctly specified, the approximation error in PAL can only decrease the expected log-likelihood.
+We tested this on simulated data, for which the model of \cite{wwr} is correctly specified.
+We focus on the version of the model with overdispersion in both the population dynamics and the measurement process.
+This version of the model, called OvOv, was found by both \cite{wwr} and \cite{stocks} to be the best choice to fit the data.
+For this simulation study, the particle filter out-performs PAL (Figure~1) which supports hypothesis 2, above.
+Consequently, we looked for model misspecification by comparing the conditional log-likelihood estimates at each time point, as described in the Methods section.
 
@@
 
-%%%%%%%%% 222222222 %%%%%%%%%%5
+%%%%%%%%% 222222222 %%%%%%%%%%
 
@@
 
-\ed{BRIEFLY DESCRIBE HERE ANY DETAILS NEEDED TO DESCRIBE THE RESULTS. COMMENT ON CODE AVAILABILITY.}
-
-To evaluate whether SMC performs as well as PAL in terms of both computational time and likelihood identifiability, we replicated the model using the R package \textbf{pomp} \citep{pomppackagepaper}.
-First, we verified that the two models are essentially identical.
-Here, we simulated data from the two models 1000 times at the MLE shown in Table \ref{tab:mlebywwr} and compared the summary statistics of 1000 simulated time series from two models using the t-test.
-Summary statistics including the mean, median, and variance for all three age groups were tested.
-The results showed no significant differences between the two models from which the data were simulated, confirming that both the latent process model and the observation model are essentially consistent with each other, given that the observations were directly simulated.
-From Table \ref{tab:mlebywwr}, it is evident that the OvOv model, with overdispersion in both the dynamic system and the measurement model, yields the lowest AIC.
-We therefore focus on this version of the model.
+We first replicated the model of \citet{wwr} in the R package \textbf{pomp} \citep{pomppackagepaper}.
+Comparing summary statistics of simulations confirms that both model versions are essentially identical, as described by \citet{hao24}.
+We therefore implement PAL using the code from \citet{wwr}, comparing directly with the particle filter implemented using \textbf{pomp}.
 
 This article has focused on likelihood evaluation, but the inferences presented also require the likelihood estimate to be maximized.
-\cite{stocks} used iterated filtering to maximize the particle filter likelihood estimate, whereas \cite{wwr} used direct numerical maximization.
+\cite{wwr} used direct numerical maximization, whereas \cite{stocks} used iterated filtering to maximize the particle filter likelihood estimate implemented using the \texttt{mif2} algorithm in \textbf{pomp}.
 We follow the approaches adopted by these two papers.
 The optimization method is not directly pertinent to the comparison of the filtering methods, so we do not discuss optimization in further detail here.
-
-Our workflow is fully reproducible. An extended description of our reanalysis of \cite{wwr} is provided by \cite{hao24}.
+An extended description of our reanalysis of \cite{wwr} is provided by \cite{hao24}.
 That thesis also contains some additional results reinforcing the conclusions presented in this article.
+The source code for this article is available at \url{https://github.com:ionides/pal-vs-pf} and archived at \ed{ZENODO - TODO}.
 

### 2024-08-19 ms.Rnw
minor edits on greatlakes

@@
       pfilter(
-        pfilter_obj[[i]], Np = NP_MIF, save.states = TRUE, params = params_wwr
+        pfilter_obj[[i]], Np = NP_MIF, save.states = FALSE, params = params_wwr
       )

### 2024-08-19 rotavirus_normq.cpp
a few edits

@@ -322,7 +322,7 @@ List rotavirus_SMC_qropxi( arma::vec init_dist,
   
   Rcpp::NumericVector ll_storage_R = Rcpp::wrap(ll_storage);
   
-  List L = List::create(Named("log_lik") = log_lik, Named("xi_particles") = xiresamp_particles, Named("q_particles") = qresamp_particles, Named("ess") = ess, Named("ll_storage") = ll_storage_R);
+  List L = List::create(Named("log_lik") = log_lik, Named("ll_storage") = ll_storage_R);
  // List L = List::create(Named("log_lik") = log_lik, Named("xi_particles") = xiresamp_particles, Named("q_particles") = qresamp_particles);
   
   return(L);

### 2024-08-21 ms.Rnw
a few edits
table moving
ms on gl

@@
 
-RUN_LEVEL = 1
+RUN_LEVEL = 3
 NP_MIF       = switch(RUN_LEVEL, 10,   5000,  50000)
@@
 
-\begin{table}[ht] % The placement specifier can be [h!tbp]
-\centering % Centers the table
-\caption{log-likelihood at the MLE of OvOv model for the real rotavirus data, computed using two filtering methods, employed 36 replicates and 50,000 particles} 
-\label{tab:ovovrealdata} % For referencing this table elsewhere in your text
-\begin{tabular}{|c c c|} 
- \hline
-  & PAL & SMC  \\
-  \hline
- OvOv & \Sexpr{round(palfilter_lik_list_wwr$log_lik,2)} & $\Sexpr{round(logmeanexp(logLik(pfilter_wwr),2))}$ \\ 
- \hline
-\end{tabular}
-\end{table}
-
 <<sim-100, echo=F, warning=F, message=F>>=
@@
 
+\begin{table}[ht] % The placement specifier can be [h!tbp]
+\centering % Centers the table
+\caption{log-likelihood at the MLE of OvOv model for the real rotavirus data, computed using two filtering methods, employed 36 replicates and 50,000 particles} 
+\label{tab:ovovrealdata} % For referencing this table elsewhere in your text
+\begin{tabular}{|c c c|} 
+ \hline
+  & PAL & SMC  \\
+  \hline
+ OvOv & \Sexpr{round(palfilter_lik_list_wwr$log_lik,2)} & $\Sexpr{round(logmeanexp(logLik(pfilter_wwr),2))}$ \\ 
+ \hline
+\end{tabular}
+\end{table}
+
 <<plot2-cond-loglik, echo=FALSE, fig.height=4, fig.width=5.5, fig.cap='Conditional log-likelihoods computed using two methods for the real rotavirus dataset, with the SMC method replicated 36 times due to its high variances.
@@
 
+\yize{ON AVERAGE PAL IS 25.3 LOG UNITS LOWER THAN PFILTER AND SHOULD WE MENTION THAT WE USED THE SIMULATED DATA WITHOUT ZEROS? \\ 
+THE TABLE SEEMS TO BE TALL AND THIN. SHOULD WE PUT IT IN A APPENDIX?}
+
 We first replicated the model of \citet{wwr} in the R package \textbf{pomp} \citep{pomppackagepaper}.

### 2024-08-21 ms.Rnw
gl results

@@
 
-\yize{ON AVERAGE PAL IS 25.3 LOG UNITS LOWER THAN PFILTER AND SHOULD WE MENTION THAT WE USED THE SIMULATED DATA WITHOUT ZEROS? \\ 
-THE TABLE SEEMS TO BE TALL AND THIN. SHOULD WE PUT IT IN A APPENDIX?}
+\yize{ON AVERAGE PAL IS 25.3 LOG UNITS LOWER THAN PFILTER AND SHOULD WE MENTION THAT WE USED THE SIMULATED DATA WITHOUT ZEROS?}
 

### 2024-08-22 choose_n_fits.R
measlespkg attribution

@@ -1,7 +1,19 @@
-# Choose top n fits
+#' Choose top n best results and make them to be the starting value for the next round.
+#'
+#' @param el Estimated values from `pfiler`.
+#' @param top_n_fits Number of best fits in terms of log-likelihood one wants to keep.
+#' @param ncores Number of starting values possibly equal to the number of cores being used.
+#'
+#' @return A matrix with ncores number of rows and estimated parameters being in each row.
+#' @export
+#'
+#' @source <https://github.com/AJAbkemeier/measlespkg/tree/main/R>
+#'
+#' @examples
+#' choose_n_fits(el, 12, 36)
+#' 
 
 choose_n_fits = function(el, top_n_fits, ncores){
-  ### Next-round code
   fit_mat <- el
   score_total = fit_mat$logLik
   ranking_total = order(score_total, decreasing = TRUE)[1:top_n_fits]
@@ -12,7 +24,6 @@ choose_n_fits = function(el, top_n_fits, ncores){
   
   recycle_vec = sort(rep_len(1:top_n_fits, ncores))
   full_best_fit <- best_fits[recycle_vec, ] 
-  
   
   starting_values <- vector(ncores, mode="list")
   

### 2024-08-22 eval_logLik_pomp.R

@@ -1,4 +1,18 @@
-#pfilter for a list of mif2 results
+#' To evaluate likelhoods of a list of pomp objects using pfilter with replications. 
+#'
+#' @param model_obj_list List of pomp results.
+#' @param nreps Number of replications for each pfilter. 
+#' @param np_pf Number of particles in pfilter.
+#' @param seed  Set seed.
+#' 
+#' @return A matrix with ncores number of rows and estimated parameters being in each row.
+#' @export
+#'
+#' @source <https://github.com/AJAbkemeier/measlespkg/tree/main/R>
+#'
+#' @examples
+#' eval_logLik_pomp(model_obj_list, 36, 1000, 123)
+#' 
 eval_logLik_pomp = function(model_obj_list, nreps, np_pf, seed){
   pf_logLik_frame = data.frame(
     logLik = rep(0, length(model_obj_list)),
@@ -8,9 +22,7 @@ eval_logLik_pomp = function(model_obj_list, nreps, np_pf, seed){
   )
   
   for(i in seq_along(model_obj_list)){
-    registerDoFuture()
     plan(multicore, workers = nreps)
-    
     foreach::foreach(
       j = 1:nreps,
       .options.future=list(seed=seed),

### 2024-08-22 ms.Rnw
comment on initial conditions, presumably delete Figure~2 later

@@
 We continued our investigation to establish the cause of this.
-We discovered that WWR (unlike SBH) used a poorly fitting choice of initial conditions for the latent process in January 2001 (their time 0).
+We discovered that the initial conditions for the latent process in January 2001 (their time 0) were estimated by SBH. However, in WWR, they were assumed fixed and made the model fitted poorly.
 

### 2024-08-23 ms.Rnw
plenty of edits, but more to do

@@
 % changes the text to double-line spacing
-\doublespacing
+% \doublespacing
 
@@
 
-\newcommand{\ed}[1]{\textcolor{red}{[EI:#1]}}
-\newcommand{\yize}[1]{\textcolor{green}{[YH:#1]}}
-\newcommand{\aaron}[1]{\textcolor{blue}{[AA:#1]}}
+\newcommand{\ed}[1]{\textcolor{red}{[EI: #1]}}
+\newcommand{\yize}[1]{\textcolor{green}{[YH: #1]}}
+\newcommand{\aaron}[1]{\textcolor{blue}{[AA: #1]}}
 
@@
 
-\title{The Poisson approximate likelihood compared to the particle filter}
+\title{Poisson approximate likelihood compared to the particle filter}
 \author{Yize Hao, Aaron A. Abkemeier and Edward L. Ionides}
@@
 A novel Poisson approximate likelihood (PAL) filter was introduced by \cite{wwr}.
-PAL employs a Poisson approximation to conditional densities, offering a fast and consistent approximation to the likelihood function for a certain subset of partially observed Markov process models.
+PAL employs a Poisson approximation to conditional densities, offering a fast approximation to the likelihood function for a certain subset of partially observed Markov process models.
 A central piece of evidence for PAL is the comparison in Table~1 of \cite{wwr}, which claims a large improvement for PAL over a standard particle filter algorithm.
@@
 rm(list = ls())     # clear objects
-
 library("knitr")
@@
   tidy = TRUE,
+  cache = TRUE,
+#  cache = FALSE,  # set cache = FALSE when editing the code
 # message = FALSE,
@@
 WWR were given the opportunity to submit a correction, after we shared the results of our investigation with them, but they declined.
-The theory developed by \cite{wwr} shows that PAL has some potentially useful scaling properties, but the numerical results in Table~1 appear to show much stronger performance than a standard particle filter on an example of scientific interest but moderate size.
+The theory developed by WWR shows that PAL has some potentially useful scaling properties, but the numerical results in Table~1 appear to show much stronger performance than a standard particle filter on an example of scientific interest but moderate size.
 Our task here is to correct the error in Table~1 so that researchers considering whether to implement Poisson approximate likelihood (PAL) are appropriately informed about its benefits.
@@
 SBH found strong evidence for the importance of overdispersion in a stochastic dynamic model for their epidemiological data.
-This is significant because earlier research on population dynamics largely avoided consideration of overdispersion, perhaps because of the lack of statistical methodology to fit such models. 
-The conclusions of SBH hinge on a comparison of likelihoods, and so the results of WWR discredit those conclusions by indicating that SBH used an inaccurate filter to calculate their likelihoods.
-An important consequence of correcting Table~1 is that the results of SBH stand undiminished.
+This is significant because earlier research on population dynamics largely avoided consideration of overdispersion, perhaps because of the lack of avaialable statistical methodology to fit such models.
+The conclusions of SBH hinge on a comparison of likelihoods, and so the results of WWR discredit those conclusions by indicating that SBH based their reasoning on inaccurately computed likelihoods.
+An important consequence of correcting Table~1 of WWR is that the results of SBH stand undiminished.
 
 SBH and WWR each fitted three different rotavirus models.
-The first has equidispersion (i.e., no overdispersion) in the measurement model and the dynamic model, and is called
-EqEq.
+The first has equidispersion (i.e., no overdispersion) in the measurement model and the dynamic model, and is called EqEq.
 The second, EqOv, includes overdispersion in only the measurement model.
 The third, OvOv, includes overdispersion in both these model components.
-We focus on OvOv, which both SBH and WWR find to be the best fitting model. 
-However, WWR report in Table~1 that their EqOv model fits better than the SBH OvOv model, from which one might conclude that the use of PAL is more important than the decision of whether to include overdispersion in the process model.
+We focus on OvOv, which WWR found to be the best fitting model.
 
-We show that most of the apparent advantage for PAL, compared to a particle filter, arises because \cite{wwr} used a different scaling of the data from \cite{stocks}.
+We show that most of the apparent advantage for PAL on the OvOv model, compared to a particle filter, arises because WWR used a different scaling of the data from SBH.
 Two models for the same data can properly be compared by their likelihood, even if the models have entirely different structures.
@@
 Presenting a direct comparison of a likelihood for the data in grams with a likelihood for the data in kilograms would evidently be inappropriate.
-
-\cite{stocks} fitted their model to a dataset derived by dividing the original reported count data by an estimated reporting rate, to put their data on the scale of the actual number of cases in the population, whereas \cite{wwr} fitted directly to the report data.
-The reporting rate used by \cite{stocks} varied over time, but was generally around $7\%$.
-On approximately 1200 data points, this corresponds to a discrepancy of around $-1200\log(0.07) \approx 3200$ log-likelihood units, largely explaining the difference reported in Table~1 and interpreted by \cite{wwr} as evidence supporting PAL.
-The comparison can be corrected either by applying the method of SBH to the data of WWR or vice versa.
-Since the method of SBH is applicable to a more general class of models, and supported by published software, it was convenient to apply the SBH method to the model and data of WWR.
-The large discrepancy in log-likelihood disappears at this point by recomputing the likelihood of the model using PAL and particle filter separately (see Table~\ref{tab:ovovrealdata}).
-This re-analysis does, however, show a discrepancy between two methods.
-We continued our investigation to establish the cause of this.
-We discovered that the initial conditions for the latent process in January 2001 (their time 0) were estimated by SBH. However, in WWR, they were assumed fixed and made the model fitted poorly.
 
@@
 
-<<plot1-100-sim-compare,echo=FALSE, fig.height=4, fig.width=5.5, fig.cap='Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model.
-The particle filter and PAL used 50,000 particles.
-The red line corresponds to equality of the two estimates.
-On average, the particle filter likelihood estimate is $7.7$ log units higher than the PAL.'>>=
-data <- data.frame(
-  pfilter = pfilter_ovov_sim_100, 
-  palfilter = palfilter_ovov_sim_100
-)
-
-ggplot(data, aes(x = pfilter, y = palfilter)) +
-geom_point(shape = 20, size = 0.3) +  # pch=20, cex=0.3 equivalent
-geom_abline(intercept = 0, slope = 1, color = "red") +
-labs(
-title = "PAL vs SMC: OvOv with 100 simulations",
-x = "pfilter",
-y = "palfilter"
-) +
-theme_minimal()
-@
-
 <<pfilter_wwr, echo=FALSE, message=FALSE, warning=FALSE>>=
@@
 \centering % Centers the table
-\caption{log-likelihood at the MLE of OvOv model for the real rotavirus data, computed using two filtering methods, employed 36 replicates and 50,000 particles} 
+\caption{AIC for the OvOv rotavirus model, computed using two filtering methods.
+PAL is the Poisson approximate likelihood, implemented using the code of WWR.
+PF is the particle filter, implemented using the code of \citet{pomppackagepaper}.
+The first two lines are from WWR, Table~1, and lines 3--5 are our own computations.
+We used 50000 particle for both PF and PAL.
+PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL.
+\ed{SHOULD WE HAVE A LINE FOR PAL AT THE IF2 MLE FROM LINE 5? PERHAPS WE CAN'T READILY MAXIMIZE PAL FOR THIS MODEL (OR CAN WE?) BUT SOME SORT OF COMPARISON WOULD BE HELPFUL}
+} 
 \label{tab:ovovrealdata} % For referencing this table elsewhere in your text
-\begin{tabular}{|c c c|} 
+\begin{tabular}{lllc} 
   \hline
-  & PAL & SMC  \\
+  & Method & Result & AIC  \\
   \hline
- OvOv & \Sexpr{round(palfilter_lik_list_wwr$log_lik,2)} & $\Sexpr{round(logmeanexp(logLik(pfilter_wwr),2))}$ \\ 
+  1. & PAL & Table~1 of WWR & 13778.08
+  \\
+  2. & PF & Table~1 of WWR, originally from SBH &  20134.38
+  \\
+  3. & PAL & Model, data and MLE from WWR &
+    \Sexpr{myround(-2*palfilter_lik_list_wwr$log_lik + 2*7 ,2)} 
+  \\
+  4. & PF & Model, data and MLE from WWR &
+    \Sexpr{myround(-2*logmeanexp(logLik(pfilter_wwr)+ 2*7),2)}
+  \\
+  5. & PF & Data from WWR, estimated initial values &
+     \Sexpr{myround(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}
+  \\
  \hline
@@
 
-<<plot2-cond-loglik, echo=FALSE, fig.height=4, fig.width=5.5, fig.cap='Conditional log-likelihoods computed using two methods for the real rotavirus dataset, with the SMC method replicated 36 times due to its high variances.
-The blue line represents overlapping results from 36 SMC computations, while the red line is derived from PAL.
-The main sources of likelihood shortfall of SMC are at time points $t=1, 2, 3, 11, 81, 194$, and $325-333$.'>>=
+SBH fitted their model to a dataset derived by dividing the original reported count data by an estimated reporting rate, to put their data on the scale of the actual number of cases in the population, whereas WWR fitted directly to the report data.
+The reporting rate used by SBH varied over time, but was generally around $7\%$.
+On approximately 1200 data points, this corresponds to a discrepancy of around $-1200\log(0.07) \approx 3200$ log-likelihood units, largely explaining the difference reported in Table~1 and interpreted by WWR as evidence supporting PAL.
+The comparison can be corrected either by applying the method of SBH to the data of WWR or vice versa.
+Since the method of SBH is applicable to a more general class of models, and supported by published software, it was convenient to apply the SBH method to the model and data of WWR.
+The large discrepancy in log-likelihood disappears at this point by recomputing the likelihood of the model using PAL and particle filter separately (see Table~\ref{tab:ovovrealdata}).
+This re-analysis does, however, show a discrepancy between two methods.
+We continued our investigation to establish the cause of this.
+We discovered that the initial conditions for the latent process in January 2001 (their time 0) were estimated by SBH. However, in WWR, they were assumed fixed and made the model fitted poorly.
+
+<<plot2-cond-loglik, echo=FALSE, fig.height=4, fig.width=5.5, out.width="4.5in",fig.align="center", fig.cap='Conditional log-likelihoods computed using PAL (red) and PF (blue) for the rotavirus model and MLE from WWR.
+We used 50000 particle for both PF and PAL.
+PF averaged over 36 replications to reduce the Monte Carlo variance, but this step was not necessary for PAL.
+The main sources of likelihood shortfall of PF are at time points $t=1, 2, 3, 11, 81, 194$, and $325-333$.'>>=
 
@@
   labs(
-    title = "Conditional Log Likelihood Over Time",
-    x = "Time",
-    y = "Conditional Log Likelihood",
+    title = NULL,
+    x = "Time (weeks)",
+    y = "Conditional Log-likelihood",
     color = "Method"
@@
 
-Figure~2 indicates that the worst conditional log-likelihood values, estimated using the particle filter, arise at the start of the time series.
-This could be a problem with data collection, or some subtle issue with the science of the epidemiological system, but a simple possibility is that the initial values of the latent state process might be poorly specified.
-For this model, the initial values were fixed based on scientific reasoning, rather than being estimated.
-If we instead estimate the initial values, we obtain the results in Table~\ref{tab:mlebywwr}.
-
 \begin{table}[htbp]
-    \caption{Maximum likelihood estimation of parameters by PAL \citep{wwr}, and by mif2 algorithm \citep{pomppackagepaper} for the OvOv model.
-The initial distribution parameter \(\lambda_0 = (S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}, R_{30})\) are assumed to be \textbf{fixed} in \cite{wwr}.
-The iterated filtering estimates were obtained by using 3 rounds of refinement, with 100 filtering iterations, 50,000 particles, and 36 replicates in each round with the top 12 best fits in terms of likelihood chosen to be the starting value for the next round.}
+    \caption{Maximum likelihood estimation of parameters for the OvOv model of WWR by PAL (taken from WWR), and by iterated particle filtering implemented using {\rm \texttt{mif2}} in {\rm \texttt{pomp}} \citep{pomppackagepaper}.
+The initial distribution parameter \(\lambda_0 = (S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}, R_{30})\) are assumed to be fixed in \cite{wwr}.
+The iterated particle filtering estimates were obtained by using 3 rounds of refinement, with 100 filtering iterations, 50,000 particles, and 36 replicates in each round with the top 12 best fits in terms of likelihood chosen to be the starting value for the next round.}
     \label{tab:mlebywwr}
@@
 
+<<plot1-100-sim-compare,echo=FALSE, fig.height=4, fig.width=5.5, out.width="4.5in",fig.align="center",fig.cap='Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model.
+Simulations with one or more zero counts were disqualified since they resulted in errors for the PAL implementation.
+We used 50000 particle for both PF and PAL.
+PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL.
+The red line corresponds to equality of the two estimates.'>>=
+data <- data.frame(
+  pfilter = pfilter_ovov_sim_100, 
+  palfilter = palfilter_ovov_sim_100
+)
+
+ggplot(data, aes(x = pfilter, y = palfilter)) +
+geom_point(shape = 20, size = 0.3) +  # pch=20, cex=0.3 equivalent
+geom_abline(intercept = 0, slope = 1, color = "red") +
+labs(
+  title=NULL,
+  x = "PF",
+  y = "PAL"
+) +
+theme_minimal()
+@
+
+Figure~\ref{fig:plot2-cond-loglik} indicates that the worst conditional log-likelihood values, estimated using the particle filter, arise at the start of the time series.
+This could be a problem with data collection, or some subtle issue with the science of the epidemiological system, but a simple possibility is that the initial values of the latent state process might be poorly specified.
+For this model, the initial values were fixed based on scientific reasoning, rather than being estimated.
+If we instead estimate the initial values, we obtain the results in Table~\ref{tab:mlebywwr}.
+
+\ed{SINCE THE MEAN DURATION OF IMMUNITY IS 1 YR, HAVING $31\times 10^6$ in $R_{2,0}$ IMPLIES THAT ABOUT HALF OF ALL SUSCEPTIBLE INDIVIDUALS GET ROTAVIRUS EVERY YEAR ACCORDING TO THE MODEL.
+IS THAT THE CASE FOR THE MODEL?
+FOR THE DATA, THE PER-CAPITA RATE IN THIS AGE GROUP IS RATER LOW.
+THE SIMPLEST WAY TO AVOID THIS PROBLEM WOULD BE TO DO WHAT SBH DID (DESCRIBED IN THEIR ONLINE APPENDIX C) - EFFECTIVELY, DRAWING $X_0$ FROM THE STATIONARY DISTRIBUTION GIVEN THE MODEL.
+THIS HAS THE ADVANTAGE OF INCLUDING NO NEW PARAMETERS AND BEING CONSISTENT WITH SBH.
+WE DON'T KNOW IN ADVANCE EXACTLY HOW SUCCESSFUL IT WOULD BE...}
+
 \section*{Discussion}
@@
 We tested this on simulated data, for which the model of \cite{wwr} is correctly specified.
-We focus on the version of the model with overdispersion in both the population dynamics and the measurement process.
-This version of the model, called OvOv, was found by both \cite{wwr} and \cite{stocks} to be the best choice to fit the data.
-For this simulation study, the particle filter out-performs PAL (Figure~1) which supports hypothesis 2, above.
+For this simulation study, the particle filter out-performs PAL (Figure~\ref{fig:plot1-100-sim-compare}) which supports hypothesis 2, above.
+On average, the particle filter likelihood estimate is $\Sexpr{myround(mean(pfilter_ovov_sim_100)-mean(palfilter_ovov_sim_100),1)}$ log units higher than the PAL.
 Consequently, we looked for model misspecification by comparing the conditional log-likelihood estimates at each time point, as described in the Methods section.
@@
 
-\yize{ON AVERAGE PAL IS 25.3 LOG UNITS LOWER THAN PFILTER AND SHOULD WE MENTION THAT WE USED THE SIMULATED DATA WITHOUT ZEROS?}
-
-We first replicated the model of \citet{wwr} in the R package \textbf{pomp} \citep{pomppackagepaper}.
+We first replicated the model of WWR in the R package \texttt{pomp} \citep{pomppackagepaper}.
 Comparing summary statistics of simulations confirms that both model versions are essentially identical, as described by \citet{hao24}.
-We therefore implement PAL using the code from \citet{wwr}, comparing directly with the particle filter implemented using \textbf{pomp}.
+We therefore implement PAL using the code from \citet{wwr}, comparing directly with the particle filter implemented using \texttt{pomp}.
 
 This article has focused on likelihood evaluation, but the inferences presented also require the likelihood estimate to be maximized.
-\cite{wwr} used direct numerical maximization, whereas \cite{stocks} used iterated filtering to maximize the particle filter likelihood estimate implemented using the \texttt{mif2} algorithm in \textbf{pomp}.
+WWR used direct numerical maximization, whereas SBH used iterated filtering to maximize the particle filter likelihood estimate implemented using the \texttt{mif2} algorithm in \texttt{pomp}.
 We follow the approaches adopted by these two papers.
 The optimization method is not directly pertinent to the comparison of the filtering methods, so we do not discuss optimization in further detail here.
-An extended description of our reanalysis of \cite{wwr} is provided by \cite{hao24}.
+An extended description of our reanalysis of WWR is provided by \cite{hao24}.
 That thesis also contains some additional results reinforcing the conclusions presented in this article.

### 2024-08-30 ms.Rnw
maximize with SBH's init; add PAL filter on est mle

@@
 
-RUN_LEVEL = 3
-NP_MIF       = switch(RUN_LEVEL, 10,   5000,  50000)
+RUN_LEVEL = 1
+NP_MIF       = switch(RUN_LEVEL, 100,   5000,  50000)
 NMIF         = switch(RUN_LEVEL, 5,    20,    100)
@@
 
+rinit_stocks <- Csnippet("
+  S1=5.09484e+06;
+  S2=5.73856e+07;
+  S3=1.96976e+07;
+  I1=2871;
+  I2=639;
+  I3=174;
+  R1=124410;
+  R2=57072;
+  R3=9578;
+  H1 = 0;
+  H2 = 0;
+  H3 = 0;
+")
+
 # x <- c(3876549,57139612,19573727,30351,871,2550,1315221,302852,131092)
@@
 )
+
+sir_sbh_init <- pomp(
+  sir_wwr,
+  t0 = 1-6*52,
+  rinit = rinit_stocks,
+  statenames = statenames
+) 
 @
 
-% Maximization Round 1
+% Maximization Round 1 - Full rinit
 
@@
 
-bake(file = paste0(out_dir, "mif2_maximization/round_01/mifs_global_01.rds"),{ 
+bake(file = paste0(out_dir, "mif2_maximization/full/round_01/mifs_global_01.rds"),{ 
   plan(multicore, workers = ncores)
@@
 
-bake(file = paste0(out_dir, "mif2_maximization/round_01/el_01.rds"),{
+bake(file = paste0(out_dir, "mif2_maximization/full/round_01/el_01.rds"),{
   eval_logLik_pomp(mifs_global_01, ncores, NP_MIF, 4248930)
@@
 
-% Continue: Maximization Round 2
+% Continue: Maximization Round 2 - Full rinit
 
@@
 
-bake(file = paste0(out_dir,"mif2_maximization/round_02/mifs_global_02.rds"),{ 
+bake(file = paste0(out_dir,"mif2_maximization/full/round_02/mifs_global_02.rds"),{ 
   plan(multicore, workers = ncores)
@@
 
-bake(file = paste0(out_dir, "mif2_maximization/round_02/el_02.rds"),{
+bake(file = paste0(out_dir, "mif2_maximization/full/round_02/el_02.rds"),{
   eval_logLik_pomp(mifs_global_02, ncores, NP_MIF, seed=4248930)
@@
 
-% Continue: Maximization Round 3
+% Continue: Maximization Round 3 - Full rinit
 
@@
 
-bake(file = paste0(out_dir, "mif2_maximization/round_03/mifs_global_03.rds"),{ 
+bake(file = paste0(out_dir, "mif2_maximization/full/round_03/mifs_global_03.rds"),{ 
   plan(multicore, workers = ncores)
@@
 
-bake(file = paste0(out_dir,"mif2_maximization/round_03/el_03.rds"),{
+bake(file = paste0(out_dir,"mif2_maximization/full/round_03/el_03.rds"),{
   eval_logLik_pomp(mifs_global_03, ncores, NP_MIF, seed = 4248930)
@@
 est_mle_vec <- el_03[which.max(el_03$logLik),]
+@
+
+% Maximization Round 1 - Stocks rinit
+
+<<maximization1-Stocks_rinit, echo=F, warning=F, message=F>>=
+set.seed(123)
+sir_box_sbh <- list(
+  beta1 = c(10, 15),
+  beta2 = c(0.2, 0.4),
+  beta3 = c(0.3, 0.5),
+  phi = c(0.01, 0.3),
+  beta11 = c(0.1, 0.2),
+  sigma_q = c(0.001, 0.1),
+  sigma_xi = c(65, 70),
+  gamma = c(1, 1),
+  delta1 = c(1/(5*52), 1/(5*52)),
+  delta2 = c(1/(55*52), 1/(55*52)),
+  alpha = c(1/(78.86912*52), 1/(78.86912*52)),
+  mu = c(0, 0),
+  N = c(82372825, 82372825),
+  omega = c(1/(1*52), 1/(1*52))
+)
+
+design_matrix_sbh <- runif_design(
+  lower = sapply(sir_box_sbh, `[`, 1),
+  upper = sapply(sir_box_sbh, `[`, 2),
+  n = ncores
+)
+
+bake(file = paste0(out_dir, "mif2_maximization/SBH_init/round_01/mifs_global_sbh_01.rds"),{ 
+  plan(multicore, workers = ncores)
+  foreach(i = 1:ncores, .options.future = list(seed = 652643293)) %dofuture% {
+    pomp::mif2(
+      sir_sbh_init,
+      Np = NP_MIF,
+      cooling.fraction.50 = 0.5,
+      rw.sd = rw_sd(
+        beta1 = 0.01, beta2 = 0.01, beta3 = 0.01, beta11 = 0.01, 
+        phi = 0.01, sigma_q = 0.01, sigma_xi = 0.01
+      ),
+      cooling.type = "geometric",
+      Nmif = NMIF,
+      params = design_matrix_sbh[i, ]
+    ) 
+  }
+}) -> mifs_global_sbh_01
+
+bake(file = paste0(out_dir, "mif2_maximization/SBH_init/round_01/sbh_el_01.rds"),{
+  eval_logLik_pomp(mifs_global_sbh_01, ncores, NP_MIF, 4248930)
+}) -> sbh_el_01
+
+@
+
+% Continue: Maximization Round 2 - Stocks rinit
+
+<<maximization2-Stocks_rinit, echo=F, warning=F, message=F>>=
+set.seed(1234)
+sd_02 = 2/3
+starting_values_sbh_02 <- choose_n_fits(sbh_el_01, top_n_fits, ncores)
+
+bake(file = paste0(out_dir,"mif2_maximization/SBH_init/round_02/mifs_global_sbh_02.rds"),{ 
+  plan(multicore, workers = ncores)
+  foreach(i = 1:ncores, .options.future = list(seed = 652643293)) %dofuture% {
+    pomp::mif2(
+      mifs_global_sbh_01[[1]],
+      rw.sd = rw_sd(
+        beta1 = 0.01*sd_02, beta2 = 0.01*sd_02, beta3 = 0.01*sd_02,
+        beta11 = 0.01*sd_02, phi = 0.01*sd_02,
+        sigma_q = 0.01*sd_02, sigma_xi = 0.01*sd_02
+      ),
+      params = starting_values_sbh_02[[i]]
+    ) 
+  }
+}) ->  mifs_global_sbh_02
+
+bake(file = paste0(out_dir, "mif2_maximization/SBH_init/round_02/sbh_el_02.rds"),{
+  eval_logLik_pomp(mifs_global_sbh_02, ncores, NP_MIF, seed=4248930)
+}) -> sbh_el_02
+@
+
+% Continue: Maximization Round 3 - Stocks rinit
+
+<<maximization3-Stocks_rinit, echo=F, warning=F, message=F>>=
+set.seed(12345)
+sd_03 = (2/3)^2
+starting_values_sbh_03 <- choose_n_fits(sbh_el_02, top_n_fits, ncores)
+
+bake(file = paste0(out_dir, "mif2_maximization/SBH_init/round_03/mifs_global_sbh_03.rds"),{ 
+  plan(multicore, workers = ncores)
+  foreach(i = 1:ncores, .options.future = list(seed = 652643293)) %dofuture% {
+    pomp::mif2(
+      mifs_global_sbh_02[[1]],
+      rw.sd = rw_sd(
+        beta1 = 0.01*sd_03, beta2 = 0.01*sd_03, beta3 = 0.01*sd_03,
+        beta11 = 0.01*sd_03, phi = 0.01*sd_03,
+        sigma_q = 0.01*sd_03, sigma_xi = 0.01*sd_03
+      ),
+      params = starting_values_sbh_03[[i]]
+    ) 
+  }
+}) -> mifs_global_sbh_03
+
+bake(file = paste0(out_dir,"mif2_maximization/SBH_init/round_03/sbh_el_03.rds"),{
+  eval_logLik_pomp(mifs_global_sbh_03, ncores, NP_MIF, seed = 4248930)
+}) ->  sbh_el_03
+@
+
+<<estimated-mle_stocks_init, echo=FALSE, message=FALSE, warning=FALSE>>=
+# Can be changed if more rounds needed
+est_mle_vec_sbh <- sbh_el_03[which.max(sbh_el_03$logLik),]
 @
@@
   ovov_sim_computation <- list()
-
   i <- 1
@@
 
+<<PAL-MLE-from-mif2, echo=FALSE, message=FALSE, warning=FALSE>>=
+prop <- 420
+realdat <- as.matrix(dat[,-1])
+init_dist_mif2_mle <- c(
+  pop*as.numeric(est_mle_vec["S10"]), 
+  pop*as.numeric(est_mle_vec["I10"]), 
+  pop*as.numeric(est_mle_vec["R10"]),
+  pop*as.numeric(est_mle_vec["S20"]),
+  pop*as.numeric(est_mle_vec["R20"]),
+  pop*as.numeric(est_mle_vec["I20"]),
+  pop*as.numeric(est_mle_vec["S30"]),
+  pop*as.numeric(est_mle_vec["R30"]),
+  pop*as.numeric(est_mle_vec["I30"])
+)
+
+initial_guess_mif2_mle <- c(
+  as.numeric(est_mle_vec["beta1"]), 
+  as.numeric(est_mle_vec["beta2"]), 
+  as.numeric(est_mle_vec["beta3"]),
+  as.numeric(est_mle_vec["phi"]),
+  as.numeric(est_mle_vec["beta11"]),
+  as.numeric(est_mle_vec["sigma_q"]),
+  as.numeric(est_mle_vec["sigma_xi"])
+)
+
+bake(file = paste0(out_dir,"pfilter/original_dataset/palfilter_mif2_mle.rds"),{
+ set.seed(1223)
+ rotavirus_SMC_qropxi(
+    init_dist = init_dist_mif2_mle, 
+    y_obs = realdat, 
+    m = 9, 
+    regular_params = c(
+      initial_guess_mif2_mle[1], 
+      initial_guess_mif2_mle[2], 
+      initial_guess_mif2_mle[3], 
+      initial_guess_mif2_mle[4],
+      initial_guess_mif2_mle[5]
+    ), 
+    gamma_par = c(initial_guess_mif2_mle[7], 1/initial_guess_mif2_mle[7]), 
+    norm_par = c(0.07, initial_guess_mif2_mle[6]), 
+    prop = t(prop), 
+    n_particles = NP_MIF, 
+    ncores = ncores
+  ) 
+}) -> palfilter_mif2_mle
+@
+
 \begin{table}[ht] % The placement specifier can be [h!tbp]
@@
      \Sexpr{myround(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}
+  \\
+  6. & PAL & Data from WWR, estimated initial values &
+     \Sexpr{myround(-2*palfilter_mif2_mle$log_lik + 2*7 ,2)}
   \\
@@
 \begin{center}    
-    \begin{tabular}{lcc}  % Left-aligned for the first column, centered for the other two columns
+    \begin{tabular}{lccc}  % Left-aligned for the first column, centered for the other two columns
       \toprule
-      Parameter      & PAL      & iterated filtering \\
+      Parameter      & PAL      & iterated filtering & iterated filtering (stocks)\\
       \midrule
-      \(\beta_1\)    & 11.48     & $\Sexpr{round(est_mle_vec["beta1"],2)}$\\
-      \(\beta_2\)    & 0.25      & $\Sexpr{round(est_mle_vec["beta2"],2)}$\\
-      \(\beta_3\)    & 0.35      & $\Sexpr{round(est_mle_vec["beta3"],2)}$\\
-      \(\phi\)       & 0.14      & $\Sexpr{round(est_mle_vec["phi"],2)}$\\
-      \(\rho\)       & 0.16      & $\Sexpr{round(est_mle_vec["beta11"],2)}$\\
-      \(\sigma^2_q\) & 0.021     & $\Sexpr{round(est_mle_vec["sigma_q"],3)}$\\
-      \(\sigma_\xi\) & 66.89     & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$\\
-      \(S_{10}\)     & 3876549   & $\Sexpr{round(pop*est_mle_vec["S10"],0)}$\\
-      \(I_{10}\)     & 30351     & $\Sexpr{round(pop*est_mle_vec["I10"],0)}$\\
-      \(R_{10}\)     & 1315221   & $\Sexpr{round(pop*est_mle_vec["R10"],0)}$\\
-      \(S_{20}\)     & 57139612  & $\Sexpr{round(pop*est_mle_vec["S20"],0)}$\\
-      \(I_{20}\)     & 871       & $\Sexpr{round(pop*est_mle_vec["I20"],0)}$\\
-      \(R_{20}\)     & 302852    & $\Sexpr{round(pop*est_mle_vec["R20"],0)}$\\
-      \(S_{30}\)     & 19573727  & $\Sexpr{round(pop*est_mle_vec["S30"],0)}$\\
-      \(I_{30}\)     & 2550      & $\Sexpr{round(pop*est_mle_vec["I30"],0)}$\\
-      \(R_{30}\)     & 131092    & $\Sexpr{round(pop*est_mle_vec["R30"],0)}$\\
+      \(\beta_1\)    & 11.48     & $\Sexpr{round(est_mle_vec["beta1"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta1"],2)}$\\
+      \(\beta_2\)    & 0.25      & $\Sexpr{round(est_mle_vec["beta2"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta2"],2)}$\\
+      \(\beta_3\)    & 0.35      & $\Sexpr{round(est_mle_vec["beta3"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta3"],2)}$\\
+      \(\phi\)       & 0.14      & $\Sexpr{round(est_mle_vec["phi"],2)}$   & $\Sexpr{round(est_mle_vec_sbh["phi"],2)}$\\
+      \(\rho\)       & 0.16      & $\Sexpr{round(est_mle_vec["beta11"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta11"],2)}$\\
+      \(\sigma^2_q\) & 0.021     & $\Sexpr{round(est_mle_vec["sigma_q"],3)}$ & $\Sexpr{round(est_mle_vec_sbh["sigma_q"],3)}$\\
+      \(\sigma_\xi\) & 66.89     & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$ & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$\\
+      \(S_{10}\)     & 3876549   & $\Sexpr{round(pop*est_mle_vec["S10"],0)}$  & $\Sexpr{round(pop*est_mle_vec_sbh["S10"],0)}$\\
+      \(I_{10}\)     & 30351     & $\Sexpr{round(pop*est_mle_vec["I10"],0)}$  & $\Sexpr{round(pop*est_mle_vec_sbh["I10"],0)}$\\
+      \(R_{10}\)     & 1315221   & $\Sexpr{round(pop*est_mle_vec["R10"],0)}$  & $\Sexpr{round(pop*est_mle_vec_sbh["R10"],0)}$\\
+      \(S_{20}\)     & 57139612  & $\Sexpr{round(pop*est_mle_vec["S20"],0)}$  & $\Sexpr{round(pop*est_mle_vec_sbh["S20"],0)}$\\
+      \(I_{20}\)     & 871       & $\Sexpr{round(pop*est_mle_vec["I20"],0)}$  & $\Sexpr{round(pop*est_mle_vec_sbh["I20"],0)}$\\
+      \(R_{20}\)     & 302852    & $\Sexpr{round(pop*est_mle_vec["R20"],0)}$  & $\Sexpr{round(pop*est_mle_vec_sbh["R20"],0)}$\\
+      \(S_{30}\)     & 19573727  & $\Sexpr{round(pop*est_mle_vec["S30"],0)}$  & $\Sexpr{round(pop*est_mle_vec_sbh["S30"],0)}$\\
+      \(I_{30}\)     & 2550      & $\Sexpr{round(pop*est_mle_vec["I30"],0)}$  & $\Sexpr{round(pop*est_mle_vec_sbh["I30"],0)}$\\
+      \(R_{30}\)     & 131092    & $\Sexpr{round(pop*est_mle_vec["R30"],0)}$  & $\Sexpr{round(pop*est_mle_vec_sbh["R30"],0)}$\\
       \midrule
-      \textbf{AIC}  & 13778.08  & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$\\
+      \textbf{AIC}  & 13778.08  & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$ & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec_sbh["logLik"]),2)}$\\
       \bottomrule

### 2024-08-30 ms_01.tex
output_01

--- 2024-08-23 ms.Rnw
@@
-\documentclass[10pt]{article}
+\documentclass[10pt]{article}\usepackage[]{graphicx}\usepackage[]{xcolor}
+% maxwidth is the original width if it is less than linewidth
+% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlsng}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hldef}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+\let\hlipl\hlkwb
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 \usepackage{color}
@@
 \bibliographystyle{apalike}
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 
-<<setup,echo=FALSE,message=FALSE,warning=FALSE>>=
-rm(list = ls())     # clear objects
-library("knitr")
-opts_knit$set(concordance=TRUE)
-opts_chunk$set(
-  tidy = TRUE,
-  cache = TRUE,
-#  cache = FALSE,  # set cache = FALSE when editing the code
-# message = FALSE,
-# warning = FALSE,
-  tidy.opts = list(
-    keep.blank.line = FALSE
-  ),
-  comment = "",
-  echo = FALSE,
-  results = FALSE,
-  dev.args = list(
-    bg = "transparent",
-    pointsize = 9
-  ),
-  fig.path = "figure/"
-)
-
-myround <- function(x, digits = 1) {
-  # taken from the broman package
-  if (digits < 1)
-    stop("This is intended for the case digits >= 1.")
-  if (length(digits) > 1) {
-    digits <- digits[1]
-    warning("Using only digits[1]")
-  }
-  tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
-  zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
-  tmp[tmp == paste0("-", zero)] <- zero
-  tmp
-}
-
-graphics.off()      # close graphics windows
-library(pomp)
-library(ggplot2)
-library(scales)
-library(foreach)
-library(doParallel)
-library(doFuture)
-library(dplyr)
-library(Rcpp)
-library(RcppArmadillo)
-sourceCpp('rotavirus_normq.cpp')
-source("choose_n_fits.R")
-source("eval_logLik_pomp.R")
-
-RUN_LEVEL = 3
-NP_MIF       = switch(RUN_LEVEL, 10,   5000,  50000)
-NMIF         = switch(RUN_LEVEL, 5,    20,    100)
-ncores       = switch(RUN_LEVEL, 4,    10,    36)
-top_n_fits   = switch(RUN_LEVEL, 1,    2,     12)
-sim_rep      = switch(RUN_LEVEL, 5,    20,    100) #sim_ovov
-
-#population number
-pop <- 82372825
-time <- c(1:416)
-
-out_dir <- paste0("output_0",RUN_LEVEL, "/")
-@
-
 % OvOv model is here
 
-<<OvOv-model, echo=F, warning=F, message=F>>=
-# measurement model 
-dmeas <- Csnippet("
-  lik =  dbinom(cases1, H1, q1, 1) +
-  dbinom(cases2, H2, q2, 1) +
-  dbinom(cases3, H3, q3, 1);"
-)
-rmeas <-  Csnippet("
-  cases1 = rbinom(H1, q1);
-  cases2 = rbinom(H2, q2);
-  cases3 = rbinom(H3, q3);
-")
+% Maximization Round 1 - Full rinit
 
-rproc <- Csnippet("
-  int I = I1 + I2 + I3;
-  int trans_S1[3], trans_S2[3], trans_S3[2], trans_I1[3], trans_I2[3], 
-  trans_I3[2], trans_R1[3], trans_R2[3], trans_R3[2];
+% Continue: Maximization Round 2 - Full rinit
 
-  double prob_S1[3], prob_I1[3], prob_R1[3], prob_S2[3], prob_I2[3], prob_R2[3],
-  prob_S3[2], prob_I3[2], prob_R3[2];
+% Continue: Maximization Round 3 - Full rinit
 
-  double xi = rgamma(sigma_xi, 1/sigma_xi);
+% Maximization Round 1 - Stocks rinit
 
-  double kappa = (1 + beta11*cos(2*3.141593*t/52 + phi)) * xi;
+% Continue: Maximization Round 2 - Stocks rinit
 
-  // Define rate
-  prob_S1[0] = 1 - exp(-dt*beta1*kappa*I/N); // 0->1
-  prob_S1[1] = 1 - exp(-delta1*dt);
-  prob_S1[2] = exp(-delta1*dt) + exp(-dt*beta1*kappa*I/N) - 1;
-
-  prob_I1[0] = 1 - exp(-gamma*dt);
-  prob_I1[1] = 1 - exp(-delta1*dt);
-  prob_I1[2] = exp(-gamma*dt)+exp(-delta1*dt) - 1;
-
-  prob_R1[0] = 1 - exp(-omega*dt);  // E_1,t this goes back to S_1,(t+1)
-  prob_R1[1] = 1 - exp(-delta1*dt);
-  prob_R1[2] = exp(-omega*dt) + exp(-delta1*dt) - 1;
-
-  prob_S2[0] = 1 - exp(-dt*beta2*kappa*I/N);
-  prob_S2[1] = 1 - exp(-delta2*dt);
-  prob_S2[2] = exp(-delta2*dt) + exp(-dt*beta2*kappa*I/N) - 1;
-
-  prob_I2[0] = 1 - exp(-dt*gamma);
-  prob_I2[1] = 1 - exp(-dt*delta2);
-  prob_I2[2] = exp(-dt*gamma)+exp(-dt*delta2) - 1;
-
-  prob_R2[0] = 1 - exp(-dt*omega);  // E_1,t this goes back to S_1,(t+1)
-  prob_R2[1] = 1 - exp(-dt*delta2);
-  prob_R2[2] = exp(-dt*omega) + exp(-dt*delta2) - 1;
-
-  // For Age Group (3): Die first before transition;
-
-  int S3mD, I3mD, R3mD;
-
-  S3mD = rbinom(S3, 1-dt*mu); // S3 minus Death: mu is the death rate, so it's 1-mu here
-  I3mD = rbinom(I3, 1-dt*mu);
-  R3mD = rbinom(R3, 1-dt*mu);
-
-  prob_S3[0] = 1-exp(-dt*beta3*kappa*I/N);
-  prob_S3[1] = exp(-dt*beta3*kappa*I/N);
-
-  prob_I3[0] = 1 - exp(-dt*gamma);
-  prob_I3[1] = exp(-dt*gamma);
-
-  prob_R3[0] = 1 - exp(-dt*omega);
-  prob_R3[1] = exp(-dt*omega);
-
-  // Transition
-  // B: S->I
-  // C: I->R
-  // F: Aging: (1)->(2)->(3)
-  // E: R->S
-  // D: Death
-  //// Note: Here S_1, S_2... are all old values from (t-1)
-  rmultinom(S1, &prob_S1, 3, &trans_S1); // B, F, S-B-F
-  rmultinom(I1, &prob_I1, 3, &trans_I1); // C, F, I-C-F
-  rmultinom(R1, &prob_R1, 3, &trans_R1); // E, F, R-E-F
-
-  rmultinom(S2, &prob_S2, 3, &trans_S2); // B, F, S-B-F
-  rmultinom(I2, &prob_I2, 3, &trans_I2); // C, F, I-C-F
-  rmultinom(R2, &prob_R2, 3, &trans_R2); // E, F, R-E-F
-
-  rmultinom(S3mD, &prob_S3, 2, &trans_S3); // B, (S-D)-B
-  rmultinom(I3mD, &prob_I3, 2, &trans_I3); // C, (I-D)-C
-  rmultinom(R3mD, &prob_R3, 2, &trans_R3); // E, (R-D)-E
-
-  S1 = trans_S1[2] + trans_R1[0] + rpois(4*1025.7); // Include Birth
-  I1 = trans_I1[2] + trans_S1[0];
-  R1 = trans_R1[2] + trans_I1[0];
-
-  S2 = trans_S2[2] + trans_R2[0] + trans_S1[1]; // Include Aging
-  I2 = trans_I2[2] + trans_S2[0] + trans_I1[1];
-  R2 = trans_R2[2] + trans_I2[0] + trans_R1[1];
-
-  S3 = trans_S3[1] + trans_R3[0] + trans_S2[1]; // Include Aging
-  I3 = trans_I3[1] + trans_S3[0] + trans_I2[1];
-  R3 = trans_R3[1] + trans_I3[0] + trans_R2[1];
-
-  //Accumvar
-  H1 += trans_S1[0];
-  H2 += trans_S2[0];
-  H3 += trans_S3[0];
-
-  q1 = -1; 
-  while(q1 < 0 || q1 > 1){
-    q1 = rnorm(0.07, sigma_q);
-  }
-
-  q2 = -1; 
-  while(q2 < 0 || q2 > 1){
-    q2 = rnorm(0.07, sigma_q);
-  }
-
-  q3 = -1; 
-  while(q3 < 0 || q3 > 1){
-    q3 = rnorm(0.07, sigma_q);
-  }
-")
-
-# full-estimate-rinit
-rinit_full <- Csnippet("
-  double m = N/(S10+I10+R10+S20+I20+R20+S30+I30+R30);
-  I1=nearbyint(m*I10);
-  I2=nearbyint(m*I20);
-  I3=nearbyint(m*I30);
-  S1=nearbyint(m*S10);
-  S2=nearbyint(m*S20);
-  S3=nearbyint(m*S30);
-  R1=nearbyint(m*R10);
-  R2=nearbyint(m*R20);
-  R3=nearbyint(m*R30);
-  H1 = 0;
-  H2 = 0;
-  H3 = 0;
-")
-
-# WWR's rinit
-rinit_wwr <- Csnippet("
-  S1=3876549;
-  S2=57139612;
-  S3=19573727;
-  I1=30351;
-  I2=871;
-  I3=2550;
-  R1=1315221;
-  R2=302852;
-  R3=131092;
-  H1 = 0;
-  H2 = 0;
-  H3 = 0;
-")
-
-# x <- c(3876549,57139612,19573727,30351,871,2550,1315221,302852,131092)
-# x/sum(x)
-
-# define parameters (fixed)
-params_fixed <- c(
-  gamma = 1, 
-  delta1 = 1/(5*52),
-  delta2 = 1/(55*52), 
-  alpha = 1/(78.86912*52), 
-  mu = 0, 
-  N = 82372825, 
-  omega = 1/(1*52)
-)
-# Set to MLE
-params_wwr <- c(
-  params_fixed, 
-  "beta1" = 11.48,
-  "beta2" = 0.25,
-  "beta3" = 0.35,
-  "phi" = 0.14,
-  "beta11" = 0.16,
-  "sigma_q" = 0.021,
-  "sigma_xi" = 66.89
-)
-
-params_full <- c(
-  params_wwr,
-  "S10" = 4.706102e-02,
-  "S20" = 6.936707e-01,
-  "S30" = 2.376236e-01,
-  "I10" = 3.684589e-04,
-  "I20" = 1.057388e-05,
-  "I30" = 3.095681e-05,
-  "R10" = 1.596669e-02,
-  "R20" = 3.676601e-03,
-  "R30" = 1.591447e-03
-)
-
-pt_full <- pomp::parameter_trans(
-  log = c("beta1", "beta2", "beta3", "sigma_q", "sigma_xi"),
-  logit = c("beta11"),
-  barycentric = c("S10","I10","R10",
-                  "S20","I20","R20",
-                  "S30","I30","R30"),
-  toEst = pomp::Csnippet("T_phi = logit(phi/(M_2PI));"),
-  fromEst = pomp::Csnippet("phi = M_2PI*expit(T_phi);")
-)
-
-pt_wwr <- pomp::parameter_trans(
-  log = c("beta1", "beta2", "beta3", "sigma_q", "sigma_xi"),
-  logit = c("beta11"),
-  toEst = pomp::Csnippet("T_phi = logit(phi/(M_2PI));"),
-  fromEst = pomp::Csnippet("phi = M_2PI*expit(T_phi);")
-)
-
-dat <- read.table("real_rotavirus_metadata.txt") |>
-  arrange(time)
-
-statenames <- c(
-  "S1", "I1", "R1", "H1",
-  "S2", "I2", "R2", "H2",
-  "S3", "I3", "R3", "H3",
-  "q1", "q2", "q3"
-)
-
-sir_wwr <- pomp(
-  data = dat,
-  times = "time",
-  t0 = 0,
-  dmeasure = dmeas,
-  rmeasure = rmeas,
-  rprocess = discrete_time(step.fun = rproc, delta.t = 1/4),
-  statenames = statenames,
-  paramnames = names(params_wwr),
-  accumvars = c("H1", "H2", "H3"),
-  rinit = rinit_wwr,
-  partrans = pt_wwr,
-  params = params_wwr
-)
-
-sir_full <- pomp(
-  sir_wwr,
-  rinit = rinit_full,
-  params = params_full,
-  partrans = pt_full,
-  statenames = statenames,
-  paramnames = names(params_full)
-)
-@
-
-% Maximization Round 1
-
-<<maximization1, echo=F, warning=F, message=F>>=
-set.seed(123)
-sir_box <- list(
-  beta1 = c(10, 15),
-  beta2 = c(0.2, 0.4),
-  beta3 = c(0.3, 0.5),
-  phi = c(0.01, 0.3),
-  beta11 = c(0.1, 0.2),
-  sigma_q = c(0.001, 0.1),
-  sigma_xi = c(65, 70),
-  S10 = c(0.01, 0.05),
-  I10 = c(0.0001, 0.0005),
-  R10 = c(0.01, 0.05),
-  S20 = c(0.01, 0.7),
-  I20 = c(0.000001, 0.00002),
-  R20 = c(0.001, 0.005),
-  S30 = c(0.01, 0.5),
-  I30 = c(0.000001, 0.0001),
-  R30 = c(0.0001, 0.005),
-  gamma = c(1, 1),
-  delta1 = c(1/(5*52), 1/(5*52)),
-  delta2 = c(1/(55*52), 1/(55*52)),
-  alpha = c(1/(78.86912*52), 1/(78.86912*52)),
-  mu = c(0, 0),
-  N = c(82372825, 82372825),
-  omega = c(1/(1*52), 1/(1*52))
-)
-
-design_matrix <- runif_design(
-  lower = sapply(sir_box, `[`, 1),
-  upper = sapply(sir_box, `[`, 2),
-  n = ncores
-)
-
-bake(file = paste0(out_dir, "mif2_maximization/round_01/mifs_global_01.rds"),{ 
-  plan(multicore, workers = ncores)
-  foreach(i = 1:ncores, .options.future = list(seed = 652643293)) %dofuture% {
-    pomp::mif2(
-      sir_full,
-      Np = NP_MIF,
-      cooling.fraction.50 = 0.5,
-      rw.sd = rw_sd(
-        beta1 = 0.01, beta2 = 0.01, beta3 = 0.01, beta11 = 0.01, 
-        phi = 0.01, sigma_q = 0.01, sigma_xi = 0.01,
-        S10 = ivp(0.24), I10 = ivp(0.24), R10 = ivp(0.24),
-        S20 = ivp(0.24), I20 = ivp(0.24), R20 = ivp(0.24),
-        S30 = ivp(0.24), I30 = ivp(0.24), R30 = ivp(0.24)
-      ),
-      cooling.type = "geometric",
-      Nmif = NMIF,
-      params = design_matrix[i, ]
-    ) 
-  }
-}) -> mifs_global_01
-
-bake(file = paste0(out_dir, "mif2_maximization/round_01/el_01.rds"),{
-  eval_logLik_pomp(mifs_global_01, ncores, NP_MIF, 4248930)
-}) ->  el_01
-
-@
-
-% Continue: Maximization Round 2
-
-<<maximization2, echo=F, warning=F, message=F>>=
-set.seed(1234)
-sd_02 = 2/3
-starting_values_02 <- choose_n_fits(el_01, top_n_fits, ncores)
-
-bake(file = paste0(out_dir,"mif2_maximization/round_02/mifs_global_02.rds"),{ 
-  plan(multicore, workers = ncores)
-  foreach(i = 1:ncores, .options.future = list(seed = 652643293)) %dofuture% {
-    pomp::mif2(
-      mifs_global_01[[1]],
-      rw.sd = rw_sd(
-        beta1 = 0.01*sd_02, beta2 = 0.01*sd_02, beta3 = 0.01*sd_02,
-        beta11 = 0.01*sd_02, phi = 0.01*sd_02,
-        sigma_q = 0.01*sd_02, sigma_xi = 0.01*sd_02,
-        S10 = ivp(0.24)*sd_02, I10 = ivp(0.24)*sd_02, R10 = ivp(0.24)*sd_02,
-        S20 = ivp(0.24)*sd_02, I20 = ivp(0.24)*sd_02, R20 = ivp(0.24)*sd_02,
-        S30 = ivp(0.24)*sd_02, I30 = ivp(0.24)*sd_02, R30 = ivp(0.24)*sd_02
-      ),
-      params = starting_values_02[[i]]
-    ) 
-  }
-}) ->  mifs_global_02
-
-bake(file = paste0(out_dir, "mif2_maximization/round_02/el_02.rds"),{
-  eval_logLik_pomp(mifs_global_02, ncores, NP_MIF, seed=4248930)
-}) ->  el_02
-@
-
-% Continue: Maximization Round 3
-
-<<maximization3, echo=F, warning=F, message=F>>=
-set.seed(12345)
-sd_03 = (2/3)^2
-starting_values_03 <- choose_n_fits(el_02, top_n_fits, ncores)
-
-bake(file = paste0(out_dir, "mif2_maximization/round_03/mifs_global_03.rds"),{ 
-  plan(multicore, workers = ncores)
-  foreach(i = 1:ncores, .options.future = list(seed = 652643293)) %dofuture% {
-    pomp::mif2(
-      mifs_global_02[[1]],
-      rw.sd = rw_sd(
-        beta1 = 0.01*sd_03, beta2 = 0.01*sd_03, beta3 = 0.01*sd_03,
-        beta11 = 0.01*sd_03, phi = 0.01*sd_03,
-        sigma_q = 0.01*sd_03, sigma_xi = 0.01*sd_03,
-        S10 = ivp(0.24)*sd_03, I10 = ivp(0.24)*sd_03, R10 = ivp(0.24)*sd_03,
-        S20 = ivp(0.24)*sd_03, I20 = ivp(0.24)*sd_03, R20 = ivp(0.24)*sd_03,
-        S30 = ivp(0.24)*sd_03, I30 = ivp(0.24)*sd_03, R30 = ivp(0.24)*sd_03
-      ),
-      params = starting_values_03[[i]]
-    ) 
-  }
-}) -> mifs_global_03
-
-bake(file = paste0(out_dir,"mif2_maximization/round_03/el_03.rds"),{
-  eval_logLik_pomp(mifs_global_03, ncores, NP_MIF, seed = 4248930)
-}) ->  el_03
-@
-
-<<estimated-mle, echo=FALSE, message=FALSE, warning=FALSE>>=
-# Can be changed if more rounds needed
-est_mle_vec <- el_03[which.max(el_03$logLik),]
-@
+% Continue: Maximization Round 3 - Stocks rinit
 
@@
 
-<<sim-100, echo=F, warning=F, message=F>>=
-bake(file = paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"),{
-  ovov_sim_computation <- list()
-
-  i <- 1
-  repeat {
-    # Simulation
-    sim_data <- simulate(sir_wwr, format = "data.frame") |>
-      subset(select=c("time", "cases1", "cases2", "cases3"))
-
-    if (!any(sim_data == 0)) {
-      ovov_sim_computation[[length(ovov_sim_computation) + 1]] <- sim_data
-      i <- i + 1  
-    } else {
-      print(paste("Condition not met for iteration", i, "- retrying"))
-    }
-
-    if (length(ovov_sim_computation) >= sim_rep) break  
-  }
-  ovov_sim_computation
-}) -> ovov_sim_100
-@
-
-<<pfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
-bake(file = paste0(out_dir, "pfilter/simulated_dataset/pfilter_ovov_sim_100.rds"),{
-  pfilter_obj <- list()
-  pfilter_ovov_sim_computation <- c()
-
-  for(i in 1:length(ovov_sim_100)){
-    pomp(
-      data = ovov_sim_100[[i]],
-      times = "time",
-      t0 = 0,
-      dmeasure = dmeas,
-      rmeasure = rmeas,
-      rprocess = discrete_time(step.fun = rproc, delta.t = 1/4),
-      statenames = statenames,
-      paramnames = names(params_wwr),
-      accumvars = c("H1", "H2", "H3"),
-      rinit = rinit_wwr,
-      partrans = pt_wwr,
-      params = params_wwr
-    )  -> pfilter_obj[[i]]
-
-    plan(multicore, workers = ncores)
-
-    foreach(j = 1:ncores, .combine = c,
-      .options.future = list(seed = 998468235L)
-    ) %dofuture% {
-      pfilter(
-        pfilter_obj[[i]], Np = NP_MIF, save.states = FALSE, params = params_wwr
-      )
-    } -> pfs
-    pfilter_ovov_sim_computation[i] <- logmeanexp(logLik(pfs))  
-  }
-  pfilter_ovov_sim_computation
-}) ->  pfilter_ovov_sim_100
-@
-
-<<palfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
-prop <- 420
-init_dist <- c(
-  3876549, 30351, 1315221, 57139612, 871, 302852, 19573727, 2550, 131092
-)
-initial_guess <- c(11.48, 0.25, 0.35, 0.14, 0.16, 0.021, 66.89)
-
-bake(file = paste0(out_dir, "pfilter/simulated_dataset/palfilter_ovov_sim_100.rds"),{
-  plan(multicore, workers = ncores)
-  foreach(
-    i = 1:length(ovov_sim_100), .combine = c, .options.future = list(seed = 998468235L)
-  ) %dofuture% {
-    pal_sim_dat <- ovov_sim_100[[i]][,-1] |> as.matrix()
-    lik_list <- rotavirus_SMC_qropxi(
-      init_dist = init_dist, 
-      y_obs = pal_sim_dat, 
-      m = 9, 
-      regular_params = c(
-        initial_guess[1], initial_guess[2], initial_guess[3], initial_guess[4],
-        initial_guess[5]
-      ), 
-      gamma_par = c(initial_guess[7], 1/initial_guess[7]), 
-      norm_par = c(0.07, initial_guess[6]), 
-      prop = t(prop), 
-      n_particles = NP_MIF,
-      ncores = ncores
-    )
-    lik_list$log_lik
-  }
-}) -> palfilter_ovov_sim_100
-
-# lik_list$ll_storage |> sum()
-@
-
-<<pfilter_wwr, echo=FALSE, message=FALSE, warning=FALSE>>=
-bake(file = paste0(out_dir, "pfilter/original_dataset/pfilter_wwr.rds"),{
-  plan(multicore, workers = ncores)
-  foreach(
-    i = 1:ncores, .combine = c, .options.future = list(seed = 998468235L)
-  ) %dofuture% {
-      pfilter(sir_wwr, Np = NP_MIF, params = params_wwr)
-    } 
-}) ->  pfilter_wwr
-
-bake(file = paste0(out_dir, "pfilter/original_dataset/pfilter_wwr_cond_lik.rds"),{
-  pfilter_wwr_cond_lik_computation <- list()
-  for(k in 1:ncores){
-    pfilter_wwr_cond_lik_computation[[k]] <- pfilter_wwr[[k]]@cond.logLik
-  }
-  pfilter_wwr_cond_lik_computation
-}) -> pfilter_wwr_cond_lik
-@
-
-<<palfilter_wwr, echo=FALSE>>=
-realdat <- as.matrix(dat[,-1])
-
-bake(file = paste0(out_dir,"pfilter/original_dataset/palfilter_lik_list_wwr.rds"),{
- set.seed(1223)
- rotavirus_SMC_qropxi(
-    init_dist = init_dist, 
-    y_obs = realdat, 
-    m = 9, 
-    regular_params = c(
-      initial_guess[1], initial_guess[2], initial_guess[3], initial_guess[4],
-      initial_guess[5]
-    ), 
-    gamma_par = c(initial_guess[7], 1/initial_guess[7]), 
-    norm_par = c(0.07, initial_guess[6]), 
-    prop = t(prop), 
-    n_particles = NP_MIF, 
-    ncores = ncores
-  ) 
-}) -> palfilter_lik_list_wwr
-@
-
 \begin{table}[ht] % The placement specifier can be [h!tbp]
@@
   3. & PAL & Model, data and MLE from WWR &
-    \Sexpr{myround(-2*palfilter_lik_list_wwr$log_lik + 2*7 ,2)} 
+    13814.57 
   \\
   4. & PF & Model, data and MLE from WWR &
-    \Sexpr{myround(-2*logmeanexp(logLik(pfilter_wwr)+ 2*7),2)}
+    25785.52
   \\
   5. & PF & Data from WWR, estimated initial values &
-     \Sexpr{myround(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}
+     143875.53
+  \\
+  6. & PAL & Data from WWR, estimated initial values &
+     15376.67
   \\
@@
 
-<<plot2-cond-loglik, echo=FALSE, fig.height=4, fig.width=5.5, out.width="4.5in",fig.align="center", fig.cap='Conditional log-likelihoods computed using PAL (red) and PF (blue) for the rotavirus model and MLE from WWR.
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+
+{\centering \includegraphics[width=4.5in]{figure/plot2-cond-loglik-1}
+
+}
+
+\caption[Conditional log-likelihoods computed using PAL (red) and PF (blue) for the rotavirus model and MLE from WWR]{Conditional log-likelihoods computed using PAL (red) and PF (blue) for the rotavirus model and MLE from WWR.
 We used 50000 particle for both PF and PAL.
 PF averaged over 36 replications to reduce the Monte Carlo variance, but this step was not necessary for PAL.
-The main sources of likelihood shortfall of PF are at time points $t=1, 2, 3, 11, 81, 194$, and $325-333$.'>>=
+The main sources of likelihood shortfall of PF are at time points $t=1, 2, 3, 11, 81, 194$, and $325-333$.}\label{fig:plot2-cond-loglik}
+\end{figure}
 
-# Combine the data into a single dataframe for SMC with grouping variable
-data_list <- lapply(1:ncores, function(i) {
-  data.frame(
-    time = time, 
-    cond_loglik = pfilter_wwr_cond_lik[[i]], 
-    method = "SMC",
-    group = i
-  )
-})
-
-# Combine all the SMC data frames into one
-df_smc <- bind_rows(data_list)
-
-# Create the PAL data frame
-df_pal <- data.frame(
-  time = time, 
-  cond_loglik = palfilter_lik_list_wwr$ll_storage, 
-  method = "PAL",
-  group = 1  # Single group for PAL
-)
-
-# Combine both data frames
-df <- bind_rows(df_smc, df_pal)
-
-# Plot with ggplot
-ggplot(df, aes(x = time, y = cond_loglik, color = method)) +
-  geom_line(data = df %>% filter(method == "SMC"), aes(group = group), color = "blue", alpha = 0.7) +
-  geom_line(data = df %>% filter(method == "PAL"), aes(group = group), color = "red") +
-  theme_minimal() +
-  labs(
-    title = NULL,
-    x = "Time (weeks)",
-    y = "Conditional Log-likelihood",
-    color = "Method"
-  ) +
-  theme(legend.position = "bottom")
-@
+\end{knitrout}
 
@@
 \begin{center}    
-    \begin{tabular}{lcc}  % Left-aligned for the first column, centered for the other two columns
+    \begin{tabular}{lccc}  % Left-aligned for the first column, centered for the other two columns
       \toprule
-      Parameter      & PAL      & iterated filtering \\
+      Parameter      & PAL      & iterated filtering & iterated filtering (stocks)\\
       \midrule
-      \(\beta_1\)    & 11.48     & $\Sexpr{round(est_mle_vec["beta1"],2)}$\\
-      \(\beta_2\)    & 0.25      & $\Sexpr{round(est_mle_vec["beta2"],2)}$\\
-      \(\beta_3\)    & 0.35      & $\Sexpr{round(est_mle_vec["beta3"],2)}$\\
-      \(\phi\)       & 0.14      & $\Sexpr{round(est_mle_vec["phi"],2)}$\\
-      \(\rho\)       & 0.16      & $\Sexpr{round(est_mle_vec["beta11"],2)}$\\
-      \(\sigma^2_q\) & 0.021     & $\Sexpr{round(est_mle_vec["sigma_q"],3)}$\\
-      \(\sigma_\xi\) & 66.89     & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$\\
-      \(S_{10}\)     & 3876549   & $\Sexpr{round(pop*est_mle_vec["S10"],0)}$\\
-      \(I_{10}\)     & 30351     & $\Sexpr{round(pop*est_mle_vec["I10"],0)}$\\
-      \(R_{10}\)     & 1315221   & $\Sexpr{round(pop*est_mle_vec["R10"],0)}$\\
-      \(S_{20}\)     & 57139612  & $\Sexpr{round(pop*est_mle_vec["S20"],0)}$\\
-      \(I_{20}\)     & 871       & $\Sexpr{round(pop*est_mle_vec["I20"],0)}$\\
-      \(R_{20}\)     & 302852    & $\Sexpr{round(pop*est_mle_vec["R20"],0)}$\\
-      \(S_{30}\)     & 19573727  & $\Sexpr{round(pop*est_mle_vec["S30"],0)}$\\
-      \(I_{30}\)     & 2550      & $\Sexpr{round(pop*est_mle_vec["I30"],0)}$\\
-      \(R_{30}\)     & 131092    & $\Sexpr{round(pop*est_mle_vec["R30"],0)}$\\
+      \(\beta_1\)    & 11.48     & $13.53$ & $16.53$\\
+      \(\beta_2\)    & 0.25      & $0.42$ & $0.25$\\
+      \(\beta_3\)    & 0.35      & $0.24$ & $0.18$\\
+      \(\phi\)       & 0.14      & $0.23$   & $0.27$\\
+      \(\rho\)       & 0.16      & $0.11$ & $0.13$\\
+      \(\sigma^2_q\) & 0.021     & $0.052$ & $0.109$\\
+      \(\sigma_\xi\) & 66.89     & $24.83$ & $24.83$\\
+      \(S_{10}\)     & 3876549   & $2490623$  & $5094840$\\
+      \(I_{10}\)     & 30351     & $3599$  & $2871$\\
+      \(R_{10}\)     & 1315221   & $3035890$  & $124410$\\
+      \(S_{20}\)     & 57139612  & $55481369$  & $57385600$\\
+      \(I_{20}\)     & 871       & $222$  & $639$\\
+      \(R_{20}\)     & 302852    & $42911$  & $57072$\\
+      \(S_{30}\)     & 19573727  & $21155579$  & $19697600$\\
+      \(I_{30}\)     & 2550      & $442$  & $174$\\
+      \(R_{30}\)     & 131092    & $162190$  & $9578$\\
       \midrule
-      \textbf{AIC}  & 13778.08  & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$\\
+      \textbf{AIC}  & 13778.08  & $143875.53$ & $444098.92$\\
       \bottomrule
@@
 
-<<plot1-100-sim-compare,echo=FALSE, fig.height=4, fig.width=5.5, out.width="4.5in",fig.align="center",fig.cap='Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model.
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
+
+{\centering \includegraphics[width=4.5in]{figure/plot1-100-sim-compare-1}
+
+}
+
+\caption[Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model]{Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model.
 Simulations with one or more zero counts were disqualified since they resulted in errors for the PAL implementation.
@@
 PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL.
-The red line corresponds to equality of the two estimates.'>>=
-data <- data.frame(
-  pfilter = pfilter_ovov_sim_100, 
-  palfilter = palfilter_ovov_sim_100
-)
+The red line corresponds to equality of the two estimates.}\label{fig:plot1-100-sim-compare}
+\end{figure}
 
-ggplot(data, aes(x = pfilter, y = palfilter)) +
-geom_point(shape = 20, size = 0.3) +  # pch=20, cex=0.3 equivalent
-geom_abline(intercept = 0, slope = 1, color = "red") +
-labs(
-  title=NULL,
-  x = "PF",
-  y = "PAL"
-) +
-theme_minimal()
-@
+\end{knitrout}
 
@@
 For this simulation study, the particle filter out-performs PAL (Figure~\ref{fig:plot1-100-sim-compare}) which supports hypothesis 2, above.
-On average, the particle filter likelihood estimate is $\Sexpr{myround(mean(pfilter_ovov_sim_100)-mean(palfilter_ovov_sim_100),1)}$ log units higher than the PAL.
+On average, the particle filter likelihood estimate is $-1027.3$ log units higher than the PAL.
 Consequently, we looked for model misspecification by comparing the conditional log-likelihood estimates at each time point, as described in the Methods section.

### 2024-08-31 ms.Rnw
table minor modification
runlevel 3

@@
 
-RUN_LEVEL = 1
-NP_MIF       = switch(RUN_LEVEL, 100,   5000,  50000)
+RUN_LEVEL = 3
+NP_MIF       = switch(RUN_LEVEL, 10,   5000,  50000)
 NMIF         = switch(RUN_LEVEL, 5,    20,    100)
@@
       \(\sigma_\xi\) & 66.89     & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$ & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$\\
-      \(S_{10}\)     & 3876549   & $\Sexpr{round(pop*est_mle_vec["S10"],0)}$  & $\Sexpr{round(pop*est_mle_vec_sbh["S10"],0)}$\\
-      \(I_{10}\)     & 30351     & $\Sexpr{round(pop*est_mle_vec["I10"],0)}$  & $\Sexpr{round(pop*est_mle_vec_sbh["I10"],0)}$\\
-      \(R_{10}\)     & 1315221   & $\Sexpr{round(pop*est_mle_vec["R10"],0)}$  & $\Sexpr{round(pop*est_mle_vec_sbh["R10"],0)}$\\
-      \(S_{20}\)     & 57139612  & $\Sexpr{round(pop*est_mle_vec["S20"],0)}$  & $\Sexpr{round(pop*est_mle_vec_sbh["S20"],0)}$\\
-      \(I_{20}\)     & 871       & $\Sexpr{round(pop*est_mle_vec["I20"],0)}$  & $\Sexpr{round(pop*est_mle_vec_sbh["I20"],0)}$\\
-      \(R_{20}\)     & 302852    & $\Sexpr{round(pop*est_mle_vec["R20"],0)}$  & $\Sexpr{round(pop*est_mle_vec_sbh["R20"],0)}$\\
-      \(S_{30}\)     & 19573727  & $\Sexpr{round(pop*est_mle_vec["S30"],0)}$  & $\Sexpr{round(pop*est_mle_vec_sbh["S30"],0)}$\\
-      \(I_{30}\)     & 2550      & $\Sexpr{round(pop*est_mle_vec["I30"],0)}$  & $\Sexpr{round(pop*est_mle_vec_sbh["I30"],0)}$\\
-      \(R_{30}\)     & 131092    & $\Sexpr{round(pop*est_mle_vec["R30"],0)}$  & $\Sexpr{round(pop*est_mle_vec_sbh["R30"],0)}$\\
+      \(S_{10}\)     & 3876549   & $\Sexpr{round(pop*est_mle_vec["S10"],0)}$  & $5094840$\\
+      \(I_{10}\)     & 30351     & $\Sexpr{round(pop*est_mle_vec["I10"],0)}$  & $2871$\\
+      \(R_{10}\)     & 1315221   & $\Sexpr{round(pop*est_mle_vec["R10"],0)}$  & $124410$\\
+      \(S_{20}\)     & 57139612  & $\Sexpr{round(pop*est_mle_vec["S20"],0)}$  & $57385600$\\
+      \(I_{20}\)     & 871       & $\Sexpr{round(pop*est_mle_vec["I20"],0)}$  & $639$\\
+      \(R_{20}\)     & 302852    & $\Sexpr{round(pop*est_mle_vec["R20"],0)}$  & $57072$\\
+      \(S_{30}\)     & 19573727  & $\Sexpr{round(pop*est_mle_vec["S30"],0)}$  & $19697600$\\
+      \(I_{30}\)     & 2550      & $\Sexpr{round(pop*est_mle_vec["I30"],0)}$  & $174$\\
+      \(R_{30}\)     & 131092    & $\Sexpr{round(pop*est_mle_vec["R30"],0)}$  & $9578$\\
       \midrule

### 2024-08-31 ms.Rnw
minor mod

@@
 
-RUN_LEVEL = 3
+RUN_LEVEL = 2
 NP_MIF       = switch(RUN_LEVEL, 10,   5000,  50000)

### 2024-08-31 ms.Rnw
more editing

@@
   tidy = TRUE,
-  cache = TRUE,
-#  cache = FALSE,  # set cache = FALSE when editing the code
+# cache = TRUE,
+  cache = FALSE,  # set cache = FALSE when editing the code
 # message = FALSE,

### 2024-09-03 ms.Rnw
modify with sbh init - add NA

--- 2024-08-31 ms.Rnw
@@
   paramnames = names(params_full)
-)
-
-sir_sbh_init <- pomp(
-  sir_wwr,
-  t0 = 1-6*52,
-  rinit = rinit_stocks,
-  statenames = statenames
 )
@@
 set.seed(123)
+
+dat %>%
+  rbind(data.frame(time=0,cases1=NA,cases2=NA,cases3=NA)) %>%
+  arrange(time) -> dat_sbh
+
+sir_sbh_init <- pomp(
+  sir_wwr,
+  data = dat_sbh,
+  times = "time",
+  t0 = 1-6*52,
+  rinit = rinit_stocks,
+  statenames = statenames
+)
+
 sir_box_sbh <- list(
@@
 }) -> sbh_el_01
-
 @

### 2024-09-04 ms.Rnw
minor correction

--- 2024-09-03 ms.Rnw
@@
 dmeas <- Csnippet("
+  if(ISNA(cases1)) {
+    lik = (give_log) ? 0 : 1;
+  } else {
       lik =  dbinom(cases1, H1, q1, 1) +
       dbinom(cases2, H2, q2, 1) +
-  dbinom(cases3, H3, q3, 1);"
-)
+      dbinom(cases3, H3, q3, 1);
+    }
+  lik = (give_log) ? lik : exp(lik);"
+)
+
 rmeas <-  Csnippet("
@@
 sir_sbh_init <- pomp(
-  sir_wwr,
   data = dat_sbh,
@@
   t0 = 1-6*52,
+  dmeasure = dmeas,
+  rmeasure = rmeas,
+  rprocess = discrete_time(step.fun = rproc, delta.t = 1/4),
+  statenames = statenames,
+  paramnames = names(params_wwr),
+  accumvars = c("H1", "H2", "H3"),
   rinit = rinit_stocks,
-  statenames = statenames
+  partrans = pt_wwr,
+  params = params_wwr
 )

### 2024-09-10 ms.Rnw
another 2 rounds

@@
 
+<<maximization4-Stocks_rinit, echo=F, warning=F, message=F>>=
+set.seed(12345)
+sd_04 = (2/3)^3
+starting_values_sbh_04 <- choose_n_fits(sbh_el_03, top_n_fits, ncores)
+
+bake(file = paste0(out_dir, "mif2_maximization/SBH_init/round_04/mifs_global_sbh_04.rds"),{ 
+  plan(multicore, workers = ncores)
+  foreach(i = 1:ncores, .options.future = list(seed = 652643293)) %dofuture% {
+    pomp::mif2(
+      mifs_global_sbh_03[[1]],
+      rw.sd = rw_sd(
+        beta1 = 0.01*sd_04, beta2 = 0.01*sd_04, beta3 = 0.01*sd_04,
+        beta11 = 0.01*sd_04, phi = 0.01*sd_04,
+        sigma_q = 0.01*sd_04, sigma_xi = 0.01*sd_04
+      ),
+      params = starting_values_sbh_04[[i]]
+    ) 
+  }
+}) -> mifs_global_sbh_04
+
+bake(file = paste0(out_dir,"mif2_maximization/SBH_init/round_04/sbh_el_04.rds"),{
+  eval_logLik_pomp(mifs_global_sbh_04, ncores, NP_MIF, seed = 4248930)
+}) ->  sbh_el_04
+@
+
+<<maximization5-Stocks_rinit, echo=F, warning=F, message=F>>=
+set.seed(12345)
+sd_05 = (2/3)^4
+starting_values_sbh_05 <- choose_n_fits(sbh_el_04, top_n_fits, ncores)
+
+bake(file = paste0(out_dir, "mif2_maximization/SBH_init/round_05/mifs_global_sbh_05.rds"),{ 
+  plan(multicore, workers = ncores)
+  foreach(i = 1:ncores, .options.future = list(seed = 652643293)) %dofuture% {
+    pomp::mif2(
+      mifs_global_sbh_04[[1]],
+      rw.sd = rw_sd(
+        beta1 = 0.01*sd_05, beta2 = 0.01*sd_05, beta3 = 0.01*sd_05,
+        beta11 = 0.01*sd_05, phi = 0.01*sd_05,
+        sigma_q = 0.01*sd_05, sigma_xi = 0.01*sd_05
+      ),
+      params = starting_values_sbh_05[[i]]
+    ) 
+  }
+}) -> mifs_global_sbh_05
+
+bake(file = paste0(out_dir,"mif2_maximization/SBH_init/round_05/sbh_el_05.rds"),{
+  eval_logLik_pomp(mifs_global_sbh_05, ncores, NP_MIF, seed = 4248930)
+}) ->  sbh_el_05
+@
+
 <<estimated-mle_stocks_init, echo=FALSE, message=FALSE, warning=FALSE>>=
 # Can be changed if more rounds needed
-est_mle_vec_sbh <- sbh_el_03[which.max(sbh_el_03$logLik),]
+est_mle_vec_sbh <- sbh_el_05[which.max(sbh_el_05$logLik),]
 @
@@
 WE DON'T KNOW IN ADVANCE EXACTLY HOW SUCCESSFUL IT WOULD BE...}
+\yize{ADD ANOTHER ROUND OF MIF WITH SBH INITIALIZATION \\ PLOT THE COND LIK AT NEW EST \\ MAXIMIZE/ FILTER WITH PAL ON THEIR MLE. I DOUBT THEIR RESULTS CAN BE EXACTLY REPRODUCED.}
 

### 2024-09-12 ms.Rnw
another two rounds

@@
 
+<<maximization6-Stocks_rinit, echo=F, warning=F, message=F>>=
+set.seed(12345)
+sd_06 = (2/3)^5
+starting_values_sbh_06 <- choose_n_fits(sbh_el_05, top_n_fits, ncores)
+
+bake(file = paste0(out_dir, "mif2_maximization/SBH_init/round_06/mifs_global_sbh_06.rds"),{ 
+  plan(multicore, workers = ncores)
+  foreach(i = 1:ncores, .options.future = list(seed = 652643293)) %dofuture% {
+    pomp::mif2(
+      mifs_global_sbh_05[[1]],
+      rw.sd = rw_sd(
+        beta1 = 0.01*sd_06, beta2 = 0.01*sd_06, beta3 = 0.01*sd_06,
+        beta11 = 0.01*sd_06, phi = 0.01*sd_06,
+        sigma_q = 0.01*sd_06, sigma_xi = 0.01*sd_06
+      ),
+      params = starting_values_sbh_06[[i]]
+    ) 
+  }
+}) -> mifs_global_sbh_06
+
+bake(file = paste0(out_dir,"mif2_maximization/SBH_init/round_06/sbh_el_06.rds"),{
+  eval_logLik_pomp(mifs_global_sbh_06, ncores, NP_MIF, seed = 4248930)
+}) ->  sbh_el_06
+@
+
+<<maximization7-Stocks_rinit, echo=F, warning=F, message=F>>=
+set.seed(12345)
+sd_07 = (2/3)^6
+starting_values_sbh_07 <- choose_n_fits(sbh_el_06, top_n_fits, ncores)
+
+bake(file = paste0(out_dir, "mif2_maximization/SBH_init/round_07/mifs_global_sbh_07.rds"),{ 
+  plan(multicore, workers = ncores)
+  foreach(i = 1:ncores, .options.future = list(seed = 652643293)) %dofuture% {
+    pomp::mif2(
+      mifs_global_sbh_06[[1]],
+      rw.sd = rw_sd(
+        beta1 = 0.01*sd_07, beta2 = 0.01*sd_07, beta3 = 0.01*sd_07,
+        beta11 = 0.01*sd_07, phi = 0.01*sd_07,
+        sigma_q = 0.01*sd_07, sigma_xi = 0.01*sd_07
+      ),
+      params = starting_values_sbh_07[[i]]
+    ) 
+  }
+}) -> mifs_global_sbh_07
+
+bake(file = paste0(out_dir,"mif2_maximization/SBH_init/round_07/sbh_el_07.rds"),{
+  eval_logLik_pomp(mifs_global_sbh_07, ncores, NP_MIF, seed = 4248930)
+}) ->  sbh_el_07
+@
+
 <<estimated-mle_stocks_init, echo=FALSE, message=FALSE, warning=FALSE>>=
 # Can be changed if more rounds needed
-est_mle_vec_sbh <- sbh_el_05[which.max(sbh_el_05$logLik),]
+est_mle_vec_sbh <- sbh_el_07[which.max(sbh_el_07$logLik),]
 @
@@
       \(\sigma^2_q\) & 0.021     & $\Sexpr{round(est_mle_vec["sigma_q"],3)}$ & $\Sexpr{round(est_mle_vec_sbh["sigma_q"],3)}$\\
-      \(\sigma_\xi\) & 66.89     & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$ & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$\\
+      \(\sigma_\xi\) & 66.89     & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["sigma_xi"],2)}$\\
       \(S_{10}\)     & 3876549   & $\Sexpr{round(pop*est_mle_vec["S10"],0)}$  & $5094840$\\

### 2024-09-12 ms.Rnw
edits incorporating AIC for WWR

--- 2024-09-10 ms.Rnw
@@
 On simulations where the model is correctly specified, the particle filter outperforms PAL.
-If a poorly fitting specification of initial value parameters is amended, the particle filter also outperforms PAL on the actual data.
+
 \end{abstract}
@@
   # taken from the broman package
-  if (digits < 1)
-    stop("This is intended for the case digits >= 1.")
   if (length(digits) > 1) {
@@
 
-RUN_LEVEL = 3
+RUN_LEVEL = 1
 NP_MIF       = switch(RUN_LEVEL, 10,   5000,  50000)
@@
 WWR were given the opportunity to submit a correction, after we shared the results of our investigation with them, but they declined.
-The theory developed by WWR shows that PAL has some potentially useful scaling properties, but the numerical results in Table~1 appear to show much stronger performance than a standard particle filter on an example of scientific interest but moderate size.
-Our task here is to correct the error in Table~1 so that researchers considering whether to implement Poisson approximate likelihood (PAL) are appropriately informed about its benefits.
+The theory developed by WWR shows that PAL has some potentially useful scaling properties, but the numerical results in their Table~1 appear to show much stronger performance than a standard particle filter on an example of scientific interest but moderate size.
+We present a correction of this error so that researchers considering whether to implement Poisson approximate likelihood (PAL) are appropriately informed about its benefits.
 
@@
 SBH found strong evidence for the importance of overdispersion in a stochastic dynamic model for their epidemiological data.
-This is significant because earlier research on population dynamics largely avoided consideration of overdispersion, perhaps because of the lack of avaialable statistical methodology to fit such models.
+This is significant because most earlier research on population dynamics avoided consideration of overdispersion, perhaps due to the lack of available statistical methodology for fitting overdispersed nonlinear stochastic dynamic models.
 The conclusions of SBH hinge on a comparison of likelihoods, and so the results of WWR discredit those conclusions by indicating that SBH based their reasoning on inaccurately computed likelihoods.
@@
 SBH and WWR each fitted three different rotavirus models.
-The first has equidispersion (i.e., no overdispersion) in the measurement model and the dynamic model, and is called EqEq.
+The first has equidispersion (i.e., no overdispersion) in the measurement model and the dynamic model, and is called EqEq by WWR.
 The second, EqOv, includes overdispersion in only the measurement model.
@@
 
+<<benchmark,echo=FALSE>>=
+arma_benchmark <- function(x,order=c(2,0,1)){
+  arima(log(x + 1), order = order)$loglik - sum(log(x+1))
+}
+
+sum(apply(realdat,2, function(x)  arima(log(x + 1), order =c(2,0,1))$aic + 2 * sum(log(x+1))))
+
+sbh_data <- read.csv("sbh_data.txt",sep=" ",skip=3)[,-1]
+sbh_benchmark <- sum(apply(sbh_data,2, function(x)
+  arima(log(x + 1), order =c(2,0,1))$aic + 2 * sum(log(x+1))
+))
+
+wwr_benchmark <- sum(apply(realdat,2, function(x)
+  arima(log(x + 1), order =c(2,0,1))$aic + 2 * sum(log(x+1))
+))
+
+tab_signif <- 0
+@
+
 \begin{table}[ht] % The placement specifier can be [h!tbp]
@@
 PAL is the Poisson approximate likelihood, implemented using the code of WWR.
-PF is the particle filter, implemented using the code of \citet{pomppackagepaper}.
+PF is the particle filter, implemented using the pomp R package \citet{pomppackagepaper}.
 The first two lines are from WWR, Table~1, and lines 3--5 are our own computations.
@@
 PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL.
-\ed{SHOULD WE HAVE A LINE FOR PAL AT THE IF2 MLE FROM LINE 5? PERHAPS WE CAN'T READILY MAXIMIZE PAL FOR THIS MODEL (OR CAN WE?) BUT SOME SORT OF COMPARISON WOULD BE HELPFUL}
+PF was maximized using iterated filtering, and PAL was maximized using coordinate gradient descent.
 } 
 \label{tab:ovovrealdata} % For referencing this table elsewhere in your text
-\begin{tabular}{lllc} 
+\begin{tabular}{lllr} 
   \hline
@@
   \hline
-  1. & PAL & Table~1 of WWR & 13778.08
+  1. & PF & Table~1 of WWR, originally from SBH &  \Sexpr{myround(20134.38,tab_signif)}
   \\
-  2. & PF & Table~1 of WWR, originally from SBH &  20134.38
+  2. & PAL & Table~1 of WWR & \Sexpr{myround(13778.08,tab_signif)}
   \\
-  3. & PAL & Model, data and MLE from WWR &
-    \Sexpr{myround(-2*palfilter_lik_list_wwr$log_lik + 2*7 ,2)} 
+  3. & PAL & Data, model and parameters from WWR &
+    \Sexpr{myround(-2*palfilter_lik_list_wwr$log_lik + 2*7 ,tab_signif)} 
   \\
-  4. & PF & Model, data and MLE from WWR &
-    \Sexpr{myround(-2*logmeanexp(logLik(pfilter_wwr)+ 2*7),2)}
+  4. & PF & Data, model and parameters matching 3. &
+    \Sexpr{myround(-2*logmeanexp(logLik(pfilter_wwr)+ 2*7),tab_signif)}
   \\
-  5. & PF & Data from WWR, estimated initial values &
-     \Sexpr{myround(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}
+  5. & PF & Changing initial values &
+     \Sexpr{myround(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),tab_signif)}
   \\
-  6. & PAL & Data from WWR, estimated initial values &
-     \Sexpr{myround(-2*palfilter_mif2_mle$log_lik + 2*7 ,2)}
+  6. & PAL & Data, model and parameters matching 5.  &
+     \Sexpr{myround(-2*palfilter_mif2_mle$log_lik + 2*7 ,tab_signif)}
+  \\   
+  7. & ARMA & Sec.~7.3.2 of WWR & 23043
+  \\
+  8. & ARMA & Data from WWR &
+    \Sexpr{myround(wwr_benchmark,tab_signif)} 
   \\
@@
 We continued our investigation to establish the cause of this.
-We discovered that the initial conditions for the latent process in January 2001 (their time 0) were estimated by SBH. However, in WWR, they were assumed fixed and made the model fitted poorly.
-
-<<plot2-cond-loglik, echo=FALSE, fig.height=4, fig.width=5.5, out.width="4.5in",fig.align="center", fig.cap='Conditional log-likelihoods computed using PAL (red) and PF (blue) for the rotavirus model and MLE from WWR.
+
+Inspection of log-likelihood anomalies \citep{wheeler24,li24,hao24} showed that the initial conditions for the latent process in January 2001 were fixed at values which were incompatible with the trajectory of the data early in the time series.
+By contrast, SBH fixed the initial conditions 6 years before the first measurement, giving time for the system to reach its equilibrum distribution.
+Line 5 of Table~\ref{tab:ovovrealdata} shows that this improvement enables PF to reach close to the values attained by PAL.
+Further, at the MLE obtained by maximizing PF, we found that PF beats PAL (line 6).
+
+WWR compare their fitted models with a log-ARMA(2,1) benchmark AIC value of 23043 (line 7), inferring that all the mechanistic models possessing overdispersion have better statistical fit than a simple loglinear time series model.
+However, this value corresponds to a log-ARMA(2,1) model fitted not to WWR's data but to SBH's data.
+Thus, line 7 can properly compared to line 1 but not to any other line of  Table~\ref{tab:ovovrealdata}. 
+We independently fitted a log-ARMA(2,1) model to the SBH data and obtained a similar AIC value (\Sexpr{myround(sbh_benchmark,tab_signif)}).
+Carrying out the same computation for the data fitted by WWR gives a log-ARMA benchmark AIC value of \Sexpr{myround(wwr_benchmark,tab_signif)} (line 8).
+Thus, the statistical fit of the mechanistic model considered by WWR is inferior to a simple log-ARMA model.
+This holds for all the variants in lines 2--6 of  Table~\ref{tab:ovovrealdata}, regardless of whether PF or PAL is used.
+The goal of mechanistic modeling is not necessarily to beat a simple statistical benchmark, but this can be an indication that additional model development could be worthwhile \citep{wheeler24}.
+A correct interpretation of Table~\ref{tab:ovovrealdata} is therefore very different to the conclusions drawn by WWR from comparing line 2 inappropriately to lines 1 and 7.
+
+<<plot1-100-sim-compare,echo=FALSE, fig.height=4, fig.width=5.5, out.width="4.5in",fig.align="center",fig.cap='Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model.
+Simulations with one or more zero counts were disqualified since they resulted in errors for the PAL implementation.
 We used 50000 particle for both PF and PAL.
-PF averaged over 36 replications to reduce the Monte Carlo variance, but this step was not necessary for PAL.
-The main sources of likelihood shortfall of PF are at time points $t=1, 2, 3, 11, 81, 194$, and $325-333$.'>>=
-
-# Combine the data into a single dataframe for SMC with grouping variable
-data_list <- lapply(1:ncores, function(i) {
-  data.frame(
-    time = time, 
-    cond_loglik = pfilter_wwr_cond_lik[[i]], 
-    method = "SMC",
-    group = i
-  )
-})
-
-# Combine all the SMC data frames into one
-df_smc <- bind_rows(data_list)
-
-# Create the PAL data frame
-df_pal <- data.frame(
-  time = time, 
-  cond_loglik = palfilter_lik_list_wwr$ll_storage, 
-  method = "PAL",
-  group = 1  # Single group for PAL
-)
-
-# Combine both data frames
-df <- bind_rows(df_smc, df_pal)
-
-# Plot with ggplot
-ggplot(df, aes(x = time, y = cond_loglik, color = method)) +
-  geom_line(data = df %>% filter(method == "SMC"), aes(group = group), color = "blue", alpha = 0.7) +
-  geom_line(data = df %>% filter(method == "PAL"), aes(group = group), color = "red") +
-  theme_minimal() +
+PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL.
+The red line corresponds to equality of the two estimates.'>>=
+data <- data.frame(
+  pfilter = pfilter_ovov_sim_100, 
+  palfilter = palfilter_ovov_sim_100
+)
+
+ggplot(data, aes(x = pfilter, y = palfilter)) +
+geom_point(shape = 20, size = 0.3) +  # pch=20, cex=0.3 equivalent
+geom_abline(intercept = 0, slope = 1, color = "red") +
 labs(
-    title = NULL,
-    x = "Time (weeks)",
-    y = "Conditional Log-likelihood",
-    color = "Method"
+  title=NULL,
+  x = "PF",
+  y = "PAL"
 ) +
-  theme(legend.position = "bottom")
-@
+theme_minimal()
+@
+
+In the presence of model misspecification, it becomes harder to compare likelihood evaluation methods.
+A likelihood approximation, such as PAL, may potentially obtain a higher value than the exact likelihood if it compensates for the misspecification.
+Log-likelihood is a proper scoring rule for forecasts \cite{gneiting2}, and both the particle filter and PAL construct their log-likelihood estimates via a sequence of one-step forecasts.
+Therefore, if the model is correctly specified, the approximation error in PAL can only decrease the expected log-likelihood.
+We tested this on simulated data for which the model of WWR is correctly specified.
+For this simulation study, the particle filter out-performs PAL (Figure~\ref{fig:plot1-100-sim-compare}).
+On average, the particle filter likelihood estimate is $\Sexpr{myround(mean(pfilter_ovov_sim_100)-mean(palfilter_ovov_sim_100),1)}$ log units higher than the PAL.
+We know from the benchmark AIC value in Table~\ref{tab:ovovrealdata}. line 8, that there is substantial model misspecification.
+Therefore, this real-data example may not be well suited to comparing the filtering skill of PAL and PF.
+
+We conclude that PAL is a potentially useful algorithm, with some favorable theoretical properties.
+However, the corrected evidence does not indicate an advantage for using PAL in situations where the particle filter is effective.
+
+An extended description of our reanalysis of WWR is provided by \cite{hao24}.
+That thesis also contains some additional results reinforcing the conclusions presented in this article.
+The source code for this article is available at \url{https://github.com:ionides/pal-vs-pf} and archived at \ed{ZENODO - TODO}.
+
+\bibliography{bib-pal}
+
+\clearpage
+
+\end{document}
+
+\section{Extra material excluded in the final version}
+
+%%%%%%%%% 222222222 %%%%%%%%%%
+
+\section*{Methods}
+
+We first replicated the model of WWR in the R package \texttt{pomp} \citep{pomppackagepaper}.
+Comparing summary statistics of simulations confirms that both model versions are essentially identical, as described by \citet{hao24}.
+We therefore implement PAL using the code from \citet{wwr}, comparing directly with the particle filter implemented using \texttt{pomp}.
+
+This article has focused on likelihood evaluation, but the inferences presented also require the likelihood estimate to be maximized.
+WWR used direct numerical maximization, whereas SBH used iterated filtering to maximize the particle filter likelihood estimate implemented using the \texttt{mif2} algorithm in \texttt{pomp}.
+We follow the approaches adopted by these two papers.
+The optimization method is not directly pertinent to the comparison of the filtering methods, so we do not discuss optimization in further detail here.
 
@@
 
-<<plot1-100-sim-compare,echo=FALSE, fig.height=4, fig.width=5.5, out.width="4.5in",fig.align="center",fig.cap='Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model.
-Simulations with one or more zero counts were disqualified since they resulted in errors for the PAL implementation.
+<<plot2-cond-loglik, echo=FALSE, fig.height=4, fig.width=5.5, out.width="4.5in",fig.align="center", fig.cap='Conditional log-likelihoods computed using PAL (red) and PF (blue) for the rotavirus model and MLE from WWR.
 We used 50000 particle for both PF and PAL.
-PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL.
-The red line corresponds to equality of the two estimates.'>>=
-data <- data.frame(
-  pfilter = pfilter_ovov_sim_100, 
-  palfilter = palfilter_ovov_sim_100
-)
-
-ggplot(data, aes(x = pfilter, y = palfilter)) +
-geom_point(shape = 20, size = 0.3) +  # pch=20, cex=0.3 equivalent
-geom_abline(intercept = 0, slope = 1, color = "red") +
+PF averaged over 36 replications to reduce the Monte Carlo variance, but this step was not necessary for PAL.
+The main sources of likelihood shortfall of PF are at time points $t=1, 2, 3, 11, 81, 194$, and $325-333$.'>>=
+
+# Combine the data into a single dataframe for SMC with grouping variable
+data_list <- lapply(1:ncores, function(i) {
+  data.frame(
+    time = time, 
+    cond_loglik = pfilter_wwr_cond_lik[[i]], 
+    method = "SMC",
+    group = i
+  )
+})
+
+# Combine all the SMC data frames into one
+df_smc <- bind_rows(data_list)
+
+# Create the PAL data frame
+df_pal <- data.frame(
+  time = time, 
+  cond_loglik = palfilter_lik_list_wwr$ll_storage, 
+  method = "PAL",
+  group = 1  # Single group for PAL
+)
+
+# Combine both data frames
+df <- bind_rows(df_smc, df_pal)
+
+# Plot with ggplot
+ggplot(df, aes(x = time, y = cond_loglik, color = method)) +
+  geom_line(data = df %>% filter(method == "SMC"), aes(group = group), color = "blue", alpha = 0.7) +
+  geom_line(data = df %>% filter(method == "PAL"), aes(group = group), color = "red") +
+  theme_minimal() +
   labs(
-  title=NULL,
-  x = "PF",
-  y = "PAL"
+    title = NULL,
+    x = "Time (weeks)",
+    y = "Conditional Log-likelihood",
+    color = "Method"
   ) +
-theme_minimal()
+  theme(legend.position = "bottom")
 @
@@
 If we instead estimate the initial values, we obtain the results in Table~\ref{tab:mlebywwr}.
-
-\ed{SINCE THE MEAN DURATION OF IMMUNITY IS 1 YR, HAVING $31\times 10^6$ in $R_{2,0}$ IMPLIES THAT ABOUT HALF OF ALL SUSCEPTIBLE INDIVIDUALS GET ROTAVIRUS EVERY YEAR ACCORDING TO THE MODEL.
-IS THAT THE CASE FOR THE MODEL?
-FOR THE DATA, THE PER-CAPITA RATE IN THIS AGE GROUP IS RATER LOW.
-THE SIMPLEST WAY TO AVOID THIS PROBLEM WOULD BE TO DO WHAT SBH DID (DESCRIBED IN THEIR ONLINE APPENDIX C) - EFFECTIVELY, DRAWING $X_0$ FROM THE STATIONARY DISTRIBUTION GIVEN THE MODEL.
-THIS HAS THE ADVANTAGE OF INCLUDING NO NEW PARAMETERS AND BEING CONSISTENT WITH SBH.
-WE DON'T KNOW IN ADVANCE EXACTLY HOW SUCCESSFUL IT WOULD BE...}
-\yize{ADD ANOTHER ROUND OF MIF WITH SBH INITIALIZATION \\ PLOT THE COND LIK AT NEW EST \\ MAXIMIZE/ FILTER WITH PAL ON THEIR MLE. I DOUBT THEIR RESULTS CAN BE EXACTLY REPRODUCED.}
-
-\section*{Discussion}
-
-PAL provides an approximate filter, whereas the particle filter provides a Monte~Carlo estimate of an exact filter.
-Possible explanations for PAL obtaining a higher log-likelihood than a particle filter include:
-\begin{enumerate}
-\item The particle filter could have high Monte Carlo variance.
-This results in negative bias in its log-likilihood estimate due to Jensen's inequality, since the particle filter provides an unbiased likelihood estimate.
-\item If the model is misspecified, the approximation error in PAL could lead to a higher log-likelihood estimate than that of the actual misspecified model.
-\end{enumerate}
-Log-likelihood is a proper scoring rule for forecasts \cite{gneiting2}, and both the particle filter and PAL construct their log-likelihood estimates via a sequence of one-step forecasts.
-Therefore, if the model is correctly specified, the approximation error in PAL can only decrease the expected log-likelihood.
-We tested this on simulated data, for which the model of \cite{wwr} is correctly specified.
-For this simulation study, the particle filter out-performs PAL (Figure~\ref{fig:plot1-100-sim-compare}) which supports hypothesis 2, above.
-On average, the particle filter likelihood estimate is $\Sexpr{myround(mean(pfilter_ovov_sim_100)-mean(palfilter_ovov_sim_100),1)}$ log units higher than the PAL.
-Consequently, we looked for model misspecification by comparing the conditional log-likelihood estimates at each time point, as described in the Methods section.
-
-We conclude that PAL is a potentially useful algorithm, with some favorable theoretical properties.
-However, the corrected evidence does not indicate an advantage for using PAL in situations where the particle filter is effective.
-
-%%%%%%%%% 222222222 %%%%%%%%%%
-
-\section*{Methods}
-
-We first replicated the model of WWR in the R package \texttt{pomp} \citep{pomppackagepaper}.
-Comparing summary statistics of simulations confirms that both model versions are essentially identical, as described by \citet{hao24}.
-We therefore implement PAL using the code from \citet{wwr}, comparing directly with the particle filter implemented using \texttt{pomp}.
-
-This article has focused on likelihood evaluation, but the inferences presented also require the likelihood estimate to be maximized.
-WWR used direct numerical maximization, whereas SBH used iterated filtering to maximize the particle filter likelihood estimate implemented using the \texttt{mif2} algorithm in \texttt{pomp}.
-We follow the approaches adopted by these two papers.
-The optimization method is not directly pertinent to the comparison of the filtering methods, so we do not discuss optimization in further detail here.
-An extended description of our reanalysis of WWR is provided by \cite{hao24}.
-That thesis also contains some additional results reinforcing the conclusions presented in this article.
-The source code for this article is available at \url{https://github.com:ionides/pal-vs-pf} and archived at \ed{ZENODO - TODO}.
-
-\bibliography{bib-pal}
-
-\end{document}

### 2024-09-12 choose_n_fits.R
measlespkg attribution

@@ -1,16 +1,16 @@
-#' Choose top n best results and make them to be the starting value for the next round.
+#' Choose top n best results and make them to be the starting values for the next round.
 #'
 #' @param el Estimated values from `pfiler`.
-#' @param top_n_fits Number of best fits in terms of log-likelihood one wants to keep.
+#' @param top_n_fits Number of the best fits in terms of the log-likelihood one wants to keep.
 #' @param ncores Number of starting values possibly equal to the number of cores being used.
 #'
-#' @return A matrix with ncores number of rows and estimated parameters being in each row.
+#' @return A matrix with ncores rows and estimated parameter vector in each row.
 #' @export
 #'
 #' @source <https://github.com/AJAbkemeier/measlespkg/tree/main/R>
 #'
 #' @examples
-#' choose_n_fits(el, 12, 36)
+#' choose_n_fits(el, top_n_fits=12, ncores=36)
 #' 
 
 choose_n_fits = function(el, top_n_fits, ncores){

### 2024-09-12 eval_logLik_pomp.R

@@ -2,16 +2,16 @@
 #'
 #' @param model_obj_list List of pomp results.
 #' @param nreps Number of replications for each pfilter. 
-#' @param np_pf Number of particles in pfilter.
+#' @param np_pf Number of particles in the pfilter.
 #' @param seed  Set seed.
 #' 
-#' @return A matrix with ncores number of rows and estimated parameters being in each row.
+#' @return A matrix with ncores rows and estimated parameter vector in each row.
 #' @export
 #'
 #' @source <https://github.com/AJAbkemeier/measlespkg/tree/main/R>
 #'
 #' @examples
-#' eval_logLik_pomp(model_obj_list, 36, 1000, 123)
+#' eval_logLik_pomp(model_obj_list, nreps=36, np_pf=1000, seed=123)
 #' 
 eval_logLik_pomp = function(model_obj_list, nreps, np_pf, seed){
   pf_logLik_frame = data.frame(

### 2024-09-14 ms.Rnw

@@
 
+\begin{table}[htbp]
+    \caption{Maximum likelihood estimation of parameters for the OvOv model of WWR by PAL (taken from WWR), and by iterated particle filtering implemented using {\rm \texttt{mif2}} in {\rm \texttt{pomp}} \citep{pomppackagepaper}.
+The initial distribution parameter \(\lambda_0 = (S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}, R_{30})\) are assumed to be fixed in \cite{wwr}.
+The iterated particle filtering estimates were obtained by using 3 rounds of refinement, with 100 filtering iterations, 50,000 particles, and 36 replicates in each round with the top 12 best fits in terms of likelihood chosen to be the starting value for the next round.}
+    \label{tab:mlebywwr}
+\begin{center}    
+    \begin{tabular}{lccc}  % Left-aligned for the first column, centered for the other two columns
+      \toprule
+      Parameter      & PAL      & iterated filtering & iterated filtering (stocks)\\
+      \midrule
+      \(\beta_1\)    & 11.48     & $\Sexpr{round(est_mle_vec["beta1"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta1"],2)}$\\
+      \(\beta_2\)    & 0.25      & $\Sexpr{round(est_mle_vec["beta2"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta2"],2)}$\\
+      \(\beta_3\)    & 0.35      & $\Sexpr{round(est_mle_vec["beta3"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta3"],2)}$\\
+      \(\phi\)       & 0.14      & $\Sexpr{round(est_mle_vec["phi"],2)}$   & $\Sexpr{round(est_mle_vec_sbh["phi"],2)}$\\
+      \(\rho\)       & 0.16      & $\Sexpr{round(est_mle_vec["beta11"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta11"],2)}$\\
+      \(\sigma^2_q\) & 0.021     & $\Sexpr{round(est_mle_vec["sigma_q"],3)}$ & $\Sexpr{round(est_mle_vec_sbh["sigma_q"],3)}$\\
+      \(\sigma_\xi\) & 66.89     & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["sigma_xi"],2)}$\\
+      \(S_{10}\)     & 3876549   & $\Sexpr{round(pop*est_mle_vec["S10"],0)}$  & $5094840$\\
+      \(I_{10}\)     & 30351     & $\Sexpr{round(pop*est_mle_vec["I10"],0)}$  & $2871$\\
+      \(R_{10}\)     & 1315221   & $\Sexpr{round(pop*est_mle_vec["R10"],0)}$  & $124410$\\
+      \(S_{20}\)     & 57139612  & $\Sexpr{round(pop*est_mle_vec["S20"],0)}$  & $57385600$\\
+      \(I_{20}\)     & 871       & $\Sexpr{round(pop*est_mle_vec["I20"],0)}$  & $639$\\
+      \(R_{20}\)     & 302852    & $\Sexpr{round(pop*est_mle_vec["R20"],0)}$  & $57072$\\
+      \(S_{30}\)     & 19573727  & $\Sexpr{round(pop*est_mle_vec["S30"],0)}$  & $19697600$\\
+      \(I_{30}\)     & 2550      & $\Sexpr{round(pop*est_mle_vec["I30"],0)}$  & $174$\\
+      \(R_{30}\)     & 131092    & $\Sexpr{round(pop*est_mle_vec["R30"],0)}$  & $9578$\\
+      \midrule
+      \textbf{AIC}  & 13778.08  & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$ & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec_sbh["logLik"]),2)}$\\
+      \bottomrule
+    \end{tabular}
+\end{center}
+\end{table}
+
 Figure~\ref{fig:plot2-cond-loglik} indicates that the worst conditional log-likelihood values, estimated using the particle filter, arise at the start of the time series.

### 2024-09-14 ms.Rnw
compile to pdf, sbh_data

@@
 <<benchmark,echo=FALSE>>=
-arma_benchmark <- function(x,order=c(2,0,1)){
-  arima(log(x + 1), order = order)$loglik - sum(log(x+1))
-}
-
-sum(apply(realdat,2, function(x)  arima(log(x + 1), order =c(2,0,1))$aic + 2 * sum(log(x+1))))
-
-sbh_data <- read.csv("sbh_data.txt",sep=" ",skip=3)[,-1]
+set.seed(123)
+
+sbh_data <- read.csv("sbh_data.txt",sep=" ")[,-c(1)]
 sbh_benchmark <- sum(apply(sbh_data,2, function(x)
@@
 
-wwr_benchmark <- sum(apply(realdat,2, function(x)
+wwr_benchmark <- sum(apply(dat[,-1],2, function(x)
   arima(log(x + 1), order =c(2,0,1))$aic + 2 * sum(log(x+1))
@@
   5. & PF & Changing initial values &
-     \Sexpr{myround(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),tab_signif)}
+     \Sexpr{myround(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec_sbh["logLik"]),tab_signif)}
   \\

### 2024-09-14 ms.Rnw
final output

@@
 # Can be changed if more rounds needed
-est_mle_vec_sbh <- sbh_el_07[which.max(sbh_el_07$logLik),]
+est_mle_vec_sbh <- sbh_el_06[which.max(sbh_el_06$logLik),]
 @
@@
   4. & PF & Data, model and parameters matching 3. &
-    \Sexpr{myround(-2*logmeanexp(logLik(pfilter_wwr)+ 2*7),tab_signif)}
+    \Sexpr{myround(-2*logmeanexp(logLik(pfilter_wwr)) + 2*7,tab_signif)}
   \\
-  5. & PF & Changing initial values &
-     \Sexpr{myround(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec_sbh["logLik"]),tab_signif)}
+  5. & PF & Using the initial values of SBH &
+     \Sexpr{myround(-2*est_mle_vec_sbh["logLik"] + 2*7,tab_signif)}
   \\

### 2024-09-15 ms.Rnw
more edits

--- 2024-09-14 ms.Rnw
@@
 Alternatively, the improvement in likelihood found by \cite{wwr} compared to \cite{stocks} might indicate a flaw in the model or numerical methods for that previous work.
-We show that neither of these conclusions is valid, and the comparison of log-likelihood values made by \cite{wwr} is flawed because the PAL calculations were carried out using a dataset scaled differently from the previous study.
-If PAL and the particle filter are used on this new scale, the superficial advantage of PAL largely disappears.
+We show that the comparison of log-likelihood values made by \cite{wwr} is flawed because the PAL calculations were carried out using a dataset scaled differently from the previous study.
+If PAL and the particle filter are applied to the same data, the advantage claimed for PAL disappears.
 On simulations where the model is correctly specified, the particle filter outperforms PAL.
@@
 WWR were given the opportunity to submit a correction, after we shared the results of our investigation with them, but they declined.
-The theory developed by WWR shows that PAL has some potentially useful scaling properties, but the numerical results in their Table~1 appear to show much stronger performance than a standard particle filter on an example of scientific interest but moderate size.
-We present a correction of this error so that researchers considering whether to implement Poisson approximate likelihood (PAL) are appropriately informed about its benefits.
-
-Table~1 of WWR reanalyzes the model and data of \citet{stocks} (henceforth, SBH) for which the likelihood was calculated using a particle filter.
-SBH found strong evidence for the importance of overdispersion in a stochastic dynamic model for their epidemiological data.
+The theory developed by WWR shows that their Poisson Approximate Likelihood (PAL) method has some potentially useful scaling properties.
+This theory is supported by numerical results, in their Table~1, which erroneously claim to show that PAL has substantially stronger performance than a particle filter (PF) on an example of scientific interest but moderate size.
+We present corrected results so that researchers considering whether to implement PAL are appropriately informed about its benefits.
+
+Table~1 of WWR reanalyzes the model and data of \citet{stocks} (henceforth, SBH) for which the likelihood was calculated using PF.
+SBH found clear evidence for the importance of overdispersion in a stochastic dynamic model for their epidemiological data.
 This is significant because most earlier research on population dynamics avoided consideration of overdispersion, perhaps due to the lack of available statistical methodology for fitting overdispersed nonlinear stochastic dynamic models.
@@
 The third, OvOv, includes overdispersion in both these model components.
-We focus on OvOv, which WWR found to be the best fitting model.
-
-We show that most of the apparent advantage for PAL on the OvOv model, compared to a particle filter, arises because WWR used a different scaling of the data from SBH.
+We focus on OvOv, which WWR and SBH both found to be the best fitting model.
+
+We show that most of the apparent advantage for PAL over PF, on the OvOv model, arises because WWR used a different scaling of the data from SBH.
 Two models for the same data can properly be compared by their likelihood, even if the models have entirely different structures.
-One can make allowance for the number of estimated parameters using a quantity such as Akaike's information criterion \citep{Akaike1974ANL}.
-However, if data are rescaled, a correction is required to make likelihoods comparable.
+Allowance for the number of estimated parameters can be made using a quantity such as Akaike's information criterion \citep{Akaike1974ANL}.
+However, if data are rescaled, an adjustment is required to make likelihoods comparable.
 For example, if one model describes a dataset in grams and another describes it in kilograms, then the latter model will earn an increased log-likelihood of $\log(10^3)$ for each data point simply because of the change in scale.
@@
 \begin{table}[ht] % The placement specifier can be [h!tbp]
-\centering % Centers the table
+\centering 
 \caption{AIC for the OvOv rotavirus model, computed using two filtering methods.
 PAL is the Poisson approximate likelihood, implemented using the code of WWR.
-PF is the particle filter, implemented using the pomp R package \citet{pomppackagepaper}.
-The first two lines are from WWR, Table~1, and lines 3--5 are our own computations.
-We used 50000 particle for both PF and PAL.
-PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL.
-PF was maximized using iterated filtering, and PAL was maximized using coordinate gradient descent.
-} 
-\label{tab:ovovrealdata} % For referencing this table elsewhere in your text
+PF is the particle filter, implemented using the pomp R package \citep{pomppackagepaper}.
+Lines 1, 2 and 7 are from WWR, and the remainder are our own computations.
+We used $5\times 10^4$ particle for both PF and PAL.
+PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL due to its lower Monte~Carlo variance.
+PF was maximized using iterated filtering, following the approach of SBH.
+PAL was maximized using coordinate gradient descent, using the code of WWR.
+} 
+\label{tab:ovovrealdata}
 \begin{tabular}{lllr} 
@@
 SBH fitted their model to a dataset derived by dividing the original reported count data by an estimated reporting rate, to put their data on the scale of the actual number of cases in the population, whereas WWR fitted directly to the report data.
-The reporting rate used by SBH varied over time, but was generally around $7\%$.
-On approximately 1200 data points, this corresponds to a discrepancy of around $-1200\log(0.07) \approx 3200$ log-likelihood units, largely explaining the difference reported in Table~1 and interpreted by WWR as evidence supporting PAL.
-The comparison can be corrected either by applying the method of SBH to the data of WWR or vice versa.
+The reporting rate used by SBH varied over time and location, but was generally around $7\%$.
+On $3\times 416$ data points, this corresponds to a discrepancy of $-1248 \, \log(0.07) \approx 3300$ log-likelihood units, largely explaining the difference interpreted by WWR as evidence supporting PAL.
+The comparison between PAL and PF can be corrected by applying the method of SBH to the model and data of WWR, or vice versa.
 Since the method of SBH is applicable to a more general class of models, and supported by published software, it was convenient to apply the SBH method to the model and data of WWR.
-The large discrepancy in log-likelihood disappears at this point by recomputing the likelihood of the model using PAL and particle filter separately (see Table~\ref{tab:ovovrealdata}).
-This re-analysis does, however, show a discrepancy between two methods.
-We continued our investigation to establish the cause of this.
-
-Inspection of log-likelihood anomalies \citep{wheeler24,li24,hao24} showed that the initial conditions for the latent process in January 2001 were fixed at values which were incompatible with the trajectory of the data early in the time series.
-By contrast, SBH fixed the initial conditions 6 years before the first measurement, giving time for the system to reach its equilibrum distribution.
-Line 5 of Table~\ref{tab:ovovrealdata} shows that this improvement enables PF to reach close to the values attained by PAL.
-Further, at the MLE obtained by maximizing PF, we found that PF beats PAL (line 6).
-
-WWR compare their fitted models with a log-ARMA(2,1) benchmark AIC value of 23043 (line 7), inferring that all the mechanistic models possessing overdispersion have better statistical fit than a simple loglinear time series model.
-However, this value corresponds to a log-ARMA(2,1) model fitted not to WWR's data but to SBH's data.
+The large discrepancy in log-likelihood disappears when recomputing the likelihood using PF for the model fitted via PAL (Table~\ref{tab:ovovrealdata}, comparing lines 3 and 4).
+Some discrepancy remains, and we continued our investigation to establish the cause of this.
+
+Inspection of log-likelihood anomalies \citep{wheeler24,li24} showed that the initial conditions for the latent process in January 2001 were fixed at values which were incompatible with the trajectory of the data early in the time series.
+By contrast, SBH fixed their initial conditions 6 years before the first measurement, giving time for the system to reach its equilibrum distribution.
+Line 5 of Table~\ref{tab:ovovrealdata} incorporates this into the model of WWR and shows that this improvement enables PF to reach the values attained by PAL.
+Further, at the MLE obtained by maximizing PF, we found that PF beats PAL (comparing lines 5 and 6).
+
+WWR compared their fitted models with a log-ARMA(2,1) benchmark AIC value of 23043 (line 7).
+From this high AIC value, they inferred that all the mechanistic models possessing overdispersion have better statistical fit than a simple loglinear time series model.
+However, this AIC value corresponds to SBH's data, not the data fitted by WWR.
 Thus, line 7 can properly compared to line 1 but not to any other line of  Table~\ref{tab:ovovrealdata}. 
@@
 This holds for all the variants in lines 2--6 of  Table~\ref{tab:ovovrealdata}, regardless of whether PF or PAL is used.
-The goal of mechanistic modeling is not necessarily to beat a simple statistical benchmark, but this can be an indication that additional model development could be worthwhile \citep{wheeler24}.
-A correct interpretation of Table~\ref{tab:ovovrealdata} is therefore very different to the conclusions drawn by WWR from comparing line 2 inappropriately to lines 1 and 7.
-
-<<plot1-100-sim-compare,echo=FALSE, fig.height=4, fig.width=5.5, out.width="4.5in",fig.align="center",fig.cap='Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model.
-Simulations with one or more zero counts were disqualified since they resulted in errors for the PAL implementation.
-We used 50000 particle for both PF and PAL.
-PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL.
+The goal of mechanistic modeling is not necessarily to beat a simple statistical benchmark, but falling far below a simple statistical benchmark is an indication that additional model development could be worthwhile \citep{wheeler24}.
+A correct interpretation of Table~\ref{tab:ovovrealdata} is therefore very different to the conclusions drawn by WWR, who compared line 2 inappropriately to lines 1 and 7.
+
+<<plot1-100-sim-compare,echo=FALSE, fig.height=4, fig.width=5.5, out.width="4.5in",fig.align="center",fig.cap='Log-likelihood computed using PAL and PF for 100 randomly simulated datasets using the model and parameter values of WWR.
+Simulations with one or more zero counts were disqualified since they resulted in errors for the PAL implementation of WWR.
+We used $5\times 10^4$ particle for both PF and PAL.
+PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL due to its lower Monte~Carlo variance.
 The red line corresponds to equality of the two estimates.'>>=
@@
 
-In the presence of model misspecification, it becomes harder to compare likelihood evaluation methods.
-A likelihood approximation, such as PAL, may potentially obtain a higher value than the exact likelihood if it compensates for the misspecification.
+In the presence of model misspecification, it becomes difficult to compare likelihood evaluation methods.
+A likelihood approximation, such as PAL, may potentially obtain a higher value than the exact likelihood if it compensates for model misspecification.
 Log-likelihood is a proper scoring rule for forecasts \cite{gneiting2}, and both the particle filter and PAL construct their log-likelihood estimates via a sequence of one-step forecasts.
@@
 We know from the benchmark AIC value in Table~\ref{tab:ovovrealdata}. line 8, that there is substantial model misspecification.
-Therefore, this real-data example may not be well suited to comparing the filtering skill of PAL and PF.
-
-We conclude that PAL is a potentially useful algorithm, with some favorable theoretical properties.
-However, the corrected evidence does not indicate an advantage for using PAL in situations where the particle filter is effective.
-
-An extended description of our reanalysis of WWR is provided by \cite{hao24}.
-That thesis also contains some additional results reinforcing the conclusions presented in this article.
+
+It is currently unclear why the model of SBH beats the log-ARMA benchmark for the data they fitted to, whereas the model of WWR fits more poorly than the log-ARMA model for its corresponding data.
+Additional rounds of model development are required to resolve this, for which it is desirable to employ statistical methods that are broadly applicable both in theory and practice.
+Particle filter methods meet this criterion since they have the plug-and-play property \citep{breto09,he10} which is not posessed by PAL.
+Although WWR have shown that PAL is a potentially useful algorithm with some favorable theoretical properties, the corrected evidence does not indicate an advantage for using PAL in situations where the particle filter is effective.
+
+Also contained dditional results reinforcing the conclusions presented in this article.
 The source code for this article is available at \url{https://github.com:ionides/pal-vs-pf} and archived at \ed{ZENODO - TODO}.
+An extended description of our methods, together with additional numerical results, is provided by \cite{hao24}.
 
@@
 \end{document}
-
-\section{Extra material excluded in the final version}
-
-%%%%%%%%% 222222222 %%%%%%%%%%
-
-\section*{Methods}
-
-We first replicated the model of WWR in the R package \texttt{pomp} \citep{pomppackagepaper}.
-Comparing summary statistics of simulations confirms that both model versions are essentially identical, as described by \citet{hao24}.
-We therefore implement PAL using the code from \citet{wwr}, comparing directly with the particle filter implemented using \texttt{pomp}.
-
-This article has focused on likelihood evaluation, but the inferences presented also require the likelihood estimate to be maximized.
-WWR used direct numerical maximization, whereas SBH used iterated filtering to maximize the particle filter likelihood estimate implemented using the \texttt{mif2} algorithm in \texttt{pomp}.
-We follow the approaches adopted by these two papers.
-The optimization method is not directly pertinent to the comparison of the filtering methods, so we do not discuss optimization in further detail here.
-
-\begin{table}[htbp]
-    \caption{Maximum likelihood estimation of parameters for the OvOv model of WWR by PAL (taken from WWR), and by iterated particle filtering implemented using {\rm \texttt{mif2}} in {\rm \texttt{pomp}} \citep{pomppackagepaper}.
-The initial distribution parameter \(\lambda_0 = (S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}, R_{30})\) are assumed to be fixed in \cite{wwr}.
-The iterated particle filtering estimates were obtained by using 3 rounds of refinement, with 100 filtering iterations, 50,000 particles, and 36 replicates in each round with the top 12 best fits in terms of likelihood chosen to be the starting value for the next round.}
-    \label{tab:mlebywwr}
-\begin{center}    
-    \begin{tabular}{lccc}  % Left-aligned for the first column, centered for the other two columns
-      \toprule
-      Parameter      & PAL      & iterated filtering & iterated filtering (stocks)\\
-      \midrule
-      \(\beta_1\)    & 11.48     & $\Sexpr{round(est_mle_vec["beta1"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta1"],2)}$\\
-      \(\beta_2\)    & 0.25      & $\Sexpr{round(est_mle_vec["beta2"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta2"],2)}$\\
-      \(\beta_3\)    & 0.35      & $\Sexpr{round(est_mle_vec["beta3"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta3"],2)}$\\
-      \(\phi\)       & 0.14      & $\Sexpr{round(est_mle_vec["phi"],2)}$   & $\Sexpr{round(est_mle_vec_sbh["phi"],2)}$\\
-      \(\rho\)       & 0.16      & $\Sexpr{round(est_mle_vec["beta11"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta11"],2)}$\\
-      \(\sigma^2_q\) & 0.021     & $\Sexpr{round(est_mle_vec["sigma_q"],3)}$ & $\Sexpr{round(est_mle_vec_sbh["sigma_q"],3)}$\\
-      \(\sigma_\xi\) & 66.89     & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$ & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$\\
-      \(S_{10}\)     & 3876549   & $\Sexpr{round(pop*est_mle_vec["S10"],0)}$  & $5094840$\\
-      \(I_{10}\)     & 30351     & $\Sexpr{round(pop*est_mle_vec["I10"],0)}$  & $2871$\\
-      \(R_{10}\)     & 1315221   & $\Sexpr{round(pop*est_mle_vec["R10"],0)}$  & $124410$\\
-      \(S_{20}\)     & 57139612  & $\Sexpr{round(pop*est_mle_vec["S20"],0)}$  & $57385600$\\
-      \(I_{20}\)     & 871       & $\Sexpr{round(pop*est_mle_vec["I20"],0)}$  & $639$\\
-      \(R_{20}\)     & 302852    & $\Sexpr{round(pop*est_mle_vec["R20"],0)}$  & $57072$\\
-      \(S_{30}\)     & 19573727  & $\Sexpr{round(pop*est_mle_vec["S30"],0)}$  & $19697600$\\
-      \(I_{30}\)     & 2550      & $\Sexpr{round(pop*est_mle_vec["I30"],0)}$  & $174$\\
-      \(R_{30}\)     & 131092    & $\Sexpr{round(pop*est_mle_vec["R30"],0)}$  & $9578$\\
-      \midrule
-      \textbf{AIC}  & 13778.08  & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$ & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec_sbh["logLik"]),2)}$\\
-      \bottomrule
-    \end{tabular}
-\end{center}
-\end{table}
-
-<<plot2-cond-loglik, echo=FALSE, fig.height=4, fig.width=5.5, out.width="4.5in",fig.align="center", fig.cap='Conditional log-likelihoods computed using PAL (red) and PF (blue) for the rotavirus model and MLE from WWR.
-We used 50000 particle for both PF and PAL.
-PF averaged over 36 replications to reduce the Monte Carlo variance, but this step was not necessary for PAL.
-The main sources of likelihood shortfall of PF are at time points $t=1, 2, 3, 11, 81, 194$, and $325-333$.'>>=
-
-# Combine the data into a single dataframe for SMC with grouping variable
-data_list <- lapply(1:ncores, function(i) {
-  data.frame(
-    time = time, 
-    cond_loglik = pfilter_wwr_cond_lik[[i]], 
-    method = "SMC",
-    group = i
-  )
-})
-
-# Combine all the SMC data frames into one
-df_smc <- bind_rows(data_list)
-
-# Create the PAL data frame
-df_pal <- data.frame(
-  time = time, 
-  cond_loglik = palfilter_lik_list_wwr$ll_storage, 
-  method = "PAL",
-  group = 1  # Single group for PAL
-)
-
-# Combine both data frames
-df <- bind_rows(df_smc, df_pal)
-
-# Plot with ggplot
-ggplot(df, aes(x = time, y = cond_loglik, color = method)) +
-  geom_line(data = df %>% filter(method == "SMC"), aes(group = group), color = "blue", alpha = 0.7) +
-  geom_line(data = df %>% filter(method == "PAL"), aes(group = group), color = "red") +
-  theme_minimal() +
-  labs(
-    title = NULL,
-    x = "Time (weeks)",
-    y = "Conditional Log-likelihood",
-    color = "Method"
-  ) +
-  theme(legend.position = "bottom")
-@
-
-\begin{table}[htbp]
-    \caption{Maximum likelihood estimation of parameters for the OvOv model of WWR by PAL (taken from WWR), and by iterated particle filtering implemented using {\rm \texttt{mif2}} in {\rm \texttt{pomp}} \citep{pomppackagepaper}.
-The initial distribution parameter \(\lambda_0 = (S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}, R_{30})\) are assumed to be fixed in \cite{wwr}.
-The iterated particle filtering estimates were obtained by using 3 rounds of refinement, with 100 filtering iterations, 50,000 particles, and 36 replicates in each round with the top 12 best fits in terms of likelihood chosen to be the starting value for the next round.}
-    \label{tab:mlebywwr}
-\begin{center}    
-    \begin{tabular}{lccc}  % Left-aligned for the first column, centered for the other two columns
-      \toprule
-      Parameter      & PAL      & iterated filtering & iterated filtering (stocks)\\
-      \midrule
-      \(\beta_1\)    & 11.48     & $\Sexpr{round(est_mle_vec["beta1"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta1"],2)}$\\
-      \(\beta_2\)    & 0.25      & $\Sexpr{round(est_mle_vec["beta2"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta2"],2)}$\\
-      \(\beta_3\)    & 0.35      & $\Sexpr{round(est_mle_vec["beta3"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta3"],2)}$\\
-      \(\phi\)       & 0.14      & $\Sexpr{round(est_mle_vec["phi"],2)}$   & $\Sexpr{round(est_mle_vec_sbh["phi"],2)}$\\
-      \(\rho\)       & 0.16      & $\Sexpr{round(est_mle_vec["beta11"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta11"],2)}$\\
-      \(\sigma^2_q\) & 0.021     & $\Sexpr{round(est_mle_vec["sigma_q"],3)}$ & $\Sexpr{round(est_mle_vec_sbh["sigma_q"],3)}$\\
-      \(\sigma_\xi\) & 66.89     & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["sigma_xi"],2)}$\\
-      \(S_{10}\)     & 3876549   & $\Sexpr{round(pop*est_mle_vec["S10"],0)}$  & $5094840$\\
-      \(I_{10}\)     & 30351     & $\Sexpr{round(pop*est_mle_vec["I10"],0)}$  & $2871$\\
-      \(R_{10}\)     & 1315221   & $\Sexpr{round(pop*est_mle_vec["R10"],0)}$  & $124410$\\
-      \(S_{20}\)     & 57139612  & $\Sexpr{round(pop*est_mle_vec["S20"],0)}$  & $57385600$\\
-      \(I_{20}\)     & 871       & $\Sexpr{round(pop*est_mle_vec["I20"],0)}$  & $639$\\
-      \(R_{20}\)     & 302852    & $\Sexpr{round(pop*est_mle_vec["R20"],0)}$  & $57072$\\
-      \(S_{30}\)     & 19573727  & $\Sexpr{round(pop*est_mle_vec["S30"],0)}$  & $19697600$\\
-      \(I_{30}\)     & 2550      & $\Sexpr{round(pop*est_mle_vec["I30"],0)}$  & $174$\\
-      \(R_{30}\)     & 131092    & $\Sexpr{round(pop*est_mle_vec["R30"],0)}$  & $9578$\\
-      \midrule
-      \textbf{AIC}  & 13778.08  & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$ & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec_sbh["logLik"]),2)}$\\
-      \bottomrule
-    \end{tabular}
-\end{center}
-\end{table}
-
-Figure~\ref{fig:plot2-cond-loglik} indicates that the worst conditional log-likelihood values, estimated using the particle filter, arise at the start of the time series.
-This could be a problem with data collection, or some subtle issue with the science of the epidemiological system, but a simple possibility is that the initial values of the latent state process might be poorly specified.
-For this model, the initial values were fixed based on scientific reasoning, rather than being estimated.
-If we instead estimate the initial values, we obtain the results in Table~\ref{tab:mlebywwr}.

### 2024-09-15 ms.Rnw

@@
 \usepackage{makecell}
-% \usepackage{threeparttable} % For table notes
 \usepackage{array} % For table formatting
@@
 <<benchmark,echo=FALSE>>=
-arma_benchmark <- function(x,order=c(2,0,1)){
-  arima(log(x + 1), order = order)$loglik - sum(log(x+1))
-}
-
-sum(apply(realdat,2, function(x)  arima(log(x + 1), order =c(2,0,1))$aic + 2 * sum(log(x+1))))
-
-sbh_data <- read.csv("sbh_data.txt",sep=" ",skip=3)[,-1]
+set.seed(123)
+
+sbh_data <- read.csv("sbh_data.txt",sep=" ",skip=3)[,-c(1)]
 sbh_benchmark <- sum(apply(sbh_data,2, function(x)
@@
   4. & PF & Data, model and parameters matching 3. &
-    \Sexpr{myround(-2*logmeanexp(logLik(pfilter_wwr)+ 2*7),tab_signif)}
+    \Sexpr{myround(-2*logmeanexp(logLik(pfilter_wwr)) + 2*7,tab_signif)}
   \\
-  5. & PF & Changing initial values &
-     \Sexpr{myround(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),tab_signif)}
+  5. & PF & Using the initial values of SBH &
+     \Sexpr{myround(-2*est_mle_vec_sbh["logLik"] + 2*7,tab_signif)}
   \\

### 2024-09-15 ms.Rnw
recalculated table 1. changed to sbh and pfilter at mle with sbh init

--- 2024-09-14 ms.Rnw
@@
 
-<<PAL-MLE-from-mif2, echo=FALSE, message=FALSE, warning=FALSE>>=
+<<simulate-initial-for-palfilter>>=
+n_sim <- 1000
+
+sim_sbh_init_for_pal <- simulate(
+  sir_sbh_init,
+  nsim=n_sim,
+  params=est_mle_vec_sbh,
+  seed = 123456
+)
+
+s1_t1_vec <- c()
+i1_t1_vec <- c()
+r1_t1_vec <- c()
+s2_t1_vec <- c()
+i2_t1_vec <- c()
+r2_t1_vec <- c()
+s3_t1_vec <- c()
+i3_t1_vec <- c()
+r3_t1_vec <- c()
+
+for(i in 1:n_sim){
+  s1_t1_vec[i] <- sim_sbh_init_for_pal[[i]]@states["S1",2]
+  i1_t1_vec[i] <- sim_sbh_init_for_pal[[i]]@states["I1",2]
+  r1_t1_vec[i] <- sim_sbh_init_for_pal[[i]]@states["R1",2]
+
+  s2_t1_vec[i] <- sim_sbh_init_for_pal[[i]]@states["S2",2]
+  i2_t1_vec[i] <- sim_sbh_init_for_pal[[i]]@states["I2",2]
+  r2_t1_vec[i] <- sim_sbh_init_for_pal[[i]]@states["R2",2]
+
+  s3_t1_vec[i] <- sim_sbh_init_for_pal[[i]]@states["S3",2]
+  i3_t1_vec[i] <- sim_sbh_init_for_pal[[i]]@states["I3",2]
+  r3_t1_vec[i] <- sim_sbh_init_for_pal[[i]]@states["R3",2]
+}
+
+init_dist_sbh_mle <- c(
+  S1=round(mean(s1_t1_vec),0),
+  I1=round(mean(i1_t1_vec),0),
+  R1=round(mean(r1_t1_vec),0),
+  S2=round(mean(s2_t1_vec),0),
+  I2=round(mean(i2_t1_vec),0),
+  R2=round(mean(r2_t1_vec),0),
+  S3=round(mean(s3_t1_vec),0),
+  I3=round(mean(i3_t1_vec),0),
+  R3=round(mean(r3_t1_vec),0)
+)
+@
+
+<<palfilter-at-mle_sbh, echo=FALSE, message=FALSE, warning=FALSE>>=
 prop <- 420
-realdat <- as.matrix(dat[,-1])
-init_dist_mif2_mle <- c(
-  pop*as.numeric(est_mle_vec["S10"]), 
-  pop*as.numeric(est_mle_vec["I10"]), 
-  pop*as.numeric(est_mle_vec["R10"]),
-  pop*as.numeric(est_mle_vec["S20"]),
-  pop*as.numeric(est_mle_vec["R20"]),
-  pop*as.numeric(est_mle_vec["I20"]),
-  pop*as.numeric(est_mle_vec["S30"]),
-  pop*as.numeric(est_mle_vec["R30"]),
-  pop*as.numeric(est_mle_vec["I30"])
-)
 
 initial_guess_mif2_mle <- c(
-  as.numeric(est_mle_vec["beta1"]), 
-  as.numeric(est_mle_vec["beta2"]), 
-  as.numeric(est_mle_vec["beta3"]),
-  as.numeric(est_mle_vec["phi"]),
-  as.numeric(est_mle_vec["beta11"]),
-  as.numeric(est_mle_vec["sigma_q"]),
-  as.numeric(est_mle_vec["sigma_xi"])
+  as.numeric(est_mle_vec_sbh["beta1"]), 
+  as.numeric(est_mle_vec_sbh["beta2"]), 
+  as.numeric(est_mle_vec_sbh["beta3"]),
+  as.numeric(est_mle_vec_sbh["phi"]),
+  as.numeric(est_mle_vec_sbh["beta11"]),
+  as.numeric(est_mle_vec_sbh["sigma_q"]),
+  as.numeric(est_mle_vec_sbh["sigma_xi"])
 )
@@
  rotavirus_SMC_qropxi(
-    init_dist = init_dist_mif2_mle, 
+    init_dist = init_dist_sbh_mle, 
     y_obs = realdat, 
@@
 
-wwr_benchmark <- sum(apply(dat[,-1],2, function(x)
+wwr_benchmark <- sum(apply(realdat,2, function(x)
   arima(log(x + 1), order =c(2,0,1))$aic + 2 * sum(log(x+1))
@@
 
+<<PAL-MLE-from-mif2, echo=FALSE, message=FALSE, warning=FALSE>>=
+prop <- 420
+realdat <- as.matrix(dat[,-1])
+init_dist_mif2_mle <- c(
+  pop*as.numeric(est_mle_vec["S10"]), 
+  pop*as.numeric(est_mle_vec["I10"]), 
+  pop*as.numeric(est_mle_vec["R10"]),
+  pop*as.numeric(est_mle_vec["S20"]),
+  pop*as.numeric(est_mle_vec["R20"]),
+  pop*as.numeric(est_mle_vec["I20"]),
+  pop*as.numeric(est_mle_vec["S30"]),
+  pop*as.numeric(est_mle_vec["R30"]),
+  pop*as.numeric(est_mle_vec["I30"])
+)
+
+initial_guess_mif2_mle <- c(
+  as.numeric(est_mle_vec["beta1"]), 
+  as.numeric(est_mle_vec["beta2"]), 
+  as.numeric(est_mle_vec["beta3"]),
+  as.numeric(est_mle_vec["phi"]),
+  as.numeric(est_mle_vec["beta11"]),
+  as.numeric(est_mle_vec["sigma_q"]),
+  as.numeric(est_mle_vec["sigma_xi"])
+)
+
+bake(file = paste0(out_dir,"pfilter/original_dataset/palfilter_mif2_mle.rds"),{
+ set.seed(1223)
+ rotavirus_SMC_qropxi(
+    init_dist = init_dist_mif2_mle, 
+    y_obs = realdat, 
+    m = 9, 
+    regular_params = c(
+      initial_guess_mif2_mle[1], 
+      initial_guess_mif2_mle[2], 
+      initial_guess_mif2_mle[3], 
+      initial_guess_mif2_mle[4],
+      initial_guess_mif2_mle[5]
+    ), 
+    gamma_par = c(initial_guess_mif2_mle[7], 1/initial_guess_mif2_mle[7]), 
+    norm_par = c(0.07, initial_guess_mif2_mle[6]), 
+    prop = t(prop), 
+    n_particles = NP_MIF, 
+    ncores = ncores
+  ) 
+}) -> palfilter_mif2_mle
+@
+
 <<plot2-cond-loglik, echo=FALSE, fig.height=4, fig.width=5.5, out.width="4.5in",fig.align="center", fig.cap='Conditional log-likelihoods computed using PAL (red) and PF (blue) for the rotavirus model and MLE from WWR.

### 2024-09-15 ms.Rnw

--- 2024-09-15 ms.Rnw
@@
 \end{document}
+
+\section{Extra material excluded in the final version}
+
+%%%%%%%%% 222222222 %%%%%%%%%%
+
+\section*{Methods}
+
+We first replicated the model of WWR in the R package \texttt{pomp} \citep{pomppackagepaper}.
+Comparing summary statistics of simulations confirms that both model versions are essentially identical, as described by \citet{hao24}.
+We therefore implement PAL using the code from \citet{wwr}, comparing directly with the particle filter implemented using \texttt{pomp}.
+
+This article has focused on likelihood evaluation, but the inferences presented also require the likelihood estimate to be maximized.
+WWR used direct numerical maximization, whereas SBH used iterated filtering to maximize the particle filter likelihood estimate implemented using the \texttt{mif2} algorithm in \texttt{pomp}.
+We follow the approaches adopted by these two papers.
+The optimization method is not directly pertinent to the comparison of the filtering methods, so we do not discuss optimization in further detail here.
+
+\begin{table}[htbp]
+    \caption{Maximum likelihood estimation of parameters for the OvOv model of WWR by PAL (taken from WWR), and by iterated particle filtering implemented using {\rm \texttt{mif2}} in {\rm \texttt{pomp}} \citep{pomppackagepaper}.
+The initial distribution parameter \(\lambda_0 = (S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}, R_{30})\) are assumed to be fixed in \cite{wwr}.
+The iterated particle filtering estimates were obtained by using 3 rounds of refinement, with 100 filtering iterations, 50,000 particles, and 36 replicates in each round with the top 12 best fits in terms of likelihood chosen to be the starting value for the next round.}
+    \label{tab:mlebywwr}
+\begin{center}    
+    \begin{tabular}{lccc}  % Left-aligned for the first column, centered for the other two columns
+      \toprule
+      Parameter      & PAL      & iterated filtering & iterated filtering (stocks)\\
+      \midrule
+      \(\beta_1\)    & 11.48     & $\Sexpr{round(est_mle_vec["beta1"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta1"],2)}$\\
+      \(\beta_2\)    & 0.25      & $\Sexpr{round(est_mle_vec["beta2"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta2"],2)}$\\
+      \(\beta_3\)    & 0.35      & $\Sexpr{round(est_mle_vec["beta3"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta3"],2)}$\\
+      \(\phi\)       & 0.14      & $\Sexpr{round(est_mle_vec["phi"],2)}$   & $\Sexpr{round(est_mle_vec_sbh["phi"],2)}$\\
+      \(\rho\)       & 0.16      & $\Sexpr{round(est_mle_vec["beta11"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta11"],2)}$\\
+      \(\sigma^2_q\) & 0.021     & $\Sexpr{round(est_mle_vec["sigma_q"],3)}$ & $\Sexpr{round(est_mle_vec_sbh["sigma_q"],3)}$\\
+      \(\sigma_\xi\) & 66.89     & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$ & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$\\
+      \(S_{10}\)     & 3876549   & $\Sexpr{round(pop*est_mle_vec["S10"],0)}$  & $5094840$\\
+      \(I_{10}\)     & 30351     & $\Sexpr{round(pop*est_mle_vec["I10"],0)}$  & $2871$\\
+      \(R_{10}\)     & 1315221   & $\Sexpr{round(pop*est_mle_vec["R10"],0)}$  & $124410$\\
+      \(S_{20}\)     & 57139612  & $\Sexpr{round(pop*est_mle_vec["S20"],0)}$  & $57385600$\\
+      \(I_{20}\)     & 871       & $\Sexpr{round(pop*est_mle_vec["I20"],0)}$  & $639$\\
+      \(R_{20}\)     & 302852    & $\Sexpr{round(pop*est_mle_vec["R20"],0)}$  & $57072$\\
+      \(S_{30}\)     & 19573727  & $\Sexpr{round(pop*est_mle_vec["S30"],0)}$  & $19697600$\\
+      \(I_{30}\)     & 2550      & $\Sexpr{round(pop*est_mle_vec["I30"],0)}$  & $174$\\
+      \(R_{30}\)     & 131092    & $\Sexpr{round(pop*est_mle_vec["R30"],0)}$  & $9578$\\
+      \midrule
+      \textbf{AIC}  & 13778.08  & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$ & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec_sbh["logLik"]),2)}$\\
+      \bottomrule
+    \end{tabular}
+\end{center}
+\end{table}
+
+<<PAL-MLE-from-mif2, echo=FALSE, message=FALSE, warning=FALSE>>=
+prop <- 420
+realdat <- as.matrix(dat[,-1])
+init_dist_mif2_mle <- c(
+  pop*as.numeric(est_mle_vec["S10"]), 
+  pop*as.numeric(est_mle_vec["I10"]), 
+  pop*as.numeric(est_mle_vec["R10"]),
+  pop*as.numeric(est_mle_vec["S20"]),
+  pop*as.numeric(est_mle_vec["R20"]),
+  pop*as.numeric(est_mle_vec["I20"]),
+  pop*as.numeric(est_mle_vec["S30"]),
+  pop*as.numeric(est_mle_vec["R30"]),
+  pop*as.numeric(est_mle_vec["I30"])
+)
+
+initial_guess_mif2_mle <- c(
+  as.numeric(est_mle_vec["beta1"]), 
+  as.numeric(est_mle_vec["beta2"]), 
+  as.numeric(est_mle_vec["beta3"]),
+  as.numeric(est_mle_vec["phi"]),
+  as.numeric(est_mle_vec["beta11"]),
+  as.numeric(est_mle_vec["sigma_q"]),
+  as.numeric(est_mle_vec["sigma_xi"])
+)
+
+bake(file = paste0(out_dir,"pfilter/original_dataset/palfilter_mif2_mle.rds"),{
+ set.seed(1223)
+ rotavirus_SMC_qropxi(
+    init_dist = init_dist_mif2_mle, 
+    y_obs = realdat, 
+    m = 9, 
+    regular_params = c(
+      initial_guess_mif2_mle[1], 
+      initial_guess_mif2_mle[2], 
+      initial_guess_mif2_mle[3], 
+      initial_guess_mif2_mle[4],
+      initial_guess_mif2_mle[5]
+    ), 
+    gamma_par = c(initial_guess_mif2_mle[7], 1/initial_guess_mif2_mle[7]), 
+    norm_par = c(0.07, initial_guess_mif2_mle[6]), 
+    prop = t(prop), 
+    n_particles = NP_MIF, 
+    ncores = ncores
+  ) 
+}) -> palfilter_mif2_mle
+@
+
+<<plot2-cond-loglik, echo=FALSE, fig.height=4, fig.width=5.5, out.width="4.5in",fig.align="center", fig.cap='Conditional log-likelihoods computed using PAL (red) and PF (blue) for the rotavirus model and MLE from WWR.
+We used 50000 particle for both PF and PAL.
+PF averaged over 36 replications to reduce the Monte Carlo variance, but this step was not necessary for PAL.
+The main sources of likelihood shortfall of PF are at time points $t=1, 2, 3, 11, 81, 194$, and $325-333$.'>>=
+
+# Combine the data into a single dataframe for SMC with grouping variable
+data_list <- lapply(1:ncores, function(i) {
+  data.frame(
+    time = time, 
+    cond_loglik = pfilter_wwr_cond_lik[[i]], 
+    method = "SMC",
+    group = i
+  )
+})
+
+# Combine all the SMC data frames into one
+df_smc <- bind_rows(data_list)
+
+# Create the PAL data frame
+df_pal <- data.frame(
+  time = time, 
+  cond_loglik = palfilter_lik_list_wwr$ll_storage, 
+  method = "PAL",
+  group = 1  # Single group for PAL
+)
+
+# Combine both data frames
+df <- bind_rows(df_smc, df_pal)
+
+# Plot with ggplot
+ggplot(df, aes(x = time, y = cond_loglik, color = method)) +
+  geom_line(data = df %>% filter(method == "SMC"), aes(group = group), color = "blue", alpha = 0.7) +
+  geom_line(data = df %>% filter(method == "PAL"), aes(group = group), color = "red") +
+  theme_minimal() +
+  labs(
+    title = NULL,
+    x = "Time (weeks)",
+    y = "Conditional Log-likelihood",
+    color = "Method"
+  ) +
+  theme(legend.position = "bottom")
+@
+
+\begin{table}[htbp]
+    \caption{Maximum likelihood estimation of parameters for the OvOv model of WWR by PAL (taken from WWR), and by iterated particle filtering implemented using {\rm \texttt{mif2}} in {\rm \texttt{pomp}} \citep{pomppackagepaper}.
+The initial distribution parameter \(\lambda_0 = (S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}, R_{30})\) are assumed to be fixed in \cite{wwr}.
+The iterated particle filtering estimates were obtained by using 3 rounds of refinement, with 100 filtering iterations, 50,000 particles, and 36 replicates in each round with the top 12 best fits in terms of likelihood chosen to be the starting value for the next round.}
+    \label{tab:mlebywwr}
+\begin{center}    
+    \begin{tabular}{lccc}  % Left-aligned for the first column, centered for the other two columns
+      \toprule
+      Parameter      & PAL      & iterated filtering & iterated filtering (stocks)\\
+      \midrule
+      \(\beta_1\)    & 11.48     & $\Sexpr{round(est_mle_vec["beta1"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta1"],2)}$\\
+      \(\beta_2\)    & 0.25      & $\Sexpr{round(est_mle_vec["beta2"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta2"],2)}$\\
+      \(\beta_3\)    & 0.35      & $\Sexpr{round(est_mle_vec["beta3"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta3"],2)}$\\
+      \(\phi\)       & 0.14      & $\Sexpr{round(est_mle_vec["phi"],2)}$   & $\Sexpr{round(est_mle_vec_sbh["phi"],2)}$\\
+      \(\rho\)       & 0.16      & $\Sexpr{round(est_mle_vec["beta11"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta11"],2)}$\\
+      \(\sigma^2_q\) & 0.021     & $\Sexpr{round(est_mle_vec["sigma_q"],3)}$ & $\Sexpr{round(est_mle_vec_sbh["sigma_q"],3)}$\\
+      \(\sigma_\xi\) & 66.89     & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["sigma_xi"],2)}$\\
+      \(S_{10}\)     & 3876549   & $\Sexpr{round(pop*est_mle_vec["S10"],0)}$  & $5094840$\\
+      \(I_{10}\)     & 30351     & $\Sexpr{round(pop*est_mle_vec["I10"],0)}$  & $2871$\\
+      \(R_{10}\)     & 1315221   & $\Sexpr{round(pop*est_mle_vec["R10"],0)}$  & $124410$\\
+      \(S_{20}\)     & 57139612  & $\Sexpr{round(pop*est_mle_vec["S20"],0)}$  & $57385600$\\
+      \(I_{20}\)     & 871       & $\Sexpr{round(pop*est_mle_vec["I20"],0)}$  & $639$\\
+      \(R_{20}\)     & 302852    & $\Sexpr{round(pop*est_mle_vec["R20"],0)}$  & $57072$\\
+      \(S_{30}\)     & 19573727  & $\Sexpr{round(pop*est_mle_vec["S30"],0)}$  & $19697600$\\
+      \(I_{30}\)     & 2550      & $\Sexpr{round(pop*est_mle_vec["I30"],0)}$  & $174$\\
+      \(R_{30}\)     & 131092    & $\Sexpr{round(pop*est_mle_vec["R30"],0)}$  & $9578$\\
+      \midrule
+      \textbf{AIC}  & 13778.08  & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$ & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec_sbh["logLik"]),2)}$\\
+      \bottomrule
+    \end{tabular}
+\end{center}
+\end{table}
+
+Figure~\ref{fig:plot2-cond-loglik} indicates that the worst conditional log-likelihood values, estimated using the particle filter, arise at the start of the time series.
+This could be a problem with data collection, or some subtle issue with the science of the epidemiological system, but a simple possibility is that the initial values of the latent state process might be poorly specified.
+For this model, the initial values were fixed based on scientific reasoning, rather than being estimated.
+If we instead estimate the initial values, we obtain the results in Table~\ref{tab:mlebywwr}.

### 2024-09-15 ms.Rnw
defer 7 to Exclude

@@
 }) ->  sbh_el_06
-@
-
-<<maximization7-Stocks_rinit, echo=F, warning=F, message=F>>=
-set.seed(12345)
-sd_07 = (2/3)^6
-starting_values_sbh_07 <- choose_n_fits(sbh_el_06, top_n_fits, ncores)
-
-bake(file = paste0(out_dir, "mif2_maximization/SBH_init/round_07/mifs_global_sbh_07.rds"),{ 
-  plan(multicore, workers = ncores)
-  foreach(i = 1:ncores, .options.future = list(seed = 652643293)) %dofuture% {
-    pomp::mif2(
-      mifs_global_sbh_06[[1]],
-      rw.sd = rw_sd(
-        beta1 = 0.01*sd_07, beta2 = 0.01*sd_07, beta3 = 0.01*sd_07,
-        beta11 = 0.01*sd_07, phi = 0.01*sd_07,
-        sigma_q = 0.01*sd_07, sigma_xi = 0.01*sd_07
-      ),
-      params = starting_values_sbh_07[[i]]
-    ) 
-  }
-}) -> mifs_global_sbh_07
-
-bake(file = paste0(out_dir,"mif2_maximization/SBH_init/round_07/sbh_el_07.rds"),{
-  eval_logLik_pomp(mifs_global_sbh_07, ncores, NP_MIF, seed = 4248930)
-}) ->  sbh_el_07
 @
@@
 
+<<maximization7-Stocks_rinit, include=FALSE>>=
+set.seed(12345)
+sd_07 = (2/3)^6
+starting_values_sbh_07 <- choose_n_fits(sbh_el_06, top_n_fits, ncores)
+
+bake(file = paste0(out_dir, "mif2_maximization/SBH_init/round_07/mifs_global_sbh_07.rds"),{ 
+  plan(multicore, workers = ncores)
+  foreach(i = 1:ncores, .options.future = list(seed = 652643293)) %dofuture% {
+    pomp::mif2(
+      mifs_global_sbh_06[[1]],
+      rw.sd = rw_sd(
+        beta1 = 0.01*sd_07, beta2 = 0.01*sd_07, beta3 = 0.01*sd_07,
+        beta11 = 0.01*sd_07, phi = 0.01*sd_07,
+        sigma_q = 0.01*sd_07, sigma_xi = 0.01*sd_07
+      ),
+      params = starting_values_sbh_07[[i]]
+    ) 
+  }
+}) -> mifs_global_sbh_07
+
+bake(file = paste0(out_dir,"mif2_maximization/SBH_init/round_07/sbh_el_07.rds"),{
+  eval_logLik_pomp(mifs_global_sbh_07, ncores, NP_MIF, seed = 4248930)
+}) ->  sbh_el_07
+@
+
 <<PAL-MLE-from-mif2, echo=FALSE, message=FALSE, warning=FALSE>>=

### 2024-09-15 ms.Rnw
some edits

@@
 PAL is the Poisson approximate likelihood, implemented using the code of WWR.
-PF is the particle filter, implemented using the pomp R package \citep{pomppackagepaper}.
+PF is the particle filter, implemented using the R package pomp \citep{pomppackagepaper}.
 Lines 1, 2 and 7 are from WWR, and the remainder are our own computations.
@@
   \\
-  5. & PF & Using the initial values of SBH &
+  5. & PF & Initial values changed to match SBH &
      \Sexpr{myround(-2*est_mle_vec_sbh["logLik"] + 2*7,tab_signif)}
@@
 However, this AIC value corresponds to SBH's data, not the data fitted by WWR.
-Thus, line 7 can properly compared to line 1 but not to any other line of  Table~\ref{tab:ovovrealdata}. 
+Thus, line 7 can properly be compared only to line 1 but not to any other line of  Table~\ref{tab:ovovrealdata}. 
 We independently fitted a log-ARMA(2,1) model to the SBH data and obtained a similar AIC value (\Sexpr{myround(sbh_benchmark,tab_signif)}).
@@
 
-Also contained dditional results reinforcing the conclusions presented in this article.
+Also contained additional results reinforcing the conclusions presented in this article.
 The source code for this article is available at \url{https://github.com:ionides/pal-vs-pf} and archived at \ed{ZENODO - TODO}.

### 2024-09-16 ms.Rnw
some edits
initial dist at t0 should be sum up to pop

@@
 
-init_dist_sbh_mle <- c(
+sbh_mean_state <- c(
   S1=round(mean(s1_t1_vec),0),
@@
 )
+
+#calculate the difference in the init dist
+diff <- sum(sbh_mean_state) - pop
+#reduce the differenc proportionally
+init_dist_sbh_mle <- round(sbh_mean_state - diff*(sbh_mean_state/sum(sbh_mean_state)),0)
+#check if init_dist == pop
+sum(init_dist_sbh_mle) == pop
+
+#sum(init_dist_sbh_mle) == pop
 @
@@
 However, this AIC value corresponds to SBH's data, not the data fitted by WWR.
-Thus, line 7 can properly be compared only to line 1 but not to any other line of  Table~\ref{tab:ovovrealdata}. 
+Thus, line 7 can be properly compared only to line 1 but not to any other line of  Table~\ref{tab:ovovrealdata}. 
 We independently fitted a log-ARMA(2,1) model to the SBH data and obtained a similar AIC value (\Sexpr{myround(sbh_benchmark,tab_signif)}).

### 2024-09-16 ms.Rnw
a more-or-less final round of edits

@@
 \usepackage{color}
-\usepackage{graphicx} % Required for inserting
-\usepackage{booktabs} % For better-quality horizontal lines
+\usepackage{graphicx} 
+\usepackage{booktabs} 
 \usepackage{multirow}
 \usepackage{makecell}
-\usepackage{array} % For table formatting
+\usepackage{array} 
 \usepackage{subcaption}
-\usepackage{setspace} % For double-line spacing
+\usepackage{setspace} 
 \usepackage{float} 
@@
 \usepackage[colorlinks=true,citecolor=blue]{hyperref}
-% Using \doublespacing in the preamble 
-% changes the text to double-line spacing
-% \doublespacing
-
-% Set page size and margins
-% Replace `letterpaper' with `a4paper' for UK/EU standard size
-\usepackage[a4paper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}
-
-\newcommand{\ed}[1]{\textcolor{red}{[EI: #1]}}
-\newcommand{\yize}[1]{\textcolor{green}{[YH: #1]}}
-\newcommand{\aaron}[1]{\textcolor{blue}{[AA: #1]}}
 
@@
 A central piece of evidence for PAL is the comparison in Table~1 of \cite{wwr}, which claims a large improvement for PAL over a standard particle filter algorithm.
-This evidence, based on a model and data from a previous scientific study by \citet{stocks}, might suggests that researchers confronted with similar models should use PAL rather than previous particle filter methods.
-Alternatively, the improvement in likelihood found by \cite{wwr} compared to \cite{stocks} might indicate a flaw in the model or numerical methods for that previous work.
-We show that the comparison of log-likelihood values made by \cite{wwr} is flawed because the PAL calculations were carried out using a dataset scaled differently from the previous study.
+This evidence, based on a model and data from a previous scientific study by \citet{stocks}, might suggests that researchers confronted with similar models should use PAL rather than particle filter methods.
+Taken at face value, this evidence also reduces the credibility of \cite{stocks} by indicating a shortcoming with the numerical methods that they used.
+However, we show that the comparison of log-likelihood values made by \cite{wwr} is flawed because their PAL calculations were carried out using a dataset scaled differently from the previous study.
 If PAL and the particle filter are applied to the same data, the advantage claimed for PAL disappears.
@@
 \end{abstract}
-
-% R starts here
 
@@
 
-# x <- c(3876549,57139612,19573727,30351,871,2550,1315221,302852,131092)
-# x/sum(x)
-
 # define parameters (fixed)
@@
 
-dat <- read.table("real_rotavirus_metadata.txt") |>
+dat <- read.table("unscaled_rotavirus_data.txt") |>
   arrange(time)
@@
 The theory developed by WWR shows that their Poisson Approximate Likelihood (PAL) method has some potentially useful scaling properties.
-This theory is supported by numerical results, in their Table~1, which erroneously claim to show that PAL has substantially stronger performance than a particle filter (PF) on an example of scientific interest but moderate size.
+This theory is supported by numerical results, in their Table~1, which erroneously claim to show that PAL has substantially stronger performance than a particle filter (PF) on an example of scientific interest.
 We present corrected results so that researchers considering whether to implement PAL are appropriately informed about its benefits.
 
-Table~1 of WWR reanalyzes the model and data of \citet{stocks} (henceforth, SBH) for which the likelihood was calculated using PF.
-SBH found clear evidence for the importance of overdispersion in a stochastic dynamic model for their epidemiological data.
+Table~1 of WWR uses a model and data adapted from \citet{stocks} (henceforth, SBH).
+SBH used PF to calculate the likelihood for a stochastic dynamic model of rotavirus transmission.
+SBH found clear evidence for the importance of including overdispersion in the model for their epidemiological data.
 This is significant because most earlier research on population dynamics avoided consideration of overdispersion, perhaps due to the lack of available statistical methodology for fitting overdispersed nonlinear stochastic dynamic models.
@@
 
-We show that most of the apparent advantage for PAL over PF, on the OvOv model, arises because WWR used a different scaling of the data from SBH.
+We show that the claimed advantage for PAL over PF, on the OvOv model, arose because WWR used a different scaling of the data from SBH.
 Two models for the same data can properly be compared by their likelihood, even if the models have entirely different structures.
@@
 Lines 1, 2 and 7 are from WWR, and the remainder are our own computations.
-We used $5\times 10^4$ particle for both PF and PAL.
+We used $5 \times 10^4$ particle for both PF and PAL.
 PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL due to its lower Monte~Carlo variance.
@@
   \\
-  3. & PAL & Data, model and parameters from WWR &
+  3. & PAL & Re-evaluating the data, model and parameters from WWR &
     \Sexpr{myround(-2*palfilter_lik_list_wwr$log_lik + 2*7 ,tab_signif)} 
@@
 The reporting rate used by SBH varied over time and location, but was generally around $7\%$.
-On $3\times 416$ data points, this corresponds to a discrepancy of $-1248 \, \log(0.07) \approx 3300$ log-likelihood units, largely explaining the difference interpreted by WWR as evidence supporting PAL.
+On $3\times 416$ data points, this corresponds to a discrepancy of $-1248 \, \log(0.07) \approx 3300$ log-likelihood units, largely explaining the difference interpreted by WWR as evidence supporting PAL (Table~\ref{tab:ovovrealdata}, comparing lines 1 and 2).
 The comparison between PAL and PF can be corrected by applying the method of SBH to the model and data of WWR, or vice versa.
-Since the method of SBH is applicable to a more general class of models, and supported by published software, it was convenient to apply the SBH method to the model and data of WWR.
+Since the method of SBH is applicable to a more general class of models, and supported by widely-used software, it was convenient to apply the SBH method to the model and data of WWR.
 The large discrepancy in log-likelihood disappears when recomputing the likelihood using PF for the model fitted via PAL (Table~\ref{tab:ovovrealdata}, comparing lines 3 and 4).
@@
 
-Inspection of log-likelihood anomalies \citep{wheeler24,li24} showed that the initial conditions for the latent process in January 2001 were fixed at values which were incompatible with the trajectory of the data early in the time series.
-By contrast, SBH fixed their initial conditions 6 years before the first measurement, giving time for the system to reach its equilibrum distribution.
-Line 5 of Table~\ref{tab:ovovrealdata} incorporates this into the model of WWR and shows that this improvement enables PF to reach the values attained by PAL.
+Inspection of log-likelihood anomalies \citep{wheeler24,li24} showed that the initial conditions for the latent process in January 2001 were fixed at values which were incompatible with the trajectory of the data early in the time series \citep[Figure~3 of][]{hao24}.
+By contrast, SBH fixed their initial conditions 6 years before the first measurement, giving time for the system to reach its equilibrium distribution.
+Line 5 of Table~\ref{tab:ovovrealdata} incorporates the SBH specification of initial values into the model of WWR.
+Comparison of lines 3 and 5 shows that this improvement enables PF to reach the values attained by PAL, and show a small improvement.
 Further, at the MLE obtained by maximizing PF, we found that PF beats PAL (comparing lines 5 and 6).
@@
 WWR compared their fitted models with a log-ARMA(2,1) benchmark AIC value of 23043 (line 7).
-From this high AIC value, they inferred that all the mechanistic models possessing overdispersion have better statistical fit than a simple loglinear time series model.
+From this high AIC value, they inferred that all the mechanistic models possessing overdispersion have better statistical fit than a simple log-linear time series model.
 However, this AIC value corresponds to SBH's data, not the data fitted by WWR.
 Thus, line 7 can be properly compared only to line 1 but not to any other line of  Table~\ref{tab:ovovrealdata}. 
-We independently fitted a log-ARMA(2,1) model to the SBH data and obtained a similar AIC value (\Sexpr{myround(sbh_benchmark,tab_signif)}).
+We refitted a log-ARMA(2,1) model to the SBH data and obtained a similar AIC value (\Sexpr{myround(sbh_benchmark,tab_signif)}).
 Carrying out the same computation for the data fitted by WWR gives a log-ARMA benchmark AIC value of \Sexpr{myround(wwr_benchmark,tab_signif)} (line 8).
@@
 Simulations with one or more zero counts were disqualified since they resulted in errors for the PAL implementation of WWR.
-We used $5\times 10^4$ particle for both PF and PAL.
+We used $5 \\times 10^4$ particle for both PF and PAL.
 PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL due to its lower Monte~Carlo variance.
@@
 A likelihood approximation, such as PAL, may potentially obtain a higher value than the exact likelihood if it compensates for model misspecification.
-Log-likelihood is a proper scoring rule for forecasts \cite{gneiting2}, and both the particle filter and PAL construct their log-likelihood estimates via a sequence of one-step forecasts.
+Log-likelihood is a proper scoring rule for forecasts \citep{gneiting2}, and both the particle filter and PAL construct their log-likelihood estimates via a sequence of one-step forecasts.
 Therefore, if the model is correctly specified, the approximation error in PAL can only decrease the expected log-likelihood.
@@
 On average, the particle filter likelihood estimate is $\Sexpr{myround(mean(pfilter_ovov_sim_100)-mean(palfilter_ovov_sim_100),1)}$ log units higher than the PAL.
-We know from the benchmark AIC value in Table~\ref{tab:ovovrealdata}. line 8, that there is substantial model misspecification.
-
-It is currently unclear why the model of SBH beats the log-ARMA benchmark for the data they fitted to, whereas the model of WWR fits more poorly than the log-ARMA model for its corresponding data.
+We know from the benchmark AIC value in Table~\ref{tab:ovovrealdata} (comparing line 8 to lines 2--6) that there is substantial model misspecification.
+
+It is currently unclear why the model of SBH beats the log-ARMA benchmark for their data, whereas the model of WWR fits more poorly than the log-ARMA model for its corresponding data.
 Additional rounds of model development are required to resolve this, for which it is desirable to employ statistical methods that are broadly applicable both in theory and practice.
-Particle filter methods meet this criterion since they have the plug-and-play property \citep{breto09,he10} which is not posessed by PAL.
+Particle filter methods meet this criterion since they have the plug-and-play property \citep{breto09,he10} which is not possessed by PAL.
 Although WWR have shown that PAL is a potentially useful algorithm with some favorable theoretical properties, the corrected evidence does not indicate an advantage for using PAL in situations where the particle filter is effective.
@@
 Also contained additional results reinforcing the conclusions presented in this article.
-The source code for this article is available at \url{https://github.com:ionides/pal-vs-pf} and archived at \ed{ZENODO - TODO}.
+The source code for this article is available at \url{https://github.com:ionides/pal-vs-pf} and archived at Zenodo.
 An extended description of our methods, together with additional numerical results, is provided by \cite{hao24}.
@@
 \end{document}
-
-\section{Extra material excluded in the final version}
-
-%%%%%%%%% 222222222 %%%%%%%%%%
-
-\section*{Methods}
-
-We first replicated the model of WWR in the R package \texttt{pomp} \citep{pomppackagepaper}.
-Comparing summary statistics of simulations confirms that both model versions are essentially identical, as described by \citet{hao24}.
-We therefore implement PAL using the code from \citet{wwr}, comparing directly with the particle filter implemented using \texttt{pomp}.
-
-This article has focused on likelihood evaluation, but the inferences presented also require the likelihood estimate to be maximized.
-WWR used direct numerical maximization, whereas SBH used iterated filtering to maximize the particle filter likelihood estimate implemented using the \texttt{mif2} algorithm in \texttt{pomp}.
-We follow the approaches adopted by these two papers.
-The optimization method is not directly pertinent to the comparison of the filtering methods, so we do not discuss optimization in further detail here.
-
-\begin{table}[htbp]
-    \caption{Maximum likelihood estimation of parameters for the OvOv model of WWR by PAL (taken from WWR), and by iterated particle filtering implemented using {\rm \texttt{mif2}} in {\rm \texttt{pomp}} \citep{pomppackagepaper}.
-The initial distribution parameter \(\lambda_0 = (S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}, R_{30})\) are assumed to be fixed in \cite{wwr}.
-The iterated particle filtering estimates were obtained by using 3 rounds of refinement, with 100 filtering iterations, 50,000 particles, and 36 replicates in each round with the top 12 best fits in terms of likelihood chosen to be the starting value for the next round.}
-    \label{tab:mlebywwr}
-\begin{center}    
-    \begin{tabular}{lccc}  % Left-aligned for the first column, centered for the other two columns
-      \toprule
-      Parameter      & PAL      & iterated filtering & iterated filtering (stocks)\\
-      \midrule
-      \(\beta_1\)    & 11.48     & $\Sexpr{round(est_mle_vec["beta1"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta1"],2)}$\\
-      \(\beta_2\)    & 0.25      & $\Sexpr{round(est_mle_vec["beta2"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta2"],2)}$\\
-      \(\beta_3\)    & 0.35      & $\Sexpr{round(est_mle_vec["beta3"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta3"],2)}$\\
-      \(\phi\)       & 0.14      & $\Sexpr{round(est_mle_vec["phi"],2)}$   & $\Sexpr{round(est_mle_vec_sbh["phi"],2)}$\\
-      \(\rho\)       & 0.16      & $\Sexpr{round(est_mle_vec["beta11"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta11"],2)}$\\
-      \(\sigma^2_q\) & 0.021     & $\Sexpr{round(est_mle_vec["sigma_q"],3)}$ & $\Sexpr{round(est_mle_vec_sbh["sigma_q"],3)}$\\
-      \(\sigma_\xi\) & 66.89     & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$ & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$\\
-      \(S_{10}\)     & 3876549   & $\Sexpr{round(pop*est_mle_vec["S10"],0)}$  & $5094840$\\
-      \(I_{10}\)     & 30351     & $\Sexpr{round(pop*est_mle_vec["I10"],0)}$  & $2871$\\
-      \(R_{10}\)     & 1315221   & $\Sexpr{round(pop*est_mle_vec["R10"],0)}$  & $124410$\\
-      \(S_{20}\)     & 57139612  & $\Sexpr{round(pop*est_mle_vec["S20"],0)}$  & $57385600$\\
-      \(I_{20}\)     & 871       & $\Sexpr{round(pop*est_mle_vec["I20"],0)}$  & $639$\\
-      \(R_{20}\)     & 302852    & $\Sexpr{round(pop*est_mle_vec["R20"],0)}$  & $57072$\\
-      \(S_{30}\)     & 19573727  & $\Sexpr{round(pop*est_mle_vec["S30"],0)}$  & $19697600$\\
-      \(I_{30}\)     & 2550      & $\Sexpr{round(pop*est_mle_vec["I30"],0)}$  & $174$\\
-      \(R_{30}\)     & 131092    & $\Sexpr{round(pop*est_mle_vec["R30"],0)}$  & $9578$\\
-      \midrule
-      \textbf{AIC}  & 13778.08  & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$ & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec_sbh["logLik"]),2)}$\\
-      \bottomrule
-    \end{tabular}
-\end{center}
-\end{table}
-
-<<maximization7-Stocks_rinit, include=FALSE>>=
-set.seed(12345)
-sd_07 = (2/3)^6
-starting_values_sbh_07 <- choose_n_fits(sbh_el_06, top_n_fits, ncores)
-
-bake(file = paste0(out_dir, "mif2_maximization/SBH_init/round_07/mifs_global_sbh_07.rds"),{ 
-  plan(multicore, workers = ncores)
-  foreach(i = 1:ncores, .options.future = list(seed = 652643293)) %dofuture% {
-    pomp::mif2(
-      mifs_global_sbh_06[[1]],
-      rw.sd = rw_sd(
-        beta1 = 0.01*sd_07, beta2 = 0.01*sd_07, beta3 = 0.01*sd_07,
-        beta11 = 0.01*sd_07, phi = 0.01*sd_07,
-        sigma_q = 0.01*sd_07, sigma_xi = 0.01*sd_07
-      ),
-      params = starting_values_sbh_07[[i]]
-    ) 
-  }
-}) -> mifs_global_sbh_07
-
-bake(file = paste0(out_dir,"mif2_maximization/SBH_init/round_07/sbh_el_07.rds"),{
-  eval_logLik_pomp(mifs_global_sbh_07, ncores, NP_MIF, seed = 4248930)
-}) ->  sbh_el_07
-@
-
-<<PAL-MLE-from-mif2, echo=FALSE, message=FALSE, warning=FALSE>>=
-prop <- 420
-realdat <- as.matrix(dat[,-1])
-init_dist_mif2_mle <- c(
-  pop*as.numeric(est_mle_vec["S10"]), 
-  pop*as.numeric(est_mle_vec["I10"]), 
-  pop*as.numeric(est_mle_vec["R10"]),
-  pop*as.numeric(est_mle_vec["S20"]),
-  pop*as.numeric(est_mle_vec["R20"]),
-  pop*as.numeric(est_mle_vec["I20"]),
-  pop*as.numeric(est_mle_vec["S30"]),
-  pop*as.numeric(est_mle_vec["R30"]),
-  pop*as.numeric(est_mle_vec["I30"])
-)
-
-initial_guess_mif2_mle <- c(
-  as.numeric(est_mle_vec["beta1"]), 
-  as.numeric(est_mle_vec["beta2"]), 
-  as.numeric(est_mle_vec["beta3"]),
-  as.numeric(est_mle_vec["phi"]),
-  as.numeric(est_mle_vec["beta11"]),
-  as.numeric(est_mle_vec["sigma_q"]),
-  as.numeric(est_mle_vec["sigma_xi"])
-)
-
-bake(file = paste0(out_dir,"pfilter/original_dataset/palfilter_mif2_mle.rds"),{
- set.seed(1223)
- rotavirus_SMC_qropxi(
-    init_dist = init_dist_mif2_mle, 
-    y_obs = realdat, 
-    m = 9, 
-    regular_params = c(
-      initial_guess_mif2_mle[1], 
-      initial_guess_mif2_mle[2], 
-      initial_guess_mif2_mle[3], 
-      initial_guess_mif2_mle[4],
-      initial_guess_mif2_mle[5]
-    ), 
-    gamma_par = c(initial_guess_mif2_mle[7], 1/initial_guess_mif2_mle[7]), 
-    norm_par = c(0.07, initial_guess_mif2_mle[6]), 
-    prop = t(prop), 
-    n_particles = NP_MIF, 
-    ncores = ncores
-  ) 
-}) -> palfilter_mif2_mle
-@
-
-<<plot2-cond-loglik, echo=FALSE, fig.height=4, fig.width=5.5, out.width="4.5in",fig.align="center", fig.cap='Conditional log-likelihoods computed using PAL (red) and PF (blue) for the rotavirus model and MLE from WWR.
-We used 50000 particle for both PF and PAL.
-PF averaged over 36 replications to reduce the Monte Carlo variance, but this step was not necessary for PAL.
-The main sources of likelihood shortfall of PF are at time points $t=1, 2, 3, 11, 81, 194$, and $325-333$.'>>=
-
-# Combine the data into a single dataframe for SMC with grouping variable
-data_list <- lapply(1:ncores, function(i) {
-  data.frame(
-    time = time, 
-    cond_loglik = pfilter_wwr_cond_lik[[i]], 
-    method = "SMC",
-    group = i
-  )
-})
-
-# Combine all the SMC data frames into one
-df_smc <- bind_rows(data_list)
-
-# Create the PAL data frame
-df_pal <- data.frame(
-  time = time, 
-  cond_loglik = palfilter_lik_list_wwr$ll_storage, 
-  method = "PAL",
-  group = 1  # Single group for PAL
-)
-
-# Combine both data frames
-df <- bind_rows(df_smc, df_pal)
-
-# Plot with ggplot
-ggplot(df, aes(x = time, y = cond_loglik, color = method)) +
-  geom_line(data = df %>% filter(method == "SMC"), aes(group = group), color = "blue", alpha = 0.7) +
-  geom_line(data = df %>% filter(method == "PAL"), aes(group = group), color = "red") +
-  theme_minimal() +
-  labs(
-    title = NULL,
-    x = "Time (weeks)",
-    y = "Conditional Log-likelihood",
-    color = "Method"
-  ) +
-  theme(legend.position = "bottom")
-@
-
-\begin{table}[htbp]
-    \caption{Maximum likelihood estimation of parameters for the OvOv model of WWR by PAL (taken from WWR), and by iterated particle filtering implemented using {\rm \texttt{mif2}} in {\rm \texttt{pomp}} \citep{pomppackagepaper}.
-The initial distribution parameter \(\lambda_0 = (S_{10}, I_{10}, R_{10}, S_{20}, I_{20}, R_{20}, S_{30}, I_{30}, R_{30})\) are assumed to be fixed in \cite{wwr}.
-The iterated particle filtering estimates were obtained by using 3 rounds of refinement, with 100 filtering iterations, 50,000 particles, and 36 replicates in each round with the top 12 best fits in terms of likelihood chosen to be the starting value for the next round.}
-    \label{tab:mlebywwr}
-\begin{center}    
-    \begin{tabular}{lccc}  % Left-aligned for the first column, centered for the other two columns
-      \toprule
-      Parameter      & PAL      & iterated filtering & iterated filtering (stocks)\\
-      \midrule
-      \(\beta_1\)    & 11.48     & $\Sexpr{round(est_mle_vec["beta1"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta1"],2)}$\\
-      \(\beta_2\)    & 0.25      & $\Sexpr{round(est_mle_vec["beta2"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta2"],2)}$\\
-      \(\beta_3\)    & 0.35      & $\Sexpr{round(est_mle_vec["beta3"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta3"],2)}$\\
-      \(\phi\)       & 0.14      & $\Sexpr{round(est_mle_vec["phi"],2)}$   & $\Sexpr{round(est_mle_vec_sbh["phi"],2)}$\\
-      \(\rho\)       & 0.16      & $\Sexpr{round(est_mle_vec["beta11"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["beta11"],2)}$\\
-      \(\sigma^2_q\) & 0.021     & $\Sexpr{round(est_mle_vec["sigma_q"],3)}$ & $\Sexpr{round(est_mle_vec_sbh["sigma_q"],3)}$\\
-      \(\sigma_\xi\) & 66.89     & $\Sexpr{round(est_mle_vec["sigma_xi"],2)}$ & $\Sexpr{round(est_mle_vec_sbh["sigma_xi"],2)}$\\
-      \(S_{10}\)     & 3876549   & $\Sexpr{round(pop*est_mle_vec["S10"],0)}$  & $5094840$\\
-      \(I_{10}\)     & 30351     & $\Sexpr{round(pop*est_mle_vec["I10"],0)}$  & $2871$\\
-      \(R_{10}\)     & 1315221   & $\Sexpr{round(pop*est_mle_vec["R10"],0)}$  & $124410$\\
-      \(S_{20}\)     & 57139612  & $\Sexpr{round(pop*est_mle_vec["S20"],0)}$  & $57385600$\\
-      \(I_{20}\)     & 871       & $\Sexpr{round(pop*est_mle_vec["I20"],0)}$  & $639$\\
-      \(R_{20}\)     & 302852    & $\Sexpr{round(pop*est_mle_vec["R20"],0)}$  & $57072$\\
-      \(S_{30}\)     & 19573727  & $\Sexpr{round(pop*est_mle_vec["S30"],0)}$  & $19697600$\\
-      \(I_{30}\)     & 2550      & $\Sexpr{round(pop*est_mle_vec["I30"],0)}$  & $174$\\
-      \(R_{30}\)     & 131092    & $\Sexpr{round(pop*est_mle_vec["R30"],0)}$  & $9578$\\
-      \midrule
-      \textbf{AIC}  & 13778.08  & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec["logLik"]),2)}$ & $\Sexpr{round(2*((length(params_wwr) - length(params_fixed)) - est_mle_vec_sbh["logLik"]),2)}$\\
-      \bottomrule
-    \end{tabular}
-\end{center}
-\end{table}
-
-Figure~\ref{fig:plot2-cond-loglik} indicates that the worst conditional log-likelihood values, estimated using the particle filter, arise at the start of the time series.
-This could be a problem with data collection, or some subtle issue with the science of the epidemiological system, but a simple possibility is that the initial values of the latent state process might be poorly specified.
-For this model, the initial values were fixed based on scientific reasoning, rather than being estimated.
-If we instead estimate the initial values, we obtain the results in Table~\ref{tab:mlebywwr}.

### 2024-09-16 ms.Rnw
Small changes to writing

@@
 A central piece of evidence for PAL is the comparison in Table~1 of \cite{wwr}, which claims a large improvement for PAL over a standard particle filter algorithm.
-This evidence, based on a model and data from a previous scientific study by \citet{stocks}, might suggests that researchers confronted with similar models should use PAL rather than particle filter methods.
+This evidence, based on a model and data from a previous scientific study by \citet{stocks}, might suggest that researchers confronted with similar models should use PAL rather than particle filter methods.
 Taken at face value, this evidence also reduces the credibility of \cite{stocks} by indicating a shortcoming with the numerical methods that they used.
@@
 
+\section{Results}
 \begin{table}[ht] % The placement specifier can be [h!tbp]
@@
 Lines 1, 2 and 7 are from WWR, and the remainder are our own computations.
-We used $5 \times 10^4$ particle for both PF and PAL.
+We used $5 \times 10^4$ particles for both PF and PAL.
 PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL due to its lower Monte~Carlo variance.
-PF was maximized using iterated filtering, following the approach of SBH.
-PAL was maximized using coordinate gradient descent, using the code of WWR.
+PF results were maximized using iterated filtering, following the approach of SBH.
+PAL results were maximized using coordinate gradient descent, using the code of WWR.
 } 
@@
 Comparison of lines 3 and 5 shows that this improvement enables PF to reach the values attained by PAL, and show a small improvement.
-Further, at the MLE obtained by maximizing PF, we found that PF beats PAL (comparing lines 5 and 6).
+Further, at the MLE obtained by maximizing with PF, we found that PF beats PAL (comparing lines 5 and 6).
 
@@
 Simulations with one or more zero counts were disqualified since they resulted in errors for the PAL implementation of WWR.
-We used $5 \\times 10^4$ particle for both PF and PAL.
+We used $5 \\times 10^4$ particles for both PF and PAL.
 PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL due to its lower Monte~Carlo variance.
@@
 
+\section{Discussion}
 It is currently unclear why the model of SBH beats the log-ARMA benchmark for their data, whereas the model of WWR fits more poorly than the log-ARMA model for its corresponding data.

### 2024-09-17 ms.Rnw
some edit

--- 2024-09-16 ms.Rnw
@@
 
-% Maximization Round 1 - Full rinit
+% Maximization Round 1 - Full rinit: also estimate initial values
 
@@
 
-% Continue: Maximization Round 2 - Full rinit
+% Round 2 - Full rinit
 
@@
 
-% Continue: Maximization Round 3 - Full rinit
+% Round 3 - Full rinit
 
@@
 
-% Maximization Round 1 - Stocks rinit
+% Maximization Round 1 - Fix initial values with Stocks' rinit
 
@@
 
-% Continue: Maximization Round 2 - Stocks rinit
+% Round 2 - Stocks rinit
 
@@
 
-% Continue: Maximization Round 3 - Stocks rinit
+% Round 3 - Stocks rinit
 
@@
 
+% Round 4 - Stocks rinit
+
 <<maximization4-Stocks_rinit, echo=F, warning=F, message=F>>=
@@
 
+% Round 5 - Stocks rinit
+
 <<maximization5-Stocks_rinit, echo=F, warning=F, message=F>>=
@@
 @
+
+% Round 6 - Stocks rinit
 
@@
 \label{tab:ovovrealdata}
-\begin{tabular}{lllr} 
+\begin{tabular}{lllllr} 
   \hline
-  & Method & Result & AIC  \\
+  & Method & Data & Model & Parameters                          & AIC  \\
   \hline
-  1. & PF & Table~1 of WWR, originally from SBH &  \Sexpr{myround(20134.38,tab_signif)}
+  1. & PF  & SBH  & SBH   & Table~1 of WWR, originally from SBH &  \Sexpr{myround(20134.38,tab_signif)}
   \\
-  2. & PAL & Table~1 of WWR & \Sexpr{myround(13778.08,tab_signif)}
+  2. & PAL & WWR  & WWR   & Table~1 of WWR                      & \Sexpr{myround(13778.08,tab_signif)}
   \\
-  3. & PAL & Re-evaluating the data, model and parameters from WWR &
+  3. & PAL & WWR  & WWR   & Re-evaluating the data, parameters from WWR &
     \Sexpr{myround(-2*palfilter_lik_list_wwr$log_lik + 2*7 ,tab_signif)} 
   \\
-  4. & PF & Data, model and parameters matching 3. &
+  4. & PF  & WWR & WWR    & Parameters matching 3. &
     \Sexpr{myround(-2*logmeanexp(logLik(pfilter_wwr)) + 2*7,tab_signif)}
   \\
-  5. & PF & Initial values changed to match SBH &
+  5. & PF  & WWR & WWR modified   & Initial values changed to match SBH &
      \Sexpr{myround(-2*est_mle_vec_sbh["logLik"] + 2*7,tab_signif)}
   \\
-  6. & PAL & Data, model and parameters matching 5.  &
+  6. & PAL & WWR & WWR modified   & Data, model and parameters matching 5.  &
      \Sexpr{myround(-2*palfilter_mif2_mle$log_lik + 2*7 ,tab_signif)}
   \\   
-  7. & ARMA & Sec.~7.3.2 of WWR & 23043
+  7. & ARMA & SBH & ARMA benchmark    & Sec.~7.3.2 of WWR & 23043
   \\
-  8. & ARMA & Data from WWR &
+  8. & ARMA & WWR & ARMA benchmark      & ARMA benchmark &
     \Sexpr{myround(wwr_benchmark,tab_signif)} 
   \\
+ \hline
+\end{tabular}
+\end{table}
+
+\begin{table}[ht] % The placement specifier can be [h!tbp]
+\centering 
+\caption{AIC for the OvOv rotavirus model, computed using two filtering methods.
+PAL is the Poisson approximate likelihood, implemented using the code of WWR.
+PF is the particle filter, implemented using the R package pomp \citep{pomppackagepaper}.
+Lines 1, 2 and 7 are from WWR, and the remainder are our own computations.
+We used $5 \times 10^4$ particle for both PF and PAL.
+PF was repeated 36 times to reduce the Monte Carlo variance, but this step was not necessary for PAL due to its lower Monte~Carlo variance.
+PF was maximized using iterated filtering, following the approach of SBH.
+PAL was maximized using coordinate gradient descent, using the code of WWR.}
+\label{tab:ovovrealdata}
+\begin{tabular}{lllllr} 
+  \hline
+  & Method   & Data  & Model & Parameters \& Source                    & AIC  \\  
+  \hline
+  1. & PF      & SBH   & SBH   & Table~1 of WWR, originally from SBH &  \Sexpr{myround(20134.38,tab_signif)} \\
+  2. & PAL     & WWR   & WWR   & Table~1 of WWR                      & \Sexpr{myround(13778.08,tab_signif)} \\
+  3. & PAL     & WWR   & WWR   & Re-evaluating the data, parameters from WWR & \Sexpr{myround(-2*palfilter_lik_list_wwr$log_lik + 2*7 ,tab_signif)} \\
+  4. & PF      & WWR   & WWR   & Parameters matching 3. & \Sexpr{myround(-2*logmeanexp(logLik(pfilter_wwr)) + 2*7,tab_signif)} \\
+  5. & PF      & WWR   & WWR modified   & Initial values changed to match SBH & \Sexpr{myround(-2*est_mle_vec_sbh["logLik"] + 2*7,tab_signif)} \\
+  6. & PAL     & WWR   & WWR modified   & Data, model and parameters matching 5. & \Sexpr{myround(-2*palfilter_mif2_mle$log_lik + 2*7 ,tab_signif)} \\
+  \hline
+     & Benchmark &  &  &  \\
+  \hline
+  7. & ARMA    & SBH   & \multicolumn{2}{c}{\centering Sec.~7.3.2 of WWR}             & 23043 \\
+  8. & ARMA    & WWR   & \multicolumn{2}{c}{\centering Re-evaluating 7. with data used by WWR}                 & \Sexpr{myround(wwr_benchmark,tab_signif)} \\
   \hline
@@
 
-Also contained additional results reinforcing the conclusions presented in this article.
 The source code for this article is available at \url{https://github.com:ionides/pal-vs-pf} and archived at Zenodo.

### 2024-09-17 ms.Rnw
edits following Aaron's suggestions

--- 2024-09-16 ms.Rnw
@@
 
-\section{Introduction}
+% \section{Introduction}
 
@@
 
-\section{Results}
+% \section{Results}
+
 \begin{table}[ht] % The placement specifier can be [h!tbp]
@@
 \label{tab:ovovrealdata}
-\begin{tabular}{lllr} 
+\begin{tabular}{llllr} 
   \hline
-  & Method & Result & AIC  \\
+  & Method & Data & Model & AIC  \\
   \hline
-  1. & PF & Table~1 of WWR, originally from SBH &  \Sexpr{myround(20134.38,tab_signif)}
+  1. & PF & Rescaled counts & SBH &  \Sexpr{myround(20134.38,tab_signif)}
   \\
-  2. & PAL & Table~1 of WWR & \Sexpr{myround(13778.08,tab_signif)}
+  2. & PAL & Counts &  WWR & \Sexpr{myround(13778.08,tab_signif)}
   \\
-  3. & PAL & Re-evaluating the data, model and parameters from WWR &
+  3. & PAL & Counts & Re-evaluating WWR at published parameters &
     \Sexpr{myround(-2*palfilter_lik_list_wwr$log_lik + 2*7 ,tab_signif)} 
   \\
-  4. & PF & Data, model and parameters matching 3. &
+  4. & PF & Counts & Model and parameters matching line 3 &
     \Sexpr{myround(-2*logmeanexp(logLik(pfilter_wwr)) + 2*7,tab_signif)}
   \\
-  5. & PF & Initial values changed to match SBH &
+  5. & PF & Counts & WWR with initial values changed to SBH &
      \Sexpr{myround(-2*est_mle_vec_sbh["logLik"] + 2*7,tab_signif)}
   \\
-  6. & PAL & Data, model and parameters matching 5.  &
+  6. & PAL & Counts & Model and parameters matching line 5  &
      \Sexpr{myround(-2*palfilter_mif2_mle$log_lik + 2*7 ,tab_signif)}
   \\   
-  7. & ARMA & Sec.~7.3.2 of WWR & 23043
+  7. & Benchmark & Rescaled counts  & log-ARMA$(2,1)$ & 23043
   \\
-  8. & ARMA & Data from WWR &
-    \Sexpr{myround(wwr_benchmark,tab_signif)} 
+  8. & Benchmark & Counts           &  log-ARMA$(2,1)$ & \Sexpr{myround(wwr_benchmark,tab_signif)} 
   \\
@@
 
-SBH fitted their model to a dataset derived by dividing the original reported count data by an estimated reporting rate, to put their data on the scale of the actual number of cases in the population, whereas WWR fitted directly to the report data.
+SBH fitted their model to a dataset of rescaled counts derived by dividing the original reported count data by an estimated reporting rate.
+Thus, SBH fitted to data on the scale of the disease incidence in the population.
+By contrast, WWR fitted directly to the reported case count data.
 The reporting rate used by SBH varied over time and location, but was generally around $7\%$.
@@
 
-\section{Discussion}
+% \section{Discussion}
+
 It is currently unclear why the model of SBH beats the log-ARMA benchmark for their data, whereas the model of WWR fits more poorly than the log-ARMA model for its corresponding data.

### 2024-09-17 ms.Rnw

@@
 PF is the particle filter, implemented using the R package pomp \citep{pomppackagepaper}.
-Lines 1, 2 and 7 are from WWR, and the remainder are our own computations.
+Lines 1, 2 and 7 are taken from WWR, and the remainder are our own computations.
+Line 3 recomputes the previously published value in Line 2, and the small difference is presumably due to rounding in Table~2 of WWR.
 We used $5 \times 10^4$ particles for both PF and PAL.
@@
 \label{tab:ovovrealdata}
-\begin{tabular}{llllr} 
+\begin{tabular}{lllllr} 
   \hline
-  & Method & Data & Model & AIC  \\
+  & Method & Data & Model & Parameters                          & AIC  \\
   \hline
-  1. & PF & Rescaled counts & SBH &  \Sexpr{myround(20134.38,tab_signif)}
+  1. & PF  & Rescaled counts  & SBH   & Table~2 of SBH &  \Sexpr{myround(20134.38,tab_signif)}
   \\
-  2. & PAL & Counts &  WWR & \Sexpr{myround(13778.08,tab_signif)}
+  2. & PAL & Counts  & WWR   & Table~2 of WWR                      & \Sexpr{myround(13778.08,tab_signif)}
   \\
-  3. & PAL & Counts & Re-evaluating WWR at published parameters &
+  3. & PAL & Counts  & WWR   & Table~2 of WWR  &
     \Sexpr{myround(-2*palfilter_lik_list_wwr$log_lik + 2*7 ,tab_signif)} 
   \\
-  4. & PF & Counts & Model and parameters matching line 3 &
+  4. & PF  & Counts & WWR    & Table~2 of WWR &
     \Sexpr{myround(-2*logmeanexp(logLik(pfilter_wwr)) + 2*7,tab_signif)}
   \\
-  5. & PF & Counts & WWR with initial values changed to SBH &
+  5. & PF  & Counts & WWR modified   & Maximum likelihood &
      \Sexpr{myround(-2*est_mle_vec_sbh["logLik"] + 2*7,tab_signif)}
   \\
-  6. & PAL & Counts & Model and parameters matching line 5  &
+  6. & PAL & Counts & WWR modified   & Matching line 5 &
      \Sexpr{myround(-2*palfilter_mif2_mle$log_lik + 2*7 ,tab_signif)}
   \\   
-  7. & Benchmark & Rescaled counts  & log-ARMA$(2,1)$ & 23043
+  7. & Benchmark & Rescaled counts & log-ARMA(2,1)    &  & 23043
   \\
-  8. & Benchmark & Counts           &  log-ARMA$(2,1)$ & \Sexpr{myround(wwr_benchmark,tab_signif)} 
+  8. & Benchmark & Counts & log-ARMA(2,1)      &  &
+    \Sexpr{myround(wwr_benchmark,tab_signif)} 
   \\
@@
 Line 5 of Table~\ref{tab:ovovrealdata} incorporates the SBH specification of initial values into the model of WWR.
-Comparison of lines 3 and 5 shows that this improvement enables PF to reach the values attained by PAL, and show a small improvement.
-Further, at the MLE obtained by maximizing with PF, we found that PF beats PAL (comparing lines 5 and 6).
+The likelihood for this modified model was then maximized using an iterated filtering procedure similar to SBH.
+Comparison of lines 3 and 5 shows that this improvement enables PF to reach the AIC values attained by PAL, and show a small improvement.
+Further, at the parameter values obtained by maximizing with PF, we found that PF beats PAL (comparing lines 5 and 6).
 

### 2024-09-17 ms.Rnw

@@
 Comparison of lines 3 and 5 shows that this improvement enables PF to reach the AIC values attained by PAL, and show a small improvement.
-Further, at the parameter values obtained by maximizing with PF, we found that PF beats PAL (comparing lines 5 and 6).
+Further, at these maximum likelihood parameter values, we found that PF beats PAL (comparing lines 5 and 6; for line 6 we initialized PAL at the average value of the stochastic initialization used in line 5).
 

### 2024-09-29 ms.Rnw
: -> /
tidying
add Zenodo link
minor changes for jrssb submission

@@
 \title{Poisson approximate likelihood compared to the particle filter}
-\author{Yize Hao, Aaron A. Abkemeier and Edward L. Ionides}
-\date{Draft compiled on \today}
+\author{Yize Hao, Aaron J. Abkemeier and Edward L. Ionides}
+\date{The University of Michigan}
 \maketitle
@@
 Filtering algorithms are fundamental for inference on partially observed stochastic dynamic systems, since they provide access to the likelihood function and hence enable likelihood-based or Bayesian inference.
-A novel Poisson approximate likelihood (PAL) filter was introduced by \cite{wwr}.
+A novel Poisson approximate likelihood (PAL) filter was introduced by Whitehouse et al.
+(2023) (henceforth, WWR).
 PAL employs a Poisson approximation to conditional densities, offering a fast approximation to the likelihood function for a certain subset of partially observed Markov process models.
-A central piece of evidence for PAL is the comparison in Table~1 of \cite{wwr}, which claims a large improvement for PAL over a standard particle filter algorithm.
-This evidence, based on a model and data from a previous scientific study by \citet{stocks}, might suggest that researchers confronted with similar models should use PAL rather than particle filter methods.
-Taken at face value, this evidence also reduces the credibility of \cite{stocks} by indicating a shortcoming with the numerical methods that they used.
-However, we show that the comparison of log-likelihood values made by \cite{wwr} is flawed because their PAL calculations were carried out using a dataset scaled differently from the previous study.
+A central piece of evidence for PAL is the comparison in Table 1 of WWR, which claims a large improvement for PAL over a standard particle filter algorithm.
+This evidence, based on a model and data from a previous scientific study, might suggest that researchers confronted with similar models should use PAL rather than particle filter methods.
+However, we show that the comparison of log-likelihood values made by WWR is flawed because their PAL calculations were carried out using a dataset scaled differently from the previous study.
 If PAL and the particle filter are applied to the same data, the advantage claimed for PAL disappears.
 On simulations where the model is correctly specified, the particle filter outperforms PAL.
-
 \end{abstract}
@@
 
-The source code for this article is available at \url{https://github.com:ionides/pal-vs-pf} and archived at Zenodo.
+The source code for this article is available on GitHub (\url{https://github.com/ionides/pal-vs-pf}) and archived at Zenodo (\url{https://zenodo.org/doi/10.5281/zenodo.13777112}).
 An extended description of our methods, together with additional numerical results, is provided by \cite{hao24}.
